### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Migrated files from Fernando's repo

---
title: A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Fernando A. Lopez
    email: fernando.lopez@upct.es
    affiliation: Universidad Politecnica de Cartagena
  - name: Tatiane Menezes
    email: tatiane.menezes@ufpe.br
    affiliation: Departamento de Economia
  - name: Renata Cavalcanti
    email: renata.vcsantos@gmail.com
    affiliation: Nucleo de Pesquisa
  - name: Maira Galdino da Rocha Pitta
    email: mgrpitta@ufpe.br
    affiliation: Nucleo de Pesquisa
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
  - code: Universidad Politecnica de Cartagena
    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
  - code: Departamento de Economia
    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
  - code: Nucleo de Pesquisa
    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress. While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees. How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature and humidity. In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems. An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process. Our results indicate that incidence of the disease is lower at higher temperatures. The evidence with respect to humidity is more mixed, with coefficients for this variable that are significant in only some equations. Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.

journal: "Geographical Analysis"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

```{r load-packages, include=FALSE}
# The data package needs to be installed from github repository:
#devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
# Also, ask Fernando if the version of `spatsur` needed for the analysis is available from CRAN or needs special distribution
#library(aemet) # Needs install from github
library(covid19env)
library(ggthemes)
library(gridExtra)
library(kableExtra)
library(lubridate)
#library(meteoland) # Ask Fernando if this package was used in the end
library(plm)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
```

```{r load-data, include=FALSE}
# Load data from package `covid19env`
data("covid19_spain")
data("provinces_spain")
```

```{r data-preparation, include=FALSE}
# Convert GDP per capita to thousands of euros:
provinces_spain <- provinces_spain %>%
  mutate(GDPpc = GDPpc/1000)

# Join provincial data to incidence data and convert to simple features:
covid19_spain <- covid19_spain %>% 
  left_join(provinces_spain %>% st_drop_geometry(),
            by = c("province", "CCAA", "ID_INE"))
```

```{r sort-communities-by-latitude, include=FALSE}
# To visualize the distribution of temperature by CCAA, we sort the communities by latitude, from north to south. We do this by community instead of province because the large number of provinces clutters the plots.

# Dissolve internal provincial boundaries to obtain a simple features object with the autonomous communities: 
ccaa.sf <-  provinces_spain %>% 
  group_by(CCAA) %>% 
  summarize(provinces = n())

# Extract coordinates of autonomous communities
ccaa.coords <- ccaa.sf %>%
  st_centroid() %>%
  st_coordinates() %>%
  as.data.frame()

# Join Y coordinate to ccaa.sf
ccaa.sf <- ccaa.sf %>% 
  mutate(long = ccaa.coords$Y)

# Sort autonomous communities from north to south
ccaa.levels <- ccaa.sf %>% 
  arrange(desc(long)) %>% select(CCAA)

ccaa.levels <- as.character(ccaa.levels$CCAA)

# Relevel autonomous communities
covid19_spain <- covid19_spain %>%
  mutate(CCAA = factor(CCAA, levels = ccaa.levels, ordered = TRUE))
```

Introduction
==========================

From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus has threatened to overrun health systems the world over. In efforts to contain the spread, numerous governments in many nations and regions have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place. There are encouraging signs that these measures have arrested the spread of the virus where they have been implemented, and have thus helped to keep a bad situation from becoming even worse [e.g., -@Lancastle2020impact]. However, this has come at a high cost, and the consequences for all spheres of the economy, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how]. As a result, there is a sense of urgency to anticipate the progression of the pandemic, in order to plan for progressive lifting of restrictions to movement and social contact [e.g., @Kissler2020projecting]. Needless to say, this has become a delicate, and politically charged, balancing act between public health and the economy [@Gong2020balance].

A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations. Earlier, diverse studies have shown the effect of temperature and humidity on the incidence of influenza [e.g., @Makinen2009cold; @Jaakkola2014decline; @Kudo2019low]. Jaakkola et al. [-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low. After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV. In this way, Casanova et al. [-@Casanova2010effects] used two surrogates, namely the gastroenteritis (TGEV) and mouse hepatitis viruses (MHV), to find that virus inactivation was more rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels. In a similar vein, but working directly with SARS-CoV, Chan et al. [-@Chan2011effects] found that viability of the virus was lost at temperatures higher than 38C and relative humidity superior to 95%.

While existing research on similar pathogens suggests that SARS-CoV is more stable and potentially easier to transmit in conditions of low temperature and low humidity, it is far from certain that this will also be the case with the novel SARS-CoV2. If such is the case, there is the possibility of easing restrictions to social contact as the weather warms; however, weeks or possibly months of costly measures could become undone if not, and the restrictions are lifted prematurely. Not surprisingly, given the stakes involved, this issue has already triggered a lively debate.

Some of what is thought about the possible seasonality of COVID-19 is based on analogies to the patterns of other known respiratory viruses. However, de Ángel Solá et al. [-@deangel2020weathering] note that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" (section 4). This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive. At a global scale, whereas de Ángel Solá et al. [-@deangel2020weathering] see less risk in the Caribean Basin, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread). In the United States, the National Academy of Sciences, Engineering, and Medicine was engaged by the Office of the Executive for guidance on this matter [see @National2020rapid]. Their conclusion summarizes the situation well (see p. 6): "Some limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV-2 pandemic unfolds could shed more light on the effects of climate on transmission."

With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature and humidity. Hence, here we report results from a spatial analysis of incidence of COVID-19 in fifty provinces in Spain. Spain is one of the countries hardest hit by the virus, and enacted lockdown measures on March 16, 2020, in response to a rapidly growing outbreak of COVID-19. We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days. We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables. These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a spatial contagion process.

The results provide evidence of the effect of temperature on the incidence of COVID-19. The evidence concerning humidity is more mixed: while the direction of the effect is negative, as anticipated, the coefficients for this variable are only significant in some of equations in the system. Our control variables also provide some intriguing insights. The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers wrestling with the question of the trajectory of the pandemic.

Please note that this paper is prepared as a reproducible research document. The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following repository:

https://github.com/paezha/covid19-environmental-correlates

Context, Data, and Methods
============

## Covid-19 in Spain

When was the first case reported? When did the country close international arrivals? When did initial measures of social distancing were implemented? The lockdown?

## Selection of Variables

Explain the rationale for selecting the variables.

For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world. Controlling for socioeconomic characteristics the objective of this paper is observe the effect of clime on COVID-19  proliferation.

Some questions are important to point out. Is not evident present a diary association between the temprature (resp. humidity) with declarate case. There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate

## Sources of Data

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases tested} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units). The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity. The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province. Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provencance of the data.

```{r descriptive-statistics, echo=FALSE}
junk <- data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

## Methods: Spatial SUR

Seemingly Unrelated Regressions are used in the context of...

The classical SUR model without spatial effects (from here, SUR-SIM) is a stack of equations as follows:

\begin{equation}
\label{eq:sur-sim}
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_T
\end{bmatrix}
=
\begin{bmatrix}
X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
\end{bmatrix}
\
\begin{bmatrix}
\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
\end{bmatrix}
\end{equation}

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.

The more relevant characteristic of a SUR model is the dependence structure among the vectors of residuals, namely:

\begin{equation}
\label{eq:sur-err}
E[\epsilon_t \epsilon'_{t'}]=\sigma_{tt'}
\end{equation}

Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$. This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time. In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ considering the first $r$ coefficients contanst.

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

\begin{equation}
y = X \beta + \epsilon
\end{equation}

\noindent where....

As is the case of cross-sectional data, it is possible identify spatial autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several Lagrange Multipliers tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, there are alternative several specifications that have been propose to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. The stack expresion for the spatial lag SUR model (SUR-SLM),

\begin{equation}
\label{eq:sur-slm}
\bf{A}y = X \beta + \epsilon \\
\epsilon =N(0,\Omega)
\end{equation}

\noindent where A =$I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.

This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Parameters of spatial dependence are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$). The correspond Wald test is available in the R package **spsur**

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).

Analysis and Results
===================

## Exploratory Data Analysis

We begin with the exploratory analysis of the data.

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages. Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

week11.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 11) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 13 - March 15") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week12.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 12) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds",
                       direction = 1) +
  ggtitle("March 16 - March 22") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 4.9, y = madrid_xy[2] + 0.6) +
  geom_segment(x = madrid_xy[1] - 3.4, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])

week13.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 13) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 23 - March 29") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid",
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week14.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 14) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 30 - April 5") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week15.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 15) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("April 6 - April 11") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, nrow = 3)
```

To visualize the distribution of temperature and humidity we aggregate the provinces by Autonomous Community. In Figure \ref{fig:descriptives-temperature} the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost. There is a relatively wide range of values both within and between provinces over the 30 day period examined. The top panel of the figure shows the distribution of mean temperature. The lowest mean temperature for a community on any given day is 2.8C, and the highest is 22.4C for a range of approximately 20 degrees. Likewise, there is a great deal of variability in humidity, as seen in the bottom panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6. The actual values for the provinces display somewhat more variability even.

```{r descriptives-temperature, echo = FALSE, fig.height=7, fig.cap="\\label{fig:descriptives-temperature} Distribution of mean temperatures and humidities in the Autonomous Communities in Spain between March 12, 2020 and April 11, 2020. The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost"}
# Summary of temperature by Autonomous Community
temp.summary <- covid19_spain %>% 
  group_by(CCAA, Date) %>%
  summarize(Value = mean(Mean_Temp)) %>%
  mutate(Variable = "Mean Temperature in Community")

# Summary of humidity by Autonomous Community
humidity.summary <- covid19_spain %>% 
  group_by(CCAA, Date) %>%
  summarize(Value = mean(Humidity)) %>%
  mutate(Variable = "Mean Humidity in Community")

# Bind tables
env.summary <- rbind(temp.summary, humidity.summary) %>%
  mutate(Variable = factor(Variable, 
                           levels = c("Mean Temperature in Community",
                                      "Mean Humidity in Community")))

# Plot
ggplot(data = env.summary, 
       aes(x = CCAA,
           y = Value)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Autonomous Community (sorted from north to south)") +
  ylab("Value") +
  facet_wrap(~ Variable, nrow = 2, 
             scales = "free_y")
```

```{r control-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)"}
gdppc.plot <- provinces_spain %>%
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc)) +
  scale_fill_distiller(name = "GDP per capita", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")

# Older adults:
older.plot <- provinces_spain %>% 
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = as.numeric(Older))) +
  scale_fill_distiller(name = "Percentage of older adults", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")

# Population density:
density.plot <- provinces_spain %>% 
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = as.numeric(Density))) +
  scale_fill_distiller(name = "Population Density", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")

grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
```

Figure \ref{fig:daily-correlations} shows the distribution of daily correlations of the independent variables with incidence of COVID-19, after log-transforming all variables. It can be seen there that the correlation of GDPpc and temperature (in its three definitions) have the strongest positive and negative correlations with incidence, respectively. Percentage of older adults displays somewhat weaker negative correlations with incidence, as does density. It can be seen that the humidity variable, in its three forms, tends to be possitively correlated with incidence of COVID-19.

```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)"}
covid19_spain %>% 
  group_by(Date) %>%
  summarize(Older = cor(log(Older), log(Incidence)),
            GDPpc = cor(log(GDPpc), log(Incidence)),
            Density = cor(log(Density), log(Incidence)),
            Humidity_lag8 = cor(Humidity_lag8, Incidence),
                        Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
                        Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
  select(-Date) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  mutate(Variable = factor(Variable,
                           levels = c("GDPpc", 
                                      "Older", 
                                      "Density", 
                                      "Humidity_lag8",
                                      "Humidity_lag11",
                                      "Humidity_lag11w",
                                      "Mean_Temp_lag8",
                                      "Mean_Temp_lag11",
                                      "Mean_Temp_lag11w"))) %>%
  ggplot(aes(x = Variable, y = Correlation)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90))
```

## SUR Models

```{r data-preparation-for-modelling, include=FALSE}
# Organize data for SUR modelling 
GPanel <- plm::pdata.frame(covid19_spain %>%
                             select(province, 
                                    Date,
                                    Incidence, 
                                    Median_Age,
                                    Male2Female,
                                    Older,
                                    GDPpc,
                                    Density,
                                    Transit,
                                    Mean_Temp_lag8,
                                    Humidity_lag8,
                                    Sunshine_Hours_lag8,
                                    Mean_Temp_lag11,
                                    Humidity_lag11,
                                    Sunshine_Hours_lag11,
                                    Mean_Temp_lag11w,
                                    Humidity_lag11w,
                                    Sunshine_Hours_lag11w),
                           c("province","Date"))

```

```{r spatial-weights, include=FALSE}
# Create spatial weights matrix: 
Wmat <- provinces_spain %>%
  #drop_na() %>%
  as("Spatial") %>%
  poly2nb(queen = FALSE) %>%
  nb2mat(zero.policy = TRUE)

Wmat <- (Wmat > 0) * 1

# Join the two provinces in Canarias
Wmat[which(provinces_spain$province == "Palmas(Las)"), 
     which(provinces_spain$province == "Santa Cruz de Tenerife")] <- 1
Wmat[which(provinces_spain$province == "Santa Cruz de Tenerife"), 
     which(provinces_spain$province == "Palmas(Las)")] <- 1

# 'Paises Catalans'
#n = 8
Wmat[which(provinces_spain$province == "Barcelona"), 
     which(provinces_spain$province == "Baleares")] <- 1
Wmat[which(provinces_spain$province == "Baleares"), 
     which(provinces_spain$province == "Barcelona")] <- 1
Wmat[which(provinces_spain$province == "Baleares"), 
     which(provinces_spain$province == "Castellon/Castello")] <- 1 
Wmat[which(provinces_spain$province == "Castellon/Castello"), 
     which(provinces_spain$province == "Baleares")] <- 1
Wmat[which(provinces_spain$province == "Baleares"), 
     which(provinces_spain$province == "Tarragona")] <- 1 
Wmat[which(provinces_spain$province == "Tarragona"), 
     which(provinces_spain$province == "Baleares")] <- 1
miW <- Wmat/rowSums(Wmat)

# Convert to listw
listw <- mat2listw(Wmat,style = "W")
```

```{r formulas, include=FALSE}
# Define formulas with three different lagged variables:
formula_lag8 <- log(Incidence) ~ 
  log(GDPpc) +
  log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag8) +
  log(Mean_Temp_lag8)

formula_lag11 <- log(Incidence) ~ 
  log(GDPpc) +
  log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag11) +
  log(Mean_Temp_lag11)

formula_lag11w <- log(Incidence) ~ 
  log(GDPpc) +
  log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag11w) +
  log(Mean_Temp_lag11w)
```

```{r model-restrictions, include=FALSE}
T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # Recall that T is the number of days, i.e., time periods, i.e., equations; add 1 to include the starting day
k <- 7 # Number of independent variables, including the constant
coef_rest <- 2 # Number of restrictions

# nrow is number of equations (time periods) minus 1, times the number of restrictions
# ncol is number of variables times number of equations
R2 <- matrix(0, nrow = (T - 1) * coef_rest, ncol = k * T)

for (i in 1:(T-1)){
  R2[i, 2] <- 1
  R2[i, (2 + i * k)] <- -1
  R2[(i + T - 1), 3] <- 1
  R2[(i + T - 1), (3 + i * k)] <- -1
  # Use if more restrictions are needed
  #R2[(i + T - 1) * 2, 4] <- 1
  #R2[(i + T - 1) * 2, (4 + i * k)] <- -1
}
b2 <- matrix(0, ncol = 21*coef_rest)
```

```{r model1-8-day-moving-average, include=FALSE}
# Model with a lagged 8-day moving average of climatic variables:
sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
                                 data=GPanel, 
                                 time = GPanel$Date, 
                                 type = "slm", 
                                 fit_method = "3sls", 
                                 listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag8)
#print(paste("Pooled R^2 = ", sur.slm_lag8$R2[1]))
```

```{r model2-11-day-moving-average, include=FALSE}
# Model with 11-day moving average of climatic variables:
sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
                                  data=GPanel, 
                                  time = GPanel$Date, 
                                  type = "slm", 
                                  fit_method = "3sls", 
                                  listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag11)
#print(paste("Pooled R^2 = ", sur.slm_lag11$R2[1]))
```

```{r model3-11-day-weighted-moving-average, include=FALSE}
# Model with 11-day weighted moving average of climatic variables:
sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
                                   data=GPanel, 
                                   time = GPanel$Date, 
                                   type = "slm", 
                                   fit_method = "3sls", 
                                   listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag11w)
#print(paste("Pooled R^2 = ", sur.slm_lag11w$R2[1]))
```

The goodness of fit of the three systems of equations is shown in Figure \ref{fig:goodness-of-fit}.

```{r goodness-of-fit, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
data.frame(Model1 = sur.slm_lag8$R2, 
           Model2 = sur.slm_lag11$R2,
           Model3 = sur.slm_lag11w$R2) %>%
  slice(2:n()) %>%
  rownames_to_column(var = "Equation") %>%
  mutate(Date = seq(ymd("2020-03-13"), 
                    ymd("2020-04-11"), 
                    by = "days")) %>%
  pivot_longer(cols = starts_with("M"), names_to = "Model", values_to = "R2") %>%
  mutate(Model = factor(Model, 
                        levels = c("Model1", "Model2", "Model3"),
                        labels = c("Model 1", "Model 2", "Model 3"))) %>%
  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Model 1" = "blue", "Model 2" = "orange", "Model 3" = "green") ) + 
  annotate(geom = "text", 
           label = c(paste("Model 1: pooled-R2=", 
                          round(sur.slm_lag8$R2[1], 4)),
                     paste("Model 2: pooled-R2=", 
                          round(sur.slm_lag11$R2[1], 4)),
                     paste("Model 3: pooled-R2=", 
                          round(sur.slm_lag11w$R2[1], 4))),
            x = as_date("2020-04-01"),
            y = c(0.3, 0.27, 0.24),
           family = "serif",
           size = 3.5) +
  theme_tufte()

```

Summary of best model.
```{r summary-best-model, echo=FALSE}
data.frame(Coefficient = sur.slm_lag11$coefficients, 
                   pval = dnorm(abs(sur.slm_lag11$coefficients/sur.slm_lag11$rest.se))) %>%
  rownames_to_column(var = "Variable") %>%
  mutate(Variable = c("Intercept", 
                      "log(GDPpc)", 
                      "log(Older)",
                      "log(Density)",
                      "Transit",
                      "log(Humidity)",
                      "log(Temperature)",
                      rep(c("Intercept",
                            "log(Density)", 
                            "Transit", 
                            "log(Humidity)", 
                            "log(Temperature)"), 29)),
         Variable = factor(Variable, 
                           levels = c("Intercept", 
                                      "log(GDPpc)", 
                                      "log(Older)", 
                                      "log(Density)",
                                      "Transit", 
                                      "log(Humidity)",
                                      "log(Temperature)"))) %>%
  group_by(Variable) %>%
  summarize(Min = min(Coefficient), 
            Mean = mean(Coefficient), 
            Max = max(Coefficient),
            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
  mutate(Note = c("Non-constrained", 
                  "Constrained", 
                  "Constrained", 
                  "Non-constrained", 
                  "Non-constrained",
                  "Non-constrained",
                  "Non-constrained")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 3,
        col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
                       "Non-constrained: coefficient was allowed to vary across equations",
                       "Constrained: coefficient as constant across equations"))
```

Discussion
===================

Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$).

```{r delta-time, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Deltas = sur.slm_lag11$deltas,
           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Deltas, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.

```{r beta-intercept-time, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
#Intercept
n = 1
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Density)$.

```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)"}
#Density:
n = 2 # Although this is the fourth coefficient, after intercept, log(GDPpc), and log(Older), we start at 2 here because the two constrained coefficients are removed from the matrix of coefficients 
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Figure \ref{fig:beta-transit-time} shows the temporal evolution of the coefficient for $Transit$.

```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit"}
n = 3
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the coefficient for $\log(Humidity)$.

```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)"}
n = 4
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.

```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)"}
n = 5
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Concluding Remarks
===================

More words go here.

Acknowledgments {#acknowledgments .unnumbered}
==========

Add acknowledgments as appropriate in final draft.

The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].

References {#references .unnumbered}
==========

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Migrated files from Fernando's repo

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Geographical Analysis} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatial analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{A spatial analysis of the environmental correlates of COVID-19 incidence
in Spain}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
   \ead{fernando.lopez@upct.es} 
    \author[Departamento de Economia]{Tatiane Menezes}
   \ead{tatiane.menezes@ufpe.br} 
    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
   \ead{renata.vcsantos@gmail.com} 
    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
   \ead{mgrpitta@ufpe.br} 
      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
St W, Hamilton, ON, L8S 4K1, Canada}
    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
30201, Cartagena, Murcia, Spain}
    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
50670-901, Brazil}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature and humidity. In
  this research, we address this question by means of a spatial analysis
  of the incidence of COVID-19 in the provinces in Spain. Use of a spatial
  Seemingly Unrelated Regressions (SUR) approach allows us to model the
  incidence of reported cases of the disease per 100,000 population, as a
  function of temperature and humidity, while controlling for GDP per
  capita, population density, percentage of older adults in the
  population, and presence of mass transit systems. Our results indicate
  that incidence of the disease is lower at higher temperatures. The
  evidence with respect to humidity is more mixed, with coefficients for
  this variable that are significant in only some equations. Our control
  variables also yield interesting insights, as population density and
  percentage of older adults display negative associations with incidence
  of COVID-19, whereas the presence of mass transit systems in the
  province is associated with a greater incidence of the disease.
  \end{abstract}

 \end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

From a small outbreak linked to a live animal market at the end of 2019
to a global pandemic in a matter of weeks, the SARS-CoV2 virus has
threatened to overrun health systems the world over. In efforts to
contain the spread, numerous governments in many nations and regions
have either recommended or mandated social distancing measures, and as
of this writing, millions of people in five continents shelter in place.
There are encouraging signs that these measures have arrested the spread
of the virus where they have been implemented, and have thus helped to
keep a bad situation from becoming even worse (e.g., 2020). However,
this has come at a high cost, and the consequences for all spheres of
the economy, social, and cultural life have been dire (e.g., Fernandes,
2020; Luo and Tsang, 2020). As a result, there is a sense of urgency to
anticipate the progression of the pandemic, in order to plan for
progressive lifting of restrictions to movement and social contact
(e.g., Kissler et al., 2020). Needless to say, this has become a
delicate, and politically charged, balancing act between public health
and the economy (Gong et al., 2020).

A salient question in the debate on how and when to ease social
distancing measures is whether the virus will display seasonal
variations. Earlier, diverse studies have shown the effect of
temperature and humidity on the incidence of influenza (e.g., Mäkainen
et al., 2009; Jaakkola et al., 2014; Kudo et al., 2019). Jaakkola et
al.~(2014), for example, found that a decrease of temperature and
absolute humidity precedes the onset of symptoms of influenza A and B
viruses by 3 days in places where the temperature is low. After the
2002-2004 outbreak of SARS, researchers also began to investigate the
relationship between these factors and SARS-CoV. In this way, Casanova
et al.~(2010) used two surrogates, namely the gastroenteritis (TGEV) and
mouse hepatitis viruses (MHV), to find that virus inactivation was more
rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of
humidity, these researchers reported that survival of the virus was
lower at moderate relative humidity levels. In a similar vein, but
working directly with SARS-CoV, Chan et al.~(2011) found that viability
of the virus was lost at temperatures higher than 38C and relative
humidity superior to 95\%.

While existing research on similar pathogens suggests that SARS-CoV is
more stable and potentially easier to transmit in conditions of low
temperature and low humidity, it is far from certain that this will also
be the case with the novel SARS-CoV2. If such is the case, there is the
possibility of easing restrictions to social contact as the weather
warms; however, weeks or possibly months of costly measures could become
undone if not, and the restrictions are lifted prematurely. Not
surprisingly, given the stakes involved, this issue has already
triggered a lively debate.

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, whereas de Ángel Solá et al.~(2020) see less risk in the Caribean
Basin, Coelho et al.~(Coelho et al., 2020) warn that at least during the
exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. To further complicate matters, much of the relevant work
has yet to be peer-reviewed and therefore is open to change (see for
example the challenge of Harbert et al. (2020) to Araujo and Naimi
(2020)). In the United States, the National Academy of Sciences,
Engineering, and Medicine was engaged by the Office of the Executive for
guidance on this matter (see National Academies of Sciences, Engineering
and Medicine, 2020). Their conclusion summarizes the situation well (see
p.~6): ``Some limited data support a potential waning of cases in warmer
and more humid seasons, yet none are without major
limitations\ldots Additional studies as the SARS-CoV-2 pandemic unfolds
could shed more light on the effects of climate on transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature and
humidity. Hence, here we report results from a spatial analysis of
incidence of COVID-19 in fifty provinces in Spain. Spain is one of the
countries hardest hit by the virus, and enacted lockdown measures on
March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
We combine data on reported cases of the disease with metereological
information, to create a spatio-temporal dataset covering a period of 30
days. We then join this dataset with provincial-level economic and
demographic information to act as controls to our key environmental
variables. These data are analyzed using a spatial Seemingly Unrelated
Regressions (SUR) approach, which allows us to account for spatial
autocorrelation.

The results provide evidence of the effect of temperature on the
incidence of COVID-19, as \textbf{NOTE ABOUT THE MAGNITUDE OF THE
EFFECT}. The evidence concerning humidity is more mixed: while the
direction of the effect is negative, as anticipated, the coefficients
for this variable are only significant in some of equations in the
system. Our control variables also provide some intriguing insights. The
results of this analysis provide support to the hypothesis of
seasonality of the novel SARS-CoV2, and should be of interest to public
health officials and policy makers wrestling with the question of the
trajectory of the pandemic.

Please note that this paper is prepared as a reproducible research
document. The source R Markdown document, as well as all data and code
needed to reproduce/review/extend the analysis can be obtained from the
following repository:

\begin{quote}
\begin{quote}
Include link to repository
\end{quote}
\end{quote}

\hypertarget{context-data-and-methods}{%
\section{Context, Data, and Methods}\label{context-data-and-methods}}

\hypertarget{covid-19-in-spain}{%
\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}

When was the first case reported? When did the country close
international arrivals? When did initial measures of social distancing
were implemented? The lockdown?

\hypertarget{selection-of-variables}{%
\subsection{Selection of Variables}\label{selection-of-variables}}

Explain the rationale for selecting the variables.

For example: The literature about COVID-19 suggested that population
density is the one of the most important proliferate cause of these
viscous, however this ill spread with different intensity at big cities
of the world. Controlling for socioeconomic characteristics the
objective of this paper is observe the effect of clime on COVID-19
proliferation.

Some questions are important to point out. Is not evident present a
diary association between the temprature (resp. humidity) with declarate
case. There is evidences (ref) that a time lag is necessary between the
day that the individuo inffect and the case is diagnosticate

\hypertarget{sources-of-data}{%
\subsection{Sources of Data}\label{sources-of-data}}

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain \footnote{cases tested} at the provincial
level (NUTIII in eurostat terminology) from March 13th to April 11th (30
days and 50 spatial units). The Spanish National Government publishes
periodic updates at the regional level (NUTII) and the information is
also released at the provincial level as part of a collaborative
project. This information is compiled from several sources, mainly the
official web pages of the Spanish regional goverments. In addition, we
consider two sets of explanatory variables. The first one, and the focus
of this research, is set of two environmental variables, namely
temperature and humidity. The second set provides some relevant controls
for multivariate analysis, and refers to economic and demographic
attributes of the province. Table \ref{tab:descriptive-statistics} shows
the descriptive statistics and the provencance of the data.

\hypertarget{methods-spatial-sur}{%
\subsection{Methods: Spatial SUR}\label{methods-spatial-sur}}

Seemingly Unrelated Regressions are used in the context of\ldots{}

The classical SUR model without spatial effects (from here, SUR-SIM) is
a stack of equations as follows:

\begin{equation}
\label{eq:sur-sim}
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_T
\end{bmatrix}
=
\begin{bmatrix}
X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
\end{bmatrix}
\
\begin{bmatrix}
\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
\end{bmatrix}
\end{equation}

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in the
province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
independent variables, \(X_i=(X^i_{st})\);
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
residuals.

The more relevant characteristic of a SUR model is the dependence
structure among the vectors of residuals, namely:

\begin{equation}
\label{eq:sur-err}
E[\epsilon_t \epsilon'_{t'}]=\sigma_{tt'}
\end{equation}

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
posible to impose restrinctions considering some \(\beta\) coefficients
as being constant over time. In this case, we can reformulate the
coefficients expression of
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
considering the first \(r\) coefficients contanst.

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

\begin{equation}
y = X \beta + \epsilon
\end{equation}

\noindent where\ldots.

As is the case of cross-sectional data, it is possible identify spatial
autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several
Lagrange Multipliers tests have been developed to test the null
hypothesis of spatial independence (see López et al., 2014). When the
null hypothesis is rejected, there are alternative several
specifications that have been propose to include spatial effects
(Anselin, 1988, see also 2016). In this paper we consider a spatial SUR
model that incorporates a spatial lag of the dependent variable as an
explanatory factor. The stack expresion for the spatial lag SUR model
(SUR-SLM),

\begin{equation}
\label{eq:sur-slm}
\bf{A}y = X \beta + \epsilon \\
\epsilon =N(0,\Omega)
\end{equation}

\noindent where A =\(I_{TN}-\bf{\Lambda} \otimes W\) with
\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).

This specification assumes that incidence in a province (\(y_st\)) at
time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
of incidence in neighbouring provinces. Parameters of spatial dependence
are estimated for each time period \(\lambda_t\) and identifies the
intensity and the sign of the impacts of neighbourhood. It is possible
test the null hypotheis of identtical levels of spatial dependence
(\(\lambda_i=\lambda_j, \forall i,j\)). The correspond Wald test is
available in the R package \textbf{spsur}

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)).

\hypertarget{analysis-and-results}{%
\section{Analysis and Results}\label{analysis-and-results}}

\hypertarget{exploratory-data-analysis}{%
\subsection{Exploratory Data Analysis}\label{exploratory-data-analysis}}

We begin with the exploratory analysis of the data.

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of the disease, as well as its temporal
progression in weekly averages. Our analysis begins on March 13, which
is the first date when every province had reported at least one case of
COVID-19. It can be seen that the highest incidence at this early date
was in the provice of Álava, in the North of Spain. While not the most
populous province, with a population of only 331,549, Álava has the
highest GDP per capita of all provinces. Vitoria, its main city, is the
capital of the Basque country and has been the focus of efforts to
develop a ``Global Basque City'' (Meijers et al., 2008), along with San
Sebastian and Bilbao. The other early focus of the pandemic in Spain was
Madrid, which is the most populous province in the country and has the
second highest GDP per capita after Álava. The early outbreaks in these
two provinces can be traced throughout the progression of the pandemic
over time, although by the end of the period under consideration, other
provinces had matched and even surpased their incidence rates, including
Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
to the south.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/weekly-average-incidence-map-1.pdf}
\caption{\label{fig:weekly-average-incidence-map}Mean weekly incidence
of COVID-19 by province, in reported cases by 100,000 people (map omits
Islas Canarias)}
\end{figure}

To visualize the distribution of temperature and humidity we aggregate
the provinces by Autonomous Community. In Figure
\ref{fig:descriptives-temperature} the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. There is a relatively wide range of
values both within and between provinces over the 30 day period
examined. The top panel of the figure shows the distribution of mean
temperature. The lowest mean temperature for a community on any given
day is 2.8C, and the highest is 22.4C for a range of approximately 20
degrees. Likewise, there is a great deal of variability in humidity, as
seen in the bottom panel of the figure, where the lowest mean humidity
for any community is 48.3 and the highest is 99.6. The actual values for
the provinces display somewhat more variability even.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/descriptives-temperature-1.pdf}
\caption{\label{fig:descriptives-temperature} Distribution of mean
temperatures and humidities in the Autonomous Communities in Spain
between March 12, 2020 and April 11, 2020. The Autonomous Communities
have been sorted by latitude, with communities to the left being the
northermost, and to the right the southernmost}
\end{figure}

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/control-map-1.pdf}
\caption{\label{fig:control-map}Spatial distribution of control
variables by province (maps omits Islas Canarias)}
\end{figure}

Figure \ref{fig:daily-correlations} shows the distribution of daily
correlations of the independent variables with incidence of COVID-19,
after log-transforming all variables. It can be seen there that the
correlation of GDPpc and temperature (in its three definitions) have the
strongest positive and negative correlations with incidence,
respectively. Percentage of older adults displays somewhat weaker
negative correlations with incidence, as does density. It can be seen
that the humidity variable, in its three forms, tends to be possitively
correlated with incidence of COVID-19.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/daily-correlations-1.pdf}
\caption{\label{fig:daily-correlations}Distribution of daily
correlations of the independent variables with daily incidence of
COVID-19 (all variables have been log-transformed)}
\end{figure}

\hypertarget{sur-models}{%
\subsection{SUR Models}\label{sur-models}}

The goodness of fit of the three systems of equations is shown in Figure
\ref{fig:goodness-of-fit}.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/goodness-of-fit-1.pdf}
\caption{\label{fig:goodness-of-fit} Goodness of fit of the SUR systems:
by date and pooled}
\end{figure}

Summary of best model.

\begin{table}

\caption{\label{tab:summary-best-model}\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Estimates} & \multicolumn{3}{c}{Significance} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
Variable & Min & Mean & Max & p > 0.10 & 0.10 <= p < 0.05 & p <= 0.05 & Note\\
\midrule
\rowcolor{gray!6}  Intercept & 7.175 & 10.008 & 13.845 & 0 & 0 & 30 & Non-constrained\\
log(GDPpc) & 0.513 & 0.513 & 0.513 & 0 & 1 & 0 & Constrained\\
\rowcolor{gray!6}  log(Older) & -0.783 & -0.783 & -0.783 & 0 & 0 & 1 & Constrained\\
log(Density) & -0.217 & -0.092 & 0.188 & 19 & 1 & 10 & Non-constrained\\
\rowcolor{gray!6}  Transit & 0.300 & 0.518 & 0.595 & 11 & 10 & 9 & Non-constrained\\
\addlinespace
log(Humidity) & -1.176 & -0.372 & 0.342 & 17 & 4 & 9 & Non-constrained\\
\rowcolor{gray!6}  log(Temperature) & -2.048 & -1.431 & -0.989 & 0 & 0 & 30 & Non-constrained\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{Significance: This is the number of coefficients with p-values as indicated}\\
\multicolumn{8}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{8}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}}
\end{table}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Figure \ref{fig:delta-time} shows the temporal evolution of the spatial
autocorrelation coefficient (\(\rho\)).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/delta-time-1.pdf}
\caption{\label{fig:delta-time}Temporal evolution of spatial
autocorrelation coefficient}
\end{figure}

Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the
intercept.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-intercept-time-1.pdf}
\caption{\label{fig:beta-intercept-time}Temporal evolution of intercept}
\end{figure}

Figure \ref{fig:beta-density-time} shows the temporal evolution of the
coefficient for \(\log(Density)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-density-time-1.pdf}
\caption{\label{fig:beta-density-time}Temporal evolution of coefficient
for log(Density)}
\end{figure}

Figure \ref{fig:beta-transit-time} shows the temporal evolution of the
coefficient for \(Transit\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-transit-time-1.pdf}
\caption{\label{fig:beta-transit-time}Temporal evolution of coefficient
for Transit}
\end{figure}

Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the
coefficient for \(\log(Humidity)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-humidity-time-1.pdf}
\caption{\label{fig:beta-humidity-time}Temporal evolution of coefficient
for log(Humidity)}
\end{figure}

Figure \ref{fig:beta-density-time} shows the temporal evolution of the
coefficient for \(\log(Temperature)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-temperature-time-1.pdf}
\caption{\label{fig:beta-temperature-time}Temporal evolution of
coefficient for log(Temperature)}
\end{figure}

\hypertarget{concluding-remarks}{%
\section{Concluding Remarks}\label{concluding-remarks}}

More words go here.

\hypertarget{acknowledgments}{%
\section*{Acknowledgments}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgments}

Add acknowledgments as appropriate in final draft.

The following \texttt{R} packages were used in the course of this
investigation and the authors wish to acknowledge their developers:
\texttt{aemet} {[}{]}, \texttt{ggthemes} (Arnold, 2019),
\texttt{gridExtra} (Auguie, 2017), \texttt{kableExtra} (Zhu, 2019),
\texttt{knitr} (Xie, 2015, 2014), \texttt{lubridate} (Grolemund and
Wickham, 2011), \texttt{plm} (Millo, 2017), \texttt{rticles} (Allaire et
al., 2020), \texttt{sf} (Pebesma, 2018), \texttt{spdep} (Bivand et al.,
2013), spsur (Angulo et al., 2020) \texttt{tidyverse} (Wickham et al.,
2019), \texttt{units} (Pebesma et al., 2016).

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Allaire2020}{}%
Allaire, J., Xie, Y., R Foundation, Wickham, H., Journal of Statistical
Software, Vaidyanathan, R., Association for Computing Machinery,
Boettiger, C., Elsevier, Broman, K., Mueller, K., Quast, B., Pruim, R.,
Marwick, B., Wickham, C., Keyes, O., Yu, M., Emaasit, D., Onkelinx, T.,
Gasparini, A., Desautels, M.-A., Leutnant, D., MDPI, Taylor and Francis,
Ögreden, O., Hance, D., Nüst, D., Uvesten, P., Campitelli, E.,
Muschelli, J., Kamvar, Z.N., Ross, N., Cannoodt, R., Luguern, D.,
Kaplan, D.M., 2020. Rticles: Article formats for r markdown.

\leavevmode\hypertarget{ref-Angulo2020spsur}{}%
Angulo, A., Lopez, F.A., Minguez, R., Mur, J., 2020. Spsur: Spatial
seemingly unrelated regression models.

\leavevmode\hypertarget{ref-Anselin1988spatial}{}%
Anselin, L., 1988. Spatial econometrics: Methods and models, Studies in
operational regional science. Kluwer Academic Publishers, Dordrecht.

\leavevmode\hypertarget{ref-Anselin2016estimation}{}%
Anselin, L., 2016. Estimation and testing in the spatial seemingly
unrelated regression (sur). Geoda Center for Geospatial Analysis;
Computation, Arizona State University. Working Paper 2016-01.

\leavevmode\hypertarget{ref-Araujo2020spread}{}%
Araujo, M.B., Naimi, B., 2020. Spread of sars-cov-2 coronavirus likely
to be constrained by climate. medRxiv.

\leavevmode\hypertarget{ref-Arnold2019}{}%
Arnold, J.B., 2019. Ggthemes: Extra themes, scales and geoms for
'ggplot2'.

\leavevmode\hypertarget{ref-Auguie2017gridextra}{}%
Auguie, B., 2017. GridExtra: Miscellaneous functions for "grid"
graphics.

\leavevmode\hypertarget{ref-deangel2020weathering}{}%
Ángel Solá, D.E. de, Wang, L., Vázquez, M., Lázaro, P.A.M., 2020.
Weathering the pandemic: How the caribbean basin can use viral and
environmental patterns to predict, prepare and respond to covid‐19.
Journal of Medical Virology.

\leavevmode\hypertarget{ref-Bivand2013}{}%
Bivand, R.S., Pebesma, E., Gomez-Rubio, V., 2013. Applied spatial data
analysis with R, second edition. Springer, NY.

\leavevmode\hypertarget{ref-Casanova2010effects}{}%
Casanova, L.M., Jeon, S., Rutala, W.A., Weber, D.J., Sobsey, M.D., 2010.
Effects of air temperature and relative humidity on coronavirus survival
on surfaces. Appl. Environ. Microbiol. 76, 2712--2717.

\leavevmode\hypertarget{ref-Chan2011effects}{}%
Chan, K., Peiris, J., Lam, S., Poon, L., Yuen, K., Seto, W., 2011. The
effects of temperature and relative humidity on the viability of the
sars coronavirus. Advances in virology 2011.

\leavevmode\hypertarget{ref-Coelho2020exponential}{}%
Coelho, M.T.P., Rodrigues, J.F.M., Medina, A.M., Scalco, P., Terribile,
L.C., Vilela, B., Diniz-Filho, J.A.F., Dobrovolski, R., 2020.
Exponential phase of covid19 expansion is not driven by climate at
global scale. medRxiv.

\leavevmode\hypertarget{ref-Fernandes2020economic}{}%
Fernandes, N., 2020. Economic effects of coronavirus outbreak (covid-19)
on the world economy. Available at SSRN 3557504.

\leavevmode\hypertarget{ref-Gong2020balance}{}%
Gong, B., Zhang, S., Yuan, L., Chen, K.Z., 2020. A balance act:
Minimizing economic loss while controlling novel coronavirus pneumonia.
Journal of Chinese Governance 1--20.

\leavevmode\hypertarget{ref-Grolemund2011dates}{}%
Grolemund, G., Wickham, H., 2011. Dates and times made easy with
lubridate. Journal of Statistical Software 40, 1--25.

\leavevmode\hypertarget{ref-Harbert2020spatial}{}%
Harbert, R.S., Cunningham, S.W., Tessler, M., 2020. Spatial modeling
cannot currently differentiate sars-cov-2 coronavirus and human
distributions on the basis of climate in the united states. medRxiv.

\leavevmode\hypertarget{ref-Jaakkola2014decline}{}%
Jaakkola, K., Saukkoriipi, A., Jokelainen, J., Juvonen, R., Kauppila,
J., Vainio, O., Ziegler, T., Rönkkö, E., Jaakkola, J.J., Ikäheimo, T.M.,
2014. Decline in temperature and humidity increases the occurrence of
influenza in cold climate. Environmental Health 13, 22.

\leavevmode\hypertarget{ref-Kissler2020projecting}{}%
Kissler, S.M., Tedijanto, C., Goldstein, E., Grad, Y.H., Lipsitch, M.,
2020. Projecting the transmission dynamics of sars-cov-2 through the
postpandemic period. Science eabb5793.
doi:\href{https://doi.org/10.1126/science.abb5793}{10.1126/science.abb5793}

\leavevmode\hypertarget{ref-Kudo2019low}{}%
Kudo, E., Song, E., Yockey, L.J., Rakib, T., Wong, P.W., Homer, R.J.,
Iwasaki, A., 2019. Low ambient humidity impairs barrier function and
innate resistance against influenza infection. Proceedings of the
National Academy of Sciences 116, 10905--10910.

\leavevmode\hypertarget{ref-Lancastle2020impact}{}%
Lancastle, N.M., 2020. Is the impact of social distancing on coronavirus
growth rates effective across different settings? A non-parametric and
local regression approach to test and compare the growth rate. medRxiv.

\leavevmode\hypertarget{ref-Lopez2014spatial}{}%
López, F.A., Mur, J., Angulo, A., 2014. Spatial model selection
strategies in a sur framework. The case of regional productivity in eu.
The Annals of Regional Science 53, 197--220.

\leavevmode\hypertarget{ref-Luo2020how}{}%
Luo, S., Tsang, K.P., 2020. How much of china and world gdp has the
coronavirus reduced? Available at SSRN 3543760.

\leavevmode\hypertarget{ref-Makinen2009cold}{}%
Mäkainen, T.M., Juvonen, R., Jokelainen, J., Harju, T.H., Peitso, A.,
Bloigu, A., Silvennoinen-Kassinen, S., Leinonen, M., Hassi, J., 2009.
Cold temperature and low humidity are associated with increased
occurrence of respiratory tract infections. Respiratory medicine 103,
456--462.

\leavevmode\hypertarget{ref-Meijers2008strategic}{}%
Meijers, E., Hoekstra, J., Aguado, R., 2008. Strategic planning for city
networks: The emergence of a basque global city? International Planning
Studies 13, 239--259.
doi:\href{https://doi.org/10.1080/13563470802521440}{10.1080/13563470802521440}

\leavevmode\hypertarget{ref-Millo2017robust}{}%
Millo, G., 2017. Robust standard error estimators for panel models: A
unifying approach. Journal of Statistical Software 82, 1--27.
doi:\href{https://doi.org/10.18637/jss.v082.i03}{10.18637/jss.v082.i03}

\leavevmode\hypertarget{ref-Minguez2019}{}%
Mínguez, R., López, F., Mur, J., 2019. ML versus iv estimates of spatial
sur models: Evidence from the case of airbnb in madrid urban area. The
Annals of Regional Science 1--35.

\leavevmode\hypertarget{ref-National2020rapid}{}%
National Academies of Sciences, Engineering and Medicine, 2020. Rapid
expert consultation on sars-cov-2 survival in relation to temperature
and humidity and potential for seasonality for the covid-19 pandemic
(april 7, 2020). The National Academies Press, Washington, DC.
doi:\href{https://doi.org/doi:10.17226/25771}{doi:10.17226/25771}

\leavevmode\hypertarget{ref-Pebesma2018}{}%
Pebesma, E., 2018. Simple Features for R: Standardized Support for
Spatial Vector Data. The R Journal 10, 439--446.
doi:\href{https://doi.org/10.32614/RJ-2018-009}{10.32614/RJ-2018-009}

\leavevmode\hypertarget{ref-Pebesma2016}{}%
Pebesma, E., Mailund, T., Hiebert, J., 2016. Measurement units in R. R
Journal 8, 486--494.
doi:\href{https://doi.org/10.32614/RJ-2016-061}{10.32614/RJ-2016-061}

\leavevmode\hypertarget{ref-Wickham2019}{}%
Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L.D., François,
R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen,
T.L., Miller, E., Bache, S.M., Müller, K., Ooms, J., Robinson, D.,
Seidel, D.P., Spinu, V., Takahashi, K., Vaughan, D., Wilke, C., Woo, K.,
Yutani, H., 2019. Welcome to the tidyverse. Journal of Open Source
Software 4, 1686.
doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}

\leavevmode\hypertarget{ref-Xie2014}{}%
Xie, Y., 2014. Knitr: A comprehensive tool for reproducible research in
R, in: Stodden, V., Leisch, F., Peng, R.D. (Eds.), Implementing
Reproducible Computational Research. Chapman; Hall/CRC.

\leavevmode\hypertarget{ref-Xie2015}{}%
Xie, Y., 2015. Dynamic documents with R and knitr, 2nd ed. Chapman;
Hall/CRC, Boca Raton, Florida.

\leavevmode\hypertarget{ref-Zhu2019}{}%
Zhu, H., 2019. KableExtra: Construct complex table with 'kable' and pipe
syntax.

\end{document}

### 2020-04-18 README.Rmd
Migrated files from Fernando's repo

--- 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 ---
-title: A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
-author:
-  - name: Antonio Paez
-    email: paezha@mcmaster.ca
-    affiliation: McMaster University
-    footnote: 1
-  - name: Fernando A. Lopez
-    email: fernando.lopez@upct.es
-    affiliation: Universidad Politecnica de Cartagena
-  - name: Tatiane Menezes
-    email: tatiane.menezes@ufpe.br
-    affiliation: Departamento de Economia
-  - name: Renata Cavalcanti
-    email: renata.vcsantos@gmail.com
-    affiliation: Nucleo de Pesquisa
-  - name: Maira Galdino da Rocha Pitta
-    email: mgrpitta@ufpe.br
-    affiliation: Nucleo de Pesquisa
-address:
-  - code: McMaster University
-    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
-  - code: Universidad Politecnica de Cartagena
-    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
-  - code: Departamento de Economia
-    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
-  - code: Nucleo de Pesquisa
-    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
-footnote:
-  - code: 1
-    text: "Corresponding Author"
-abstract: |
+output: github_document
+always_allow_html: true
+bibliography: References.bib
+---
+
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+# A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
+
+Antonio Paez (McMaster University)  
+Fernando A. Lopez (Universidad Politecnica de Cartagena)  
+Tatiane Menezes (Universidade Federal de Pernambuco)  
+Renata Cavalcanti (Universidade Federal de Pernambuco)  
+Maira Galdino da Rocha Pitta (Universidade Federal de Pernambuco)
+
+## Abstract
+
 Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress.
@@
 
-journal: "Geographical Analysis"
-date: "`r Sys.Date()`"
-bibliography: References.bib
-#linenumbers: true
-#numbersections: true
-csl: elsevier-harvard.csl
-output: rticles::elsevier_article
----
+## Keywords
+
+SARS-CoV2  
+COVID-19  
+Temperature  
+Humidity  
+Spatial SUR  
+Contagion  
+Spain
 
 ```{r load-packages, include=FALSE}
-# The data package needs to be installed from github repository:
-#devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
+# The data package needs to be installed from github repository: remember to include the instructions to install
 # Also, ask Fernando if the version of `spatsur` needed for the analysis is available from CRAN or needs special distribution
@@
 
-Introduction
-==========================
+## Introduction
 
@@
 
-Context, Data, and Methods
-============
+## Acknowledgments
 
-## Covid-19 in Spain
+The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
 
-When was the first case reported? When did the country close international arrivals? When did initial measures of social distancing were implemented? The lockdown?
-
-## Selection of Variables
-
-Explain the rationale for selecting the variables.
-
-For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world.
-Controlling for socioeconomic characteristics the objective of this paper is observe the effect of clime on COVID-19  proliferation.
-
-Some questions are important to point out.
-Is not evident present a diary association between the temprature (resp.
-humidity) with declarate case.
-There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate
-
-## Sources of Data
-
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases tested} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units).
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project.
-This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments.
-In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity.
-The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province.
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provencance of the data.
-
-```{r descriptive-statistics, echo=FALSE}
-junk <- data.frame(Variable = c("COVID-19 Incidence",
-                        "GDPpc", 
-                        "Older", 
-                        "Population Density",
-                        "Mean Temperature",
-                        "Humidity"),
-           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
-                    "GDP per capita in €1,000s",
-                    "Percentage of people aged 65 and older in the province",
-                    "Population density in the province in people per sq.km",
-                    "Mean temperature in province by date, in C",
-                    "Relative humidity in province by date"),
-           Min = c(min(covid19_spain$Incidence),
-                   min(provinces_spain$GDPpc),
-                   min(provinces_spain$Older),
-                   min(provinces_spain$Density),
-                   min(covid19_spain$Mean_Temp),
-                   min(covid19_spain$Humidity)),
-           Mean = c(mean(covid19_spain$Incidence),
-                   mean(provinces_spain$GDPpc),
-                   mean(provinces_spain$Older),
-                   mean(provinces_spain$Density),
-                   mean(covid19_spain$Mean_Temp),
-                   mean(covid19_spain$Humidity)),
-           Max = c(max(covid19_spain$Incidence),
-                   max(provinces_spain$GDPpc),
-                   max(provinces_spain$Older),
-                   max(provinces_spain$Density),
-                   max(covid19_spain$Mean_Temp),
-                   max(covid19_spain$Humidity)),
-           SD = c(sd(covid19_spain$Incidence),
-                   sd(provinces_spain$GDPpc),
-                   sd(provinces_spain$Older),
-                   sd(provinces_spain$Density),
-                   sd(covid19_spain$Mean_Temp),
-                   sd(covid19_spain$Humidity)),
-           Source = c("Montera34",
-                      "INE",
-                      "INE",
-                      "INE",
-                      "AEMET",
-                      "MAPA")) %>%
-  kable("latex",
-        booktabs = TRUE,
-        digits = 2,
-        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  column_spec(2, width = "15em") %>%
-  column_spec(7, width = "5em") %>%
-  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
-                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
-                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
-                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
-```
-
-## Methods: Spatial SUR
-
-Seemingly Unrelated Regressions are used in the context of...
-
-The classical SUR model without spatial effects (from here, SUR-SIM) is a stack of equations as follows:
-
-\begin{equation}
-\label{eq:sur-sim}
-\begin{bmatrix}
-y_1 \\ y_2 \\ \vdots \\ y_T
-\end{bmatrix}
-=
-\begin{bmatrix}
-X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
-\end{bmatrix}
-\
-\begin{bmatrix}
-\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
-\end{bmatrix}
-+
-\begin{bmatrix}
-\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
-\end{bmatrix}
-\end{equation}
-
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
-
-The more relevant characteristic of a SUR model is the dependence structure among the vectors of residuals, namely:
-
-\begin{equation}
-\label{eq:sur-err}
-E[\epsilon_t \epsilon'_{t'}]=\sigma_{tt'}
-\end{equation}
-
-Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$.
-This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ considering the first $r$ coefficients contanst.
-
-Equation \ref{eq:sur-sim}) can be rewriten in compact form:
-
-\begin{equation}
-y = X \beta + \epsilon
-\end{equation}
-
-\noindent where....
-
-As is the case of cross-sectional data, it is possible identify spatial autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several Lagrange Multipliers tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
-When the null hypothesis is rejected, there are alternative several specifications that have been propose to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
-In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor.
-The stack expresion for the spatial lag SUR model (SUR-SLM),
-
-\begin{equation}
-\label{eq:sur-slm}
-\bf{A}y = X \beta + \epsilon \\
-\epsilon =N(0,\Omega)
-\end{equation}
-
-\noindent where A =$I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
-
-This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Parameters of spatial dependence are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
-The correspond Wald test is available in the R package **spsur**
-
-The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
-
-Analysis and Results
-===================
-
-## Exploratory Data Analysis
-
-We begin with the exploratory analysis of the data.
-
-Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages.
-Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19.
-It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain.
-While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces.
-Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao.
-The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava.
-The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.
-
-```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE}
-# Retrieve the coordinates of Madrid
-madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()
-
-week11.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 11) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("March 13 - March 15") +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 4.9, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-week12.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 12) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds",
-                       direction = 1) +
-  ggtitle("March 16 - March 22") +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 4.9, y = madrid_xy[2] + 0.6) +
-  geom_segment(x = madrid_xy[1] - 3.4, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
-
-week13.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 13) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("March 23 - March 29") +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid",
-               x = madrid_xy[1] - 4.9, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-week14.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 14) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("March 30 - April 5") +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 4.9, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-week15.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 15) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("April 6 - April 11") +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 4.9, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, nrow = 3)
-```
-
-To visualize the distribution of temperature and humidity we aggregate the provinces by Autonomous Community.
-In Figure \ref{fig:descriptives-temperature} the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost.
-There is a relatively wide range of values both within and between provinces over the 30 day period examined.
-The top panel of the figure shows the distribution of mean temperature.
-The lowest mean temperature for a community on any given day is 2.8C, and the highest is 22.4C for a range of approximately 20 degrees.
-Likewise, there is a great deal of variability in humidity, as seen in the bottom panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
-The actual values for the provinces display somewhat more variability even.
-
-```{r descriptives-temperature, echo = FALSE, fig.height=7, fig.cap="\\label{fig:descriptives-temperature} Distribution of mean temperatures and humidities in the Autonomous Communities in Spain between March 12, 2020 and April 11, 2020.
-The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost"}
-# Summary of temperature by Autonomous Community
-temp.summary <- covid19_spain %>% 
-  group_by(CCAA, Date) %>%
-  summarize(Value = mean(Mean_Temp)) %>%
-  mutate(Variable = "Mean Temperature in Community")
-
-# Summary of humidity by Autonomous Community
-humidity.summary <- covid19_spain %>% 
-  group_by(CCAA, Date) %>%
-  summarize(Value = mean(Humidity)) %>%
-  mutate(Variable = "Mean Humidity in Community")
-
-# Bind tables
-env.summary <- rbind(temp.summary, humidity.summary) %>%
-  mutate(Variable = factor(Variable, 
-                           levels = c("Mean Temperature in Community",
-                                      "Mean Humidity in Community")))
-
-# Plot
-ggplot(data = env.summary, 
-       aes(x = CCAA,
-           y = Value)) +
-  geom_boxplot() +
-  theme_tufte() +
-  theme(axis.text.x = element_text(angle = 90)) +
-  xlab("Autonomous Community (sorted from north to south)") +
-  ylab("Value") +
-  facet_wrap(~ Variable, nrow = 2, 
-             scales = "free_y")
-```
-
-```{r control-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)"}
-gdppc.plot <- provinces_spain %>%
-  filter(CCAA != "Canarias") %>%
-  ggplot() +
-  geom_sf(aes(fill = GDPpc)) +
-  scale_fill_distiller(name = "GDP per capita", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom")
-
-# Older adults:
-older.plot <- provinces_spain %>% 
-  filter(CCAA != "Canarias") %>%
-  ggplot() +
-  geom_sf(aes(fill = as.numeric(Older))) +
-  scale_fill_distiller(name = "Percentage of older adults", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom")
-
-# Population density:
-density.plot <- provinces_spain %>% 
-  filter(CCAA != "Canarias") %>%
-  ggplot() +
-  geom_sf(aes(fill = as.numeric(Density))) +
-  scale_fill_distiller(name = "Population Density", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom")
-
-grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
-```
-
-Figure \ref{fig:daily-correlations} shows the distribution of daily correlations of the independent variables with incidence of COVID-19, after log-transforming all variables.
-It can be seen there that the correlation of GDPpc and temperature (in its three definitions) have the strongest positive and negative correlations with incidence, respectively.
-Percentage of older adults displays somewhat weaker negative correlations with incidence, as does density.
-It can be seen that the humidity variable, in its three forms, tends to be possitively correlated with incidence of COVID-19.
-
-```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)"}
-covid19_spain %>% 
-  group_by(Date) %>%
-  summarize(Older = cor(log(Older), log(Incidence)),
-            GDPpc = cor(log(GDPpc), log(Incidence)),
-            Density = cor(log(Density), log(Incidence)),
-            Humidity_lag8 = cor(Humidity_lag8, Incidence),
-                        Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
-                        Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
-            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
-            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
-            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
-  select(-Date) %>%
-  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
-  mutate(Variable = factor(Variable,
-                           levels = c("GDPpc", 
-                                      "Older", 
-                                      "Density", 
-                                      "Humidity_lag8",
-                                      "Humidity_lag11",
-                                      "Humidity_lag11w",
-                                      "Mean_Temp_lag8",
-                                      "Mean_Temp_lag11",
-                                      "Mean_Temp_lag11w"))) %>%
-  ggplot(aes(x = Variable, y = Correlation)) +
-  geom_boxplot() +
-  theme_tufte() +
-  theme(axis.text.x = element_text(angle = 90))
-```
-
-## SUR Models
-
-```{r data-preparation-for-modelling, include=FALSE}
-# Organize data for SUR modelling 
-GPanel <- plm::pdata.frame(covid19_spain %>%
-                             select(province, 
-                                    Date,
-                                    Incidence, 
-                                    Median_Age,
-                                    Male2Female,
-                                    Older,
-                                    GDPpc,
-                                    Density,
-                                    Transit,
-                                    Mean_Temp_lag8,
-                                    Humidity_lag8,
-                                    Sunshine_Hours_lag8,
-                                    Mean_Temp_lag11,
-                                    Humidity_lag11,
-                                    Sunshine_Hours_lag11,
-                                    Mean_Temp_lag11w,
-                                    Humidity_lag11w,
-                                    Sunshine_Hours_lag11w),
-                           c("province","Date"))
-
-```
-
-```{r spatial-weights, include=FALSE}
-# Create spatial weights matrix: 
-Wmat <- provinces_spain %>%
-  #drop_na() %>%
-  as("Spatial") %>%
-  poly2nb(queen = FALSE) %>%
-  nb2mat(zero.policy = TRUE)
-
-Wmat <- (Wmat > 0) * 1
-
-# Join the two provinces in Canarias
-Wmat[which(provinces_spain$province == "Palmas(Las)"), 
-     which(provinces_spain$province == "Santa Cruz de Tenerife")] <- 1
-Wmat[which(provinces_spain$province == "Santa Cruz de Tenerife"), 
-     which(provinces_spain$province == "Palmas(Las)")] <- 1
-
-# 'Paises Catalans'
-#n = 8
-Wmat[which(provinces_spain$province == "Barcelona"), 
-     which(provinces_spain$province == "Baleares")] <- 1
-Wmat[which(provinces_spain$province == "Baleares"), 
-     which(provinces_spain$province == "Barcelona")] <- 1
-Wmat[which(provinces_spain$province == "Baleares"), 
-     which(provinces_spain$province == "Castellon/Castello")] <- 1 
-Wmat[which(provinces_spain$province == "Castellon/Castello"), 
-     which(provinces_spain$province == "Baleares")] <- 1
-Wmat[which(provinces_spain$province == "Baleares"), 
-     which(provinces_spain$province == "Tarragona")] <- 1 
-Wmat[which(provinces_spain$province == "Tarragona"), 
-     which(provinces_spain$province == "Baleares")] <- 1
-miW <- Wmat/rowSums(Wmat)
-
-# Convert to listw
-listw <- mat2listw(Wmat,style = "W")
-```
-
-```{r formulas, include=FALSE}
-# Define formulas with three different lagged variables:
-formula_lag8 <- log(Incidence) ~ 
-  log(GDPpc) +
-  log(Older) +
-  log(Density) +
-  Transit +
-  log(Humidity_lag8) +
-  log(Mean_Temp_lag8)
-
-formula_lag11 <- log(Incidence) ~ 
-  log(GDPpc) +
-  log(Older) +
-  log(Density) +
-  Transit +
-  log(Humidity_lag11) +
-  log(Mean_Temp_lag11)
-
-formula_lag11w <- log(Incidence) ~ 
-  log(GDPpc) +
-  log(Older) +
-  log(Density) +
-  Transit +
-  log(Humidity_lag11w) +
-  log(Mean_Temp_lag11w)
-```
-
-```{r model-restrictions, include=FALSE}
-T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # Recall that T is the number of days, i.e., time periods, i.e., equations; add 1 to include the starting day
-k <- 7 # Number of independent variables, including the constant
-coef_rest <- 2 # Number of restrictions
-
-# nrow is number of equations (time periods) minus 1, times the number of restrictions
-# ncol is number of variables times number of equations
-R2 <- matrix(0, nrow = (T - 1) * coef_rest, ncol = k * T)
-
-for (i in 1:(T-1)){
-  R2[i, 2] <- 1
-  R2[i, (2 + i * k)] <- -1
-  R2[(i + T - 1), 3] <- 1
-  R2[(i + T - 1), (3 + i * k)] <- -1
-  # Use if more restrictions are needed
-  #R2[(i + T - 1) * 2, 4] <- 1
-  #R2[(i + T - 1) * 2, (4 + i * k)] <- -1
-}
-b2 <- matrix(0, ncol = 21*coef_rest)
-```
-
-```{r model1-8-day-moving-average, include=FALSE}
-# Model with a lagged 8-day moving average of climatic variables:
-sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
-                                 data=GPanel, 
-                                 time = GPanel$Date, 
-                                 type = "slm", 
-                                 fit_method = "3sls", 
-                                 listw=  listw,
-                                 R = R2,
-                                 b = b2)
-#summary(sur.slm_lag8)
-#print(paste("Pooled R^2 = ", sur.slm_lag8$R2[1]))
-```
-
-```{r model2-11-day-moving-average, include=FALSE}
-# Model with 11-day moving average of climatic variables:
-sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
-                                  data=GPanel, 
-                                  time = GPanel$Date, 
-                                  type = "slm", 
-                                  fit_method = "3sls", 
-                                  listw=  listw,
-                                 R = R2,
-                                 b = b2)
-#summary(sur.slm_lag11)
-#print(paste("Pooled R^2 = ", sur.slm_lag11$R2[1]))
-```
-
-```{r model3-11-day-weighted-moving-average, include=FALSE}
-# Model with 11-day weighted moving average of climatic variables:
-sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
-                                   data=GPanel, 
-                                   time = GPanel$Date, 
-                                   type = "slm", 
-                                   fit_method = "3sls", 
-                                   listw=  listw,
-                                 R = R2,
-                                 b = b2)
-#summary(sur.slm_lag11w)
-#print(paste("Pooled R^2 = ", sur.slm_lag11w$R2[1]))
-```
-
-The goodness of fit of the three systems of equations is shown in Figure \ref{fig:goodness-of-fit}.
-
-```{r goodness-of-fit, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
-data.frame(Model1 = sur.slm_lag8$R2, 
-           Model2 = sur.slm_lag11$R2,
-           Model3 = sur.slm_lag11w$R2) %>%
-  slice(2:n()) %>%
-  rownames_to_column(var = "Equation") %>%
-  mutate(Date = seq(ymd("2020-03-13"), 
-                    ymd("2020-04-11"), 
-                    by = "days")) %>%
-  pivot_longer(cols = starts_with("M"), names_to = "Model", values_to = "R2") %>%
-  mutate(Model = factor(Model, 
-                        levels = c("Model1", "Model2", "Model3"),
-                        labels = c("Model 1", "Model 2", "Model 3"))) %>%
-  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("Model 1" = "blue", "Model 2" = "orange", "Model 3" = "green") ) + 
-  annotate(geom = "text", 
-           label = c(paste("Model 1: pooled-R2=", 
-                          round(sur.slm_lag8$R2[1], 4)),
-                     paste("Model 2: pooled-R2=", 
-                          round(sur.slm_lag11$R2[1], 4)),
-                     paste("Model 3: pooled-R2=", 
-                          round(sur.slm_lag11w$R2[1], 4))),
-            x = as_date("2020-04-01"),
-            y = c(0.3, 0.27, 0.24),
-           family = "serif",
-           size = 3.5) +
-  theme_tufte()
-
-```
-
-Summary of best model.
-```{r summary-best-model, echo=FALSE}
-data.frame(Coefficient = sur.slm_lag11$coefficients, 
-                   pval = dnorm(abs(sur.slm_lag11$coefficients/sur.slm_lag11$rest.se))) %>%
-  rownames_to_column(var = "Variable") %>%
-  mutate(Variable = c("Intercept", 
-                      "log(GDPpc)", 
-                      "log(Older)",
-                      "log(Density)",
-                      "Transit",
-                      "log(Humidity)",
-                      "log(Temperature)",
-                      rep(c("Intercept",
-                            "log(Density)", 
-                            "Transit", 
-                            "log(Humidity)", 
-                            "log(Temperature)"), 29)),
-         Variable = factor(Variable, 
-                           levels = c("Intercept", 
-                                      "log(GDPpc)", 
-                                      "log(Older)", 
-                                      "log(Density)",
-                                      "Transit", 
-                                      "log(Humidity)",
-                                      "log(Temperature)"))) %>%
-  group_by(Variable) %>%
-  summarize(Min = min(Coefficient), 
-            Mean = mean(Coefficient), 
-            Max = max(Coefficient),
-            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
-            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
-            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
-  mutate(Note = c("Non-constrained", 
-                  "Constrained", 
-                  "Constrained", 
-                  "Non-constrained", 
-                  "Non-constrained",
-                  "Non-constrained",
-                  "Non-constrained")) %>%
-  kable("latex",
-        booktabs = TRUE,
-        digits = 3,
-        col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
-        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
-  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
-                       "Non-constrained: coefficient was allowed to vary across equations",
-                       "Constrained: coefficient as constant across equations"))
-```
-
-Discussion
-===================
-
-Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$).
-
-```{r delta-time, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Deltas = sur.slm_lag11$deltas,
-           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Deltas, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.
-
-```{r beta-intercept-time, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
-#Intercept
-n = 1
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Density)$.
-
-```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)"}
-#Density:
-n = 2 # Although this is the fourth coefficient, after intercept, log(GDPpc), and log(Older), we start at 2 here because the two constrained coefficients are removed from the matrix of coefficients 
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-transit-time} shows the temporal evolution of the coefficient for $Transit$.
-
-```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit"}
-n = 3
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the coefficient for $\log(Humidity)$.
-
-```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)"}
-n = 4
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.
-
-```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)"}
-n = 5
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Concluding Remarks
-===================
-
-More words go here.
-
-Acknowledgments {#acknowledgments .unnumbered}
-==========
-
-Add acknowledgments as appropriate in final draft.
-
-The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
-
-References {#references .unnumbered}
-==========
+## References

### 2020-04-18 covid19env/data_processing/01-Data-Preprocessing.Rmd
Migrated files from Fernando's repo

---
title: "COVID-19-Panel-Spain"
author: "TURURU"
date: "4/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, collapse=TRUE, message=FALSE}
#library(meteoland)
#library(geosphere)
library(ggthemes)
library(lubridate)
library(readr)
library(sf)
library(tidyverse)
library(units)
#library(aemet)
#library(spdep)
#library(spsur)
#library(spatialreg)
#library(systemfit)
#library(plm)
#library(splm)
```

# Los datos Spain

Algunas notas sobre los datos:

* **Los datos son provisionales**

* Las cifras oficiales de Diagnosticados/Muertes estan disponibles a nivel de Comunidad Autónoma (CCAA). 
Aunque sean oficiales hay dudas sobre ellas.
Los informes están disponibles aquí: 
https://www.mscbs.gob.es/en/profesionales/saludPublica/ccayes/alertasActual/nCov-China/situacionActual.htm

* Algunas regiones (CCAA) ofrencen datos a nivel de provincia, otras no.

* Los datos a nivel provincial, principalmente por el esfuerzo de la televisón española (RTVE).
https://www.rtve.es/noticias/20200407/mapa-del-coronavirus-espana/2004681.shtml

* Estos datos junto con otros recogidos de diversas fuentes están recogidos en GitHub en la web de datalista: https://github.com/datadista/datasets/tree/master/COVID%2019

* Los datos climatológicos (temperaturas, etc) son oficiales 
http://www.aemet.es/

* Los datos de poblacion tb son oficiales del Instituto Nacional de Estadística
http://www.ine.es/

## Lectura datos COVID-19 Spain

Principales características:

* **N=50 provincias**
* **T=22 días, desde el 14 de marzo hasta el 4 de abril**
* **Desde el día 14 Marzo hasta el 4 de Abril** EL CONFINAMIENTO SE INICIÓ EL 16 DE MARZO
* **Empezamos el 14 de Marzo porque es el primer dia con al menos 1 caso en todas las provincias**

```{r}
## Lectura de Datos COVID-19
BD <- read.table(file="COVID-19-Diagnosticados-Extendido.csv",sep = ";",header = TRUE)
BD$X <- NULL
BD$X.1 <- NULL
```

Pivot table:
```{r}
BD <- BD %>%
  rename(province = `ï..Provincia`)

covid19_spain <- BD %>%
  pivot_longer(starts_with("D"),
               names_to = "Date",
               values_to = "Cases") %>%
  mutate(Date = rep(seq(dmy("13-03-2020"), 
                        dmy("13-04-2020"), 
                        by = "days"), 
                    50))
```

## Read GDP per capita

Read data
```{r}
## Read file
GDPpc <- read_csv("PIBpc_Spain.csv", col_names = TRUE) %>%
  select(ID_INE, gdppc16)
```

## Read population and demographic data

```{r}
population.sf <- st_read("Shp/Provincias.shp") %>%
  st_drop_geometry()
```

## Join provincial data to table with COVID-19 cases

Join the tables and convert to simple features:
```{r}
covid19_spain <- covid19_spain %>%
  left_join(population.sf %>% 
              select(ID_INE, POPULATION), 
            by = "ID_INE")
```

Calculate the incidence of COVID-19 cases (cases 100,000 hab):
```{r}
covid19_spain <- covid19_spain %>%
  mutate(Incidence = Cases * 100000/POPULATION) %>%
  select(-POPULATION)
```

## Read sf with provinces

```{r}
provinces_spain <- sf::st_read("Shp/shapefiles_provincias_espana.shp") #
```

Join provincial-level data to sf:
```{r}
provinces_spain <- provinces_spain %>%
  transmute(ID_INE = cartodb_id, geometry) %>%
  left_join(covid19_spain %>% 
              filter(Date == "2020-03-14") %>%
              select(province, 
                     CCAA, 
                     ID_INE),
            by = "ID_INE") %>%
  left_join(population.sf,
            by = "ID_INE") %>%
  left_join(GDPpc, 
            by = "ID_INE")
```

## Select and rename variables

Rename dataframe:
```{r}
provinces_spain <- provinces_spain  %>%
  mutate(Density = POPULATION/units::set_units(st_area(provinces_spain), km^2)) %>% 
  select(province, 
         CCAA,
         ID_INE,
         Population = POPULATION,
         Density,
         Older = MAY_65, 
         Median_Age = EDAD_MED, 
         Male2Female = HOM_MUJ,
         GDPpc = gdppc16,
         Transit = METRO,
         Altitude = ALTITUD,
         Coast = COSTA,
         Meteo_Station = ESTACION_M,
         Area = AREA,
         geometry)
```

## Lectura datos climáticos

* Se ha usado el paquete **meteoland** para obtener las temperaturas de distintas estaciones meteorológicas (el data frame stations lista las estaciones). Para cada provincia de ha elegido una estación meteorológica que fuera representativa y que contuviera datos de temperaturas (Max/Min/Media), Pluviosidad, y horas de sol.

En general estas estaciones no recogen información sobre la humedad relativa (HR). Para poder incluir en el modelo la HR se han considerado dos opciones. La primera alternativa es la estimación en base a las temperaturas máximas y mínimas (ver nota en 'COVID-19-Datos-Climáticos-Spain.R'). la segunda alternativa es recoger datos sobre la altitud de cada provincia (datos de la capital de provincia en wikipedia). La altitud es una proxy de la humedad ya que a mayor altitud el clima es más seco y el grado de humedad menor.

## Process climate data

Load climate data (ignore the humidity data in this file; it was a calculated variable):
```{r}
load("COVID-19-Datos_Climaticos_Spain-Extendido.RData")
```

Read humidity data
```{r}
## Read file
Hum <- read_csv("COVID-19-Humedad-Spain.csv", col_names = TRUE) %>%
  rename(provincia = Provincia) %>% 
  select(provincia:D11Abr)
```

Prepare climatic factors:
```{r}
# Maximum temperature
TMax <- TMax %>% select(-ID, -provincia) %>%
  pivot_longer(starts_with("T"), 
               names_to = "Date",
               values_to = "Max_Temp") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"), 
                        by = "days"), 
                    50))

# Minimum temperature
TMin <- TMin %>% select(-ID, -provincia) %>%
  pivot_longer(starts_with("T"), 
               names_to = "Date",
               values_to = "Min_Temp") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"), 
                        by = "days"), 
                    50))

# Mean temperature
TMed <- TMed %>% select(-ID, -provincia) %>%
  pivot_longer(starts_with("T"), 
               names_to = "Date",
               values_to = "Mean_Temp") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"), 
                        by = "days"), 
                    50))

# Hours of sunshine 
Sun <- Sun %>% select(-ID, -provincia) %>%
  pivot_longer(starts_with("S"), 
               names_to = "Date",
               values_to = "Sunshine_Hours") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"), 
                        by = "days"), 
                    50))

# Precipitation 
Plu <- Plu %>% select(-ID, -provincia) %>%
  pivot_longer(c(starts_with("P")), 
               names_to = "Date",
               values_to = "Precipitation") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"),
                        by = "days"), 
                    50)) %>%
  replace_na(list(Precipitation = 0)) # Replace two missing values in Malaga on 2020-03-01 and 2020-03-02

# Humidity 
Hum <- Hum %>% select(-provincia) %>%
  pivot_longer(c(starts_with("D")), 
               names_to = "Date",
               values_to = "Humidity") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"),
                        by = "days"), 
                    50))
```

Join all climatic factors:
```{r}
climatic_vars <- TMax %>%
  left_join(TMin, by = c("ID_INE", "Date")) %>%
  left_join(TMed, by = c("ID_INE", "Date")) %>%
  left_join(Sun, by = c("ID_INE", "Date")) %>%
  left_join(Plu, by = c("ID_INE", "Date"))  %>%
  left_join(Hum, by = c("ID_INE", "Date"))
```

It is important to note that someone who is diagnosed on day "date" was likely infected between 2 and 12 days earlier. Therefore, the climatic variables need to be with respect to the likely time of contagion. For this reason, we calculate a lagged average using the variable from date-minus-12 to date-minus-5 days. These variables are called "lag8".
```{r}
# Lagged mean temperature: 8 day average
TMed_lag8 <- TMed %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Mean_Temp, 12) + 
                      lag(.x$Mean_Temp, 11)+ 
                      lag(.x$Mean_Temp, 10)+ 
                      lag(.x$Mean_Temp, 9)+ 
                      lag(.x$Mean_Temp, 8)+ 
                      lag(.x$Mean_Temp, 7)+ 
                      lag(.x$Mean_Temp, 6)+ 
                      lag(.x$Mean_Temp, 5))/8) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Mean_Temp_lag8 = value)

# Lagged sunshine hours: 8 day average
Sun_lag8 <- Sun %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Sunshine_Hours, 12) + 
                      lag(.x$Sunshine_Hours, 11)+ 
                      lag(.x$Sunshine_Hours, 10)+ 
                      lag(.x$Sunshine_Hours, 9)+ 
                      lag(.x$Sunshine_Hours, 8)+ 
                      lag(.x$Sunshine_Hours, 7)+ 
                      lag(.x$Sunshine_Hours, 6)+ 
                      lag(.x$Sunshine_Hours, 5))/8) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Sunshine_Hours_lag8 = value)

# Lagged precipitation: 8 day average
Plu_lag8 <- Plu %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Precipitation, 12) + 
                      lag(.x$Precipitation, 11)+ 
                      lag(.x$Precipitation, 10)+ 
                      lag(.x$Precipitation, 9)+ 
                      lag(.x$Precipitation, 8)+ 
                      lag(.x$Precipitation, 7)+ 
                      lag(.x$Precipitation, 6)+ 
                      lag(.x$Precipitation, 5))/8) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Precipitation_lag8 = value)

# Lagged humidity: 8 day average
Hum_lag8 <- Hum %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Humidity, 12) + 
                      lag(.x$Humidity, 11)+ 
                      lag(.x$Humidity, 10)+ 
                      lag(.x$Humidity, 9)+ 
                      lag(.x$Humidity, 8)+ 
                      lag(.x$Humidity, 7)+ 
                      lag(.x$Humidity, 6)+ 
                      lag(.x$Humidity, 5))/8) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Humidity_lag8 = value)
```

A variation of this is a lagged 11-day average, from date-minus-12 to date-minus-2 days. According to [this research](https://annals.org/aim/fullarticle/2762808/incubation-period-coronavirus-disease-2019-covid-19-from-publicly-reported) on the incubation period, this is the latency period of 95% of people who develop symptoms. Call this "lag11":
```{r}
# Lagged mean temperature: 8 day average
TMed_lag11 <- TMed %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Mean_Temp, 12) + 
                      lag(.x$Mean_Temp, 11) + 
                      lag(.x$Mean_Temp, 10) + 
                      lag(.x$Mean_Temp, 9) + 
                      lag(.x$Mean_Temp, 8) + 
                      lag(.x$Mean_Temp, 7) + 
                      lag(.x$Mean_Temp, 6) + 
                      lag(.x$Mean_Temp, 5)+ 
                      lag(.x$Mean_Temp, 4)+ 
                      lag(.x$Mean_Temp, 3)+ 
                      lag(.x$Mean_Temp, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Mean_Temp_lag11 = value)

# Lagged sunshine hours: 8 day average
Sun_lag11 <- Sun %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Sunshine_Hours, 12) + 
                      lag(.x$Sunshine_Hours, 11) + 
                      lag(.x$Sunshine_Hours, 10) + 
                      lag(.x$Sunshine_Hours, 9) + 
                      lag(.x$Sunshine_Hours, 8) + 
                      lag(.x$Sunshine_Hours, 7) + 
                      lag(.x$Sunshine_Hours, 6) + 
                      lag(.x$Sunshine_Hours, 5) + 
                      lag(.x$Sunshine_Hours, 4) + 
                      lag(.x$Sunshine_Hours, 3) + 
                      lag(.x$Sunshine_Hours, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Sunshine_Hours_lag11 = value)

# Lagged precipitation: 8 day average
Plu_lag11 <- Plu %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Precipitation, 12) + 
                      lag(.x$Precipitation, 11) + 
                      lag(.x$Precipitation, 10) + 
                      lag(.x$Precipitation, 9) + 
                      lag(.x$Precipitation, 8) + 
                      lag(.x$Precipitation, 7) + 
                      lag(.x$Precipitation, 6) + 
                      lag(.x$Precipitation, 5) + 
                      lag(.x$Precipitation, 4) + 
                      lag(.x$Precipitation, 3) + 
                      lag(.x$Precipitation, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Precipitation_lag11 = value)

# Lagged humidity: 8 day average
Hum_lag11 <- Hum %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Humidity, 12) + 
                      lag(.x$Humidity, 11) + 
                      lag(.x$Humidity, 10) + 
                      lag(.x$Humidity, 9) + 
                      lag(.x$Humidity, 8) + 
                      lag(.x$Humidity, 7) + 
                      lag(.x$Humidity, 6) + 
                      lag(.x$Humidity, 5) + 
                      lag(.x$Humidity, 4) + 
                      lag(.x$Humidity, 3) + 
                      lag(.x$Humidity, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Humidity_lag11 = value)
```

Yet another variation of this is a weighted average 11-day lag, from date-minus-12 to date-minus-2 days, with the weights given by the distribution in Lauer et al. (2020). First, calculate the weights using the parameters reported by Lauer et al.:
```{r}
# Calculate the p-values for the temporal weights
tweights <- data.frame(x = seq(2, 12, 1), p = dlnorm(seq(2, 12, 1), meanlog = 1.621, sdlog = 0.418))

# Sum, to normalize to one
ptotal <- sum(tweights$p)

# Weights are the normalized p-values
tweights <- tweights %>%
  mutate(w = p/ptotal)

tweights <- rbind(data.frame(x = 1, p = 0, w = 0), tweights)
```

Call this "lag11w":
```{r}
# Lagged mean temperature: 8 day average
TMed_lag11w <- TMed %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Mean_Temp, 12) * tweights$w[12] + 
                      lag(.x$Mean_Temp, 11) * tweights$w[11]+ 
                      lag(.x$Mean_Temp, 10) * tweights$w[10]+ 
                      lag(.x$Mean_Temp, 9) * tweights$w[9]+ 
                      lag(.x$Mean_Temp, 8) * tweights$w[8]+ 
                      lag(.x$Mean_Temp, 7) * tweights$w[7]+ 
                      lag(.x$Mean_Temp, 6) * tweights$w[6]+ 
                      lag(.x$Mean_Temp, 5) * tweights$w[5]+ 
                      lag(.x$Mean_Temp, 4) * tweights$w[4]+ 
                      lag(.x$Mean_Temp, 3) * tweights$w[3]+ 
                      lag(.x$Mean_Temp, 2) * tweights$w[2])) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Mean_Temp_lag11w = value)

# Lagged sunshine hours: 8 day average
Sun_lag11w <- Sun %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Sunshine_Hours, 12) * tweights$w[12] + 
                      lag(.x$Sunshine_Hours, 11) * tweights$w[11]+ 
                      lag(.x$Sunshine_Hours, 10) * tweights$w[10]+ 
                      lag(.x$Sunshine_Hours, 9) * tweights$w[9]+ 
                      lag(.x$Sunshine_Hours, 8) * tweights$w[8]+ 
                      lag(.x$Sunshine_Hours, 7) * tweights$w[7]+ 
                      lag(.x$Sunshine_Hours, 6) * tweights$w[6]+ 
                      lag(.x$Sunshine_Hours, 5) * tweights$w[5]+ 
                      lag(.x$Sunshine_Hours, 4) * tweights$w[4]+ 
                      lag(.x$Sunshine_Hours, 3) * tweights$w[3]+ 
                      lag(.x$Sunshine_Hours, 2) * tweights$w[2])) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Sunshine_Hours_lag11w = value)

# Lagged precipitation: 8 day average
Plu_lag11w <- Plu %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Precipitation, 12) * tweights$w[12] + 
                      lag(.x$Precipitation, 11) * tweights$w[11]+ 
                      lag(.x$Precipitation, 10) * tweights$w[10]+ 
                      lag(.x$Precipitation, 9) * tweights$w[9]+ 
                      lag(.x$Precipitation, 8) * tweights$w[8]+ 
                      lag(.x$Precipitation, 7) * tweights$w[7]+ 
                      lag(.x$Precipitation, 6) * tweights$w[6]+ 
                      lag(.x$Precipitation, 5) * tweights$w[5]+ 
                      lag(.x$Precipitation, 4) * tweights$w[4]+ 
                      lag(.x$Precipitation, 3) * tweights$w[3]+ 
                      lag(.x$Precipitation, 2) * tweights$w[2])) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Precipitation_lag11w = value)

# Lagged humidity: 8 day average
Hum_lag11w <- Hum %>% 
  group_by(ID_INE) %>%
  group_modify(~ {((lag(.x$Humidity, 12) * tweights$w[12] + 
                      lag(.x$Humidity, 11) * tweights$w[11]+ 
                      lag(.x$Humidity, 10) * tweights$w[10]+ 
                      lag(.x$Humidity, 9) * tweights$w[9]+ 
                      lag(.x$Humidity, 8) * tweights$w[8]+ 
                      lag(.x$Humidity, 7) * tweights$w[7]+ 
                      lag(.x$Humidity, 6) * tweights$w[6]+ 
                      lag(.x$Humidity, 5) * tweights$w[5]+ 
                      lag(.x$Humidity, 4) * tweights$w[4]+ 
                      lag(.x$Humidity, 3) * tweights$w[3]+ 
                      lag(.x$Humidity, 2) * tweights$w[2])) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(ID_INE, 
            Date = rep(seq(dmy("01-03-2020"),
                           dmy("11-04-2020"), 
                           by = "days"), 
                       50), 
            Humidity_lag11w = value)
```

Join lagged variables to climatic factors:
```{r}
climatic_vars <- climatic_vars %>%
  left_join(TMed_lag8, by = c("ID_INE", "Date")) %>%
  left_join(Sun_lag8, by = c("ID_INE", "Date")) %>%
  left_join(Plu_lag8, by = c("ID_INE", "Date")) %>%
  left_join(Hum_lag8, by = c("ID_INE", "Date")) %>%
  left_join(TMed_lag11, by = c("ID_INE", "Date")) %>%
  left_join(Sun_lag11, by = c("ID_INE", "Date")) %>%
  left_join(Plu_lag11, by = c("ID_INE", "Date")) %>%
  left_join(Hum_lag11, by = c("ID_INE", "Date")) %>%
  left_join(TMed_lag11w, by = c("ID_INE", "Date")) %>%
  left_join(Sun_lag11w, by = c("ID_INE", "Date")) %>%
  left_join(Plu_lag11w, by = c("ID_INE", "Date")) %>%
  left_join(Hum_lag11w, by = c("ID_INE", "Date"))
```

## Join COVID-19 cases and climatic variables

Join the dataframes:
```{r}
covid19_spain <- covid19_spain %>%
  full_join(climatic_vars, by = c("ID_INE", "Date"))
```

Summary of dataframe:
```{r}
summary(covid19_spain)
```

## Plot two control variables Ratio Male/Female; Median_Age

Ratio of male to female in province:
```{r}
ggplot(data = provinces_spain) + 
  geom_sf(aes(fill = Male2Female)) + 
  scale_fill_distiller(palette = "Blues", direction = 1) + 
  theme_tufte()
```

Median age of population in province:
```{r}
ggplot(data = provinces_spain) + 
  geom_sf(aes(fill = Median_Age)) + 
  scale_fill_distiller(palette = "Blues", direction = 1) + 
  theme_tufte()
```

### Incidence progress

Plot the progression of the incidence of COVID-19 cases by province:
```{r}
p1 <- ggplot(data = covid19_spain %>%
               filter(Date == "2020-03-19") %>% 
         left_join(provinces_spain, by = "ID_INE") %>% 
         st_as_sf()) + 
  geom_sf(aes(fill = Incidence)) + 
  scale_fill_distiller(palette = "Reds", direction = 1) + 
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") + 
  facet_wrap(~ Date, nrow = 3)

p2 <- ggplot(data = covid19_spain %>%
               filter(Date == "2020-03-28") %>% 
         left_join(provinces_spain, by = "ID_INE") %>% 
         st_as_sf()) + 
  geom_sf(aes(fill = Incidence)) + 
  scale_fill_distiller(palette = "Reds", direction = 1) + 
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") + 
  facet_wrap(~ Date, nrow = 3)

p3 <- ggplot(data = covid19_spain %>%
               filter(Date == "2020-04-03") %>% 
         left_join(provinces_spain, by = "ID_INE") %>% 
         st_as_sf()) + 
  geom_sf(aes(fill = Incidence)) + 
  scale_fill_distiller(palette = "Reds", direction = 1) + 
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") +
  facet_wrap(~ Date, nrow = 3)

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

## Save data

Filter missing records:
```{r}
covid19_spain <- covid19_spain %>%
  filter(!is.na(province)) %>% # Remove records that are missing data on the province
  filter(!is.na(Max_Temp)) # Remove records that are missing climatic variables

provinces_spain <- provinces_spain %>%
  drop_na()
```

Summary after cleaning:
```{r}
summary(covid19_spain)
```

Summary after cleaning:
```{r}
summary(provinces_spain)
```

Save:
```{r}
# Dataframe
save(covid19_spain, file = "covid19_spain.RData", compress = "xz")

# Simple features for cartography
save(provinces_spain, file = "provinces_spain.RData", compress = "xz")
```

### 2020-04-18 covid19env/R/covid19env.R

@@ -1,0 +1,104 @@
+#' covid19env: A package with datasets for analysis of environmental correlates of COVID-19.
+#'
+#' This package includes datasets used for the analysis
+#' of the climatic correlates of COVID-19 with their
+#' respective documentation.
+#'
+#' @docType package
+#' @name covid19env
+#' @author Fernando A. Lopez, Antonio Paez, School of Geography and Earth Sciences, McMaster University \email{paezha@@mcmaster.ca}
+#' @references \url{https://github.com/paezha/Reproducible-Research-Workflow}
+NULL
+
+#' COVID-19 Data for Spain.
+#'
+#' A dataset containing COVID-19 information for Spain, including
+#' climatic correlates and control variables.
+#'
+#' @format A dataframe with 1500 rows and 34 variables:
+#' \describe{
+#'   \item{province}{Names of provinces in Spain as factor}
+#'   \item{CCAA}{Names of Autonomous Communities in Spain as factor}
+#'   \item{ID_INE}{National Institute of Statistics unique identifier of the provinces}
+#'   \item{Date}{Date}
+#'   \item{Cases}{Number of COVID-19 cases reported in the province by date}
+#'   \item{Incidence}{Incidence of COVID-19 cases per 100,000 population in the province by date}
+#'   \item{Max_Temp}{Maximum temperature registered by the meteorological station representative of the province in Celsius}
+#'   \item{Min_Temp}{Minimum temperature registered by the meteorological station representative of the province in Celsius}
+#'   \item{Mean_Temp}{Mean temperature registered by the meteorological station representative of the province in Celsius}
+#'   \item{Sunshine_Hours}{Hours of sunshine registered by the meteorological station representative of the province in hours}
+#'   \item{Precipitation}{Precipitation registered by the meteorological station representative of the province}
+#'   \item{Humidity}{Calculated humidity}
+#'   \item{Mean_Temp_lag8}{8-day average of the mean temperature, calculated using the values between date-minus-12-days and date-minus-5-days in Celsius}
+#'   \item{Sunshine_Hours_lag8}{8-day average of sunshine hours, calculated using the values between date-minus-12-days and date-minus-5-days in hours}
+#'   \item{Precipitation_lag8}{8-day average of the precipitation, calculated using the values between date-minus-12-days and date-minus-5-days}
[…]

### 2020-04-18 covid19env/data_processing/COVID-19-Datos-Climaticos-Spain-v2.R

@@ -1,0 +1,153 @@
+###########################################
+##
+## Lectura de datos climáticos en AEMET
+##
+##
+###########################################
+
+setwd("~/Dropbox/COVID-19")
+
+# Lectura Paquetes
+library(meteoland)
+library(geosphere)
+library(sf)
+library(aemet)
+library(weathermetrics)
+
+load("~/Dropbox/COVID-19/COVID-19-Datos_Climaticos_Spain.RData")
+
+# API key for AEMET
+apikey <- "eyJhbGciOiJIUzI1NiJ9.eyJzdWIiOiJmOGw1aDlAZ21haWwuY29tIiwianRpIjoiNzQ4NzU1ZTgtZDBjYi00MzE3LThmMTUtNzcwZDE0ZGNiNDg0IiwiaXNzIjoiQUVNRVQiLCJpYXQiOjE1ODYxODI2MzIsInVzZXJJZCI6Ijc0ODc1NWU4LWQwY2ItNDMxNy04ZjE1LTc3MGQxNGRjYjQ4NCIsInJvbGUiOiIifQ.Ez8SgcBusQ-514k9vDtEXL29wbRi-cGIYkxnXl5AVv0"
+
+# List of dates
+mydates <- seq(as.Date("2020/3/01"), as.Date("2020/4/04"), "day")
+
+## Lee las estaciones de meteorologicas
+stations <- aemet::aemet_stations(apikey)
+
+
+# Selecciono una estación meteorológoci por provincia
+# El criterio para la seleccion es que tenga datos de temperatura y horas de sol
+# El listado de estaciones está en el fichero Provincias.shp
+
+provincias.sf <- sf::st_read("Shp/Provincias.shp") #
+
+
+Tem <- list()
+for (i in 30:34){
+  Tem[[i]] <- downloadAEMEThistorical(apikey, mydates, station_id=provincias.sf$ESTACION[i])[[1]]
+  print(i)
+  Sys.sleep(2)
+}
+
+load("~/Dropbox/COVID-19/COVID-19-Datos_Climaticos_Spain.RData")
+
+## Ordenacion de datos
+
+# Temperatura media, max, pluviosidad y horas de sol de cada provincia
+
+# Temperaturas media/max por provincia y dia
+TMed <- matrix(0,nrow = 50, ncol = 35)
+TMax <- matrix(0,nrow = 50, ncol = 35)
[…]

### 2020-04-18 covid19env/data_processing/COVID-19-Datos-Climaticos-Spain.R

@@ -1,0 +1,148 @@
+###########################################
+##
+## Lectura de datos climáticos en AEMET
+##
+##
+###########################################
+
+setwd("~/Dropbox/COVID-19")
+
+# Lectura Paquetes
+library(meteoland)
+library(geosphere)
+library(sf)
+library(aemet)
+library(weathermetrics)
+
+load("~/Dropbox/COVID-19/COVID-19-Datos_Climaticos_Spain.RData")
+
+# API key for AEMET
+apikey <- "eyJhbGciOiJIUzI1NiJ9.eyJzdWIiOiJmOGw1aDlAZ21haWwuY29tIiwianRpIjoiNzQ4NzU1ZTgtZDBjYi00MzE3LThmMTUtNzcwZDE0ZGNiNDg0IiwiaXNzIjoiQUVNRVQiLCJpYXQiOjE1ODYxODI2MzIsInVzZXJJZCI6Ijc0ODc1NWU4LWQwY2ItNDMxNy04ZjE1LTc3MGQxNGRjYjQ4NCIsInJvbGUiOiIifQ.Ez8SgcBusQ-514k9vDtEXL29wbRi-cGIYkxnXl5AVv0"
+
+# List of dates
+mydates <- seq(as.Date("2020/3/01"), as.Date("2020/4/11"), "day")
+
+## Lee las estaciones de meteorologicas
+stations <- aemet::aemet_stations(apikey)
+
+
+# Selecciono una estación meteorológoci por provincia
+# El criterio para la seleccion es que tenga datos de temperatura y horas de sol
+# El listado de estaciones está en el fichero Provincias.shp
+
+provincias.sf <- sf::st_read("Shp/Provincias.shp") #
+
+
+Tem <- list()
+for (i in 44:50){
+  Tem[[i]] <- downloadAEMEThistorical(apikey, mydates, station_id=provincias.sf$ESTACION[i])[[1]]
+  print(i)
+}
+
+load("~/Dropbox/COVID-19/COVID-19-Datos_Climaticos_Spain.RData")
+
+## Ordenacion de datos
+
+# Temperatura media, max, pluviosidad y horas de sol de cada provincia
+
+# Temperaturas media/max por provincia y dia
+TMed <- matrix(0,nrow = 50, ncol = 42)
+TMax <- matrix(0,nrow = 50, ncol = 42)
+TMin <- matrix(0,nrow = 50, ncol = 42)
[…]

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Updated README.md

@@
 # The data package needs to be installed from github repository:
-#devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
-# Also, ask Fernando if the version of `spatsur` needed for the analysis is available from CRAN or needs special distribution
-#library(aemet) # Needs install from github
+# devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
+
+# This version of spsur may be needed:
+# Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install
+
 library(covid19env)
@@
 ```{r descriptive-statistics, echo=FALSE}
-junk <- data.frame(Variable = c("COVID-19 Incidence",
+data.frame(Variable = c("COVID-19 Incidence",
                         "GDPpc", 

### 2020-04-18 README.Rmd
Updated README.md

--- 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 
-journal: "Geographical Analysis"
-date: "`r Sys.Date()`"
-bibliography: References.bib
-#linenumbers: true
-#numbersections: true
-csl: elsevier-harvard.csl
-output: rticles::elsevier_article
----
+## Keywords
+
+SARS-CoV2  
+COVID-19  
+Temperature  
+Humidity  
+Spatial SUR  
+Contagion  
+Spain
 
@@
 
-Context, Data, and Methods
-============
+## Context, Data, and Methods
 
-## Covid-19 in Spain
+### Covid-19 in Spain
 
@@
 
-## Selection of Variables
+### Selection of Variables
 
@@
 
-## Sources of Data
+### Sources of Data
 
@@
                       "MAPA")) %>%
-  kable("latex",
+  kable("html",
         booktabs = TRUE,
@@
         caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  column_spec(2, width = "15em") %>%
-  column_spec(7, width = "5em") %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
   footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
@@
 
-## Methods: Spatial SUR
+### Methods: Spatial SUR
 
@@
 
-## SUR Models
+### SUR Models
 
@@
                   "Non-constrained")) %>%
-  kable("latex",
+  kable("html",
         booktabs = TRUE,
@@
         caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
   add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
@@
 
-Discussion
-===================
+## Discussion
 
@@
 
-Concluding Remarks
-===================
+## Concluding Remarks
 
@@
 
-Acknowledgments {#acknowledgments .unnumbered}
-==========
+## Acknowledgments
 
-Add acknowledgments as appropriate in final draft.
+The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
 
-The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
-
-References {#references .unnumbered}
-==========
+## References

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
cronologia

@@
 
+For a better interpretation of the results of our analysis is necessary to know the chronology of the epidemic outbreak in Spain. This subsection presents a brief time line of COVID-19.
+
+The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the 
+Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia).
+The epidemic was growing slowly at the beginning but in March 11th the World Health Organization (WHO) declared officially the pandemic.
+This declaration was a break point and on March 13th the Spanish National Government declare the "Emergency Status".
+At this date, Spain have been declared 4209 cases of COVID-19 (mostly of cases, 1990, in Madrid).
+Only essential services (e.g.
+food, health) and some economic subsector of industry and construction could to maintain activity.
+On March 17th Spain closes its lands borders and only allows entry to nationals and residents.
+From March 30th to April 12th (April 10th-12th Easter) lockdown of the Spanish people was hardened and only essential activities were allowed.
+
 When was the first case reported? When did the country close international arrivals? When did initial measures of social distancing were implemented? The lockdown?
@@
 
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases tested} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units).
+Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases positively tested.
+Exclude sintomatic cases diagnostic by doctor with-out Reacción en Cadena de la Polimersa (PCR) test} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units).
 The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project.
@@
 
-Seemingly Unrelated Regressions are used in the context of...
+The spatial Seemingly Unrelated Regressions (SUR) methodology has been a popular method to estimate are used in the context of...
+
+The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods.
+The basis of this approach is well-known since the initial works of @zellner1962 and now is a popular methodology included in several Econometrics textbook (e.g.
+@greene2003).
+To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p.
+141).
+From this milestone, a huge researches used this methodology in spatial framework.
+@rey1999,@fingleton2007multi, @lauridsen2010, @legallo2006, @lopez2017spatial are well-known examples in this line.
 
@@
 The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
+
+Another alternative methodologies could be use.
+By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g.
+Elhorst 2014, Cap.
+4), but those models are more dont take account correlation between errors.
+Therefore, a spatial SUR approach is more reasonable for our purpose.
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
cronologia

@@
   capita, population density, percentage of older adults in the
-  population, and presence of mass transit systems. Our results indicate
-  that incidence of the disease is lower at higher temperatures. The
-  evidence with respect to humidity is more mixed, with coefficients for
-  this variable that are significant in only some equations. Our control
-  variables also yield interesting insights, as population density and
-  percentage of older adults display negative associations with incidence
-  of COVID-19, whereas the presence of mass transit systems in the
-  province is associated with a greater incidence of the disease.
+  population, and presence of mass transit systems. An interesting aspect
+  of the spatial SUR approach is that it models incidence as a contagion
+  process. Our results indicate that incidence of the disease is lower at
+  higher temperatures. The evidence with respect to humidity is more
+  mixed, with coefficients for this variable that are significant in only
+  some equations. Our control variables also yield interesting insights,
+  as population density and percentage of older adults display negative
+  associations with incidence of COVID-19, whereas the presence of mass
+  transit systems in the province is associated with a greater incidence
+  of the disease.\\
   \end{abstract}
@@
 and more humid seasons, yet none are without major
-limitations\ldots Additional studies as the SARS-CoV-2 pandemic unfolds
+limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic unfolds
 could shed more light on the effects of climate on transmission.''
@@
 variables. These data are analyzed using a spatial Seemingly Unrelated
-Regressions (SUR) approach, which allows us to account for spatial
-autocorrelation.
+Regressions (SUR) approach, which allows us to model incidence of
+COVID-19 as a spatial contagion process.
 
 The results provide evidence of the effect of temperature on the
-incidence of COVID-19, as \textbf{NOTE ABOUT THE MAGNITUDE OF THE
-EFFECT}. The evidence concerning humidity is more mixed: while the
-direction of the effect is negative, as anticipated, the coefficients
-for this variable are only significant in some of equations in the
-system. Our control variables also provide some intriguing insights. The
-results of this analysis provide support to the hypothesis of
-seasonality of the novel SARS-CoV2, and should be of interest to public
-health officials and policy makers wrestling with the question of the
-trajectory of the pandemic.
+incidence of COVID-19. The evidence concerning humidity is more mixed:
+while the direction of the effect is negative, as anticipated, the
+coefficients for this variable are only significant in some of equations
+in the system. Our control variables also provide some intriguing
+insights. The results of this analysis provide support to the hypothesis
+of seasonality of the novel SARS-CoV2, and should be of interest to
+public health officials and policy makers wrestling with the question of
+the trajectory of the pandemic.
 
@@
 
-\begin{quote}
-\begin{quote}
-Include link to repository
-\end{quote}
-\end{quote}
+https://github.com/paezha/covid19-environmental-correlates
 
@@
 \subsection{Covid-19 in Spain}\label{covid-19-in-spain}}
+
+For a better interpretation of the results of our analysis is necessary
+to know the chronology of the epidemic outbreak in Spain. This
+subsection presents a brief time line of COVID-19.
+
+The first reported case of COVID-19 in Spain was on January 31th, were a
+German tourist in the Canary Islands was positively tested, but the
+first not imported case was diagnostic on February 27th in Sevilla
+province (Andalusia). The epidemic was growing slowly at the beginning
+but in March 11th the World Health Organization (WHO) declared
+officially the pandemic. This declaration was a break point and on March
+13th the Spanish National Government declare the ``Emergency Status''.
+At this date, Spain have been declared 4209 cases of COVID-19 (mostly of
+cases, 1990, in Madrid). Only essential services (e.g.~food, health) and
+some economic subsector of industry and construction could to maintain
+activity. On March 17th Spain closes its lands borders and only allows
+entry to nationals and residents. From March 30th to April 12th (April
+10th-12th Easter) lockdown of the Spanish people was hardened and only
+essential activities were allowed.
 
@@
 Our dataset includes information about the daily number of cases of
-COVID-19 reported in Spain \footnote{cases tested} at the provincial
-level (NUTIII in eurostat terminology) from March 13th to April 11th (30
-days and 50 spatial units). The Spanish National Government publishes
-periodic updates at the regional level (NUTII) and the information is
-also released at the provincial level as part of a collaborative
-project. This information is compiled from several sources, mainly the
-official web pages of the Spanish regional goverments. In addition, we
-consider two sets of explanatory variables. The first one, and the focus
-of this research, is set of two environmental variables, namely
-temperature and humidity. The second set provides some relevant controls
-for multivariate analysis, and refers to economic and demographic
-attributes of the province. Table \ref{tab:descriptive-statistics} shows
-the descriptive statistics and the provencance of the data.
+COVID-19 reported in Spain
+\footnote{cases positively tested. Exclude sintomatic cases diagnostic by doctor with-out Reacción en Cadena de la Polimersa (PCR) test}
+at the provincial level (NUTIII in eurostat terminology) from March 13th
+to April 11th (30 days and 50 spatial units). The Spanish National
+Government publishes periodic updates at the regional level (NUTII) and
+the information is also released at the provincial level as part of a
+collaborative project. This information is compiled from several
+sources, mainly the official web pages of the Spanish regional
+goverments. In addition, we consider two sets of explanatory variables.
+The first one, and the focus of this research, is set of two
+environmental variables, namely temperature and humidity. The second set
+provides some relevant controls for multivariate analysis, and refers to
+economic and demographic attributes of the province. Table
+\ref{tab:descriptive-statistics} shows the descriptive statistics and
+the provencance of the data.
+
+\begin{table}
+
+\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
+\toprule
+Variable & Note & Min & Mean & Max & SD & Source\\
+\midrule
+\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
+GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
+\rowcolor{gray!6}  Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
+Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
+\rowcolor{gray!6}  Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
+\addlinespace
+Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
+\bottomrule
+\multicolumn{7}{l}{\textit{Note: }}\\
+\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
+\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
+\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
+\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
+\end{tabular}}
+\end{table}
 
@@
 
-Seemingly Unrelated Regressions are used in the context of\ldots{}
+The spatial Seemingly Unrelated Regressions (SUR) methodology has been a
+popular method to estimate are used in the context of\ldots{}
+
+The Seemingly Unrelated Regression equations model (SUR from now on) is
+a multivariate econometric formulation used in different fields when the
+data set is organized as cross-sections for different time periods. The
+basis of this approach is well-known since the initial works of
+({\textbf{???}}) and now is a popular methodology included in several
+Econometrics textbook (e.g.~({\textbf{???}})). To our knowledge, Anselin
+(1988) was the first text that introduced the term spatial SUR in
+reference to a model made of `an equation for each time period, which is
+estimated for a cross section of spatial units' (p.~141). From this
+milestone, a huge researches used this methodology in spatial framework.
+({\textbf{???}}),({\textbf{???}}), ({\textbf{???}}), ({\textbf{???}}),
+({\textbf{???}}) are well-known examples in this line.
 
@@
 
-\noindent where\ldots.
+\noindent where\ldots{}.
 
@@
 al. (2014)) or instrumental variables (Mínguez et al. (2019)).
+
+Another alternative methodologies could be use. By example, a dynamic
+spatial panel methodology with fixed spatial an temporal effects
+(e.g.~Elhorst 2014, Cap. 4), but those models are more dont take account
+correlation between errors. Therefore, a spatial SUR approach is more
+reasonable for our purpose.
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Experimenting with equations in README.md

@@
 
-The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the 
-Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia).
+The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia).
 The epidemic was growing slowly at the beginning but in March 11th the World Health Organization (WHO) declared officially the pandemic.
@@
 From March 30th to April 12th (April 10th-12th Easter) lockdown of the Spanish people was hardened and only essential activities were allowed.
-
-When was the first case reported? When did the country close international arrivals? When did initial measures of social distancing were implemented? The lockdown?
 
@@
 
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases positively tested.
-Exclude sintomatic cases diagnostic by doctor with-out Reacción en Cadena de la Polimersa (PCR) test} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units).
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project.
+Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
+This gives a temporal coverage of 30 days and spatial coverage of 50 units.
+For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
+The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**].
 This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments.
 In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity.
-The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province.
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provencance of the data.
+The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food).
+The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute).
+Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.
 
@@
 
-The spatial Seemingly Unrelated Regressions (SUR) methodology has been a popular method to estimate are used in the context of...
-
 The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods.
-The basis of this approach is well-known since the initial works of @zellner1962 and now is a popular methodology included in several Econometrics textbook (e.g.
-@greene2003).
+The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient] and now is a popular methodology included in several Econometrics textbook [e.g., @Greene2003econometric].
 To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p.
@@
 From this milestone, a huge researches used this methodology in spatial framework.
-@rey1999,@fingleton2007multi, @lauridsen2010, @legallo2006, @lopez2017spatial are well-known examples in this line.
+Rey and Montouri [-@Rey1999us], Fingleton [-@Fingleton2007multi], Lauridsen et al.
+[-@Lauridsen2010spatiotemporal], Le Gallo and Dall’Erba [-@Legallo2006evaluating], and Lopez et al.
+[-@Lopez2017spatial] are well-known examples in this line.
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Experimenting with equations in README.md

@@
   transit systems in the province is associated with a greater incidence
-  of the disease.\\
+  of the disease.
   \end{abstract}
@@
 and more humid seasons, yet none are without major
-limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic unfolds
+limitations\ldots Additional studies as the SARS-CoV-2 pandemic unfolds
 could shed more light on the effects of climate on transmission.''
@@
 
-When was the first case reported? When did the country close
-international arrivals? When did initial measures of social distancing
-were implemented? The lockdown?
-
 \hypertarget{selection-of-variables}{%
@@
 Our dataset includes information about the daily number of cases of
-COVID-19 reported in Spain
-\footnote{cases positively tested. Exclude sintomatic cases diagnostic by doctor with-out Reacción en Cadena de la Polimersa (PCR) test}
-at the provincial level (NUTIII in eurostat terminology) from March 13th
-to April 11th (30 days and 50 spatial units). The Spanish National
-Government publishes periodic updates at the regional level (NUTII) and
-the information is also released at the provincial level as part of a
-collaborative project. This information is compiled from several
-sources, mainly the official web pages of the Spanish regional
-goverments. In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two
-environmental variables, namely temperature and humidity. The second set
-provides some relevant controls for multivariate analysis, and refers to
-economic and demographic attributes of the province. Table
-\ref{tab:descriptive-statistics} shows the descriptive statistics and
-the provencance of the data.
+COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
+terminology) for the period between March 13th and April 11th,
+inclusive. This gives a temporal coverage of 30 days and spatial
+coverage of 50 units. For our purposes, we consider positive cases
+reported, but excluding symptomatic cases diagnosed by a doctor without
+a Polymerase Chain Reaction (PCR) test. The Spanish National Government
+publishes periodic updates at the regional level (NUTII) and the
+information is also released at the provincial level as part of a
+collaborative project {[}\textbf{by whom?}{]}. This information is
+compiled from several sources, mainly the official web pages of the
+Spanish regional goverments. In addition, we consider two sets of
+explanatory variables. The first one, and the focus of this research, is
+set of two environmental variables, namely temperature and humidity,
+which are collected from official sources (i.g., AEMET, the state
+meteorology agency, and MAPA, the ministry of agriculture, fisheries,
+and food). The second set provides some relevant controls for
+multivariate analysis, and refers to economic and demographic attributes
+of the province (also collected from official sources, i.e., INE, the
+national statistics institute). Table \ref{tab:descriptive-statistics}
+shows the descriptive statistics and the provenance of the data.
 
@@
 
-The spatial Seemingly Unrelated Regressions (SUR) methodology has been a
-popular method to estimate are used in the context of\ldots{}
-
 The Seemingly Unrelated Regression equations model (SUR from now on) is
@@
 data set is organized as cross-sections for different time periods. The
-basis of this approach is well-known since the initial works of
-({\textbf{???}}) and now is a popular methodology included in several
-Econometrics textbook (e.g.~({\textbf{???}})). To our knowledge, Anselin
-(1988) was the first text that introduced the term spatial SUR in
-reference to a model made of `an equation for each time period, which is
-estimated for a cross section of spatial units' (p.~141). From this
-milestone, a huge researches used this methodology in spatial framework.
-({\textbf{???}}),({\textbf{???}}), ({\textbf{???}}), ({\textbf{???}}),
-({\textbf{???}}) are well-known examples in this line.
+basis of this approach is well-known since the initial works of Zellner
+(1962) and now is a popular methodology included in several Econometrics
+textbook (e.g., Greene, 2003). To our knowledge, Anselin (1988) was the
+first text that introduced the term spatial SUR in reference to a model
+made of `an equation for each time period, which is estimated for a
+cross section of spatial units' (p.~141). From this milestone, a huge
+researches used this methodology in spatial framework. Rey and Montouri
+(1999), Fingleton ({\textbf{???}}), Lauridsen et al.~(2010), Le Gallo
+and Dall'Erba (2006), and Lopez et al.~(2017) are well-known examples in
+this line.
 
@@
 
-\noindent where\ldots{}.
+\noindent where\ldots.
 
@@
 
+\leavevmode\hypertarget{ref-Greene2003econometric}{}%
+Greene, W.H., 2003. Econometric analysis. Pearson Education India.
+
 \leavevmode\hypertarget{ref-Grolemund2011dates}{}%
@@
 local regression approach to test and compare the growth rate. medRxiv.
+
+\leavevmode\hypertarget{ref-Lauridsen2010spatiotemporal}{}%
+Lauridsen, J., Bech, M., López, F., Maté, M., 2010. A spatiotemporal
+analysis of public pharmaceutical expenditure. The Annals of Regional
+Science 44, 299--314.
+
+\leavevmode\hypertarget{ref-Legallo2006evaluating}{}%
+Le Gallo, J., Dall'Erba, S., 2006. Evaluating the temporal and spatial
+heterogeneity of the european convergence process, 1980--1999. Journal
+of Regional Science 46, 269--288.
+
+\leavevmode\hypertarget{ref-Lopez2017spatial}{}%
+López, F.A., Martínez-Ortiz, P.J., Cegarra-Navarro, J.-G., 2017. Spatial
+spillovers in public expenditure on a municipal level in spain. The
+Annals of Regional Science 58, 39--65.
 
@@
 
+\leavevmode\hypertarget{ref-Rey1999us}{}%
+Rey, S.J., Montouri, B.D., 1999. US regional income convergence: A
+spatial econometric perspective. Regional studies 33, 143--156.
+
 \leavevmode\hypertarget{ref-Wickham2019}{}%
@@
 
+\leavevmode\hypertarget{ref-Zellner1962efficient}{}%
+Zellner, A., 1962. An efficient method of estimating seemingly unrelated
+regressions and tests for aggregation bias. Journal of the American
+statistical Association 57, 348--368.
+
 \leavevmode\hypertarget{ref-Zhu2019}{}%

### 2020-04-18 README.Rmd
Experimenting with equations in README.md

@@
 
-\begin{equation}
-\label{eq:sur-sim}
-\begin{bmatrix}
-y_1 \\ y_2 \\ \vdots \\ y_T
-\end{bmatrix}
-=
-\begin{bmatrix}
-X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
-\end{bmatrix}
-\
-\begin{bmatrix}
-\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
-\end{bmatrix}
-+
-\begin{bmatrix}
-\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
-\end{bmatrix}
-\end{equation}
+<a href="https://www.codecogs.com/eqnedit.php?latex=\begin{equation}&space;\label{eq:sur-sim}&space;\begin{bmatrix}&space;y_1&space;\\&space;y_2&space;\\&space;\vdots&space;\\&space;y_T&space;\end{bmatrix}&space;=&space;\begin{bmatrix}&space;X_1&space;&&space;0&space;&&space;\cdots&space;&&space;0&space;\\&space;0&space;&&space;X_2&space;&&space;\cdots&space;&&space;0&space;\\&space;\vdots&space;&&space;\vdots&space;&&space;\ddots&space;&&space;\vdots&space;\\&space;0&space;&&space;0&space;&&space;\cdots&space;&&space;X_T&space;\end{bmatrix}&space;\&space;\begin{bmatrix}&space;\beta_1&space;\\&space;\beta_1&space;\\&space;\vdots&space;\\&space;\beta_T&space;\end{bmatrix}&space;&plus;&space;\begin{bmatrix}&space;\epsilon_1&space;\\&space;\epsilon_2&space;\\&space;\vdots&space;\\&space;\epsilon_T&space;\end{bmatrix}&space;\end{equation}" target="_blank"><img src="https://latex.codecogs.com/gif.latex?\begin{equation}&space;\label{eq:sur-sim}&space;\begin{bmatrix}&space;y_1&space;\\&space;y_2&space;\\&space;\vdots&space;\\&space;y_T&space;\end{bmatrix}&space;=&space;\begin{bmatrix}&space;X_1&space;&&space;0&space;&&space;\cdots&space;&&space;0&space;\\&space;0&space;&&space;X_2&space;&&space;\cdots&space;&&space;0&space;\\&space;\vdots&space;&&space;\vdots&space;&&space;\ddots&space;&&space;\vdots&space;\\&space;0&space;&&space;0&space;&&space;\cdots&space;&&space;X_T&space;\end{bmatrix}&space;\&space;\begin{bmatrix}&space;\beta_1&space;\\&space;\beta_1&space;\\&space;\vdots&space;\\&space;\beta_T&space;\end{bmatrix}&space;&plus;&space;\begin{bmatrix}&space;\epsilon_1&space;\\&space;\epsilon_2&space;\\&space;\vdots&space;\\&space;\epsilon_T&space;\end{bmatrix}&space;\end{equation}" title="\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}" /></a>
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Successfully embeded math in README.md

@@
   kable("latex",
+        #"html",
         booktabs = TRUE,
@@
   kable_styling(latex_options = c("striped", "scale_down")) %>%
+  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
   column_spec(2, width = "15em") %>%
@@
 
+<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+<!--
+![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
+-->
+
 \noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
@@
 \end{equation}
+
+<!--
+![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
+-->
 
@@
 \end{equation}
+
+<!--
+![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
+-->
 
@@
 \end{equation}
+
+<!--
+![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
+-->
 
@@
   kable_styling(latex_options = c("striped", "scale_down")) %>%
+  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
   add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Edits to Context and Methods

@@
 
-For a better interpretation of the results of our analysis is necessary to know the chronology of the epidemic outbreak in Spain. This subsection presents a brief time line of COVID-19.
-
-The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia).
-The epidemic was growing slowly at the beginning but in March 11th the World Health Organization (WHO) declared officially the pandemic.
-This declaration was a break point and on March 13th the Spanish National Government declare the "Emergency Status".
-At this date, Spain have been declared 4209 cases of COVID-19 (mostly of cases, 1990, in Madrid).
-Only essential services (e.g.
-food, health) and some economic subsector of industry and construction could to maintain activity.
-On March 17th Spain closes its lands borders and only allows entry to nationals and residents.
-From March 30th to April 12th (April 10th-12th Easter) lockdown of the Spanish people was hardened and only essential activities were allowed.
+The first reported case of COVID-19 in Spain was on January 31th, when a German tourist in the Canary Islands tested positive for the virus.
+However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
+In a short period of time, after this relatively slow start, the outbreak flared.
+By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic.
+This declaration marked a turning point for the public in Spain too.
+As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe, after Italy.
+In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th.
+As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g.
+food, health) and some economic subsectors of industry and construction.
+A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents.
+The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed.
+During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
 
@@
 Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
-This gives a temporal coverage of 30 days and spatial coverage of 50 units.
 For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
@@
 
+The spatial and temporal coverage of the data is as follows.
+Our dataset begins on March 13, which is the first date when every province had reported at least one case of COVID-19, and continues until April 11, for a period of 30 days.
+The unit of analysis is the province.
+Provinces are the equivalent of states, and are embedded in Autonomous Communities.
+As an example, Cataluña is an Autonomous Community and consists of four provinces, namely Barcelona, Gerona, Lerida, and Tarragona.
+The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}.
+The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the US) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the US).
+While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is inconsistent at smaller geographies, or cases are not reported at that level at all.
+The analysis must therefore be considered ecological.
+
 ```{r descriptive-statistics, echo=FALSE}
 data.frame(Variable = c("COVID-19 Incidence",
+                        "Area",
                         "GDPpc", 
@@
            Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
+                    "Area of province in sq.km",
                     "GDP per capita in €1,000s",
@@
            Min = c(min(covid19_spain$Incidence),
+                   min(provinces_spain %>% 
+                         st_area() %>% 
+                         set_units(km^2)), # Convert area from sq.m to sq.km
                    min(provinces_spain$GDPpc),
@@
            Mean = c(mean(covid19_spain$Incidence),
+                   mean(provinces_spain %>% 
+                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
                    mean(provinces_spain$GDPpc),
@@
            Max = c(max(covid19_spain$Incidence),
+                   max(provinces_spain %>% 
+                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
                    max(provinces_spain$GDPpc),
@@
            SD = c(sd(covid19_spain$Incidence),
+                   sd(provinces_spain %>% 
+                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
                    sd(provinces_spain$GDPpc),
@@
            Source = c("Montera34",
+                      "INE",
                       "INE",
@@
 
+An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease.
+Lauer et al.
+[-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days).
+The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days).
+For this reason, we judge it best to use lagged values of the environmental variables.
+We test different time lags as follows.
+We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*).
+Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*).
+Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*).
+The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al.
+[-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418.
+With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.
+
 ## Methods: Spatial SUR
 
-The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods.
-The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient] and now is a popular methodology included in several Econometrics textbook [e.g., @Greene2003econometric].
-To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p.
+The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consisits of cross-sections for different time periods.
+The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient], and has become a popular methodology included in several econometrics textbook [e.g., @Greene2003econometric].
+To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective.
+In his landmark text, Anselin discussed a model made of "an equation for each time period, which is estimated for a cross section of spatial units" (p.
 141).
-From this milestone, a huge researches used this methodology in spatial framework.
-Rey and Montouri [-@Rey1999us], Fingleton [-@Fingleton2007multi], Lauridsen et al.
-[-@Lauridsen2010spatiotemporal], Le Gallo and Dall’Erba [-@Legallo2006evaluating], and Lopez et al.
-[-@Lopez2017spatial] are well-known examples in this line.
-
-The classical SUR model without spatial effects (from here, SUR-SIM) is a stack of equations as follows:
+From this milestone, a large body of research has developed to extend the classical SUR into a spatial framework [e.g., @Rey1999us; @Lauridsen2010spatiotemporal; @Legallo2006evaluating; @Lopez2017spatial].
+
+The classical SUR model without spatial effects (hereafter, SUR-SIM) is a stack of equations as follows:
 
@@
 
-<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in province $s\text{ }s=1,...,N$) on day $t\text{ }(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is an $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the usual vector of random residuals.
+
+<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
 <!--
 ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
+
+\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) on day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the usual vector of random residuals.
 -->
 
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
-
-The more relevant characteristic of a SUR model is the dependence structure among the vectors of residuals, namely:
+A key feature of the SUR framework is the dependence structure among the vectors of random residuals, which takes the following form:
 
@@
 
+Note that this specification is very flexible, since it allows the coefficients $\beta_{it}$ to vary in order to modulate the effect of $X^i_{.t}$ on $y_t$.
+A less flexible if more parsimonious model can be obtained by imposing restrictions to some $\beta$ coefficients to fix them across equations.
+This is equivalent to specifying an effect that is constant in time.
+In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant.
+
 <!--
 ![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
+
+Note that this specification is very flexible, since it allows the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) to vary in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
+A less flexible if more parsimonious model can be obtained by imposing restrictions to some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients to fix them across equations.
+This is equivalent to specifying an effect that is constant in time.
+In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant.
 -->
 
-Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$.
-This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ considering the first $r$ coefficients contanst.
-
-Equation \ref{eq:sur-sim}) can be rewriten in compact form:
+Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:
 
@@
 
+\noindent where....
+
 <!--
 ![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
+
+\noindent where....
 -->
 
-\noindent where....
-
-As is the case of cross-sectional data, it is possible identify spatial autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several Lagrange Multipliers tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
-When the null hypothesis is rejected, there are alternative several specifications that have been propose to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
+As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
+When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
 In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor.
+A sp
+
 The stack expresion for the spatial lag SUR model (SUR-SLM),
@@
 
+\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.
+
 <!--
 ![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
+
+\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
 -->
 
-\noindent where A =$I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
-
 This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Parameters of spatial dependence are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
-The correspond Wald test is available in the R package **spsur**
+Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
+The correspond Wald test is available in the R package `spsur`.
+
+<!--
+This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
+Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
+The correspond Wald test is available in the R package `spsur`.
+-->
 
@@
 Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages.
-Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19.
 It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain.
@@
 ```{r model1-8-day-moving-average, include=FALSE}
-# Model with a lagged 8-day moving average of climatic variables:
+# Model with a lagged 8-day moving average of environmental variables:
 sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
@@
 ```{r model2-11-day-moving-average, include=FALSE}
-# Model with 11-day moving average of climatic variables:
+# Model with 11-day moving average of environmental variables:
 sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
@@
 ```{r model3-11-day-weighted-moving-average, include=FALSE}
-# Model with 11-day weighted moving average of climatic variables:
+# Model with 11-day weighted moving average of environmental variables:
 sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Edits to Context and Methods

@@
 
-For a better interpretation of the results of our analysis is necessary
-to know the chronology of the epidemic outbreak in Spain. This
-subsection presents a brief time line of COVID-19.
-
-The first reported case of COVID-19 in Spain was on January 31th, were a
-German tourist in the Canary Islands was positively tested, but the
-first not imported case was diagnostic on February 27th in Sevilla
-province (Andalusia). The epidemic was growing slowly at the beginning
-but in March 11th the World Health Organization (WHO) declared
-officially the pandemic. This declaration was a break point and on March
-13th the Spanish National Government declare the ``Emergency Status''.
-At this date, Spain have been declared 4209 cases of COVID-19 (mostly of
-cases, 1990, in Madrid). Only essential services (e.g.~food, health) and
-some economic subsector of industry and construction could to maintain
-activity. On March 17th Spain closes its lands borders and only allows
-entry to nationals and residents. From March 30th to April 12th (April
-10th-12th Easter) lockdown of the Spanish people was hardened and only
-essential activities were allowed.
+The first reported case of COVID-19 in Spain was on January 31th, when a
+German tourist in the Canary Islands tested positive for the virus.
+However, it was still a few weeks before the first domestic case was
+reported, on February 27th in Sevilla province (Andalusia). In a short
+period of time, after this relatively slow start, the outbreak flared.
+By March 11th the World Health Organization (WHO) declared COVID-19
+officially a pandemic. This declaration marked a turning point for the
+public in Spain too. As of March 13th, the number of cases of COVID-19
+reported in Spain was 4,473, with a majority of cases (1,990)
+concentrated in Madrid: these numbers were at the time the worst
+outbreak in Europe, after Italy. In response to the situation, on March
+13th the Spanish National Government declared a state of emergency. As
+part of the state of emergency restrictions to most activities were
+imposed, with the exception of essential services (e.g.~food, health)
+and some economic subsectors of industry and construction. A few days
+later, on March 17th, Spain closed its lands borders to allow entry only
+to returnee nationals and permanent residents. The lockdown was further
+hardened between March 30th and April 12th (including the Easter weekend
+of April 10th-12th) and during this period only essential activities
+were allowed. During this period, there was a dramatic reduction in
+overall mobility, both within provinces
+\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}
+as between
+\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
 
@@
 terminology) for the period between March 13th and April 11th,
-inclusive. This gives a temporal coverage of 30 days and spatial
-coverage of 50 units. For our purposes, we consider positive cases
-reported, but excluding symptomatic cases diagnosed by a doctor without
-a Polymerase Chain Reaction (PCR) test. The Spanish National Government
-publishes periodic updates at the regional level (NUTII) and the
-information is also released at the provincial level as part of a
-collaborative project {[}\textbf{by whom?}{]}. This information is
-compiled from several sources, mainly the official web pages of the
-Spanish regional goverments. In addition, we consider two sets of
-explanatory variables. The first one, and the focus of this research, is
-set of two environmental variables, namely temperature and humidity,
-which are collected from official sources (i.g., AEMET, the state
-meteorology agency, and MAPA, the ministry of agriculture, fisheries,
-and food). The second set provides some relevant controls for
-multivariate analysis, and refers to economic and demographic attributes
-of the province (also collected from official sources, i.e., INE, the
-national statistics institute). Table \ref{tab:descriptive-statistics}
-shows the descriptive statistics and the provenance of the data.
+inclusive. For our purposes, we consider positive cases reported, but
+excluding symptomatic cases diagnosed by a doctor without a Polymerase
+Chain Reaction (PCR) test. The Spanish National Government publishes
+periodic updates at the regional level (NUTII) and the information is
+also released at the provincial level as part of a collaborative project
+{[}\textbf{by whom?}{]}. This information is compiled from several
+sources, mainly the official web pages of the Spanish regional
+goverments. In addition, we consider two sets of explanatory variables.
+The first one, and the focus of this research, is set of two
+environmental variables, namely temperature and humidity, which are
+collected from official sources (i.g., AEMET, the state meteorology
+agency, and MAPA, the ministry of agriculture, fisheries, and food). The
+second set provides some relevant controls for multivariate analysis,
+and refers to economic and demographic attributes of the province (also
+collected from official sources, i.e., INE, the national statistics
+institute). Table \ref{tab:descriptive-statistics} shows the descriptive
+statistics and the provenance of the data.
+
+The spatial and temporal coverage of the data is as follows. Our dataset
+begins on March 13, which is the first date when every province had
+reported at least one case of COVID-19, and continues until April 11,
+for a period of 30 days. The unit of analysis is the province. Provinces
+are the equivalent of states, and are embedded in Autonomous
+Communities. As an example, Cataluña is an Autonomous Community and
+consists of four provinces, namely Barcelona, Gerona, Lerida, and
+Tarragona. The size of the provinces is relatively large, as seen in
+Table \ref{tab:descriptive-statistics}. The smallest province is
+\(1,978.12km^2\) (this is smaller than Rhode Island in the US) and the
+largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
+in the US). While this is a relatively large degree of spatial
+aggregation, reporting on COVID-19 is inconsistent at smaller
+geographies, or cases are not reported at that level at all. The
+analysis must therefore be considered ecological.
 
@@
 \rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
-GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
-\rowcolor{gray!6}  Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
-Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
-\rowcolor{gray!6}  Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
+Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4805.98 & INE\\
+\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
+Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
+\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
 \addlinespace
-Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
+Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
+\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
 \bottomrule
@@
 
+An important aspect of working with environmental data such as
+temperature and humidity is the incubation period of the disease. Lauer
+et al.~(2020) report for the case of COVID-19 a median incubation period
+of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The
+vast majority of cases (95\%) develop symptoms between 2.6 days (CI, 2.1
+to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we
+judge it best to use lagged values of the environmental variables. We
+test different time lags as follows. We consider a lagged 8-day average,
+from date-minus-12 to date-minus-5 days (hereafter \emph{lag8}).
+Secondly, we consider a lagged 11-day average, from date-minus-12 to
+date-minus-2 days (hereafter \emph{lag11}). Finally, to account for the
+likely duration of incubation, we consider a lagged 11-day
+\emph{weighted} average, from date-minus-12 to date-minus-2 days
+(hereafter \emph{lag11w}). The weights for this variable are calculated
+using the parameters of the log-normal distribution reported by Lauer et
+al.~(2020), i.e., a log-mean of 1.621 and a log-standard deviation of
+0.418. With these weights, the environmental variables at date-minus-2
+and date-minus-12 days are weighted as 0.041 and 0.009 respectively,
+whereas the environmental variables at date-minus-5 days are weighted as
+0.194.
+
 \hypertarget{methods-spatial-sur}{%
@@
 
-The Seemingly Unrelated Regression equations model (SUR from now on) is
-a multivariate econometric formulation used in different fields when the
-data set is organized as cross-sections for different time periods. The
-basis of this approach is well-known since the initial works of Zellner
-(1962) and now is a popular methodology included in several Econometrics
-textbook (e.g., Greene, 2003). To our knowledge, Anselin (1988) was the
-first text that introduced the term spatial SUR in reference to a model
-made of `an equation for each time period, which is estimated for a
-cross section of spatial units' (p.~141). From this milestone, a huge
-researches used this methodology in spatial framework. Rey and Montouri
-(1999), Fingleton ({\textbf{???}}), Lauridsen et al.~(2010), Le Gallo
-and Dall'Erba (2006), and Lopez et al.~(2017) are well-known examples in
-this line.
+The Seemingly Unrelated Regression equations model (SUR hereafter) is a
+multivariate econometric model used in different fields when the
+structure of the data consisits of cross-sections for different time
+periods. The basis of this approach is well-known since the initial
+works of Zellner (1962), and has become a popular methodology included
+in several econometrics textbook (e.g., Greene, 2003). To our knowledge,
+Anselin (1988) was the first author to discuss SUR from a spatial
+perspective. In his landmark text, Anselin discussed a model made of
+``an equation for each time period, which is estimated for a cross
+section of spatial units'' (p.~141). From this milestone, a large body
+of research has developed to extend the classical SUR into a spatial
+framework (e.g., Rey and Montouri, 1999; Lauridsen et al., 2010; Le
+Gallo and Dall'Erba, 2006; López et al., 2017).
 
-The classical SUR model without spatial effects (from here, SUR-SIM) is
+The classical SUR model without spatial effects (hereafter, SUR-SIM) is
 a stack of equations as follows:
@@
 
-\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
-vector, and in our case \(y_{st}\) is the incidence ratio in the
-province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
-\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
+\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is an \(N \times 1\)
+vector, and in our case \(y_{st}\) is the incidence ratio in province
+\(s\text{ }s=1,...,N\)) on day \(t\text{ }(t=1,...,T)\);
+\(X_t=(X^1,...,X^{k_t})\) is an \(N \times k_t\) matrix of the \(k_t\)
 independent variables, \(X_i=(X^i_{st})\);
 \(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
-\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
-residuals.
+\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the usual vector of
+random residuals.
 
-The more relevant characteristic of a SUR model is the dependence
-structure among the vectors of residuals, namely:
+A key feature of the SUR framework is the dependence structure among the
+vectors of random residuals, which takes the following form:
 
@@
 
-Note that this specification is very flexible, in that it allows changes
-in the coefficients \(\beta_{it}\) in order to modulate the effect of
-\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
-posible to impose restrinctions considering some \(\beta\) coefficients
-as being constant over time. In this case, we can reformulate the
+Note that this specification is very flexible, since it allows the
+coefficients \(\beta_{it}\) to vary in order to modulate the effect of
+\(X^i_{.t}\) on \(y_t\). A less flexible if more parsimonious model can
+be obtained by imposing restrictions to some \(\beta\) coefficients to
+fix them across equations. This is equivalent to specifying an effect
+that is constant in time. In this case, we can reformulate the
 coefficients expression of
 \(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
-considering the first \(r\) coefficients contanst.
+to restrict the first \(r\) coefficients to be constant.
 
-Equation \ref{eq:sur-sim}) can be rewriten in compact form:
+Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:
 
@@
 
-As is the case of cross-sectional data, it is possible identify spatial
-autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several
-Lagrange Multipliers tests have been developed to test the null
-hypothesis of spatial independence (see López et al., 2014). When the
-null hypothesis is rejected, there are alternative several
-specifications that have been propose to include spatial effects
-(Anselin, 1988, see also 2016). In this paper we consider a spatial SUR
-model that incorporates a spatial lag of the dependent variable as an
-explanatory factor. The stack expresion for the spatial lag SUR model
-(SUR-SLM),
+As is the case with cross-sectional data, it is possible to test the
+residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and
+several tests have been developed to test the null hypothesis of spatial
+independence (see López et al., 2014). When the null hypothesis is
+rejected, several alternative specifications have been proposed to
+include spatial effects (Anselin, 1988, see also 2016). In this paper we
+consider a spatial SUR model that incorporates a spatial lag of the
+dependent variable as an explanatory factor. A sp
+
+The stack expresion for the spatial lag SUR model (SUR-SLM),
 
@@
 
-\noindent where A =\(I_{TN}-\bf{\Lambda} \otimes W\) with
-\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).
+\noindent where \(A =I_{TN}-\bf{\rho} \otimes W\) with
+\(\bf{\rho} = diag(\rho_1, \cdots, \rho_T)\).
 
@@
 time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
-of incidence in neighbouring provinces. Parameters of spatial dependence
-are estimated for each time period \(\lambda_t\) and identifies the
-intensity and the sign of the impacts of neighbourhood. It is possible
-test the null hypotheis of identtical levels of spatial dependence
-(\(\lambda_i=\lambda_j, \forall i,j\)). The correspond Wald test is
-available in the R package \textbf{spsur}
+of incidence in neighbouring provinces. Coefficients for the spatially
+lagged variable are estimated for each time period \(\rho_t\) and
+identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial
+dependence (\(\rho_i=\rho_j, \forall i,j\)). The correspond Wald test is
+available in the R package \texttt{spsur}.
 
@@
 variation in the incidence of the disease, as well as its temporal
-progression in weekly averages. Our analysis begins on March 13, which
-is the first date when every province had reported at least one case of
-COVID-19. It can be seen that the highest incidence at this early date
-was in the provice of Álava, in the North of Spain. While not the most
-populous province, with a population of only 331,549, Álava has the
-highest GDP per capita of all provinces. Vitoria, its main city, is the
-capital of the Basque country and has been the focus of efforts to
-develop a ``Global Basque City'' (Meijers et al., 2008), along with San
-Sebastian and Bilbao. The other early focus of the pandemic in Spain was
-Madrid, which is the most populous province in the country and has the
-second highest GDP per capita after Álava. The early outbreaks in these
-two provinces can be traced throughout the progression of the pandemic
-over time, although by the end of the period under consideration, other
-provinces had matched and even surpased their incidence rates, including
-Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
-to the south.
+progression in weekly averages. It can be seen that the highest
+incidence at this early date was in the provice of Álava, in the North
+of Spain. While not the most populous province, with a population of
+only 331,549, Álava has the highest GDP per capita of all provinces.
+Vitoria, its main city, is the capital of the Basque country and has
+been the focus of efforts to develop a ``Global Basque City'' (Meijers
+et al., 2008), along with San Sebastian and Bilbao. The other early
+focus of the pandemic in Spain was Madrid, which is the most populous
+province in the country and has the second highest GDP per capita after
+Álava. The early outbreaks in these two provinces can be traced
+throughout the progression of the pandemic over time, although by the
+end of the period under consideration, other provinces had matched and
+even surpased their incidence rates, including Segovia and Soria to the
+north of Madrid, and Ciudad Real and Albacete to the south.
 
@@
 
+\leavevmode\hypertarget{ref-Lauer2020incubation}{}%
+Lauer, S.A., Grantz, K.H., Bi, Q., Jones, F.K., Zheng, Q., Meredith,
+H.R., Azman, A.S., Reich, N.G., Lessler, J., 2020. The incubation period
+of coronavirus disease 2019 (covid-19) from publicly reported confirmed
+cases: Estimation and application. Annals of Internal Medicine.
+doi:\href{https://doi.org/10.7326/m20-0504}{10.7326/m20-0504}
+
 \leavevmode\hypertarget{ref-Lauridsen2010spatiotemporal}{}%

### 2020-04-18 README.Rmd
Edits to Context and Methods

--- 2020-04-18 README.Rmd
@@
 COVID-19  
+Seasonality  
 Temperature  
@@
 
-## Introduction
+Introduction
+==========================
 
@@
 
-## Context, Data, and Methods
+Context, Data, and Methods
+============
 
-### Covid-19 in Spain
+## Covid-19 in Spain
 
-When was the first case reported? When did the country close international arrivals? When did initial measures of social distancing were implemented? The lockdown?
+For a better interpretation of the results of our analysis is necessary to know the chronology of the epidemic outbreak in Spain. This subsection presents a brief time line of COVID-19.
 
-### Selection of Variables
+The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia).
+The epidemic was growing slowly at the beginning but in March 11th the World Health Organization (WHO) declared officially the pandemic.
+This declaration was a break point and on March 13th the Spanish National Government declare the "Emergency Status".
+At this date, Spain have been declared 4209 cases of COVID-19 (mostly of cases, 1990, in Madrid).
+Only essential services (e.g.
+food, health) and some economic subsector of industry and construction could to maintain activity.
+On March 17th Spain closes its lands borders and only allows entry to nationals and residents.
+From March 30th to April 12th (April 10th-12th Easter) lockdown of the Spanish people was hardened and only essential activities were allowed.
+
+## Selection of Variables
 
@@
 
-### Sources of Data
+## Sources of Data
 
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases tested} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units).
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project.
+Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
+This gives a temporal coverage of 30 days and spatial coverage of 50 units.
+For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
+The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**].
 This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments.
 In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity.
-The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province.
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provencance of the data.
+The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food).
+The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute).
+Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.
 
@@
                       "MAPA")) %>%
-  kable("html",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
@@
         caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
-  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  #kable_styling(latex_options = c("striped", "scale_down")) %>%
+#kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  column_spec(2, width = "15em") %>%
+  column_spec(7, width = "5em") %>%
   footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
@@
 
-### Methods: Spatial SUR
+## Methods: Spatial SUR
 
-Seemingly Unrelated Regressions are used in the context of...
+The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods.
+The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient] and now is a popular methodology included in several Econometrics textbook [e.g., @Greene2003econometric].
+To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p.
+141).
+From this milestone, a huge researches used this methodology in spatial framework.
+Rey and Montouri [-@Rey1999us], Fingleton [-@Fingleton2007multi], Lauridsen et al.
+[-@Lauridsen2010spatiotemporal], Le Gallo and Dall’Erba [-@Legallo2006evaluating], and Lopez et al.
+[-@Lopez2017spatial] are well-known examples in this line.
 
@@
 
+<!--
 \begin{equation}
@@
 
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
+-->
+
+<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
+
+\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in the province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) the day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the vector of residuals.
 
@@
 This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ considering the first $r$ coefficients contanst.
+In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations, i.e., time periods.
+-->
+
+![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
+
+Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
+This flexibility can be reduced and it is posible to impose restrinctions considering some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
+In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations, i.e., time periods.
 
@@
 \end{equation}
+-->
+
+![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
 
@@
 
-\noindent where A =$I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
+\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.
+-->
 
+![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
+
+\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
+
+<!--
 This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Parameters of spatial dependence are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
-The correspond Wald test is available in the R package **spsur**
+Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
+The correspond Wald test is available in the R package `spsur`.
+-->
+
+This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
+Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
+The correspond Wald test is available in the R package `spsur`.
 
 The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
+
+Another alternative methodologies could be use.
+By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g.
+Elhorst 2014, Cap.
+4), but those models are more dont take account correlation between errors.
+Therefore, a spatial SUR approach is more reasonable for our purpose.
 
@@
 
-### SUR Models
+## SUR Models
 
@@
                   "Non-constrained")) %>%
-  kable("html",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
@@
         caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
+  #kable_styling(latex_options = c("striped", "scale_down")) %>%
   kable_styling(bootstrap_options = c("striped", "condensed")) %>%
@@
 
-## Discussion
+Discussion
+===================
 
@@
 
-## Concluding Remarks
+Concluding Remarks
+===================
 
@@
 
-## Acknowledgments
+Acknowledgments {#acknowledgments .unnumbered}
+==========
 
-The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
+Add acknowledgments as appropriate in final draft.
 
-## References
+The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
+
+References {#references .unnumbered}
+==========

### 2020-04-18 README.Rmd
revision bibliografia

@@
   #kable_styling(latex_options = c("striped", "scale_down")) %>%
-#kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
   column_spec(2, width = "15em") %>%
@@
 \end{equation}
-
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
 -->
 
-<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
 ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
 
-\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in the province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) the day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the vector of residuals.
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
 
@@
 \end{equation} 
+-->
+
+![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
 
@@
 This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations, i.e., time periods.
--->
-
-![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
-
-Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
-This flexibility can be reduced and it is posible to impose restrinctions considering some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations, i.e., time periods.
+In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ considering the first $r$ coefficients contanst.
 
@@
 \end{equation}
-
-\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.
 -->
@@
 
-\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
-
-<!--
+\noindent where A =$I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
+
 This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
-The correspond Wald test is available in the R package `spsur`.
--->
-
-This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
-The correspond Wald test is available in the R package `spsur`.
+Parameters of spatial dependence are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
+The correspond Wald test is available in the R package **spsur**
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
More work on the methods section

@@
 In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor.
-A sp
+Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio].
+While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid].
+For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
More work on the methods section

@@
 outbreak in Europe, after Italy. In response to the situation, on March
-13th the Spanish National Government declared a state of emergency. As
-part of the state of emergency restrictions to most activities were
-imposed, with the exception of essential services (e.g.~food, health)
-and some economic subsectors of industry and construction. A few days
-later, on March 17th, Spain closed its lands borders to allow entry only
-to returnee nationals and permanent residents. The lockdown was further
-hardened between March 30th and April 12th (including the Easter weekend
-of April 10th-12th) and during this period only essential activities
-were allowed. During this period, there was a dramatic reduction in
-overall mobility, both within provinces
-\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}
-as between
+13th the Spanish National Government declared a state of emergency, to
+go into effect on Saturday March 14th. As part of the state of emergency
+restrictions to most activities were imposed, with the exception of
+essential services (e.g.~food, health) and some economic subsectors of
+industry and construction. A few days later, on March 17th, Spain closed
+its lands borders to allow entry only to returnee nationals and
+permanent residents. The lockdown was further hardened between March
+30th and April 12th (including the Easter weekend of April 10th-12th)
+and during this period only essential activities were allowed. During
+this period, there was a dramatic reduction in overall mobility, both
+within provinces as between
 \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd

@@
 Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.
-
 journal: "Geographical Analysis"
@@
 
-```{r load-data, include=FALSE}
+```{r load-data, include=FALSE, cache=TRUE}
 # Load data from package `covid19env`
@@
 
-```{r data-preparation, include=FALSE}
+```{r data-preparation, include=FALSE, cache=TRUE}
 # Convert GDP per capita to thousands of euros:
@@
 
-```{r sort-communities-by-latitude, include=FALSE}
+```{r sort-communities-by-latitude, include=FALSE, cache=TRUE}
 # To visualize the distribution of temperature by CCAA, we sort the communities by latitude, from north to south.
@@
 
-https://github.com/paezha/covid19-environmental-correlates
+>https://github.com/paezha/covid19-environmental-correlates
 
@@
 
-```{r descriptive-statistics, echo=FALSE}
+```{r descriptive-statistics, echo=FALSE, cache=TRUE}
 data.frame(Variable = c("COVID-19 Incidence",
@@
                          st_area() %>%
-                         set_units(km^2)), # Convert area from sq.m to sq.km
+                         set_units(km^2)),
                    min(provinces_spain$GDPpc),
@@
                     mean(provinces_spain %>% 
-                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
+                         st_area() %>%
+                         set_units(km^2)),
                     mean(provinces_spain$GDPpc),
@@
                    max(provinces_spain %>% 
-                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
+                         st_area() %>%
+                         set_units(km^2)),
                    max(provinces_spain$GDPpc),
@@
            SD = c(sd(covid19_spain$Incidence),
+                  sd(provinces_spain$GDPpc),
                   sd(provinces_spain %>% 
-                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
-                   sd(provinces_spain$GDPpc),
+                         st_area() %>%
+                         set_units(km^2)),
                   sd(provinces_spain$Older),
@@
         digits = 2,
-        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
+        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
   kable_styling(latex_options = c("striped", "scale_down")) %>%
@@
 
-The classical SUR model without spatial effects (hereafter, SUR-SIM) is a stack of equations as follows:
+The classical SUR model without spatial effects (from here, SUR-SIM) is a stack of equations as follows:
 
@@
 
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in province $s\text{ }s=1,...,N$) on day $t\text{ }(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is an $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the usual vector of random residuals.
-
-<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
+
+<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
   <!--
@@
 
-\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) on day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the usual vector of random residuals.
+\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in the province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) the day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the vector of residuals.
 -->
 
-A key feature of the SUR framework is the dependence structure among the vectors of random residuals, which takes the following form:
+A key feature of the SUR model is the dependence structure among the vectors of residuals, namely:
 
@@
 
-Note that this specification is very flexible, since it allows the coefficients $\beta_{it}$ to vary in order to modulate the effect of $X^i_{.t}$ on $y_t$.
-A less flexible if more parsimonious model can be obtained by imposing restrictions to some $\beta$ coefficients to fix them across equations.
-This is equivalent to specifying an effect that is constant in time.
-In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant.
+Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$.
+This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
+In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations.
+This is equivalent to specifying some effects to be invariant over time.
 
@@
 
-Note that this specification is very flexible, since it allows the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) to vary in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
-A less flexible if more parsimonious model can be obtained by imposing restrictions to some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients to fix them across equations.
-This is equivalent to specifying an effect that is constant in time.
-In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant.
+Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
+This flexibility can be reduced and it is posible to impose restrictions to consider some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
+In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations.
+This is equivalent to specifying some effects to be invariant over time.
 -->
 
-Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:
+Equation \ref{eq:sur-sim}) can be rewriten in compact form:
 
@@
 
-\noindent where....
-
 <!--
   ![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
-
-\noindent where....
 -->
@@
 
-The stack expresion for the spatial lag SUR model (SUR-SLM),
+The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:
 
@@
 
-\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.
+\noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
 
@@
 
-\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
+\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
 -->
@@
 It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
-The correspond Wald test is available in the R package `spsur`.
+The correspond Wald test is available in the `R` package `spsur`.
 
@@
 It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
-The correspond Wald test is available in the R package `spsur`.
+The correspond Wald test is available in the `R` package `spsur`.
 -->
@@
 The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
-
 Another alternative methodologies could be use.
@@
 Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages.
+Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19.
 It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain.
@@
 
-```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE}
+```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE, cache = TRUE}
 # Retrieve the coordinates of Madrid
@@
 ```{r descriptives-temperature, echo = FALSE, fig.height=7, fig.cap="\\label{fig:descriptives-temperature} Distribution of mean temperatures and humidities in the Autonomous Communities in Spain between March 12, 2020 and April 11, 2020.
-The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost"}
+The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost", cache = TRUE}
 # Summary of temperature by Autonomous Community
@@
 
-```{r control-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)"}
+```{r control-map, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
 gdppc.plot <- provinces_spain %>%
@@
 
-```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)"}
+```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)", cache = TRUE}
 covid19_spain %>% 
@@
 
-```{r data-preparation-for-modelling, include=FALSE}
+```{r data-preparation-for-modelling, include=FALSE, cache = TRUE}
 # Organize data for SUR modelling 
@@
 
-```{r spatial-weights, include=FALSE}
+```{r spatial-weights, include=FALSE, cache = TRUE}
 # Create spatial weights matrix: 
@@
 
-```{r formulas, include=FALSE}
+```{r formulas, include=FALSE, cache = TRUE}
 # Define formulas with three different lagged variables:
@@
 
-```{r model-restrictions, include=FALSE}
+```{r model-restrictions, include=FALSE, cache = TRUE}
 T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # Recall that T is the number of days, i.e., time periods, i.e., equations; add 1 to include the starting day
@@
 
-```{r model1-8-day-moving-average, include=FALSE}
-# Model with a lagged 8-day moving average of environmental variables:
+```{r model1-8-day-moving-average, include=FALSE, cache = TRUE}
+# Model with a lagged 8-day moving average of climatic variables:
 sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
@@
 
-```{r model2-11-day-moving-average, include=FALSE}
-# Model with 11-day moving average of environmental variables:
+```{r model2-11-day-moving-average, include=FALSE, cache = TRUE}
+# Model with 11-day moving average of climatic variables:
 sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
@@
 
-```{r model3-11-day-weighted-moving-average, include=FALSE}
-# Model with 11-day weighted moving average of environmental variables:
+```{r model3-11-day-weighted-moving-average, include=FALSE, cache = TRUE}
+# Model with 11-day weighted moving average of climatic variables:
 sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
@@
 
-```{r goodness-of-fit, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
+```{r goodness-of-fit, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
 data.frame(Model1 = sur.slm_lag8$R2, 
@@
 Summary of best model.
-```{r summary-best-model, echo=FALSE}
+```{r summary-best-model, echo=FALSE, cache = TRUE}
 data.frame(Coefficient = sur.slm_lag11$coefficients, 
@@
 
-```{r delta-time, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
+```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
 data.frame(Date = seq(ymd("2020-03-13"), 
@@
 
-```{r beta-intercept-time, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
+```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
 #Intercept
@@
 
-```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)"}
+```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)", cache = TRUE}
 #Density:
@@
 
-```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit"}
+```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit", cache = TRUE}
 n = 3
@@
 
-```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)"}
+```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)", cache = TRUE}
 n = 4
@@
 
-```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)"}
+```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)", cache = TRUE}
 n = 5

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex

@@
 
-\hypertarget{covid-19-in-spain}{%
-\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}
+\#\# Covid-19 in Spain
 
@@
 
-\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics}
+\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics} Descriptive statistics}
 \centering
@@
 \rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
-Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4805.98 & INE\\
-\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
+Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
+\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4805.98 & INE\\
 Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
@@
 
-The classical SUR model without spatial effects (hereafter, SUR-SIM) is
+The classical SUR model without spatial effects (from here, SUR-SIM) is
 a stack of equations as follows:
@@
 
-\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is an \(N \times 1\)
-vector, and in our case \(y_{st}\) is the incidence ratio in province
-\(s\text{ }s=1,...,N\)) on day \(t\text{ }(t=1,...,T)\);
-\(X_t=(X^1,...,X^{k_t})\) is an \(N \times k_t\) matrix of the \(k_t\)
+\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
+vector, and in our case \(y_{st}\) is the incidence ratio in the
+province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
+\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
 independent variables, \(X_i=(X^i_{st})\);
 \(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
-\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the usual vector of
-random residuals.
+\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
+residuals.
 
-A key feature of the SUR framework is the dependence structure among the
-vectors of random residuals, which takes the following form:
+A key feature of the SUR model is the dependence structure among the
+vectors of residuals, namely:
 
@@
 
-Note that this specification is very flexible, since it allows the
-coefficients \(\beta_{it}\) to vary in order to modulate the effect of
-\(X^i_{.t}\) on \(y_t\). A less flexible if more parsimonious model can
-be obtained by imposing restrictions to some \(\beta\) coefficients to
-fix them across equations. This is equivalent to specifying an effect
-that is constant in time. In this case, we can reformulate the
-coefficients expression of
+Note that this specification is very flexible, in that it allows changes
+in the coefficients \(\beta_{it}\) in order to modulate the effect of
+\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
+posible to impose restrinctions considering some \(\beta\) coefficients
+as being constant over time. In this case, we can reformulate the
+coefficients expression
 \(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
-to restrict the first \(r\) coefficients to be constant.
+to restrict the first \(r\) coefficients to be constant across
+equations. This is equivalent to specifying some effects to be invariant
+over time.
 
-Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:
+Equation \ref{eq:sur-sim}) can be rewriten in compact form:
 
@@
 \end{equation}
-
-\noindent where\ldots.
 
@@
 consider a spatial SUR model that incorporates a spatial lag of the
-dependent variable as an explanatory factor. A sp
+dependent variable as an explanatory factor. Spatial analytical
+approaches were used to understand contagion-difussion processes in the
+case of infectious disease in general (e.g., Cliff et al., 1998) and the
+2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
+al., 2010). While we are mindful of the same caveat that the novel
+SARS-CoV2 may not follow the patterns of previous diseases, there is
+still evidence from the United States that COVID-19 displays spatial
+patterns that are consistent with a diffusion process (Desjardins et
+al., 2020). For this reason, the spatial lag model is appropriate to
+model incidence of COVID-19 geographically, since it accounts for
+potential spatial patterns that result from a process of contagion, as
+explained next.
 
-The stack expresion for the spatial lag SUR model (SUR-SLM),
+The stack expresion for the spatial lag SUR model (SUR-SLM) is as
+follows:
 
@@
 
-\noindent where \(A =I_{TN}-\bf{\rho} \otimes W\) with
-\(\bf{\rho} = diag(\rho_1, \cdots, \rho_T)\).
+\noindent where \(A =I_{TN}-\bf{\Lambda} \otimes W\) with
+\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).
 
@@
 The SUR-SLM model can be estimated using maximum likelihood (López et
-al. (2014)) or instrumental variables (Mínguez et al. (2019)).
-
-Another alternative methodologies could be use. By example, a dynamic
-spatial panel methodology with fixed spatial an temporal effects
-(e.g.~Elhorst 2014, Cap. 4), but those models are more dont take account
-correlation between errors. Therefore, a spatial SUR approach is more
-reasonable for our purpose.
+al. (2014)) or instrumental variables (Mínguez et al. (2019)). Another
+alternative methodologies could be use. By example, a dynamic spatial
+panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
+2014, Cap. 4), but those models are more dont take account correlation
+between errors. Therefore, a spatial SUR approach is more reasonable for
+our purpose.
 
@@
 variation in the incidence of the disease, as well as its temporal
-progression in weekly averages. It can be seen that the highest
-incidence at this early date was in the provice of Álava, in the North
-of Spain. While not the most populous province, with a population of
-only 331,549, Álava has the highest GDP per capita of all provinces.
-Vitoria, its main city, is the capital of the Basque country and has
-been the focus of efforts to develop a ``Global Basque City'' (Meijers
-et al., 2008), along with San Sebastian and Bilbao. The other early
-focus of the pandemic in Spain was Madrid, which is the most populous
-province in the country and has the second highest GDP per capita after
-Álava. The early outbreaks in these two provinces can be traced
-throughout the progression of the pandemic over time, although by the
-end of the period under consideration, other provinces had matched and
-even surpased their incidence rates, including Segovia and Soria to the
-north of Madrid, and Ciudad Real and Albacete to the south.
+progression in weekly averages. Our analysis begins on March 13, which
+is the first date when every province had reported at least one case of
+COVID-19. It can be seen that the highest incidence at this early date
+was in the provice of Álava, in the North of Spain. While not the most
+populous province, with a population of only 331,549, Álava has the
+highest GDP per capita of all provinces. Vitoria, its main city, is the
+capital of the Basque country and has been the focus of efforts to
+develop a ``Global Basque City'' (Meijers et al., 2008), along with San
+Sebastian and Bilbao. The other early focus of the pandemic in Spain was
+Madrid, which is the most populous province in the country and has the
+second highest GDP per capita after Álava. The early outbreaks in these
+two provinces can be traced throughout the progression of the pandemic
+over time, although by the end of the period under consideration, other
+provinces had matched and even surpased their incidence rates, including
+Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
+to the south.
 
@@
 
+\leavevmode\hypertarget{ref-Cao2010spatio}{}%
+Cao, Z., Zeng, D., Zheng, X., Wang, Q., Wang, F., Wang, J., Wang, X.,
+2010. Spatio-temporal evolution of beijing 2003 sars epidemic. Science
+China Earth Sciences 53, 1017--1028.
+doi:\href{https://doi.org/10.1007/s11430-010-0043-x}{10.1007/s11430-010-0043-x}
+
 \leavevmode\hypertarget{ref-Casanova2010effects}{}%
@@
 
+\leavevmode\hypertarget{ref-Cliff1998detecting}{}%
+Cliff, A., Haggett, P., Smallman-Raynor, M., 1998. Detecting
+space---time patterns in geocoded disease data. Cholera in london, 1854
+measles in the united states, 1962--95, in: Geomed'97. Springer, pp.
+13--42.
+
 \leavevmode\hypertarget{ref-Coelho2020exponential}{}%
@@
 global scale. medRxiv.
+
+\leavevmode\hypertarget{ref-Desjardins2020rapid}{}%
+Desjardins, M., Hohl, A., Delmelle, E., 2020. Rapid surveillance of
+covid-19 in the united states using a prospective space-time scan
+statistic: Detecting and evaluating emerging clusters. Applied Geography
+102202.
 
@@
 
+\leavevmode\hypertarget{ref-Meng2005understanding}{}%
+Meng, B., Wang, J., Liu, J., Wu, J., Zhong, E., 2005. Understanding the
+spatial diffusion process of severe acute respiratory syndrome in
+beijing. Public Health 119, 1080--1087.
+doi:\href{https://doi.org/https://doi.org/10.1016/j.puhe.2005.02.003}{https://doi.org/10.1016/j.puhe.2005.02.003}
+
 \leavevmode\hypertarget{ref-Millo2017robust}{}%

### 2020-04-18 README.Rmd
Updated README.Rmd and README.md

--- 2020-04-18 README.Rmd
@@
 
-## Sources of Data
+## Data
 
 Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
-This gives a temporal coverage of 30 days and spatial coverage of 50 units.
 For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
@@
 
-```{r descriptive-statistics, echo=FALSE}
+The spatial and temporal coverage of the data is as follows.
+Our dataset begins on March 13, which is the first date when every province had reported at least one case of COVID-19, and continues until April 11, for a period of 30 days.
+The unit of analysis is the province.
+Provinces are the equivalent of states, and are embedded in Autonomous Communities.
+As an example, Cataluña is an Autonomous Community and consists of four provinces, namely Barcelona, Gerona, Lerida, and Tarragona.
+The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}.
+The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the US) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the US).
+While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is inconsistent at smaller geographies, or cases are not reported at that level at all.
+The analysis must therefore be considered ecological.
+
+```{r descriptive-statistics, echo=FALSE, cache=TRUE}
 data.frame(Variable = c("COVID-19 Incidence",
+                        "Area", 
                         "GDPpc", 
@@
            Min = c(min(covid19_spain$Incidence),
+                   min(provinces_spain %>% 
+                         st_area() %>%
+                         set_units(km^2)),
                    min(provinces_spain$GDPpc),
@@
            Mean = c(mean(covid19_spain$Incidence),
+                    mean(provinces_spain %>% 
+                         st_area() %>%
+                         set_units(km^2)),
                     mean(provinces_spain$GDPpc),
@@
            Max = c(max(covid19_spain$Incidence),
+                   max(provinces_spain %>% 
+                         st_area() %>%
+                         set_units(km^2)),
                    max(provinces_spain$GDPpc),
@@
                   sd(provinces_spain$GDPpc),
+                  sd(provinces_spain %>% 
+                         st_area() %>%
+                         set_units(km^2)),
                   sd(provinces_spain$Older),
@@
         digits = 2,
-        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
+        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
   #kable_styling(latex_options = c("striped", "scale_down")) %>%
-#kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
   column_spec(2, width = "15em") %>%
@@
 
+An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease.
+Lauer et al.
+[-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days).
+The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days).
+For this reason, we judge it best to use lagged values of the environmental variables.
+We test different time lags as follows.
+We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*).
+Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*).
+Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*).
+The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al.
+[-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418.
+With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.
+
 ## Methods: Spatial SUR
 
-The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods.
-The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient] and now is a popular methodology included in several Econometrics textbook [e.g., @Greene2003econometric].
-To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p.
+The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consisits of cross-sections for different time periods.
+The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient], and has become a popular methodology included in several econometrics textbook [e.g., @Greene2003econometric].
+To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective.
+In his landmark text, Anselin discussed a model made of "an equation for each time period, which is estimated for a cross section of spatial units" (p.
 141).
-From this milestone, a huge researches used this methodology in spatial framework.
-Rey and Montouri [-@Rey1999us], Fingleton [-@Fingleton2007multi], Lauridsen et al.
-[-@Lauridsen2010spatiotemporal], Le Gallo and Dall’Erba [-@Legallo2006evaluating], and Lopez et al.
-[-@Lopez2017spatial] are well-known examples in this line.
+From this milestone, a large body of research has developed to extend the classical SUR into a spatial framework [e.g., @Rey1999us; @Lauridsen2010spatiotemporal; @Legallo2006evaluating; @Lopez2017spatial].
 
@@
 
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
 -->
 
-<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+
 ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
@@
 
-The more relevant characteristic of a SUR model is the dependence structure among the vectors of residuals, namely:
+A key feature of the SUR model is the dependence structure among the vectors of residuals, namely:
 
@@
 This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations, i.e., time periods.
+In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations.
+This is equivalent to specifying some effects to be invariant over time.
 -->
@@
 Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
-This flexibility can be reduced and it is posible to impose restrinctions considering some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations, i.e., time periods.
+This flexibility can be reduced and it is posible to impose restrictions to consider some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
+In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations.
+This is equivalent to specifying some effects to be invariant over time.
 
@@
 
-\noindent where....
-
-As is the case of cross-sectional data, it is possible identify spatial autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several Lagrange Multipliers tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
-When the null hypothesis is rejected, there are alternative several specifications that have been propose to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
+As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
+When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
 In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor.
-The stack expresion for the spatial lag SUR model (SUR-SLM),
+Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio].
+While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid].
+For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.
+
+The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:
 
@@
 
-\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.
+\noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
 -->
@@
 
-\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
+\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
 
@@
 It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
-The correspond Wald test is available in the R package `spsur`.
+The correspond Wald test is available in the `R` package `spsur`.
 -->
@@
 It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
-The correspond Wald test is available in the R package `spsur`.
+The correspond Wald test is available in the `R` package `spsur`.
 
 The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
-
 Another alternative methodologies could be use.
@@
 
-```{r model1-8-day-moving-average, include=FALSE}
+```{r model1-8-day-moving-average, include=FALSE, cache = TRUE}
 # Model with a lagged 8-day moving average of climatic variables:
@@
 
-```{r model2-11-day-moving-average, include=FALSE}
+```{r model2-11-day-moving-average, include=FALSE, cache = TRUE}
 # Model with 11-day moving average of climatic variables:
@@
 
-```{r model3-11-day-weighted-moving-average, include=FALSE}
+```{r model3-11-day-weighted-moving-average, include=FALSE, cache = TRUE}
 # Model with 11-day weighted moving average of climatic variables:

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Updated README.Rmd and README.md
Fixed inconsistent use of rho and lambda

@@
 
-## Sources of Data
+## Data
 
@@
 This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
+Coefficients for the spatially lagged variable are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
 The correspond Wald test is available in the `R` package `spsur`.
@@
 This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
+Coefficients for the spatially lagged variable are estimated for each time period ![$\lambda_t$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypotheis of identtical levels of spatial dependence (![$\lambda_i=\lambda_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_i%3D%5Clambda_j%2C%20%5Cforall%20i%2Cj%24)).
 The correspond Wald test is available in the `R` package `spsur`.
@@
 
-Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$).
+Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\lambda$).
 

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-bkp.Rmd
Created backup file

--- 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
   log(Humidity_lag8) +
-  log(Mean_Temp_lag8)
+  log(Mean_Temp_lag8) +
+  log(Sunshine_Hours_lag8 + 0.1)
 
@@
   log(Humidity_lag11) +
-  log(Mean_Temp_lag11)
+  log(Mean_Temp_lag11) +
+  log(Sunshine_Hours_lag11 + 0.1)
 
@@
   log(Humidity_lag11w) +
-  log(Mean_Temp_lag11w)
+  log(Mean_Temp_lag11w) +
+  log(Sunshine_Hours_lag11w + 0.1)
 ```
@@
 T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # Recall that T is the number of days, i.e., time periods, i.e., equations; add 1 to include the starting day
-k <- 7 # Number of independent variables, including the constant
+k <- 8 # Number of independent variables, including the constant
 coef_rest <- 2 # Number of restrictions
@@
 }
-b2 <- matrix(0, ncol = 21*coef_rest)
+b2 <- matrix(0, ncol = (T - 1) * coef_rest)
 ```
@@
                       "log(Temperature)",
+                      "log(Sunshine)",
                       rep(c("Intercept",
@@
                             "log(Humidity)", 
-                            "log(Temperature)"), 29)),
+                            "log(Temperature)",
+                            "log(Sunshine)"),
+                          29)),
          Variable = factor(Variable, 
@@
                                       "log(Humidity)",
-                                      "log(Temperature)"))) %>%
+                                      "log(Temperature)",
+                                      "log(Sunshine)"))) %>%
   group_by(Variable) %>%
@@
                   "Constrained", 
+                  "Non-constrained", 
                   "Non-constrained",
@@
 
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.
+Figure \ref{fig:beta-temperature-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.
 
@@
 
+Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the coefficient for $\log(Sunshine + 0.1)$.
+
+```{r beta-sunshine-time, echo=FALSE, fig.cap="\\label{fig:beta-sunshine-time}Temporal evolution of coefficient for log(Sunshine + 0.1)", cache = TRUE}
+n = 6
+data.frame(Date = seq(ymd("2020-03-13"), 
+                      ymd("2020-04-11"), 
+                      by = "days"),
+           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
+           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
+  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
+                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
+                         TRUE ~ "p > 0.10")) %>%
+  ggplot(aes(x = Date, y = Beta, color = sig)) +
+  geom_point(size = 2) +
+  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
+  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
+  theme_tufte()
+```
+
 Concluding Remarks

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Created backup file

@@
 
+\begin{quote}
 https://github.com/paezha/covid19-environmental-correlates
+\end{quote}
 
@@
 
-\#\# Covid-19 in Spain
+\hypertarget{covid-19-in-spain}{%
+\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}
 
@@
 
-\hypertarget{sources-of-data}{%
-\subsection{Sources of Data}\label{sources-of-data}}
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
 
@@
 of incidence in neighbouring provinces. Coefficients for the spatially
-lagged variable are estimated for each time period \(\rho_t\) and
+lagged variable are estimated for each time period \(\lambda_t\) and
 identifies the intensity and the sign of the impacts of neighbourhood.
 It is possible test the null hypotheis of identtical levels of spatial
-dependence (\(\rho_i=\rho_j, \forall i,j\)). The correspond Wald test is
-available in the R package \texttt{spsur}.
+dependence (\(\lambda_i=\lambda_j, \forall i,j\)). The correspond Wald
+test is available in the \texttt{R} package \texttt{spsur}.
 
@@
 \midrule
-\rowcolor{gray!6}  Intercept & 7.175 & 10.008 & 13.845 & 0 & 0 & 30 & Non-constrained\\
-log(GDPpc) & 0.513 & 0.513 & 0.513 & 0 & 1 & 0 & Constrained\\
-\rowcolor{gray!6}  log(Older) & -0.783 & -0.783 & -0.783 & 0 & 0 & 1 & Constrained\\
-log(Density) & -0.217 & -0.092 & 0.188 & 19 & 1 & 10 & Non-constrained\\
-\rowcolor{gray!6}  Transit & 0.300 & 0.518 & 0.595 & 11 & 10 & 9 & Non-constrained\\
+\rowcolor{gray!6}  Intercept & 7.370 & 10.037 & 12.968 & 0 & 0 & 30 & Non-constrained\\
+log(GDPpc) & 0.620 & 0.620 & 0.620 & 0 & 0 & 1 & Constrained\\
+\rowcolor{gray!6}  log(Older) & -0.737 & -0.737 & -0.737 & 0 & 0 & 1 & Constrained\\
+log(Density) & -0.220 & -0.097 & 0.153 & 19 & 1 & 10 & Non-constrained\\
+\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 10 & 9 & 11 & Non-constrained\\
 \addlinespace
-log(Humidity) & -1.176 & -0.372 & 0.342 & 17 & 4 & 9 & Non-constrained\\
-\rowcolor{gray!6}  log(Temperature) & -2.048 & -1.431 & -0.989 & 0 & 0 & 30 & Non-constrained\\
+log(Humidity) & -1.434 & -0.534 & -0.031 & 10 & 1 & 19 & Non-constrained\\
+\rowcolor{gray!6}  log(Temperature) & -2.014 & -1.406 & -0.929 & 0 & 0 & 30 & Non-constrained\\
+log(Sunshine) & -0.258 & 0.097 & 0.206 & 7 & 2 & 21 & Non-constrained\\
 \bottomrule
@@
 Figure \ref{fig:delta-time} shows the temporal evolution of the spatial
-autocorrelation coefficient (\(\rho\)).
+autocorrelation coefficient (\(\lambda\)).
 
@@
 
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the
-coefficient for \(\log(Temperature)\).
+Figure \ref{fig:beta-temperature-time} shows the temporal evolution of
+the coefficient for \(\log(Temperature)\).
 
@@
 coefficient for log(Temperature)}
+\end{figure}
+
+Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the
+coefficient for \(\log(Sunshine + 0.1)\).
+
+\begin{figure}
+\centering
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-sunshine-time-1.pdf}
+\caption{\label{fig:beta-sunshine-time}Temporal evolution of coefficient
+for log(Sunshine + 0.1)}
 \end{figure}

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Updated the analysis to include hours of sunshine, revised the abstract and introduction to reflect this change, recreated the README.md document

@@
 While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees.
-How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature and humidity.
+How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine.
 In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain.
@@
 An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process.
-Our results indicate that incidence of the disease is lower at higher temperatures.
-The evidence with respect to humidity is more mixed, with coefficients for this variable that are significant in only some equations.
-Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.
+Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations.
+Sunshine, in contrast, displays a positive association with incidence of the disease.
+Our control variables also yield interesting insights.
+Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
 journal: "Geographical Analysis"
@@
 This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive.
-At a global scale, whereas de Ángel Solá et al.
-[-@deangel2020weathering] see less risk in the Caribean Basin, Coelho et al.
+At a global scale, de Ángel Solá et al.
+[-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al.
 [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate.
@@
 [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States.
+Yao et al.
+[-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19.
+This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al.
+2016; Dąbrowska-Leonik et al.
+2018; Dinlen et al.
+2016; Mathyssen et al.
+2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).
 To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread).
@@
 
-With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature and humidity.
-Hence, here we report results from a spatial analysis of incidence of COVID-19 in fifty provinces in Spain.
-Spain is one of the countries hardest hit by the virus, and enacted lockdown measures on March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
+With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature, humidity, and sunshine.
+We adopt a population health approach, and report results from a spatial model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
 We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
@@
 
-The results provide evidence of the effect of temperature on the incidence of COVID-19.
-The evidence concerning humidity is more mixed: while the direction of the effect is negative, as anticipated, the coefficients for this variable are only significant in some of equations in the system.
+The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
+The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
 Our control variables also provide some intriguing insights.
-The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers wrestling with the question of the trajectory of the pandemic.
+Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
+The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
 

### 2020-04-18 README.Rmd
Updated the analysis to include hours of sunshine, revised the abstract and introduction to reflect this change, recreated the README.md document

@@
 An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process.
-Our results indicate that incidence of the disease is lower at higher temperatures.
-The evidence with respect to humidity is more mixed, with coefficients for this variable that are significant in only some equations.
-Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.
+Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations.
+Sunshine, in contrast, displays a positive association with incidence of the disease.
+Our control variables also yield interesting insights.
+Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
 
@@
 Humidity  
+Population density  
+Older adults  
 Spatial SUR  
@@
 <!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
-
 ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
@@
 \noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
--->
-
-![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
-
-\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
-
-<!--
+
 This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
@@
 -->
+
+![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
+
+\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
 

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
eso

@@
 While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees.
-How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine.
+How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature and humidity.
 In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain.
@@
 An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process.
-Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations.
-Sunshine, in contrast, displays a positive association with incidence of the disease.
-Our control variables also yield interesting insights.
-Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
+Our results indicate that incidence of the disease is lower at higher temperatures.
+The evidence with respect to humidity is more mixed, with coefficients for this variable that are significant in only some equations.
+Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.
 journal: "Geographical Analysis"
@@
 This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive.
-At a global scale, de Ángel Solá et al.
-[-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al.
+At a global scale, whereas de Ángel Solá et al.
+[-@deangel2020weathering] see less risk in the Caribean Basin, Coelho et al.
 [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate.
@@
 [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States.
-Yao et al.
-[-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19.
-This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al.
-2016; Dąbrowska-Leonik et al.
-2018; Dinlen et al.
-2016; Mathyssen et al.
-2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).
 To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread).
@@
 
-With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature, humidity, and sunshine.
-We adopt a population health approach, and report results from a spatial model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
+With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature and humidity.
+Hence, here we report results from a spatial analysis of incidence of COVID-19 in fifty provinces in Spain.
+Spain is one of the countries hardest hit by the virus, and enacted lockdown measures on March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
 We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
@@
 
-The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
-The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
+The results provide evidence of the effect of temperature on the incidence of COVID-19.
+The evidence concerning humidity is more mixed: while the direction of the effect is negative, as anticipated, the coefficients for this variable are only significant in some of equations in the system.
 Our control variables also provide some intriguing insights.
-Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
-The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
+The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers wrestling with the question of the trajectory of the pandemic.
 
@@
 
+* La temperatura puedes ser un factor pero cabría esperar que no tuviera un impacto lineal.
+Quizás una temperatua mantenida superior a X grados sea la mejor forma de incorporarla al modelo.
+En nuestro modelo en incremento en un grado de la temp de 2 a 3 grados (en una provincia fria al norte de España) tiene el mismo impacto que un incremento de 25 a 26 grados (en una provincia en las Islas Canarias)
+* En carácter insular de las Islas Canarias y de las Islas baleares no se ha tenido en cuenta
+* Baleares se ha linkado artificialmente con 3 provincias y debia de haberse dejado aislada para respetar su carácter insular.
+* La incidencia depende de un estado inicial
+* No se ha controlado por el sistema sanitario de cada provincia.
+Uno de los principales focos de cintagio han sido los hospitales y los centros de salud.
+En aquellas provincias donde se ha promovido "NO IR AL MEDICO" han presentado menor incidencia.
+
 Acknowledgments {#acknowledgments .unnumbered}

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
eso

@@
   transit systems in the province is associated with a greater incidence
-  of the disease.
+  of the disease.\\
   \end{abstract}

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Primer empujón variables

--- 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 # Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install
-
 library(covid19env)
@@
 
-Explain the rationale for selecting the variables.
-
-For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world.
-Controlling for socioeconomic characteristics the objective of this paper is observe the effect of clime on COVID-19  proliferation.
-
-Some questions are important to point out.
+Two set of variables has been selected to explain the diary incidence ratio in each Spanish province.
+The first, are climatic variables (temperature, humidity and sunshine) and its impact has been argued in the introduction section.
+The second, is a set of specific characteristics of each province.
+The objective is include in our model some control variables (socioeconomic characteristics) to capture the probable heterogeneity of each spatial unit.
+All those variables are constant along the temporal period analized in this research.
+In this subsetion we present some evidences of the expected relationship between the four control variables considered an the incidence ratio.
+
+The structure of population is a key variable on the incidence.
+Several papers related the relationship between a fast trasnmision of COVID-19 and population density (e.g.
+cumulative incidence in urban areas like NYC) and therefore a positive sign will be expected.
+Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship.
+In Spain the population is concentrate (with som exception) in the mediterranean coast (@goerlich2013population), in general with high temperature and high humidity.
+So, positive or negative sign will be expected in the model.
+
+The percentaje of older adults (over 65) is the second variable related with the population structure include like control variable.
+Unlike the population density the older people in Spain is concentrate in the core of the Iberian Peninsula (see figure \ref{plotfer1}).
+
+The per capita GDP also can generate differences between provinces.
+No direct relationship can be observed between the GDP and the ratio of incidence but the GDP is a proxy of inobservable information.
+This varaible has been included in similar papers (@wang2020high).
+For low DGPpc is more difficult maintain social distance (sing +); High DGPpc -> more economic activity -> es mas difícl mantener el confinamiento.
+
+Posibles relaciones causales
+Alto GDP -> mas movimiento de la población. Se recibió más gente de otros sitios que podría venir con el virus puesto!
+Alto GDP -> La actividad económica de la industria y la construcción solo se paralizó unos días.
+Aquellas provincias con alto GDP tienen mas astividad industrial/contrucción -> más movimimiento de la población.
+Bajo GDP -> La población con menores ingresos no puede ser tan obediente a la orden de confinamiento (suele ser menos disciplinada).
+
+All those variables will we include in log to capture non-linar realtionship.
+
+<!-- Because COVID-19 is primarily transmitted by respiratory droplets, population density might play a significant role in the acceleration of transmission.
+Puede hacer crecer mas rápidamente la transmision pero no explicar la incidencia   -->
+
+<!-- For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world.
+.
+-->
+
+<!-- Some questions are important to point out.
 Is not evident present a diary association between the temprature (resp.
 humidity) with declarate case.
-There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate
+There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate  -->
 
@@
 
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t$ is a $N \times k_t$ matrix of the $k_t$ independent variables, with associated vector of coefficients $\beta_t$,; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
 
@@
 
-Equation \ref{eq:sur-sim}) can be rewriten in compact form:
+Equation (\ref{eq:sur-sim}) can be rewriten in compact form:
 
 \begin{equation}
-y = X \beta + \epsilon
+\bf{Y} = \bf{X} \beta + \epsilon
+\label{eq:sur-sim-block}
+\end{equation}
+
+where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
+(with $K = \sum_t{k_t}$) and $\epsilon$ is a $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:
+
+\begin{equation}
+E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
 \end{equation}
@@
 
-As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
+As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
 When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
@@
 \label{eq:sur-slm}
-\bf{A}y = X \beta + \epsilon \\
-\epsilon =N(0,\Omega)
+\begin{aligned}
+\bf{AY} = \bf{X} \beta + \epsilon \\
+\epsilon =N(0,\bf{\Sigma})
+\end{aligned}
 \end{equation}
 
-\noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
+\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ with $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.
 
@@
 
-This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
+This specification assumes that incidence in a province ($y_{st}$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
+Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
+It is possible test the null hypothesis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
 The correspond Wald test is available in the `R` package `spsur`.
@@
 Elhorst 2014, Cap.
-4), but those models are more dont take account correlation between errors.
+4), but those models don't take account correlation between errors.
 Therefore, a spatial SUR approach is more reasonable for our purpose.
@@
 grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
+```
+
+```{r plotfer1, echo=FALSE,  warning=FALSE, fig.cap="\\label{Fig-quantile} Spatial distribution of Spanish density population", cache=TRUE}
+q <- quantile(provinces_spain$Density)
+provinces_spain$Quantile<- as.factor((provinces_spain$Density > q[2]) + (provinces_spain$Density > q[3]) +(provinces_spain$Density >= q[4]) + 1)
+p1 <- ggplot(data = provinces_spain) +
+  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
+  theme_bw(base_size=6) +
+  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
+q <- quantile(provinces_spain$Older)
+provinces_spain$Quantile<- as.factor((provinces_spain$Older > q[2]) + (provinces_spain$Older > q[3]) +(provinces_spain$Older >= q[4]) + 1)
+p2 <- ggplot(data = provinces_spain) +
+  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
+  theme_bw(base_size=6) +
+  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
+gridExtra::grid.arrange(p1,p2,nrow=1)
 ```
@@
 
+Limitaciones
+
+* La temperatura puedes ser un factor pero cabría esperar que no tuviera un impacto lineal.
+Por el contrario deberiamos esperar un 'punto de corte': Una temperatua mantenida superior a X grados durante siete días sea la mejor forma de incorporarla al modelo.
+En nuestro modelo (log-log) en incremento en un 1% de la temperatura se asocia con un incremento beta% en la incidencia.
+Esto es lo mismo si la temperatura es baja que si es alta.
+
+* Idem para la humedad y horas de sol
+
+* Los datos son 'provisionales'.
+Hay gran confusión sobre la incidencia real.
+La ausencia de test de diagnóstico PCRs al inicio de la pandemia (también ahora) puede desvirtuar el número de casos dianosticados.
+
+* Los datos oficiales (que tampoco son fiables) son reportados a nivel de Comunidades Autónomas.
+La recopilación de datos a nivel de provincial son el resultado de un esfuerzo colaborativo de recopilacion entre distintas fuentes (principalmente gobiernos locales).
+Nuevamente puede haber sesgos importanes.
+
+* En carácter insular de las Islas Canarias y de las Islas baleares no se ha tenido en cuenta.
+
+* Baleares se ha linkado artificialmente con 3/4 provincias y debería de haberse dejado aislada para respetar su carácter insular al definir la matriz W.
+
+* La incidencia depende de un estado inicial.
+Al inicio del estudio había provincias en las que la epidemia estaba muy desarrollada (Madrid/Alava) mientras que en otras apenas habías casos.
+Este hecho no ha sido considerado en el modelo.
+Decretar el confinamiento debe tener distintos impactos entre provincias.
+QUIZAR METER UNA VARAIBLE DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
+
+* No se ha controlado por el sistema sanitario de cada provincia.
+Uno de los principales focos de contagio han sido los hospitales y los centros de salud.
+En aquellas provincias donde se ha promovido el mensaje "NO IR AL MEDICO" han presentado menor incidencia.
+
+* HAY QUE INCLUIR INTERVALOS DE CONFIANZA EN LOS GRÁFICOS QUE HACEMOS DE LOS COEFICIENTES DEL MODELO
+
+* ¿hasta que punto las variables de control no recojen también factores climáticos? por ejemplo, la gente jóven vive en el sur de España que ha tenido menos incidencia.
+
+* TITULO: A spatial analysis of the environmental correlates ofCOVID-19 incidence in Spain ¿during the lookdown?
+
+* Las estaciones meteorológicas para la obtencion de datos climáticos has sido elegidas aleatoriamente (una para cada provincia).
+otra seleccion puede dar otros resultados.
+AQUI SE PODRÍA HACER EL ESFUERZO DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR PROVINCIA
+
+*IDEM para la humedad. idem para sunshine
+
 Acknowledgments {#acknowledgments .unnumbered}

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Primer empujón variables

@@
   movement hinges in part on the question whether SARS-CoV2 will display
-  seasonality associated with variations in temperature and humidity. In
-  this research, we address this question by means of a spatial analysis
-  of the incidence of COVID-19 in the provinces in Spain. Use of a spatial
-  Seemingly Unrelated Regressions (SUR) approach allows us to model the
-  incidence of reported cases of the disease per 100,000 population, as a
-  function of temperature and humidity, while controlling for GDP per
-  capita, population density, percentage of older adults in the
-  population, and presence of mass transit systems. An interesting aspect
-  of the spatial SUR approach is that it models incidence as a contagion
-  process. Our results indicate that incidence of the disease is lower at
-  higher temperatures. The evidence with respect to humidity is more
-  mixed, with coefficients for this variable that are significant in only
-  some equations. Our control variables also yield interesting insights,
-  as population density and percentage of older adults display negative
-  associations with incidence of COVID-19, whereas the presence of mass
-  transit systems in the province is associated with a greater incidence
-  of the disease.\\
+  seasonality associated with variations in temperature, humidity, and
+  hours of sunshine. In this research, we address this question by means
+  of a spatial analysis of the incidence of COVID-19 in the provinces in
+  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
+  allows us to model the incidence of reported cases of the disease per
+  100,000 population, as a function of temperature and humidity, while
+  controlling for GDP per capita, population density, percentage of older
+  adults in the population, and presence of mass transit systems. An
+  interesting aspect of the spatial SUR approach is that it models
+  incidence as a contagion process. Our results indicate that incidence of
+  the disease is lower at higher temperatures and higher levels of
+  humidity, although coefficients for this variable are significant only
+  in some equations. Sunshine, in contrast, displays a positive
+  association with incidence of the disease. Our control variables also
+  yield interesting insights. Higher incidence is associated with higher
+  GDP per capita and presence of mass transit systems in the province; in
+  contrast, population density and percentage of older adults display
+  negative associations with incidence of COVID-19.\\
   \end{abstract}
@@
 and indeed, the evidence in this respect is inconclusive. At a global
-scale, whereas de Ángel Solá et al.~(2020) see less risk in the Caribean
-Basin, Coelho et al.~(Coelho et al., 2020) warn that at least during the
-exponential phase, expansion of the virus is not driven by climate.
+scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
+however, Coelho et al.~(Coelho et al., 2020) warn that at least during
+the exponential phase, expansion of the virus is not driven by climate.
 Similarly, whereas Araujo and Naimi (2020) argue that spread of
@@
 distribution of the disease on the basis of climate, at least in the
-United States. To further complicate matters, much of the relevant work
-has yet to be peer-reviewed and therefore is open to change (see for
-example the challenge of Harbert et al. (2020) to Araujo and Naimi
-(2020)). In the United States, the National Academy of Sciences,
-Engineering, and Medicine was engaged by the Office of the Executive for
-guidance on this matter (see National Academies of Sciences, Engineering
-and Medicine, 2020). Their conclusion summarizes the situation well (see
-p.~6): ``Some limited data support a potential waning of cases in warmer
-and more humid seasons, yet none are without major
-limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic unfolds
-could shed more light on the effects of climate on transmission.''
+United States. Yao et al.~({\textbf{???}}), examined data from China and
+came to the conclusion that neither temperature nor ultraviolet indices
+had an association with transmission of COVID-19. This is despite
+previous research that has linked less exposure to UVB radiation to
+higher prevalence and severity of acute respiratory tract infections
+(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
+al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
+Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
+much of the relevant work has yet to be peer-reviewed and therefore is
+open to change (see for example the challenge of Harbert et al. (2020)
+to Araujo and Naimi (2020)). In the United States, the National Academy
+of Sciences, Engineering, and Medicine was engaged by the Office of the
+Executive for guidance on this matter (see National Academies of
+Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
+the situation well (see p.~6): ``Some limited data support a potential
+waning of cases in warmer and more humid seasons, yet none are without
+major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
+unfolds could shed more light on the effects of climate on
+transmission.''
 
@@
 to contribute to the knowledge basis regarding the spread of COVID-19
-and the influence of environmental factors, particularly temperature and
-humidity. Hence, here we report results from a spatial analysis of
-incidence of COVID-19 in fifty provinces in Spain. Spain is one of the
-countries hardest hit by the virus, and enacted lockdown measures on
-March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
-We combine data on reported cases of the disease with metereological
-information, to create a spatio-temporal dataset covering a period of 30
-days. We then join this dataset with provincial-level economic and
-demographic information to act as controls to our key environmental
-variables. These data are analyzed using a spatial Seemingly Unrelated
-Regressions (SUR) approach, which allows us to model incidence of
-COVID-19 as a spatial contagion process.
-
-The results provide evidence of the effect of temperature on the
-incidence of COVID-19. The evidence concerning humidity is more mixed:
-while the direction of the effect is negative, as anticipated, the
-coefficients for this variable are only significant in some of equations
-in the system. Our control variables also provide some intriguing
-insights. The results of this analysis provide support to the hypothesis
-of seasonality of the novel SARS-CoV2, and should be of interest to
-public health officials and policy makers wrestling with the question of
-the trajectory of the pandemic.
+and the influence of environmental factors, particularly temperature,
+humidity, and sunshine. We adopt a population health approach, and
+report results from a spatial model of the incidence of COVID-19 in
+fifty provinces in Spain, one of the countries hardest hit by the
+pandemic. We combine data on reported cases of the disease with
+metereological information, to create a spatio-temporal dataset covering
+a period of 30 days. We then join this dataset with provincial-level
+economic and demographic information to act as controls to our key
+environmental variables. These data are analyzed using a spatial
+Seemingly Unrelated Regressions (SUR) approach, which allows us to model
+incidence of COVID-19 as a spatial contagion process.
+
+The results provide evidence of the effect of temperature, humidity, and
+sunshine on the incidence of COVID-19. The clearest result with respect
+to these variables is a lower incidence of COVID-19 at higher
+temperatures and levels of humidity, while the opposite happens with
+respect to hours of sunshine. Our control variables also provide some
+intriguing insights. Higher incidence is associated with higher GDP per
+capita and presence of mass transit systems in the province; in
+contrast, population density and percentage of older adults display
+negative associations with incidence of COVID-19. The results of this
+analysis provide support to the hypothesis of seasonality of the novel
+SARS-CoV2, and should be of interest to public health officials and
+policy makers grappling with the question of the trajectory of the
+pandemic.
 
@@
 
-Explain the rationale for selecting the variables.
-
-For example: The literature about COVID-19 suggested that population
-density is the one of the most important proliferate cause of these
-viscous, however this ill spread with different intensity at big cities
-of the world. Controlling for socioeconomic characteristics the
-objective of this paper is observe the effect of clime on COVID-19
-proliferation.
-
-Some questions are important to point out. Is not evident present a
-diary association between the temprature (resp. humidity) with declarate
-case. There is evidences (ref) that a time lag is necessary between the
-day that the individuo inffect and the case is diagnosticate
+Two set of variables has been selected to explain the diary incidence
+ratio in each Spanish province. The first, are climatic variables
+(temperature, humidity and sunshine) and its impact has been argued in
+the introduction section. The second, is a set of specific
+characteristics of each province. The objective is include in our model
+some control variables (socioeconomic characteristics) to capture the
+probable heterogeneity of each spatial unit. All those variables are
+constant along the temporal period analized in this research. In this
+subsetion we present some evidences of the expected relationship between
+the four control variables considered an the incidence ratio.
+
+The structure of population is a key variable on the incidence. Several
+papers related the relationship between a fast trasnmision of COVID-19
+and population density (e.g.~cumulative incidence in urban areas like
+NYC) and therefore a positive sign will be expected. Note that our
+analysis is at provincial level - not a urban level - and therefore is
+necessary to rethinking the relationship. In Spain the population is
+concentrate (with som exception) in the mediterranean coast (Goerlich
+and Cantarino (2013)), in general with high temperature and high
+humidity. So, positive or negative sign will be expected in the model.
+
+The percentaje of older adults (over 65) is the second variable related
+with the population structure include like control variable. Unlike the
+population density the older people in Spain is concentrate in the core
+of the Iberian Peninsula (see figure \ref{plotfer1}).
+
+The per capita GDP also can generate differences between provinces. No
+direct relationship can be observed between the GDP and the ratio of
+incidence but the GDP is a proxy of inobservable information. This
+varaible has been included in similar papers (Wang et al. (2020)). For
+low DGPpc is more difficult maintain social distance (sing +); High
+DGPpc -\textgreater{} more economic activity -\textgreater{} es mas
+difícl mantener el confinamiento.
+
+Posibles relaciones causales Alto GDP -\textgreater{} mas movimiento de
+la población. Se recibió más gente de otros sitios que podría venir con
+el virus puesto! Alto GDP -\textgreater{} La actividad económica de la
+industria y la construcción solo se paralizó unos días. Aquellas
+provincias con alto GDP tienen mas astividad industrial/contrucción
+-\textgreater{} más movimimiento de la población. Bajo GDP
+-\textgreater{} La población con menores ingresos no puede ser tan
+obediente a la orden de confinamiento (suele ser menos disciplinada).
+
+All those variables will we include in log to capture non-linar
+realtionship.
 
@@
 vector, and in our case \(y_{st}\) is the incidence ratio in the
-province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
-\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
-independent variables, \(X_i=(X^i_{st})\);
+province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\); \(X_t\) is
+a \(N \times k_t\) matrix of the \(k_t\) independent variables, with
+associated vector of coefficients \(\beta_t\),;
 \(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
@@
 
-Equation \ref{eq:sur-sim}) can be rewriten in compact form:
+Equation (\ref{eq:sur-sim}) can be rewriten in compact form:
 
 \begin{equation}
-y = X \beta + \epsilon
+\bf{Y} = \bf{X} \beta + \epsilon
+\label{eq:sur-sim-block}
 \end{equation}
 
+where \(\bf{Y}\) is now a vector of dimension \(NT \times 1\),
+\(\bf{X}\) is a block-diagonal matrix \(NT \times K\) (with
+\(K = \sum_t{k_t}\)) and \(\epsilon\) is a \(NT \times 1\) vector. Using
+the Kronecker product notation (\(\otimes\)), the error matrix structure
+is expressed concisely as:
+
+\begin{equation}
+E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
+\end{equation}
+
 As is the case with cross-sectional data, it is possible to test the
-residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and
-several tests have been developed to test the null hypothesis of spatial
-independence (see López et al., 2014). When the null hypothesis is
-rejected, several alternative specifications have been proposed to
+residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation,
+and several tests have been developed to test the null hypothesis of
+spatial independence (see López et al., 2014). When the null hypothesis
+is rejected, several alternative specifications have been proposed to
 include spatial effects (Anselin, 1988, see also 2016). In this paper we
@@
 \label{eq:sur-slm}
-\bf{A}y = X \beta + \epsilon \\
-\epsilon =N(0,\Omega)
+\begin{aligned}
+\bf{AY} = \bf{X} \beta + \epsilon \\
+\epsilon =N(0,\bf{\Sigma})
+\end{aligned}
 \end{equation}
 
-\noindent where \(A =I_{TN}-\bf{\Lambda} \otimes W\) with
-\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).
-
-This specification assumes that incidence in a province (\(y_st\)) at
+\noindent where \(\bf{A} =I_{TN}-\bf{\Gamma} \otimes W\) with
+\(\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)\).
+
+This specification assumes that incidence in a province (\(y_{st}\)) at
 time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
 of incidence in neighbouring provinces. Coefficients for the spatially
-lagged variable are estimated for each time period \(\lambda_t\) and
+lagged variable are estimated for each time period \(\rho_t\) and
 identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial
-dependence (\(\lambda_i=\lambda_j, \forall i,j\)). The correspond Wald
-test is available in the \texttt{R} package \texttt{spsur}.
+It is possible test the null hypothesis of identtical levels of spatial
+dependence (\(\rho_i=\rho_j, \forall i,j\)). The correspond Wald test is
+available in the \texttt{R} package \texttt{spsur}.
 
@@
 panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
-2014, Cap. 4), but those models are more dont take account correlation
-between errors. Therefore, a spatial SUR approach is more reasonable for
-our purpose.
+2014, Cap. 4), but those models don't take account correlation between
+errors. Therefore, a spatial SUR approach is more reasonable for our
+purpose.
 
@@
 variables by province (maps omits Islas Canarias)}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plotfer1-1.pdf}
+\caption{\label{Fig-quantile} Spatial distribution of Spanish density
+population}
 \end{figure}
@@
 
+Limitaciones
+
+\begin{itemize}
+\item
+  La temperatura puedes ser un factor pero cabría esperar que no tuviera
+  un impacto lineal. Por el contrario deberiamos esperar un `punto de
+  corte': Una temperatua mantenida superior a X grados durante siete
+  días sea la mejor forma de incorporarla al modelo. En nuestro modelo
+  (log-log) en incremento en un 1\% de la temperatura se asocia con un
+  incremento beta\% en la incidencia. Esto es lo mismo si la temperatura
+  es baja que si es alta.
+\item
+  Idem para la humedad y horas de sol
+\item
+  Los datos son `provisionales'. Hay gran confusión sobre la incidencia
+  real. La ausencia de test de diagnóstico PCRs al inicio de la pandemia
+  (también ahora) puede desvirtuar el número de casos dianosticados.
+\item
+  Los datos oficiales (que tampoco son fiables) son reportados a nivel
+  de Comunidades Autónomas. La recopilación de datos a nivel de
+  provincial son el resultado de un esfuerzo colaborativo de
+  recopilacion entre distintas fuentes (principalmente gobiernos
+  locales). Nuevamente puede haber sesgos importanes.
+\item
+  En carácter insular de las Islas Canarias y de las Islas baleares no
+  se ha tenido en cuenta.
+\item
+  Baleares se ha linkado artificialmente con 3/4 provincias y debería de
+  haberse dejado aislada para respetar su carácter insular al definir la
+  matriz W.
+\item
+  La incidencia depende de un estado inicial. Al inicio del estudio
+  había provincias en las que la epidemia estaba muy desarrollada
+  (Madrid/Alava) mientras que en otras apenas habías casos. Este hecho
+  no ha sido considerado en el modelo. Decretar el confinamiento debe
+  tener distintos impactos entre provincias. QUIZAR METER UNA VARAIBLE
+  DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON
+  MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
+\item
+  No se ha controlado por el sistema sanitario de cada provincia. Uno de
+  los principales focos de contagio han sido los hospitales y los
+  centros de salud. En aquellas provincias donde se ha promovido el
+  mensaje ``NO IR AL MEDICO'' han presentado menor incidencia.
+\item
+  HAY QUE INCLUIR INTERVALOS DE CONFIANZA EN LOS GRÁFICOS QUE HACEMOS DE
+  LOS COEFICIENTES DEL MODELO
+\item
+  ¿hasta que punto las variables de control no recojen también factores
+  climáticos? por ejemplo, la gente jóven vive en el sur de España que
+  ha tenido menos incidencia.
+\item
+  TITULO: A spatial analysis of the environmental correlates ofCOVID-19
+  incidence in Spain ¿during the lookdown?
+\item
+  Las estaciones meteorológicas para la obtencion de datos climáticos
+  has sido elegidas aleatoriamente (una para cada provincia). otra
+  seleccion puede dar otros resultados. AQUI SE PODRÍA HACER EL ESFUERZO
+  DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR
+  PROVINCIA
+\end{itemize}
+
+*IDEM para la humedad. idem para sunshine
+
 \hypertarget{acknowledgments}{%
@@
 
+\leavevmode\hypertarget{ref-goerlich2013population}{}%
+Goerlich, F.J., Cantarino, I., 2013. A population density grid for
+spain. International Journal of Geographical Information Science 27,
+2247--2263.
+
 \leavevmode\hypertarget{ref-Gong2020balance}{}%
@@
 spatial econometric perspective. Regional studies 33, 143--156.
+
+\leavevmode\hypertarget{ref-wang2020high}{}%
+Wang, J., Tang, K., Feng, K., Lv, W., 2020. High temperature and high
+humidity reduce the transmission of covid-19. Available at SSRN 3551767.
 

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
seccion variables

@@
 
-Two set of variables has been selected to explain the diary incidence ratio in each Spanish province.
+Two set of variables has been selected to explain the daily incidence ratio in each Spanish province.
 The first, are climatic variables (temperature, humidity and sunshine) and its impact has been argued in the introduction section.
@@
 The objective is include in our model some control variables (socioeconomic characteristics) to capture the probable heterogeneity of each spatial unit.
-All those variables are constant along the temporal period analized in this research.
-In this subsetion we present some evidences of the expected relationship between the four control variables considered an the incidence ratio.
+All those variables are time constant along the temporal period analized in this research but have high spatial variability.
+In this subsetion we present some evidences of the expected relationship between the four control variables and the incidence ratio.
 
@@
 Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship.
-In Spain the population is concentrate (with som exception) in the mediterranean coast (@goerlich2013population), in general with high temperature and high humidity.
+In Spain the population is concentrate (with some exception) in the mediterranean coast (@goerlich2013population), in general with high temperature and high humidity.
 So, positive or negative sign will be expected in the model.
@@
 Unlike the population density the older people in Spain is concentrate in the core of the Iberian Peninsula (see figure \ref{plotfer1}).
-
-The per capita GDP also can generate differences between provinces.
-No direct relationship can be observed between the GDP and the ratio of incidence but the GDP is a proxy of inobservable information.
-This varaible has been included in similar papers (@wang2020high).
+Our hypothesis is that in the provinces with more older adults the social contact is lower (@prem2017projecting) and therefore the a negative correlation will be expected.
+This hypotesis is based in the idea that the agee-structured social contact matrix in Spain is similar to Italy (see @singh2020age Fig 1 pag 3)
+
+The per capita GDP (GDPpc) also can generate differences between provinces.
+No direct relationship can be observed between the GDP and the ratio of incidence but our hypothesis is that the GDPpc is a proxy of inobservable information.
+By example, provinces with high DGPpc have more transit of foreing people This variable has been included in similar papers (@wang2020high).
 For low DGPpc is more difficult maintain social distance (sing +); High DGPpc -> more economic activity -> es mas difícl mantener el confinamiento.
@@
   scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
-gridExtra::grid.arrange(p1,p2,nrow=1)
+q <- quantile(provinces_spain$GDPpc)
+provinces_spain$Quantile<- as.factor((provinces_spain$GDPpc > q[2]) + (provinces_spain$GDPpc > q[3]) +(provinces_spain$GDPpc >= q[4]) + 1)
+p3 <- ggplot(data = provinces_spain) +
+  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
+  theme_bw(base_size=6) +
+  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
+gridExtra::grid.arrange(p1,p2,p3,nrow=2)
 ```

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
impactos

@@
 ---
+
+<!-- load cache -->
+<!-- whoppeR::load_knitr_cache('/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain_cache/latex/') -->
 
@@
   scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
-gridExtra::grid.arrange(p1,p2,p3,nrow=2)
+
+provinces_spain$M.Transit<- as.factor(provinces_spain$Transit)
+p4 <- ggplot(data = provinces_spain) +
+  geom_sf(aes(fill = M.Transit),color = "black",size=.2) +
+  theme_bw(base_size=6) +
+  scale_fill_manual(values=c("white", "gray1"))
+gridExtra::grid.arrange(p1,p2,p3,p4,nrow=2)
 ```
@@
 
+```{r impacts}
+Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
+trsur <- spatialreg::trW(Wsur, type = "MC")
+impacts.sur.slm_lag11 <- impactspsur(sur.slm_lag11, tr = trsur,R = 1000)
+
+impacts.sur.slm_lag11[[1]]
+impacts.sur.slm_lag11[[2]]
+impacts.sur.slm_lag11[[3]]
+
+myimpact <- matrix(0,ncol = 5, nrow = 29)
+for (i in 2:30){
+  myimpact[i-1,] <- impacts.sur.slm_lag11[[i]]$res$direct
+}
+matplot(myimpact, type='l')
+
+```
+
 Concluding Remarks
@@
 
+* HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA DARLE RELEVANCIA A LA ESTIMACION SUR
+
 Acknowledgments {#acknowledgments .unnumbered}

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
impactos

@@
 \bibliographystyle{elsarticle-harv}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 
-Two set of variables has been selected to explain the diary incidence
+Two set of variables has been selected to explain the daily incidence
 ratio in each Spanish province. The first, are climatic variables
@@
 probable heterogeneity of each spatial unit. All those variables are
-constant along the temporal period analized in this research. In this
-subsetion we present some evidences of the expected relationship between
-the four control variables considered an the incidence ratio.
+time constant along the temporal period analized in this research but
+have high spatial variability. In this subsetion we present some
+evidences of the expected relationship between the four control
+variables and the incidence ratio.
 
@@
 necessary to rethinking the relationship. In Spain the population is
-concentrate (with som exception) in the mediterranean coast (Goerlich
+concentrate (with some exception) in the mediterranean coast (Goerlich
 and Cantarino (2013)), in general with high temperature and high
@@
 population density the older people in Spain is concentrate in the core
-of the Iberian Peninsula (see figure \ref{plotfer1}).
-
-The per capita GDP also can generate differences between provinces. No
-direct relationship can be observed between the GDP and the ratio of
-incidence but the GDP is a proxy of inobservable information. This
-varaible has been included in similar papers (Wang et al. (2020)). For
-low DGPpc is more difficult maintain social distance (sing +); High
-DGPpc -\textgreater{} more economic activity -\textgreater{} es mas
-difícl mantener el confinamiento.
+of the Iberian Peninsula (see figure \ref{plotfer1}). Our hypothesis is
+that in the provinces with more older adults the social contact is lower
+(Prem et al. (2017)) and therefore the a negative correlation will be
+expected. This hypotesis is based in the idea that the agee-structured
+social contact matrix in Spain is similar to Italy (see Singh and
+Adhikari (2020) Fig 1 pag 3)
+
+The per capita GDP (GDPpc) also can generate differences between
+provinces. No direct relationship can be observed between the GDP and
+the ratio of incidence but our hypothesis is that the GDPpc is a proxy
+of inobservable information. By example, provinces with high DGPpc have
+more transit of foreing people This variable has been included in
+similar papers (Wang et al. (2020)). For low DGPpc is more difficult
+maintain social distance (sing +); High DGPpc -\textgreater{} more
+economic activity -\textgreater{} es mas difícl mantener el
+confinamiento.
 
@@
 \end{figure}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Wsur <-}\StringTok{ }\KeywordTok{as}\NormalTok{(spdep}\OperatorTok{::}\KeywordTok{listw2mat}\NormalTok{(listw), }\StringTok{"CsparseMatrix"}\NormalTok{)}
+\NormalTok{trsur <-}\StringTok{ }\NormalTok{spatialreg}\OperatorTok{::}\KeywordTok{trW}\NormalTok{(Wsur, }\DataTypeTok{type =} \StringTok{"MC"}\NormalTok{)}
+\NormalTok{impacts.sur.slm_lag11 <-}\StringTok{ }\KeywordTok{impactspsur}\NormalTok{(sur.slm_lag11, }\DataTypeTok{tr =}\NormalTok{ trsur,}\DataTypeTok{R =} \DecValTok{1000}\NormalTok{)}
+
+\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{1}\NormalTok{]]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Impact measures (lag, trace):
+##                                       Direct   Indirect      Total
+## log(GDPpc)_1                       0.6769737  0.6361932  1.3131669
+## log(Older)_1                      -0.8042340 -0.7557874 -1.5600214
+## log(Density)_1                     0.1670754  0.1570109  0.3240864
+## Transit_1                          0.3430969  0.3224289  0.6655259
+## log(Humidity_lag11)_1             -1.5656410 -1.4713276 -3.0369685
+## log(Mean_Temp_lag11)_1            -1.7452380 -1.6401057 -3.3853437
+## log(Sunshine_Hours_lag11 + 0.1)_1 -0.2818768 -0.2648967 -0.5467734
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{2}\NormalTok{]]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Impact measures (lag, trace):
+##                                        Direct    Indirect      Total
+## log(Density)_2                     0.06263494  0.04033489  0.1029698
+## Transit_2                          0.44110065  0.28405466  0.7251553
+## log(Humidity_lag11)_2             -1.06494878 -0.68579283 -1.7507416
+## log(Mean_Temp_lag11)_2            -1.48108018 -0.95376810 -2.4348483
+## log(Sunshine_Hours_lag11 + 0.1)_2 -0.19117818 -0.12311261 -0.3142908
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{3}\NormalTok{]]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Impact measures (lag, trace):
+##                                        Direct    Indirect       Total
+## log(Density)_3                     0.02434857  0.01918949  0.04353807
+## Transit_3                          0.43185937  0.34035519  0.77221456
+## log(Humidity_lag11)_3             -0.97725083 -0.77018681 -1.74743764
+## log(Mean_Temp_lag11)_3            -1.39946579 -1.10294110 -2.50240689
+## log(Sunshine_Hours_lag11 + 0.1)_3 -0.09717791 -0.07658744 -0.17376535
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{myimpact <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DataTypeTok{ncol =} \DecValTok{5}\NormalTok{, }\DataTypeTok{nrow =} \DecValTok{29}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{2}\OperatorTok{:}\DecValTok{30}\NormalTok{)\{}
+\NormalTok{  myimpact[i}\DecValTok{-1}\NormalTok{,] <-}\StringTok{ }\NormalTok{impacts.sur.slm_lag11[[i]]}\OperatorTok{$}\NormalTok{res}\OperatorTok{$}\NormalTok{direct}
+\NormalTok{\}}
+\KeywordTok{matplot}\NormalTok{(myimpact, }\DataTypeTok{type=}\StringTok{'l'}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/impacts-1.pdf}
 
@@
 
+\leavevmode\hypertarget{ref-prem2017projecting}{}%
+Prem, K., Cook, A.R., Jit, M., 2017. Projecting social contact matrices
+in 152 countries using contact surveys and demographic data. PLoS
+computational biology 13, e1005697.
+
 \leavevmode\hypertarget{ref-Rey1999us}{}%
@@
 spatial econometric perspective. Regional studies 33, 143--156.
+
+\leavevmode\hypertarget{ref-singh2020age}{}%
+Singh, R., Adhikari, R., 2020. Age-structured impact of social
+distancing on the covid-19 epidemic in india. arXiv preprint
+arXiv:2003.12055.
 

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Continued editing section on selection of variables.

--- 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 
->https://github.com/paezha/covid19-environmental-correlates
+https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
 
@@
 
-Two set of variables has been selected to explain the diary incidence ratio in each Spanish province.
-The first, are climatic variables (temperature, humidity and sunshine) and its impact has been argued in the introduction section.
-The second, is a set of specific characteristics of each province.
-The objective is include in our model some control variables (socioeconomic characteristics) to capture the probable heterogeneity of each spatial unit.
-All those variables are constant along the temporal period analized in this research.
-In this subsetion we present some evidences of the expected relationship between the four control variables considered an the incidence ratio.
+The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
+In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
+Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
+As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
+These variables have been implicated in the viability and ease of transmission of similar viruses.
+In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.
+
+The first variable that we consider is GDP per capita.
+Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
+A18)}.
+The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when its cause may still be unclear [@Zhou2016accelerated].
+In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] remarks how Toronto's status as a global city was a vulnerability.
+In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and their potential to be further ahead in the trajectory of the pandemic.
+Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional].
+Therefore, it is possible that wealthier regions remained relatively more active even during the lockdown.
+On the other hand, it is also possible the less wealthy regions have a higher proportion of workers in manual occupations that cannot telework.
+
+The percentaje of older adults (over 65) is the second variable that we consider as a control.
+Early evidence suggest that the case rate mortality is higher at older ages [e.g.
+@Novel2020epidemiological].
+It is not clear that presence of older adults necessarily translates into higher transmission in the population.
+The public health tool of choice in containing the spread of the disease has been social distancing.
+In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
+In other words, older adults are, by preference or necessity, already in a form of social isolation.
+Social distancing during the pandemic may actually reinforce that condition for older adults, as the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age] suggests.
+Accordingly, our expectation is that in the provinces with higher proportions of older adults will tend to have similarl levels of lower social contact, particularly since age-structured social contact matrix in Spain is similar to Italy [@Prem2017projecting].
 
@@
 Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship.
-In Spain the population is concentrate (with som exception) in the mediterranean coast (@goerlich2013population), in general with high temperature and high humidity.
+In Spain the population is concentrate (with some exception) in the mediterranean coast (@Goerlich2013population), in general with high temperature and high humidity.
 So, positive or negative sign will be expected in the model.
 
-The percentaje of older adults (over 65) is the second variable related with the population structure include like control variable.
-Unlike the population density the older people in Spain is concentrate in the core of the Iberian Peninsula (see figure \ref{plotfer1}).
-
-The per capita GDP also can generate differences between provinces.
-No direct relationship can be observed between the GDP and the ratio of incidence but the GDP is a proxy of inobservable information.
-This varaible has been included in similar papers (@wang2020high).
-For low DGPpc is more difficult maintain social distance (sing +); High DGPpc -> more economic activity -> es mas difícl mantener el confinamiento.
-
-Posibles relaciones causales
-Alto GDP -> mas movimiento de la población. Se recibió más gente de otros sitios que podría venir con el virus puesto!
-Alto GDP -> La actividad económica de la industria y la construcción solo se paralizó unos días.
-Aquellas provincias con alto GDP tienen mas astividad industrial/contrucción -> más movimimiento de la población.
-Bajo GDP -> La población con menores ingresos no puede ser tan obediente a la orden de confinamiento (suele ser menos disciplinada).
-
 All those variables will we include in log to capture non-linar realtionship.
 
-<!-- Because COVID-19 is primarily transmitted by respiratory droplets, population density might play a significant role in the acceleration of transmission.
-Puede hacer crecer mas rápidamente la transmision pero no explicar la incidencia   -->
-
-<!-- For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world.
-.
--->
-
-<!-- Some questions are important to point out.
-Is not evident present a diary association between the temprature (resp.
-humidity) with declarate case.
-There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate  -->
+Transit.
 
@@
 week11.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  #filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
   ggtitle("March 13 - March 15") +
-  theme_tufte() +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
@@
   geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 4.9, 
+               x = madrid_xy[1] - 8, 
                y = madrid_xy[2] + 0.6,
                family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
+  geom_segment(x = madrid_xy[1] - 6, 
                y = madrid_xy[2] + 0.5, 
@@
 week12.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  #filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
   ggtitle("March 16 - March 22") +
-  theme_tufte() +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
@@
         plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 4.9, y = madrid_xy[2] + 0.6) +
-  geom_segment(x = madrid_xy[1] - 3.4, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
+  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 8, y = madrid_xy[2] + 0.6) +
+  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
 
 week13.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  #filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
   ggtitle("March 23 - March 29") +
-  theme_tufte() +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
@@
   geom_sf_text(label = "Madrid",
-               x = madrid_xy[1] - 4.9, 
+               x = madrid_xy[1] - 8, 
                y = madrid_xy[2] + 0.6,
                family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
+  geom_segment(x = madrid_xy[1] - 6, 
                y = madrid_xy[2] + 0.5, 
@@
 week14.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  #filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
   ggtitle("March 30 - April 5") +
-  theme_tufte() +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
@@
 week15.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  #filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
   ggtitle("April 6 - April 11") +
-  theme_tufte() +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
@@
 
-```{r control-map, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
+```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
+# GDP per capita:
+q <- quantile(provinces_spain$GDPpc)
 gdppc.plot <- provinces_spain %>%
-  filter(CCAA != "Canarias") %>%
+  mutate(GDPpc = GDPpc,
+         GDPpc = case_when(GDPpc < q[2] ~ "1st",
+                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
+                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
+                             GDPpc > q[4] ~ "4th")) %>%
+  #filter(CCAA != "Canarias") %>%
   ggplot() +
   geom_sf(aes(fill = GDPpc)) +
-  scale_fill_distiller(name = "GDP per capita", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
+  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
-        legend.position = "bottom")
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 8, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 # Older adults:
+q <- quantile(provinces_spain$Older)
 older.plot <- provinces_spain %>% 
-  filter(CCAA != "Canarias") %>%
+  mutate(Older = Older,
+         Older = case_when(Older < q[2] ~ "1st",
+                             Older > q[2] & Older <= q[3]  ~ "2nd",
+                             Older > q[3] & Older <= q[4] ~ "3rd",
+                             Older > q[4] ~ "4th")) %>%
+  #filter(CCAA != "Canarias") %>%
   ggplot() +
-  geom_sf(aes(fill = as.numeric(Older))) +
-  scale_fill_distiller(name = "Percentage of older adults", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
+  geom_sf(aes(fill = Older)) +
+  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
-        legend.position = "bottom")
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 8, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 # Population density:
+q <- as.numeric(quantile(provinces_spain$Density))
 density.plot <- provinces_spain %>% 
-  filter(CCAA != "Canarias") %>%
+  mutate(Density = as.numeric(Density),
+         Density = case_when(Density < q[2] ~ "1st",
+                             Density > q[2] & Density <= q[3]  ~ "2nd",
+                             Density > q[3] & Density <= q[4] ~ "3rd",
+                             Density > q[4] ~ "4th")) %>%
+  #filter(CCAA != "Canarias") %>%
   ggplot() +
-  geom_sf(aes(fill = as.numeric(Density))) +
-  scale_fill_distiller(name = "Population Density", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
+  geom_sf(aes(fill = Density)) +
+  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
-        legend.position = "bottom")
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 8, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
-```
-
-```{r plotfer1, echo=FALSE,  warning=FALSE, fig.cap="\\label{Fig-quantile} Spatial distribution of Spanish density population", cache=TRUE}
-q <- quantile(provinces_spain$Density)
-provinces_spain$Quantile<- as.factor((provinces_spain$Density > q[2]) + (provinces_spain$Density > q[3]) +(provinces_spain$Density >= q[4]) + 1)
-p1 <- ggplot(data = provinces_spain) +
-  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
-  theme_bw(base_size=6) +
-  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
-q <- quantile(provinces_spain$Older)
-provinces_spain$Quantile<- as.factor((provinces_spain$Older > q[2]) + (provinces_spain$Older > q[3]) +(provinces_spain$Older >= q[4]) + 1)
-p2 <- ggplot(data = provinces_spain) +
-  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
-  theme_bw(base_size=6) +
-  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
-gridExtra::grid.arrange(p1,p2,nrow=1)
 ```

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Continued editing section on selection of variables.

--- 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
@@
   contrast, population density and percentage of older adults display
-  negative associations with incidence of COVID-19.\\
+  negative associations with incidence of COVID-19.
   \end{abstract}
@@
 distribution of the disease on the basis of climate, at least in the
-United States. Yao et al.~({\textbf{???}}), examined data from China and
-came to the conclusion that neither temperature nor ultraviolet indices
-had an association with transmission of COVID-19. This is despite
-previous research that has linked less exposure to UVB radiation to
-higher prevalence and severity of acute respiratory tract infections
+United States. Yao et al.~(2020), examined data from China and came to
+the conclusion that neither temperature nor ultraviolet indices had an
+association with transmission of COVID-19. This is despite previous
+research that has linked less exposure to UVB radiation to higher
+prevalence and severity of acute respiratory tract infections
 (Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
@@
 waning of cases in warmer and more humid seasons, yet none are without
-major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
+major limitations\ldots Additional studies as the SARS-CoV-2 pandemic
 unfolds could shed more light on the effects of climate on
@@
 
-\begin{quote}
-https://github.com/paezha/covid19-environmental-correlates
-\end{quote}
+https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
 
@@
 
-Two set of variables has been selected to explain the diary incidence
-ratio in each Spanish province. The first, are climatic variables
-(temperature, humidity and sunshine) and its impact has been argued in
-the introduction section. The second, is a set of specific
-characteristics of each province. The objective is include in our model
-some control variables (socioeconomic characteristics) to capture the
-probable heterogeneity of each spatial unit. All those variables are
-constant along the temporal period analized in this research. In this
-subsetion we present some evidences of the expected relationship between
-the four control variables considered an the incidence ratio.
+The global emergence of infectious diseases is mostly driven by
+environmental, ecological, and socio-economic factors (Jones et al.,
+2008). In the case of SARS-CoV2, the ecological factors include the
+interaction between humans and wildlife. Once transmission of a disease
+begins to happen between humans, socio-economic and environmental
+factors become increasingly important. As noted in the introduction, the
+focus of the paper is on environmental variables, namely temperature,
+humidity, and sunshine. These variables have been implicated in the
+viability and ease of transmission of similar viruses. In addition to
+these variables, we also aim to use a set of controls, in the form of
+specific socio-economic and demographic characteristics of each
+province.
+
+The first variable that we consider is GDP per capita. Much has been
+said about globalization and the spread of infectious
+disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
+A18)}.
+The growth in global connections has presented a challenge to spatial
+approaches in the initial stages of disease management, when its cause
+may still be unclear (Zhou and Coleman, 2016). In reference to the
+earlier outbreak of SARS, van Wagner (Van Wagner, 2008) remarks how
+Toronto's status as a global city was a vulnerability. In our case, we
+think of GDP per capita as a marker of a region's relative position in a
+network of global cities, and their potential to be further ahead in the
+trajectory of the pandemic. Furthermore, wealthier regions also tend to
+concentrate more activities that produce non-traded goods, including
+building and construction (Hallet, 2002). Therefore, it is possible that
+wealthier regions remained relatively more active even during the
+lockdown. On the other hand, it is also possible the less wealthy
+regions have a higher proportion of workers in manual occupations that
+cannot telework.
+
+The percentaje of older adults (over 65) is the second variable related
+with the population structure include like control variable. Unlike the
+population density the older people in Spain is concentrate in the core
+of the Iberian Peninsula (see figure \ref{plotfer1}). Our hypothesis is
+that in the provinces with more older adults the social contact is lower
+(Prem et al. (2017)) and therefore the a negative correlation will be
+expected. This hypotesis is based in the idea that the agee-structured
+social contact matrix in Spain is similar to Italy (see Singh and
+Adhikari (2020) Fig 1 pag 3)
 
@@
 
-The percentaje of older adults (over 65) is the second variable related
-with the population structure include like control variable. Unlike the
-population density the older people in Spain is concentrate in the core
-of the Iberian Peninsula (see figure \ref{plotfer1}).
-
-The per capita GDP also can generate differences between provinces. No
-direct relationship can be observed between the GDP and the ratio of
-incidence but the GDP is a proxy of inobservable information. This
-varaible has been included in similar papers (Wang et al. (2020)). For
-low DGPpc is more difficult maintain social distance (sing +); High
-DGPpc -\textgreater{} more economic activity -\textgreater{} es mas
-difícl mantener el confinamiento.
-
-Posibles relaciones causales Alto GDP -\textgreater{} mas movimiento de
-la población. Se recibió más gente de otros sitios que podría venir con
-el virus puesto! Alto GDP -\textgreater{} La actividad económica de la
-industria y la construcción solo se paralizó unos días. Aquellas
-provincias con alto GDP tienen mas astividad industrial/contrucción
--\textgreater{} más movimimiento de la población. Bajo GDP
--\textgreater{} La población con menores ingresos no puede ser tan
-obediente a la orden de confinamiento (suele ser menos disciplinada).
-
 All those variables will we include in log to capture non-linar
 realtionship.
+
+Transit.
 
@@
 
+\leavevmode\hypertarget{ref-Hallet2002regional}{}%
+Hallet, M., 2002. Regional specialisation and concentration in the eu,
+in: Regional Convergence in the European Union. Springer, pp. 53--76.
+
 \leavevmode\hypertarget{ref-Harbert2020spatial}{}%
@@
 influenza in cold climate. Environmental Health 13, 22.
+
+\leavevmode\hypertarget{ref-Jones2008global}{}%
+Jones, K.E., Patel, N.G., Levy, M.A., Storeygard, A., Balk, D.,
+Gittleman, J.L., Daszak, P., 2008. Global trends in emerging infectious
+diseases. Nature 451, 990--993.
+doi:\href{https://doi.org/10.1038/nature06536}{10.1038/nature06536}
 
@@
 
-\leavevmode\hypertarget{ref-wang2020high}{}%
-Wang, J., Tang, K., Feng, K., Lv, W., 2020. High temperature and high
-humidity reduce the transmission of covid-19. Available at SSRN 3551767.
+\leavevmode\hypertarget{ref-singh2020age}{}%
+Singh, R., Adhikari, R., 2020. Age-structured impact of social
+distancing on the covid-19 epidemic in india. arXiv preprint
+arXiv:2003.12055.
+
+\leavevmode\hypertarget{ref-vanWagner2008toward}{}%
+Van Wagner, E., 2008. Toward a dialectical understanding of networked
+disease in the global city: Vulnerability, connectivity, topologies.
+Networked disease: Emerging infections in the global city 13--26.
 
@@
 
+\leavevmode\hypertarget{ref-Yao2020association}{}%
+Yao, Y., Pan, J., Liu, Z., Meng, X., Wang, W., Kan, H., Wang, W., 2020.
+No association of covid-19 transmission with temperature or uv radiation
+in chinese cities. European Respiratory Journal 2000517.
+doi:\href{https://doi.org/10.1183/13993003.00517-2020}{10.1183/13993003.00517-2020}
+
 \leavevmode\hypertarget{ref-Zellner1962efficient}{}%
@@
 
+\leavevmode\hypertarget{ref-Zhou2016accelerated}{}%
+Zhou, Y.R., Coleman, W.D., 2016. Accelerated contagion and response:
+Understanding the relationships among globalization, time, and disease.
+Globalizations 13, 285--299.
+
 \leavevmode\hypertarget{ref-Zhu2019}{}%

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex

@@
 
-The percentaje of older adults (over 65) is the second variable related
-with the population structure include like control variable. Unlike the
-population density the older people in Spain is concentrate in the core
-of the Iberian Peninsula (see figure \ref{plotfer1}). Our hypothesis is
-that in the provinces with more older adults the social contact is lower
-(Prem et al. (2017)) and therefore the a negative correlation will be
-expected. This hypotesis is based in the idea that the agee-structured
-social contact matrix in Spain is similar to Italy (see Singh and
-Adhikari (2020) Fig 1 pag 3)
+The percentaje of older adults (over 65) is the second variable that we
+consider as a control. Early evidence suggest that the case rate
+mortality is higher at older ages (e.g.~The Novel Coronavirus Pneumonia
+Emergency Response Epidemiology Team, 2020). It is not clear that
+presence of older adults necessarily translates into higher transmission
+in the population. The public health tool of choice in containing the
+spread of the disease has been social distancing. In this respect, the
+evidence from the field of transportation is that older adults tend to
+travel less frequently, for shorter distances, and have higher rates of
+immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder and
+Pinjari, 2012). In other words, older adults are, by preference or
+necessity, already in a form of social isolation. Social distancing
+during the pandemic may actually reinforce that condition for older
+adults, as the analysis of age-structured social contact in India,
+China, and Italy of Sing and Adhikari (2020) suggests. Accordingly, our
+expectation is that in the provinces with higher proportions of older
+adults will tend to have similarl levels of lower social contact,
+particularly since age-structured social contact matrix in Spain is
+similar to Italy (Prem et al., 2017).
 
@@
 
+\begin{verbatim}
+## Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
+## give correct results for longitude/latitude data
+
+## Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
+## give correct results for longitude/latitude data
+
+## Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
+## give correct results for longitude/latitude data
+\end{verbatim}
+
 \begin{figure}
 \centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/control-map-1.pdf}
-\caption{\label{fig:control-map}Spatial distribution of control
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/map-controls-1.pdf}
+\caption{\label{fig:map-controls}Spatial distribution of control
 variables by province (maps omits Islas Canarias)}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plotfer1-1.pdf}
-\caption{\label{Fig-quantile} Spatial distribution of Spanish density
-population}
 \end{figure}
@@
 \end{figure}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Wsur <-}\StringTok{ }\KeywordTok{as}\NormalTok{(spdep}\OperatorTok{::}\KeywordTok{listw2mat}\NormalTok{(listw), }\StringTok{"CsparseMatrix"}\NormalTok{)}
+\NormalTok{trsur <-}\StringTok{ }\NormalTok{spatialreg}\OperatorTok{::}\KeywordTok{trW}\NormalTok{(Wsur, }\DataTypeTok{type =} \StringTok{"MC"}\NormalTok{)}
+\NormalTok{impacts.sur.slm_lag11 <-}\StringTok{ }\KeywordTok{impactspsur}\NormalTok{(sur.slm_lag11, }\DataTypeTok{tr =}\NormalTok{ trsur,}\DataTypeTok{R =} \DecValTok{1000}\NormalTok{)}
+
+\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{1}\NormalTok{]]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Impact measures (lag, trace):
+##                                       Direct   Indirect      Total
+## log(GDPpc)_1                       0.6736035  0.6395634  1.3131669
+## log(Older)_1                      -0.8002302 -0.7597912 -1.5600214
+## log(Density)_1                     0.1662437  0.1578427  0.3240864
+## Transit_1                          0.3413888  0.3241370  0.6655259
+## log(Humidity_lag11)_1             -1.5578465 -1.4791220 -3.0369685
+## log(Mean_Temp_lag11)_1            -1.7365494 -1.6487943 -3.3853437
+## log(Sunshine_Hours_lag11 + 0.1)_1 -0.2804735 -0.2663000 -0.5467734
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{2}\NormalTok{]]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Impact measures (lag, trace):
+##                                        Direct    Indirect      Total
+## log(Density)_2                     0.06248795  0.04048188  0.1029698
+## Transit_2                          0.44006548  0.28508982  0.7251553
+## log(Humidity_lag11)_2             -1.06244958 -0.68829202 -1.7507416
+## log(Mean_Temp_lag11)_2            -1.47760442 -0.95724386 -2.4348483
+## log(Sunshine_Hours_lag11 + 0.1)_2 -0.19072953 -0.12356126 -0.3142908
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{3}\NormalTok{]]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Impact measures (lag, trace):
+##                                        Direct    Indirect       Total
+## log(Density)_3                     0.02426235  0.01927572  0.04353807
+## Transit_3                          0.43033000  0.34188455  0.77221455
+## log(Humidity_lag11)_3             -0.97379005 -0.77364759 -1.74743764
+## log(Mean_Temp_lag11)_3            -1.39450981 -1.10789707 -2.50240689
+## log(Sunshine_Hours_lag11 + 0.1)_3 -0.09683377 -0.07693158 -0.17376535
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{myimpact <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DataTypeTok{ncol =} \DecValTok{5}\NormalTok{, }\DataTypeTok{nrow =} \DecValTok{29}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{2}\OperatorTok{:}\DecValTok{30}\NormalTok{)\{}
+\NormalTok{  myimpact[i}\DecValTok{-1}\NormalTok{,] <-}\StringTok{ }\NormalTok{impacts.sur.slm_lag11[[i]]}\OperatorTok{$}\NormalTok{res}\OperatorTok{$}\NormalTok{direct}
+\NormalTok{\}}
+\KeywordTok{matplot}\NormalTok{(myimpact, }\DataTypeTok{type=}\StringTok{'l'}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/impacts-1.pdf}
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA
+  DARLE RELEVANCIA A LA ESTIMACION SUR
+\end{itemize}
+
 \hypertarget{acknowledgments}{%
@@
 
-\leavevmode\hypertarget{ref-goerlich2013population}{}%
+\leavevmode\hypertarget{ref-Goerlich2013population}{}%
 Goerlich, F.J., Cantarino, I., 2013. A population density grid for
@@
 
+\leavevmode\hypertarget{ref-Morency2011distance}{}%
+Morency, C., Páez, A., Roorda, M.J., Mercado, R.G., Farber, S., 2011.
+Distance traveled in three canadian cities: Spatial analysis from the
+perspective of vulnerable population segments. Journal of Transport
+Geography 19, 39--50.
+
 \leavevmode\hypertarget{ref-National2020rapid}{}%
@@
 
-\leavevmode\hypertarget{ref-prem2017projecting}{}%
+\leavevmode\hypertarget{ref-Prem2017projecting}{}%
 Prem, K., Cook, A.R., Jit, M., 2017. Projecting social contact matrices
@@
 
-\leavevmode\hypertarget{ref-singh2020age}{}%
+\leavevmode\hypertarget{ref-Roorda2010trip}{}%
+Roorda, M.J., Paez, A., Morency, C., Mercado, R., Farber, S., 2010. Trip
+generation of vulnerable populations in three canadian cities: A spatial
+ordered probit approach. Transportation 37, 525--548.
+doi:\href{https://doi.org/10.1007/s11116-010-9263-3}{10.1007/s11116-010-9263-3}
+
+\leavevmode\hypertarget{ref-Sikder2012immobility}{}%
+Sikder, S., Pinjari, A.R., 2012. Immobility levels and mobility
+preferences of the elderly in the united states evidence from 2009
+national household travel survey. Transportation Research Record
+137--147. doi:\href{https://doi.org/10.3141/2318-16}{10.3141/2318-16}
+
+\leavevmode\hypertarget{ref-Singh2020age}{}%
 Singh, R., Adhikari, R., 2020. Age-structured impact of social
@@
 arXiv:2003.12055.
+
+\leavevmode\hypertarget{ref-Novel2020epidemiological}{}%
+The Novel Coronavirus Pneumonia Emergency Response Epidemiology Team,
+2020. The epidemiological characteristics of an outbreak of 2019 novel
+coronavirus diseases (covid-19)---china, 2020. China CDC Weekly 2,
+113--122.
 

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Finished section on selection of variables

@@
 
-The structure of population is a key variable on the incidence. Several
-papers related the relationship between a fast trasnmision of COVID-19
+The third population variable the we consider is population density,
+since it directly affects the contact patterns and contact rates between
+individuals in a population (Hu et al., 2013). The evidence available
+suggests a positive relationship between the transmission of COVID-19
 and population density (e.g.~cumulative incidence in urban areas like
-NYC) and therefore a positive sign will be expected. Note that our
-analysis is at provincial level - not a urban level - and therefore is
-necessary to rethinking the relationship. In Spain the population is
-concentrate (with some exception) in the mediterranean coast (Goerlich
-and Cantarino (2013)), in general with high temperature and high
-humidity. So, positive or negative sign will be expected in the model.
-
-All those variables will we include in log to capture non-linar
-realtionship.
-
-Transit.
+NYC). For this reason, we anticipate a positive relationship between
+population density and the incidence of the disease.
+
+The last control variable is the presence of mass transit systems in a
+province. Every province in Spain offers some form of public
+transportation, however only five provinces have higher order systems of
+mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
+Public transportation has been hypothesized to relate to the spread of
+contagious disease by some researchers using agent-based approaches
+(Perez and Dragicevic, 2009; Wang et al., 2010), and while we find scant
+evidence of a link in the literature, the idea is intuitively appealing.
+After all, unlike the isolation that a car offers to travellers, most
+mass transit system are cauldrons of social contact.
 
@@
 \caption{\label{fig:weekly-average-incidence-map}Mean weekly incidence
-of COVID-19 by province, in reported cases by 100,000 people (map omits
-Islas Canarias)}
+of COVID-19 by province, in reported cases by 100,000 people}
 \end{figure}
@@
 
-\begin{verbatim}
-## Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
-## give correct results for longitude/latitude data
-
-## Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
-## give correct results for longitude/latitude data
-
-## Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
-## give correct results for longitude/latitude data
-\end{verbatim}
-
 \begin{figure}
@@
 \caption{\label{fig:map-controls}Spatial distribution of control
-variables by province (maps omits Islas Canarias)}
+variables by province}
 \end{figure}
@@
 
-\leavevmode\hypertarget{ref-Goerlich2013population}{}%
-Goerlich, F.J., Cantarino, I., 2013. A population density grid for
-spain. International Journal of Geographical Information Science 27,
-2247--2263.
-
 \leavevmode\hypertarget{ref-Gong2020balance}{}%
@@
 distributions on the basis of climate in the united states. medRxiv.
+
+\leavevmode\hypertarget{ref-Hu2013scaling}{}%
+Hu, H., Nigmatulina, K., Eckhoff, P., 2013. The scaling of contact rates
+with population density for the infectious disease models. Mathematical
+Biosciences 244, 125--134.
+doi:\href{https://doi.org/https://doi.org/10.1016/j.mbs.2013.04.013}{https://doi.org/10.1016/j.mbs.2013.04.013}
 
@@
 
+\leavevmode\hypertarget{ref-Perez2009agent}{}%
+Perez, L., Dragicevic, S., 2009. An agent-based approach for modeling
+dynamics of contagious disease spread. International journal of health
+geographics 8, 50.
+
 \leavevmode\hypertarget{ref-Prem2017projecting}{}%
@@
 Networked disease: Emerging infections in the global city 13--26.
+
+\leavevmode\hypertarget{ref-Wang2010gis}{}%
+Wang, J., Xiong, J., Yang, K., Peng, S., Xu, Q., 2010. Use of gis and
+agent-based modeling to simulate the spread of influenza, in: 2010 18th
+International Conference on Geoinformatics. IEEE, pp. 1--6.
 

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd

@@
 
-The structure of population is a key variable on the incidence.
-Several papers related the relationship between a fast trasnmision of COVID-19 and population density (e.g.
-cumulative incidence in urban areas like NYC) and therefore a positive sign will be expected.
-Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship.
-In Spain the population is concentrate (with some exception) in the mediterranean coast (@Goerlich2013population), in general with high temperature and high humidity.
-So, positive or negative sign will be expected in the model.
-
-All those variables will we include in log to capture non-linar realtionship.
-
-Transit.
+The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
+The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g.
+cumulative incidence in urban areas like NYC).
+For this reason, we anticipate a positive relationship between population density and the incidence of the disease.
+
+The last control variable is the presence of mass transit systems in a province.
+Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
+Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches [@Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
+After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.
 
@@
 
-```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE, cache = TRUE}
+```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people", warning=FALSE, cache = TRUE}
 # Retrieve the coordinates of Madrid
@@
 
-```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
+```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province", warning=FALSE, cache = TRUE}
 # GDP per capita:
@@
                         levels = c("Model1", "Model2", "Model3"),
-                        labels = c("Model 1", "Model 2", "Model 3"))) %>%
+                        labels = c("Model 1: lag8", "Model 2: lag11", "Model 3: lag11w"))) %>%
   ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
   geom_point(size = 2) +
-  scale_color_manual(values = c("Model 1" = "blue", "Model 2" = "orange", "Model 3" = "green") ) + 
+  scale_color_manual(values = c("Model 1: lag8" = "blue", "Model 2: lag11" = "orange", "Model 3: lag11w" = "green") ) + 
   annotate(geom = "text", 
-           label = c(paste("Model 1: pooled-R2=", 
+           label = c(paste("pooled-R2 (Model 1: lag8)=", 
                            round(sur.slm_lag8$R2[1], 4)),
-                     paste("Model 2: pooled-R2=", 
+                     paste("pooled-R2 (Model 2: lag11)=", 
                            round(sur.slm_lag11$R2[1], 4)),
-                     paste("Model 3: pooled-R2=", 
+                     paste("pooled-R2 (Model 3: lag11w)=", 
                            round(sur.slm_lag11w$R2[1], 4))),
@@
 
-```{r impacts}
+```{r impacts, echo=FALSE}
 Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Finished editing sections Data and Methods and improved the plots for the results

@@
 For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**].
-This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments.
+The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España.
+This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
 In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food).
+The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food).
 The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute).
@@
                   sd(covid19_spain$Humidity)),
-           Source = c("Montera34",
+           Source = c("ProvidencialData19",
                       "INE",
@@
   column_spec(7, width = "5em") %>%
-  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
+  footnote(general = c("ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19", 
                        "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
@@
 
-where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
-(with $K = \sum_t{k_t}$) and $\epsilon$ is a $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:
+\noindent where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
+(with $K = \sum_t{k_t}$) and $\epsilon$ is an $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:
 
@@
 ![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
+
+\noindent where ![$\bf{Y}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BY%7D%24) is now a vector of dimension ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24), ![$\bf{X}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BX%7D%24) is a block-diagonal matrix ![$NT \times K$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%20K%24)
+(with ![$K = \sum_t{k_t}$](https://render.githubusercontent.com/render/math?math=%24K%20%3D%20%5Csum_t%7Bk_t%7D%24)) and ![$\epsilon$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon%24) is an ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24) vector.
+Using the Kronecker product notation (![$\otimes$](https://render.githubusercontent.com/render/math?math=%24%5Cotimes%24)), the error matrix structure is expressed concisely as:
+
+![\begin{equation} E\[\epsilon \epsilon'\]=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'}) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5CSigma%20%5Cotimes%20I_N%20%3B%20%5C%20%5CSigma%3D(%5Csigma_%7Btt'%7D)%20%5Cend%7Bequation%7D)
 -->
@@
 
-The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:
+The stack expresion for the SUR model with a spatially lagged dependent variable (SUR-SLM) is as follows:
 
@@
 
-\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ with $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.
+\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ is the spatially lagged dependent variable, and $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.
 
 <!--
-  ![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
-
-\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
+![\begin{equation} \label{eq:sur-slm} \begin{aligned} \bf{AY} = \bf{X} \beta + \epsilon \\ \epsilon =N(0,\bf{\Sigma}) \end{aligned} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbegin%7Baligned%7D%20%5Cbf%7BAY%7D%20%3D%20%5Cbf%7BX%7D%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5Cbf%7B%5CSigma%7D)%20%5Cend%7Baligned%7D%20%5Cend%7Bequation%7D)
+
+\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
 -->
 
-This specification assumes that incidence in a province ($y_{st}$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypothesis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
+This specification assumes that outcome ($y_{st}$) at location $s$ and time $t$ is partially determined by the weighted average ($Wy_{st}$) of the outcome in neighboring provinces, with neighborhood defined by matrix $W$ of spatial weights.
+In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way.
+The coefficients of the spatially lagged variable are estimated for each time period $\rho_t$ and identify the intensity and the sign of the contagion effect.
+It is possible test the null hypothesis of identical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
 The correspond Wald test is available in the `R` package `spsur`.
@@
 <!--
-This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period ![$\lambda_t$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence (![$\lambda_i=\lambda_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_i%3D%5Clambda_j%2C%20%5Cforall%20i%2Cj%24)).
+This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights.
+In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way.
+The coefficients of the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identify the intensity and the sign of the contagion effect.
+It is possible test the null hypothesis of identical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
 The correspond Wald test is available in the `R` package `spsur`.
@@
             Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
-            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
+            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence)),
+            Sunshine_Hours_lag8 = cor(log(Sunshine_Hours_lag8 + 0.1), log(Incidence)),
+            Sunshine_Hours_lag11 = cor(log(Sunshine_Hours_lag11 + 0.1), log(Incidence)),
+            Sunshine_Hours_lag11w = cor(log(Sunshine_Hours_lag11w + 0.1), log(Incidence))) %>%
   select(-Date) %>%
@@
                                       "Mean_Temp_lag11",
-                                      "Mean_Temp_lag11w"))) %>%
+                                      "Mean_Temp_lag11w",
+                                      "Sunshine_Hours_lag8",
+                                      "Sunshine_Hours_lag11",
+                                      "Sunshine_Hours_lag11w"))) %>%
   ggplot(aes(x = Variable, y = Correlation)) +
@@
 
-```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
+```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of the spatial autocorrelation coefficient; dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
 data.frame(Date = seq(ymd("2020-03-13"), 
@@
                       by = "days"),
-           Deltas = sur.slm_lag11$deltas,
-           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
+           Beta = sur.slm_lag11$deltas,
+           sd = sur.slm_lag11$deltas.se) %>%
+  mutate(zscore = Beta/sd,
+         pval = 2 * pnorm(-abs(zscore)),
+           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
+           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
+         pval = case_when(pval <= 0.05 ~ "p <= 0.05",
+                          pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
                          TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Deltas, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.
-
-```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
-#Intercept
-n = 1
+  ggplot(aes(x = Date, y = Beta, color = pval)) +
+  geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
+  geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
+            ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+  geom_point(size = 3) +
+  geom_errorbar(aes(x = Date,
+                     ymin = CI90low,
+                    ymax = CI90up), 
+                width = 0.5) +
+  geom_linerange(aes(x = Date,
+                     ymin = CI95low,
+                    ymax = CI95up),
+                 size = 0.5) +
+  scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
+  geom_hline(yintercept = 0) +
+  theme_bw()
+```
+```{r function-for-plotting-coefficients, include=FALSE}
+# Define a function for plotting the coefficients over time including their confidence intervals and significance
+plot_coeff <- function(model = model, n = n){
   data.frame(Date = seq(ymd("2020-03-13"), 
@@
                         by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
+             Beta = matrix(model$coefficients[-c(2,3)],ncol = T)[n,],
+             sd = matrix(model$rest.se[-c(2,3)], ncol = T)[n,]) %>%
+    mutate(zscore = Beta/sd,
+           pval = 2 * pnorm(-abs(zscore)),
+           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
+           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
+           pval = case_when(pval <= 0.05 ~ "p <= 0.05",
+                            pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
                             TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
+    ggplot(aes(x = Date, y = Beta, color = pval)) +
+    geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
+    geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+    geom_point(size = 3) +
+    geom_errorbar(aes(x = Date,
+                      ymin = CI90low,
+                      ymax = CI90up), 
+                  width = 0.5) +
+    geom_linerange(aes(x = Date,
+                       ymin = CI95low,
+                       ymax = CI95up),
+                   size = 0.5) +
+    scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
+    geom_hline(yintercept = 0) +
+    theme_bw()
+  }
+```
+
+Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.
+
+```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept; dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
+#Intercept
+plot_coeff(model = sur.slm_lag11, n = 1) + 
+  ylab("Beta (Intercept)")
 ```
@@
 
-```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)", cache = TRUE}
+```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density); dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
 #Density:
-n = 2 # Although this is the fourth coefficient, after intercept, log(GDPpc), and log(Older), we start at 2 here because the two constrained coefficients are removed from the matrix of coefficients 
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
+plot_coeff(model = sur.slm_lag11, n = 2) + 
+  ylab("Beta (log(Density))")
 ```
@@
 
-```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit", cache = TRUE}
-n = 3
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
+```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit; dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+# Transit
+plot_coeff(model = sur.slm_lag11, n = 3) + 
+  ylab("Beta (Transit)")
 ```
@@
 
-```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)", cache = TRUE}
-n = 4
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
+```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity); dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+# Humidity:
+plot_coeff(model = sur.slm_lag11, n = 4) + 
+  ylab("Beta (log(Humidity))")
 ```
@@
 
-```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)", cache = TRUE}
-n = 5
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
+```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature); dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+# Temperature
+plot_coeff(model = sur.slm_lag11, n = 5) + 
+  ylab("Beta (log(Temperature))")
 ```
@@
 
-```{r beta-sunshine-time, echo=FALSE, fig.cap="\\label{fig:beta-sunshine-time}Temporal evolution of coefficient for log(Sunshine + 0.1)", cache = TRUE}
-n = 6
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-```{r impacts, echo=FALSE}
+```{r beta-sunshine-time, echo=FALSE, fig.cap="\\label{fig:beta-sunshine-time}Temporal evolution of coefficient for log(Sunshine + 0.1); dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+# Sunshine
+plot_coeff(model = sur.slm_lag11, n = 6) + 
+  ylab("Beta (log(Sunshine + 0.1))")
+```
+
+```{r impacts, echo=FALSE, include=FALSE}
 Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
@@
 }
+
 matplot(myimpact, type='l')

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Finished editing sections Data and Methods and improved the plots for the results

@@
 \bibliographystyle{elsarticle-harv}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 also released at the provincial level as part of a collaborative project
-{[}\textbf{by whom?}{]}. This information is compiled from several
-sources, mainly the official web pages of the Spanish regional
-goverments. In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two
-environmental variables, namely temperature and humidity, which are
-collected from official sources (i.g., AEMET, the state meteorology
-agency, and MAPA, the ministry of agriculture, fisheries, and food). The
+by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
+ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
+and ESRI España. This information is compiled from several sources,
+mainly the official web pages of the Spanish regional goverments, as
+documented in Centro de Datos
+Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
+In addition, we consider two sets of explanatory variables. The first
+one, and the focus of this research, is set of two environmental
+variables, namely temperature and humidity, which are collected from
+official sources (i.g., \emph{AEMET}, the state meteorology agency, and
+\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
 second set provides some relevant controls for multivariate analysis,
@@
 \midrule
-\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
+\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & ProvidencialData19\\
 Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
@@
 \multicolumn{7}{l}{\textit{Note: }}\\
-\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
+\multicolumn{7}{l}{ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19}\\
 \multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
@@
 
-where \(\bf{Y}\) is now a vector of dimension \(NT \times 1\),
+\noindent where \(\bf{Y}\) is now a vector of dimension \(NT \times 1\),
 \(\bf{X}\) is a block-diagonal matrix \(NT \times K\) (with
-\(K = \sum_t{k_t}\)) and \(\epsilon\) is a \(NT \times 1\) vector. Using
-the Kronecker product notation (\(\otimes\)), the error matrix structure
-is expressed concisely as:
+\(K = \sum_t{k_t}\)) and \(\epsilon\) is an \(NT \times 1\) vector.
+Using the Kronecker product notation (\(\otimes\)), the error matrix
+structure is expressed concisely as:
 
@@
 
-The stack expresion for the spatial lag SUR model (SUR-SLM) is as
-follows:
+The stack expresion for the SUR model with a spatially lagged dependent
+variable (SUR-SLM) is as follows:
 
@@
 
-\noindent where \(\bf{A} =I_{TN}-\bf{\Gamma} \otimes W\) with
+\noindent where \(\bf{A} =I_{TN}-\bf{\Gamma} \otimes W\) is the
+spatially lagged dependent variable, and
 \(\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)\).
 
-This specification assumes that incidence in a province (\(y_{st}\)) at
-time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
-of incidence in neighbouring provinces. Coefficients for the spatially
-lagged variable are estimated for each time period \(\rho_t\) and
-identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypothesis of identtical levels of spatial
-dependence (\(\rho_i=\rho_j, \forall i,j\)). The correspond Wald test is
-available in the \texttt{R} package \texttt{spsur}.
+This specification assumes that outcome (\(y_{st}\)) at location \(s\)
+and time \(t\) is partially determined by the weighted average
+(\(Wy_{st}\)) of the outcome in neighboring provinces, with neighborhood
+defined by matrix \(W\) of spatial weights. In other words, the
+spatially lagged dependent variable represents a process of contagion,
+where the disease in neighboring provinces can spillover in a spatial
+way. The coefficients of the spatially lagged variable are estimated for
+each time period \(\rho_t\) and identify the intensity and the sign of
+the contagion effect. It is possible test the null hypothesis of
+identical levels of spatial dependence (\(\rho_i=\rho_j, \forall i,j\)).
+The correspond Wald test is available in the \texttt{R} package
+\texttt{spsur}.
 
@@
 \includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/delta-time-1.pdf}
-\caption{\label{fig:delta-time}Temporal evolution of spatial
-autocorrelation coefficient}
+\caption{\label{fig:delta-time}Temporal evolution of the spatial
+autocorrelation coefficient; dots are the point estimates and vertical
+lines are 95\% confidence intervals. In yellow is the period after the
+declaration of the state of emergency, and in orange is the period when
+only essential activities were allowed.}
 \end{figure}
@@
 \includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-intercept-time-1.pdf}
-\caption{\label{fig:beta-intercept-time}Temporal evolution of intercept}
+\caption{\label{fig:beta-intercept-time}Temporal evolution of intercept;
+dots are the point estimates and vertical lines are 95\% confidence
+intervals. In yellow is the period after the declaration of the state of
+emergency, and in orange is the period when only essential activities
+were allowed.}
 \end{figure}
@@
 \caption{\label{fig:beta-density-time}Temporal evolution of coefficient
-for log(Density)}
+for log(Density); dots are the point estimates and vertical lines are
+95\% confidence intervals. In yellow is the period after the declaration
+of the state of emergency, and in orange is the period when only
+essential activities were allowed.}
 \end{figure}
@@
 \caption{\label{fig:beta-transit-time}Temporal evolution of coefficient
-for Transit}
+for Transit; dots are the point estimates and vertical lines are 95\%
+confidence intervals. In yellow is the period after the declaration of
+the state of emergency, and in orange is the period when only essential
+activities were allowed.}
 \end{figure}
@@
 \caption{\label{fig:beta-humidity-time}Temporal evolution of coefficient
-for log(Humidity)}
+for log(Humidity); dots are the point estimates and vertical lines are
+95\% confidence intervals. In yellow is the period after the declaration
+of the state of emergency, and in orange is the period when only
+essential activities were allowed.}
 \end{figure}
@@
 \caption{\label{fig:beta-temperature-time}Temporal evolution of
-coefficient for log(Temperature)}
+coefficient for log(Temperature); dots are the point estimates and
+vertical lines are 95\% confidence intervals. In yellow is the period
+after the declaration of the state of emergency, and in orange is the
+period when only essential activities were allowed.}
 \end{figure}
@@
 \caption{\label{fig:beta-sunshine-time}Temporal evolution of coefficient
-for log(Sunshine + 0.1)}
+for log(Sunshine + 0.1); dots are the point estimates and vertical lines
+are 95\% confidence intervals. In yellow is the period after the
+declaration of the state of emergency, and in orange is the period when
+only essential activities were allowed.}
 \end{figure}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Wsur <-}\StringTok{ }\KeywordTok{as}\NormalTok{(spdep}\OperatorTok{::}\KeywordTok{listw2mat}\NormalTok{(listw), }\StringTok{"CsparseMatrix"}\NormalTok{)}
-\NormalTok{trsur <-}\StringTok{ }\NormalTok{spatialreg}\OperatorTok{::}\KeywordTok{trW}\NormalTok{(Wsur, }\DataTypeTok{type =} \StringTok{"MC"}\NormalTok{)}
-\NormalTok{impacts.sur.slm_lag11 <-}\StringTok{ }\KeywordTok{impactspsur}\NormalTok{(sur.slm_lag11, }\DataTypeTok{tr =}\NormalTok{ trsur,}\DataTypeTok{R =} \DecValTok{1000}\NormalTok{)}
-
-\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{1}\NormalTok{]]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Impact measures (lag, trace):
-##                                       Direct   Indirect      Total
-## log(GDPpc)_1                       0.6736035  0.6395634  1.3131669
-## log(Older)_1                      -0.8002302 -0.7597912 -1.5600214
-## log(Density)_1                     0.1662437  0.1578427  0.3240864
-## Transit_1                          0.3413888  0.3241370  0.6655259
-## log(Humidity_lag11)_1             -1.5578465 -1.4791220 -3.0369685
-## log(Mean_Temp_lag11)_1            -1.7365494 -1.6487943 -3.3853437
-## log(Sunshine_Hours_lag11 + 0.1)_1 -0.2804735 -0.2663000 -0.5467734
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{2}\NormalTok{]]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Impact measures (lag, trace):
-##                                        Direct    Indirect      Total
-## log(Density)_2                     0.06248795  0.04048188  0.1029698
-## Transit_2                          0.44006548  0.28508982  0.7251553
-## log(Humidity_lag11)_2             -1.06244958 -0.68829202 -1.7507416
-## log(Mean_Temp_lag11)_2            -1.47760442 -0.95724386 -2.4348483
-## log(Sunshine_Hours_lag11 + 0.1)_2 -0.19072953 -0.12356126 -0.3142908
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{3}\NormalTok{]]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Impact measures (lag, trace):
-##                                        Direct    Indirect       Total
-## log(Density)_3                     0.02426235  0.01927572  0.04353807
-## Transit_3                          0.43033000  0.34188455  0.77221455
-## log(Humidity_lag11)_3             -0.97379005 -0.77364759 -1.74743764
-## log(Mean_Temp_lag11)_3            -1.39450981 -1.10789707 -2.50240689
-## log(Sunshine_Hours_lag11 + 0.1)_3 -0.09683377 -0.07693158 -0.17376535
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{myimpact <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DataTypeTok{ncol =} \DecValTok{5}\NormalTok{, }\DataTypeTok{nrow =} \DecValTok{29}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{2}\OperatorTok{:}\DecValTok{30}\NormalTok{)\{}
-\NormalTok{  myimpact[i}\DecValTok{-1}\NormalTok{,] <-}\StringTok{ }\NormalTok{impacts.sur.slm_lag11[[i]]}\OperatorTok{$}\NormalTok{res}\OperatorTok{$}\NormalTok{direct}
-\NormalTok{\}}
-\KeywordTok{matplot}\NormalTok{(myimpact, }\DataTypeTok{type=}\StringTok{'l'}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/impacts-1.pdf}
 

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Finished editing Analysis, improved the presentation of some figures, added alternate models for reference, for sensitivity analysis. Next step: Results and discussion, and conclusions.

@@
 ---
-title: A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
+title: A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain
 author:
@@
 
-Context, Data, and Methods
+Context and Data
 ============
@@
 A18)}.
-The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when its cause may still be unclear [@Zhou2016accelerated].
-In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] remarks how Toronto's status as a global city was a vulnerability.
-In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and their potential to be further ahead in the trajectory of the pandemic.
+The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated].
+In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect.
+In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic.
 Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional].
-Therefore, it is possible that wealthier regions remained relatively more active even during the lockdown.
-On the other hand, it is also possible the less wealthy regions have a higher proportion of workers in manual occupations that cannot telework.
+Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown.
+On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.
 
 The percentaje of older adults (over 65) is the second variable that we consider as a control.
-Early evidence suggest that the case rate mortality is higher at older ages [e.g.
+Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g.
 @Novel2020epidemiological].
-It is not clear that presence of older adults necessarily translates into higher transmission in the population.
-The public health tool of choice in containing the spread of the disease has been social distancing.
+However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection.
+The tool of choice in containing the spread of the disease has been social distancing.
 In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
-In other words, older adults are, by preference or necessity, already in a form of social isolation.
-Social distancing during the pandemic may actually reinforce that condition for older adults, as the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age] suggests.
-Accordingly, our expectation is that in the provinces with higher proportions of older adults will tend to have similarl levels of lower social contact, particularly since age-structured social contact matrix in Spain is similar to Italy [@Prem2017projecting].
+In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation.
+Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age].
+Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].
 
@@
 The last control variable is the presence of mass transit systems in a province.
-Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
-Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches [@Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
+Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g.
+metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
+Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
 After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.
@@
 
-## Methods: Spatial SUR
+Methods: the Spatial SUR Model
+===================
 
@@
 
-<!--
-![\begin{equation} \label{eq:sur-slm} \begin{aligned} \bf{AY} = \bf{X} \beta + \epsilon \\ \epsilon =N(0,\bf{\Sigma}) \end{aligned} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbegin%7Baligned%7D%20%5Cbf%7BAY%7D%20%3D%20%5Cbf%7BX%7D%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5Cbf%7B%5CSigma%7D)%20%5Cend%7Baligned%7D%20%5Cend%7Bequation%7D)
-
-\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
--->
-
 This specification assumes that outcome ($y_{st}$) at location $s$ and time $t$ is partially determined by the weighted average ($Wy_{st}$) of the outcome in neighboring provinces, with neighborhood defined by matrix $W$ of spatial weights.
@@
 <!--
+![\begin{equation} \label{eq:sur-slm} \begin{aligned} \bf{AY} = \bf{X} \beta + \epsilon \\ \epsilon =N(0,\bf{\Sigma}) \end{aligned} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbegin%7Baligned%7D%20%5Cbf%7BAY%7D%20%3D%20%5Cbf%7BX%7D%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5Cbf%7B%5CSigma%7D)%20%5Cend%7Baligned%7D%20%5Cend%7Bequation%7D)
+
+\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
+
 This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights.
@@
 The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
-Another alternative methodologies could be use.
-By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g.
-Elhorst 2014, Cap.
-4), but those models don't take account correlation between errors.
-Therefore, a spatial SUR approach is more reasonable for our purpose.
-
-Analysis and Results
+
+Analysis
 ===================
@@
 
-We begin with the exploratory analysis of the data.
-
-Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages.
+Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of COVID-19 in Spain, as well as the temporal progression of the disease in weekly averages.
 Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19.
@@
 While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces.
-Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao.
+Vitoria, its main city, is the capital of the Basque Country and has been the focus of efforts, along with San Sebastian and Bilbao, to develop a "Global Basque City" [@Meijers2008strategic].
 The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava.
@@
   ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
+  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
   scale_fill_distiller(name = "Mean Weekly Incidence", 
@@
   geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
+               x = madrid_xy[1] - 9, 
                y = madrid_xy[2] + 0.6,
@@
         plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 8, y = madrid_xy[2] + 0.6) +
+  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 9, y = madrid_xy[2] + 0.6) +
   geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
@@
   geom_sf_text(label = "Madrid",
-               x = madrid_xy[1] - 8, 
+               x = madrid_xy[1] - 9, 
                y = madrid_xy[2] + 0.6,
@@
 
-grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, nrow = 3)
-```
-
-To visualize the distribution of temperature and humidity we aggregate the provinces by Autonomous Community.
-In Figure \ref{fig:descriptives-temperature} the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost.
-There is a relatively wide range of values both within and between provinces over the 30 day period examined.
-The top panel of the figure shows the distribution of mean temperature.
-The lowest mean temperature for a community on any given day is 2.8C, and the highest is 22.4C for a range of approximately 20 degrees.
-Likewise, there is a great deal of variability in humidity, as seen in the bottom panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
-The actual values for the provinces display somewhat more variability even.
+# Create grid layout
+lay = rbind(c(1, 1, 2, 2),
+        c(1, 1, 2, 2),
+        c(3, 3, 4, 4),
+        c(3, 3, 4, 4),
+        c(NA, 5, 5, NA),
+        c(NA, 5, 5, NA))
+
+grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, layout_matrix = lay)
+```
+
+Figure \ref{fig:descriptives-temperature} shows the distribution of the environmental variables in Spain.
+For ease of visualization we aggregate the provinces by Autonomous Community.
+Each box-and-whisker in the figure represents the distribution of the variable for a community over the 30-day period.
+In the plot, the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost.
+As seen in the figure, there is a relatively wide range of values both within and between provinces over the 30-day period examined.
+The top panel of the figure shows the distribution of mean temperatures.
+The lowest mean temperature for a community on any given day was 2.8C, and the highest 22.4C, for a range of approximately 20 degrees.
+Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
+Finally, the bottom panel displays mean daily hours of sunshine in the community.
+This variable displays the most variability within communities over time, but remains relatively constant across communities.
+It is important to note that these are summaries by community, and the actual values of these variables for the provinces display somewhat more variability.
 
@@
 
+# Summary of hours of sunshine by Autonomous Community
+sunshine.summary <- covid19_spain %>% 
+  group_by(CCAA, Date) %>%
+  summarize(Value = mean(Sunshine_Hours)) %>%
+  mutate(Variable = "Mean Daily Hours of Sunshina in Community")
+
 # Bind tables
-env.summary <- rbind(temp.summary, humidity.summary) %>%
+env.summary <- rbind(temp.summary, humidity.summary, sunshine.summary) %>%
   mutate(Variable = factor(Variable, 
                            levels = c("Mean Temperature in Community",
-                                      "Mean Humidity in Community")))
+                                      "Mean Humidity in Community",
+                                      "Mean Daily Hours of Sunshina in Community")))
 
@@
   ylab("Value") +
-  facet_wrap(~ Variable, nrow = 2, 
+  facet_wrap(~ Variable, nrow = 3, 
              scales = "free_y")
 ```
+
+Figure \ref{fig:map-controls} includes three maps that display the spatial variation of our control variables, namely GDP per capita, percentage of older adults in province, population density, and presence of mass transit systems.
+As seen there, GDP per capita is higher in Madrid and the northeast part of the country, mainly in Pais Vasco and Cataluña.
+Percentage of older adults is somewhat more checkered, with high values in Madrid and other provinces in the center-west part of the country, but also in some provinces in the north.
+Outside of provinces with major cities (e.g., Madrid; Bizkaia and Gipuzkoa in Pais Vasco; Pontevedra in Galicia), population density tends to be higher in provinces along the Mediterranean coast.
+The final panel in the figure shows the five provinces in the country that have higher order mass transit systems (e.g., metro).
 
@@
   ggplot() +
-  geom_sf(aes(fill = GDPpc)) +
+  geom_sf(aes(fill = GDPpc), color = "gray69") +
   scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  ggtitle("GDP per capita") +
   theme_bw(base_size = 9) +
@@
   ggplot() +
-  geom_sf(aes(fill = Older)) +
+  geom_sf(aes(fill = Older), color = "gray69") +
   scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  ggtitle("Percentage of older adults") +
   theme_bw(base_size = 9) +
@@
   ggplot() +
-  geom_sf(aes(fill = Density)) +
+  geom_sf(aes(fill = Density), color = "gray69") +
   scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  ggtitle("Population density") +
   theme_bw(base_size = 9) +
@@
 
-grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
-```
-
-Figure \ref{fig:daily-correlations} shows the distribution of daily correlations of the independent variables with incidence of COVID-19, after log-transforming all variables.
-It can be seen there that the correlation of GDPpc and temperature (in its three definitions) have the strongest positive and negative correlations with incidence, respectively.
-Percentage of older adults displays somewhat weaker negative correlations with incidence, as does density.
-It can be seen that the humidity variable, in its three forms, tends to be possitively correlated with incidence of COVID-19.
+# Transit systems:
+transit.plot <- provinces_spain %>% 
+  mutate(Transit = factor(Transit, 
+                          levels = c(0, 1),
+                          labels = c("No",
+                                     "Yes"))) %>%
+  #filter(CCAA != "Canarias") %>%
+  ggplot() +
+  geom_sf(aes(fill = Transit), color = "gray69") +
+  scale_fill_manual(name = "Transit", values=c("#FFFEDE", "#D5610D")) +
+  ggtitle("Mass transit systems") +
+  theme_bw(base_size = 9) +
+  theme(axis.text = element_blank(),
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 9, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
+
+# Define layout for plot
+lay <- rbind(c(1, 2),
+             c(3, 4))
+
+grid.arrange(gdppc.plot, older.plot, density.plot, transit.plot, layout_matrix = lay)
+```
+
+Figure \ref{fig:daily-correlations} shows the distribution of daily simple correlations of incidence of COVID-19 with the independent variables (with the exception of Transit, which is a categorical variable).
+These correlations are calculated after log-transforming all variables.
+As previously discussed, the environmental variables have a temporal lag and were calculated different time windows.
+
+It can be seen in the figure that temperature (in its three forms) has the highest simple correlation with incidence of COVID-19.
+After temperature, GDP per capita has the highest positive correlation with the dependent variable.
+The distribution of these correlations is also quite tight over the 30-day period of the study.
+Hours of sunshine tends to have a moderately high correlation with incidence of COVID-19, however the distribution of these correlations is more spread, and in some cases strays into negative values.
+A similar thing happens with humidity, which also tends to display more day to day variation in the correlation with the dependent variable.
+The percentage of older adults shows a relatively tight distribution of day-to-day correlations, and is negative.
+Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive.
 
@@
 
+Correlation analysis in the preceding section provides some insights about the potential associations between incidence of COVID-19 and the various environmental and control variables.
+In this section we estimate three spatial SUR models to test the differences between the various temporal lags and weighting schemes for the environmental variables.
+Accordingly, we define three models: Model 1, which is estimated using the lagged 8-day averages of the environmental variables (*lag8*); Model 2, which is estimated using the lagged 11-day averages of the environmental variables (*lag11*); and finally, Model 3, which is estimated using the lagged 11-day *weighted* averages of the environmental variables (*lag11w*).
+
+To implement the SUR approach, we must define a matrix of spatial weights $W$.
+In this case, we consider neighborhoods based on adjacency, based on the well-known rook criterion (two provinces are adjacent if they share a boundary, but not if they touch only at a vertex).
+We modify this criterion in two cases.
+First, we make an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are adjacent between each other.
+And secondly, we assume that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello).
+After matrix $W$ has been specified, estimation of the model can proceed as usual.
+
+For estimation, we log-transform the dependent and quantitative independent variables.
+The only variable that is not transformed is the categorical variable for transity systems.
+A log-log transformation is appropriate to capture non-linear relationships between variables and provides a straightforward interpretation of the coefficients as percentage change.
+Furthermore, we introduce restrictions so that the coefficients of two of our control variables are constant over time, namely GDP per capita and percentage of older adults.
+We do not see an  *a priori* reason to let those two variables vary across equations, and the correlation analysis in Figure \ref{fig:daily-correlations} also suggest little temporal variation.
+In contrast, we let the spatial autocorrelation parameter, as well as the parameters of the rest of the independent variables (including the constant) to vary over time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious.
+These results are available in the source R markdown document.}.
+
 ```{r data-preparation-for-modelling, include=FALSE, cache = TRUE}
@@
 
+```{r model1-8-day-moving-average-all, include=FALSE, eval = FALSE}
+# Model with a lagged 8-day moving average of climatic variables - NO RESTRICTIONS TO THE COEFFICIENTS:
+sur.slm_lag8_all <- spsur::spsurtime(formula = formula_lag8, 
+                                 data=GPanel, 
+                                 time = GPanel$Date, 
+                                 type = "slm", 
+                                 fit_method = "3sls", 
+                                 listw=  listw)
+```
+
 ```{r model2-11-day-moving-average, include=FALSE, cache = TRUE}
@@
 
+```{r model2-11-day-moving-average-all, include=FALSE, cache = TRUE}
+# Model with 11-day moving average of climatic variables:
+sur.slm_lag11_all <- spsur::spsurtime(formula = formula_lag11, 
+                                  data=GPanel, 
+                                  time = GPanel$Date, 
+                                  type = "slm", 
+                                  fit_method = "3sls", 
+                                  listw=  listw)
+```
+
 ```{r model3-11-day-weighted-moving-average, include=FALSE, cache = TRUE}
@@
 
-The goodness of fit of the three systems of equations is shown in Figure \ref{fig:goodness-of-fit}.
+```{r model3-11-day-weighted-moving-average-all, include=FALSE, cache = TRUE}
+# Model with 11-day weighted moving average of climatic variables:
+sur.slm_lag11w_all <- spsur::spsurtime(formula = formula_lag11w, 
+                                   data=GPanel, 
+                                   time = GPanel$Date, 
+                                   type = "slm", 
+                                   fit_method = "3sls", 
+                                   listw=  listw)
+```
+
+After estimation, we compare the goodness of fit of the three SUR models.
+Figure \ref{fig:goodness-of-fit} shows the equation-level coefficient of determination $R^2$, one for each time period/day.
+As well, the overall coefficient of determination for the system is reported for each model $\text{pooled}-R^2$.
+The general trend is identical for the three models, with the goodness-of-fit improving over time and plateuing around a value of $R^2$ of 0.6.
+Model 1 (*lag8*) performs somewhat better in the first few equations/days, when the goodness-of-fit is relatively poor, and then again in the last few equations/days.
+Model 3 (*lag11w*), in contrast, does not perform well towards the end of the study period.
+The most balanced model in terms of equation-level goodness-of-fit appears to be Model 2 (*lag11*), and this impression is further supported by a slightly higher value of the $\text{pooled}-R^2$.
+The analysis using a lagged moving average of the environmental variables is in line with the incubation period reported by Lauer et al.
+[-@Lauer2020incubation], although the results do not support the use of a weighted average.
+For the remainder of the paper, we will adopt Model 2 (*lag11*) as our best model.
+In the following section we discuss the results of the analysis in more depth.
 
@@
            size = 3.5) +
-  theme_tufte()
-
-```
-
-Summary of best model.
+  theme_bw() +
+  theme(legend.position = "bottom")
+```
+
+```{r goodness-of-fit-all, eval=FALSE, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit-all} Goodness of fit of the SUR systems NO RESTRICTION TO THE COEFFICIENTS: by date and pooled"}
+data.frame(Model1 = sur.slm_lag8_all$R2, 
+           Model2 = sur.slm_lag11_all$R2,
+           Model3 = sur.slm_lag11w_all$R2) %>%
+  slice(2:n()) %>%
+  rownames_to_column(var = "Equation") %>%
+  mutate(Date = seq(ymd("2020-03-13"), 
+                    ymd("2020-04-11"), 
+                    by = "days")) %>%
+  pivot_longer(cols = starts_with("M"), names_to = "Model", values_to = "R2") %>%
+  mutate(Model = factor(Model, 
+                        levels = c("Model1", "Model2", "Model3"),
+                        labels = c("Model 1: lag8", "Model 2: lag11", "Model 3: lag11w"))) %>%
+  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
+  geom_point(size = 2) +
+  scale_color_manual(values = c("Model 1: lag8" = "blue", "Model 2: lag11" = "orange", "Model 3: lag11w" = "green") ) + 
+  annotate(geom = "text", 
+           label = c(paste("pooled-R2 (Model 1: lag8)=", 
+                           round(sur.slm_lag8_all$R2[1], 4)),
+                     paste("pooled-R2 (Model 2: lag11)=", 
+                           round(sur.slm_lag11_all$R2[1], 4)),
+                     paste("pooled-R2 (Model 3: lag11w)=", 
+                           round(sur.slm_lag11w_all$R2[1], 4))),
+           x = as_date("2020-04-01"),
+           y = c(0.3, 0.27, 0.24),
+           family = "serif",
+           size = 3.5) +
+  theme_bw() +
+  theme(legend.position = "bottom")
+```
+
+Results and Discussion
+===================
+
+Table \ref{tab:summary-best-model} presents a summary.
+
 ```{r summary-best-model, echo=FALSE, cache = TRUE}
-data.frame(Coefficient = sur.slm_lag11$coefficients, 
-           pval = dnorm(abs(sur.slm_lag11$coefficients/sur.slm_lag11$rest.se))) %>%
+# Copy best model
+best_model <- sur.slm_lag11
+
+data.frame(Coefficient = c(best_model$coefficients, 
+                           best_model$deltas), 
+           sd = c(best_model$rest.se, 
+                  best_model$deltas.se)) %>%
   rownames_to_column(var = "Variable") %>%
-  mutate(Variable = c("Intercept", 
+  mutate(zscore = Coefficient/sd,
+         pval = 2 * pnorm(-abs(zscore)),
+         Variable = c("Intercept", 
                       "log(GDPpc)", 
@@
                             "log(Sunshine)"),
-                          29)),
+                          29),
+                      rep("Spatially lagged y (rho)",
+                          30)),
          Variable = factor(Variable, 
@@
                                       "log(Temperature)",
-                                      "log(Sunshine)"))) %>%
+                                      "log(Sunshine)",
+                                      "Spatially lagged y (rho)"))) %>%
   group_by(Variable) %>%
@@
                   "Non-constrained",
+                  "Non-constrained",
                   "Non-constrained")) %>%
@@
         col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
-        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
+        caption = "\\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)") %>%
   kable_styling(latex_options = c("striped", "scale_down")) %>%
@@
 
-Discussion
-===================
-
-Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\lambda$).
-
-```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of the spatial autocorrelation coefficient; dots are the point estimates and vertical lines are 95% confidence intervals.
-In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = sur.slm_lag11$deltas,
-           sd = sur.slm_lag11$deltas.se) %>%
-  mutate(zscore = Beta/sd,
+```{r summary-best-model-all, include=FALSE, eval=FALSE}
+best_model_all <- sur.slm_lag11_all
+
+data.frame(Coefficient = c(best_model_all$coefficients, 
+                           # Copy best model
+                           best_model_all$deltas), 
+           sd = c(best_model_all$rest.se, 
+                  best_model_all$deltas.se)) %>%
+  rownames_to_column(var = "Variable") %>%
+  mutate(zscore = Coefficient/sd,
          pval = 2 * pnorm(-abs(zscore)),
-           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
-           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
-           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
-           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
-         pval = case_when(pval <= 0.05 ~ "p <= 0.05",
-                          pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = pval)) +
-  geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
-              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
-  geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
-            ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
-  geom_point(size = 3) +
-  geom_errorbar(aes(x = Date,
-                     ymin = CI90low,
-                    ymax = CI90up), 
-                width = 0.5) +
-  geom_linerange(aes(x = Date,
-                     ymin = CI95low,
-                    ymax = CI95up),
-                 size = 0.5) +
-  scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
-  geom_hline(yintercept = 0) +
-  theme_bw()
-```
+         Variable = c(rep(c("Intercept", 
+                            "log(GDPpc)", 
+                            "log(Older)",
+                            "log(Density)",
+                            "Transit",
+                            "log(Humidity)",
+                            "log(Temperature)",
+                            "log(Sunshine)"),
+                          30),
+                      rep("Spatially lagged y (rho)",
+                          30)),
+         Variable = factor(Variable, 
+                           levels = c("Intercept", 
+                                      "log(GDPpc)", 
+                                      "log(Older)", 
+                                      "log(Density)",
+                                      "Transit", 
+                                      "log(Humidity)",
+                                      "log(Temperature)",
+                                      "log(Sunshine)",
+                                      "Spatially lagged y (rho)"))) %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Coefficient), 
+            Mean = mean(Coefficient), 
+            Max = max(Coefficient),
+            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
+            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
+            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Non-constrained", 
+                  "Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained")) %>%
+  kable(#"latex",
+    "html",
+    booktabs = TRUE,
+    digits = 3,
+    col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
+    caption = "\\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)") %>%
+  #kable_styling(latex_options = c("striped", "scale_down")) %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
+  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
+                       "Non-constrained: coefficient was allowed to vary across equations",
+                       "Constrained: coefficient as constant across equations"))
+```
+
 ```{r function-for-plotting-coefficients, include=FALSE}
@@
               ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
-    geom_point(size = 3) +
+    geom_point(size = 1) +
     geom_errorbar(aes(x = Date,
@@
                       ymax = CI90up), 
-                  width = 0.5) +
+                  size = 0.6,
+                  width = 0.8) +
     geom_linerange(aes(x = Date,
@@
                        ymax = CI95up),
-                   size = 0.5) +
+                   size = 0.4) +
     scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
@@
 
-Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.
-
-```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept; dots are the point estimates and vertical lines are 95% confidence intervals.
+```{r function-extract-legend, include=FALSE}
+# Define a function to extract a legend from a ggplot
+#extract legend
+#https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs
+g_legend<-function(a.gplot){
+  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)}
+```
+
+Figure \ref{fig:results-time-1} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$) and the intercept of the model.
+
+```{r results-time-1, cache = TRUE, fig.height= 4, echo=FALSE, fig.cap="\\label{fig:results-time-1}Temporal evolution of the spatial autocorrelation coefficient (rho) and the intercept of the model; dots are the point estimates and vertical lines are 95% confidence intervals.
 In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
+rho.plot <- data.frame(Date = seq(ymd("2020-03-13"), 
+                      ymd("2020-04-11"), 
+                      by = "days"),
+           Beta = best_model$deltas,
+           sd = best_model$deltas.se) %>%
+  mutate(zscore = Beta/sd,
+         pval = 2 * pnorm(-abs(zscore)),
+           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
+           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
+         pval = case_when(pval <= 0.05 ~ "p <= 0.05",
+                          pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
+                         TRUE ~ "p > 0.10")) %>%
+  ggplot(aes(x = Date, y = Beta, color = pval)) +
+  geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
+  geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
+            ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+  geom_point(size = 1) +
+  geom_errorbar(aes(x = Date,
+                     ymin = CI90low,
+                    ymax = CI90up),
+                 size = 0.6,
+                width = 0.8) +
+  geom_linerange(aes(x = Date,
+                     ymin = CI95low,
+                    ymax = CI95up),
+                 size = 0.4) +
+  scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
+  geom_hline(yintercept = 0) +
+  theme_bw() +
+  ylab("rho") +
+  theme(legend.position = "bottom")
+
+common_legend <- g_legend(rho.plot)
+
+# Plot without legend
+rho.plot <- rho.plot +
+   theme(legend.position = "none")
+
 #Intercept
-plot_coeff(model = sur.slm_lag11, n = 1) + 
-  ylab("Beta (Intercept)")
-```
-
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Density)$.
-
-```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density); dots are the point estimates and vertical lines are 95% confidence intervals.
+intercept.plot <- plot_coeff(model = best_model, n = 1) + 
+  ylab("Beta (Intercept)") +
+  theme(legend.position = "none")
+
+# Define layout for plot
+lay <- rbind(c(1, 2),
+             c(3, 3))
+
+grid.arrange(rho.plot, intercept.plot, common_legend, 
+             layout_matrix = lay,
+             heights = c(5, 1))
+```
+
+Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ and $Transit$.
+
+```{r beta-controls-time, echo=FALSE,  fig.height= 4, fig.cap="\\label{fig:beta-controls-time}Temporal evolution of coefficient for the control variables; dots are the point estimates and vertical lines are 95% confidence intervals.
 In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
 #Density:
-plot_coeff(model = sur.slm_lag11, n = 2) + 
-  ylab("Beta (log(Density))")
-```
-
-Figure \ref{fig:beta-transit-time} shows the temporal evolution of the coefficient for $Transit$.
-
-```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit; dots are the point estimates and vertical lines are 95% confidence intervals.
+density.plot <- plot_coeff(model = best_model, n = 2) + 
+  ylab("Beta (log[Density]))") +
+  theme(legend.position = "none")
+
+# Transit
+transit.plot <- plot_coeff(model = best_model, n = 3) + 
+  ylab("Beta (Transit)") +
+  theme(legend.position = "none")
+
+# Define layout for plot
+lay <- rbind(c(1, 2),
+             c(3, 3))
+
+grid.arrange(density.plot, transit.plot, common_legend, 
+             layout_matrix = lay,
+             heights = c(5, 1))
+```
+
+Figure \ref{fig:beta-environmental-time} shows the temporal evolution of the coefficient for the three environmental variables.
+
+```{r beta-environmenta-time, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:beta-environmental-time}Temporal evolution of coefficient for the environmental variables; dots are the point estimates and vertical lines are 95% confidence intervals.
 In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
-# Transit
-plot_coeff(model = sur.slm_lag11, n = 3) + 
-  ylab("Beta (Transit)")
-```
-
-Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the coefficient for $\log(Humidity)$.
-
-```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity); dots are the point estimates and vertical lines are 95% confidence intervals.
-In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
-# Humidity:
-plot_coeff(model = sur.slm_lag11, n = 4) + 
-  ylab("Beta (log(Humidity))")
-```
-
-Figure \ref{fig:beta-temperature-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.
-
-```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature); dots are the point estimates and vertical lines are 95% confidence intervals.
-In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+# Humidity
+humidity.plot <- plot_coeff(model = best_model, n = 4) + 
+  ylab("Beta (log[Humidity])") +
+  theme(legend.position = "none")
+
 # Temperature
-plot_coeff(model = sur.slm_lag11, n = 5) + 
-  ylab("Beta (log(Temperature))")
-```
-
-Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the coefficient for $\log(Sunshine + 0.1)$.
-
-```{r beta-sunshine-time, echo=FALSE, fig.cap="\\label{fig:beta-sunshine-time}Temporal evolution of coefficient for log(Sunshine + 0.1); dots are the point estimates and vertical lines are 95% confidence intervals.
-In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+temperature.plot <- plot_coeff(model = best_model, n = 5) + 
+  ylab("Beta (log[Temperature])") +
+  theme(legend.position = "none")
+
 # Sunshine
-plot_coeff(model = sur.slm_lag11, n = 6) + 
-  ylab("Beta (log(Sunshine + 0.1))")
+sunshine.plot <- plot_coeff(model = best_model, n = 6) + 
+  ylab("Beta (log[Sunshine + 0.1])") +
+  theme(legend.position = "none")
+
+# Define layout for plot
+lay <- rbind(c(1, 1, 2, 2),
+             c(1, 1, 2, 2),
+             c(NA, 3, 3, NA),
+             c(NA, 3, 3, NA),
+             c(NA, 4, 4, NA))
+
+grid.arrange(humidity.plot, temperature.plot, sunshine.plot, common_legend, 
+             layout_matrix = lay,
+             heights = c(4, 4, 4, 4, 1))
+
 ```
@@
 trsur <- spatialreg::trW(Wsur, type = "MC")
-impacts.sur.slm_lag11 <- impactspsur(sur.slm_lag11, tr = trsur,R = 1000)
-
-impacts.sur.slm_lag11[[1]]
-impacts.sur.slm_lag11[[2]]
-impacts.sur.slm_lag11[[3]]
+impacts.best_model <- impactspsur(best_model, tr = trsur,R = 1000)
+
+impacts.best_model[[1]]
+impacts.best_model[[2]]
+impacts.best_model[[3]]
 
@@
 for (i in 2:30){
-  myimpact[i-1,] <- impacts.sur.slm_lag11[[i]]$res$direct
+  myimpact[i-1,] <- impacts.best_model[[i]]$res$direct
 }
@@
 
-* HAY QUE INCLUIR INTERVALOS DE CONFIANZA EN LOS GRÁFICOS QUE HACEMOS DE LOS COEFICIENTES DEL MODELO
-
 * ¿hasta que punto las variables de control no recojen también factores climáticos? por ejemplo, la gente jóven vive en el sur de España que ha tenido menos incidencia.
-
-* TITULO: A spatial analysis of the environmental correlates ofCOVID-19 incidence in Spain ¿during the lookdown?
 

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Finished editing Analysis, improved the presentation of some figures, added alternate models for reference, for sensitivity analysis. Next step: Results and discussion, and conclusions.

@@
             pdfauthor={},
-            pdftitle={A spatial analysis of the environmental correlates of COVID-19 incidence in Spain},
+            pdftitle={A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain},
             colorlinks=false,
@@
 
-  \title{A spatial analysis of the environmental correlates of COVID-19 incidence
-in Spain}
+  \title{A spatio-temporal analysis of the environmental correlates of COVID-19
+incidence in Spain}
     \author[McMaster University]{Antonio Paez\corref{1}}
@@
 
-\hypertarget{context-data-and-methods}{%
-\section{Context, Data, and Methods}\label{context-data-and-methods}}
+\hypertarget{context-and-data}{%
+\section{Context and Data}\label{context-and-data}}
 
@@
 The growth in global connections has presented a challenge to spatial
-approaches in the initial stages of disease management, when its cause
-may still be unclear (Zhou and Coleman, 2016). In reference to the
-earlier outbreak of SARS, van Wagner (Van Wagner, 2008) remarks how
-Toronto's status as a global city was a vulnerability. In our case, we
-think of GDP per capita as a marker of a region's relative position in a
-network of global cities, and their potential to be further ahead in the
-trajectory of the pandemic. Furthermore, wealthier regions also tend to
-concentrate more activities that produce non-traded goods, including
-building and construction (Hallet, 2002). Therefore, it is possible that
-wealthier regions remained relatively more active even during the
-lockdown. On the other hand, it is also possible the less wealthy
-regions have a higher proportion of workers in manual occupations that
-cannot telework.
+approaches in the initial stages of disease management, when the cause
+of a disease may still be unclear, but the plane has already departed
+(Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
+van Wagner (Van Wagner, 2008) chronicles how Toronto's status as a
+global city turned out to be a vulnerability in this respect. In our
+case, we think of GDP per capita as a marker of a region's relative
+position in a network of global cities, and its potential to be further
+ahead in the trajectory of the pandemic. Furthermore, wealthier regions
+also tend to concentrate more activities that produce non-traded goods,
+including building and construction (Hallet, 2002). Therefore, it is
+possible that wealthier regions remain relatively more active even
+during a lockdown. On the other hand, it is also possible that less
+wealthy regions have a higher proportion of workers in manual
+occupations that cannot telework, and therefore have more difficulties
+complying with shelter-in-place orders.
 
 The percentaje of older adults (over 65) is the second variable that we
-consider as a control. Early evidence suggest that the case rate
-mortality is higher at older ages (e.g.~The Novel Coronavirus Pneumonia
-Emergency Response Epidemiology Team, 2020). It is not clear that
-presence of older adults necessarily translates into higher transmission
-in the population. The public health tool of choice in containing the
-spread of the disease has been social distancing. In this respect, the
-evidence from the field of transportation is that older adults tend to
-travel less frequently, for shorter distances, and have higher rates of
-immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder and
-Pinjari, 2012). In other words, older adults are, by preference or
-necessity, already in a form of social isolation. Social distancing
-during the pandemic may actually reinforce that condition for older
-adults, as the analysis of age-structured social contact in India,
-China, and Italy of Sing and Adhikari (2020) suggests. Accordingly, our
-expectation is that in the provinces with higher proportions of older
-adults will tend to have similarl levels of lower social contact,
-particularly since age-structured social contact matrix in Spain is
-similar to Italy (Prem et al., 2017).
+consider as a control. Early evidence regarding COVID-19 suggests that
+the case rate mortality is higher at older ages (e.g.~The Novel
+Coronavirus Pneumonia Emergency Response Epidemiology Team, 2020).
+However, it is not clear that a relatively large population of older
+adults necessarily translates into higher transmission of the infection.
+The tool of choice in containing the spread of the disease has been
+social distancing. In this respect, the evidence from the field of
+transportation is that older adults tend to travel less frequently, for
+shorter distances, and have higher rates of immobility (e.g., Roorda et
+al., 2010; Morency et al., 2011; Sikder and Pinjari, 2012). In other
+words, many older adults are, whether by preference or otherwise,
+already in a form of social isolation. Social distancing during the
+pandemic may actually reinforce that condition for them, as suggested by
+the analysis of age-structured social contact in India, China, and Italy
+of Sing and Adhikari (2020). Accordingly, our expectation is that
+provinces with higher percentages of older adults will tend to have
+similarly low levels of social contact, particularly since the
+age-structured matrix of social contact in Spain is similar to Italy
+(Prem et al., 2017).
 
@@
 transportation, however only five provinces have higher order systems of
-mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
-Public transportation has been hypothesized to relate to the spread of
-contagious disease by some researchers using agent-based approaches
-(Perez and Dragicevic, 2009; Wang et al., 2010), and while we find scant
-evidence of a link in the literature, the idea is intuitively appealing.
-After all, unlike the isolation that a car offers to travellers, most
-mass transit system are cauldrons of social contact.
+mass mobility (e.g.~metro or subway), namely Barcelona, Madrid, Sevilla,
+Valencia, and Bizkaia. Public transportation has been hypothesized to
+relate to the spread of contagious disease by some researchers using
+agent-based approaches and simulation (e.g., Perez and Dragicevic, 2009;
+Wang et al., 2010), and while we find scant evidence of a link in the
+literature, the idea is intuitively appealing. After all, unlike the
+isolation that a car offers to travellers, most mass transit system are
+cauldrons of social contact.
 
@@
 
-\hypertarget{methods-spatial-sur}{%
-\subsection{Methods: Spatial SUR}\label{methods-spatial-sur}}
+\hypertarget{methods-the-spatial-sur-model}{%
+\section{Methods: the Spatial SUR
+Model}\label{methods-the-spatial-sur-model}}
 
@@
 The SUR-SLM model can be estimated using maximum likelihood (López et
-al. (2014)) or instrumental variables (Mínguez et al. (2019)). Another
-alternative methodologies could be use. By example, a dynamic spatial
-panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
-2014, Cap. 4), but those models don't take account correlation between
-errors. Therefore, a spatial SUR approach is more reasonable for our
-purpose.
-
-\hypertarget{analysis-and-results}{%
-\section{Analysis and Results}\label{analysis-and-results}}
+al. (2014)) or instrumental variables (Mínguez et al. (2019)).
+
+\hypertarget{analysis}{%
+\section{Analysis}\label{analysis}}
 
@@
 
-We begin with the exploratory analysis of the data.
-
 Figure \ref{fig:weekly-average-incidence-map} shows the geographical
-variation in the incidence of the disease, as well as its temporal
-progression in weekly averages. Our analysis begins on March 13, which
-is the first date when every province had reported at least one case of
-COVID-19. It can be seen that the highest incidence at this early date
-was in the provice of Álava, in the North of Spain. While not the most
-populous province, with a population of only 331,549, Álava has the
-highest GDP per capita of all provinces. Vitoria, its main city, is the
-capital of the Basque country and has been the focus of efforts to
-develop a ``Global Basque City'' (Meijers et al., 2008), along with San
-Sebastian and Bilbao. The other early focus of the pandemic in Spain was
-Madrid, which is the most populous province in the country and has the
-second highest GDP per capita after Álava. The early outbreaks in these
-two provinces can be traced throughout the progression of the pandemic
-over time, although by the end of the period under consideration, other
-provinces had matched and even surpased their incidence rates, including
-Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
-to the south.
+variation in the incidence of COVID-19 in Spain, as well as the temporal
+progression of the disease in weekly averages. Our analysis begins on
+March 13, which is the first date when every province had reported at
+least one case of COVID-19. It can be seen that the highest incidence at
+this early date was in the provice of Álava, in the North of Spain.
+While not the most populous province, with a population of only 331,549,
+Álava has the highest GDP per capita of all provinces. Vitoria, its main
+city, is the capital of the Basque Country and has been the focus of
+efforts, along with San Sebastian and Bilbao, to develop a ``Global
+Basque City'' (Meijers et al., 2008). The other early focus of the
+pandemic in Spain was Madrid, which is the most populous province in the
+country and has the second highest GDP per capita after Álava. The early
+outbreaks in these two provinces can be traced throughout the
+progression of the pandemic over time, although by the end of the period
+under consideration, other provinces had matched and even surpased their
+incidence rates, including Segovia and Soria to the north of Madrid, and
+Ciudad Real and Albacete to the south.
 
@@
 
-To visualize the distribution of temperature and humidity we aggregate
-the provinces by Autonomous Community. In Figure
-\ref{fig:descriptives-temperature} the communities have been sorted by
+Figure \ref{fig:descriptives-temperature} shows the distribution of the
+environmental variables in Spain. For ease of visualization we aggregate
+the provinces by Autonomous Community. Each box-and-whisker in the
+figure represents the distribution of the variable for a community over
+the 30-day period. In the plot, the communities have been sorted by
 latitude, so that Principado de Asturias is the northernmost community,
-and Canarias the southernmost. There is a relatively wide range of
-values both within and between provinces over the 30 day period
-examined. The top panel of the figure shows the distribution of mean
-temperature. The lowest mean temperature for a community on any given
-day is 2.8C, and the highest is 22.4C for a range of approximately 20
-degrees. Likewise, there is a great deal of variability in humidity, as
-seen in the bottom panel of the figure, where the lowest mean humidity
-for any community is 48.3 and the highest is 99.6. The actual values for
-the provinces display somewhat more variability even.
+and Canarias the southernmost. As seen in the figure, there is a
+relatively wide range of values both within and between provinces over
+the 30-day period examined. The top panel of the figure shows the
+distribution of mean temperatures. The lowest mean temperature for a
+community on any given day was 2.8C, and the highest 22.4C, for a range
+of approximately 20 degrees. Likewise, there is a great deal of
+variability in humidity, as seen in the middle panel of the figure,
+where the lowest mean humidity for any community is 48.3 and the highest
+is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
+in the community. This variable displays the most variability within
+communities over time, but remains relatively constant across
+communities. It is important to note that these are summaries by
+community, and the actual values of these variables for the provinces
+display somewhat more variability.
 
@@
 
+Figure \ref{fig:map-controls} includes three maps that display the
+spatial variation of our control variables, namely GDP per capita,
+percentage of older adults in province, population density, and presence
+of mass transit systems. As seen there, GDP per capita is higher in
+Madrid and the northeast part of the country, mainly in Pais Vasco and
+Cataluña. Percentage of older adults is somewhat more checkered, with
+high values in Madrid and other provinces in the center-west part of the
+country, but also in some provinces in the north. Outside of provinces
+with major cities (e.g., Madrid; Bizkaia and Gipuzkoa in Pais Vasco;
+Pontevedra in Galicia), population density tends to be higher in
+provinces along the Mediterranean coast. The final panel in the figure
+shows the five provinces in the country that have higher order mass
+transit systems (e.g., metro).
+
 \begin{figure}
@@
 Figure \ref{fig:daily-correlations} shows the distribution of daily
-correlations of the independent variables with incidence of COVID-19,
-after log-transforming all variables. It can be seen there that the
-correlation of GDPpc and temperature (in its three definitions) have the
-strongest positive and negative correlations with incidence,
-respectively. Percentage of older adults displays somewhat weaker
-negative correlations with incidence, as does density. It can be seen
-that the humidity variable, in its three forms, tends to be possitively
-correlated with incidence of COVID-19.
+simple correlations of incidence of COVID-19 with the independent
+variables (with the exception of Transit, which is a categorical
+variable). These correlations are calculated after log-transforming all
+variables. As previously discussed, the environmental variables have a
+temporal lag and were calculated different time windows.
+
+It can be seen in the figure that temperature (in its three forms) has
+the highest simple correlation with incidence of COVID-19. After
+temperature, GDP per capita has the highest positive correlation with
+the dependent variable. The distribution of these correlations is also
+quite tight over the 30-day period of the study. Hours of sunshine tends
+to have a moderately high correlation with incidence of COVID-19,
+however the distribution of these correlations is more spread, and in
+some cases strays into negative values. A similar thing happens with
+humidity, which also tends to display more day to day variation in the
+correlation with the dependent variable. The percentage of older adults
+shows a relatively tight distribution of day-to-day correlations, and is
+negative. Population density, in contrast, tends to be negative, but is
+relatively spread, and on some days, the simple correlation between
+density and incidence of COVID-19 is weakly positive.
 
@@
 
-The goodness of fit of the three systems of equations is shown in Figure
-\ref{fig:goodness-of-fit}.
+Correlation analysis in the preceding section provides some insights
+about the potential associations between incidence of COVID-19 and the
+various environmental and control variables. In this section we estimate
+three spatial SUR models to test the differences between the various
+temporal lags and weighting schemes for the environmental variables.
+Accordingly, we define three models: Model 1, which is estimated using
+the lagged 8-day averages of the environmental variables (\emph{lag8});
+Model 2, which is estimated using the lagged 11-day averages of the
+environmental variables (\emph{lag11}); and finally, Model 3, which is
+estimated using the lagged 11-day \emph{weighted} averages of the
+environmental variables (\emph{lag11w}).
+
+To implement the SUR approach, we must define a matrix of spatial
+weights \(W\). In this case, we consider neighborhoods based on
+adjacency, based on the well-known rook criterion (two provinces are
+adjacent if they share a boundary, but not if they touch only at a
+vertex). We modify this criterion in two cases. First, we make an
+allowance for adjacency between the two islands in the Autonomous
+Community of Canarias in the Pacific (Las Palmas and Santa Cruz de
+Tenerife), which we assume are adjacent between each other. And
+secondly, we assume that Islas Baleares in the Mediterranean are
+adjacent to three provinces in Pais Catalans (i.e., Barcelona,
+Tarragona, and Castello). After matrix \(W\) has been specified,
+estimation of the model can proceed as usual.
+
+For estimation, we log-transform the dependent and quantitative
+independent variables. The only variable that is not transformed is the
+categorical variable for transity systems. A log-log transformation is
+appropriate to capture non-linear relationships between variables and
+provides a straightforward interpretation of the coefficients as
+percentage change. Furthermore, we introduce restrictions so that the
+coefficients of two of our control variables are constant over time,
+namely GDP per capita and percentage of older adults. We do not see an
+\emph{a priori} reason to let those two variables vary across equations,
+and the correlation analysis in Figure \ref{fig:daily-correlations} also
+suggest little temporal variation. In contrast, we let the spatial
+autocorrelation parameter, as well as the parameters of the rest of the
+independent variables (including the constant) to vary over
+time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious.
+These results are available in the source R markdown document.}.
+
+After estimation, we compare the goodness of fit of the three SUR
+models. Figure \ref{fig:goodness-of-fit} shows the equation-level
+coefficient of determination \(R^2\), one for each time period/day. As
+well, the overall coefficient of determination for the system is
+reported for each model \(\text{pooled}-R^2\). The general trend is
+identical for the three models, with the goodness-of-fit improving over
+time and plateuing around a value of \(R^2\) of 0.6. Model 1
+(\emph{lag8}) performs somewhat better in the first few equations/days,
+when the goodness-of-fit is relatively poor, and then again in the last
+few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
+perform well towards the end of the study period. The most balanced
+model in terms of equation-level goodness-of-fit appears to be Model 1
+(\emph{lag11}), and this impression is further supported with a slightly
+higher value of the \(\text{pooled}-R^2\). This analysis is in line with
+the incubation period reported by Lauer et al.~(2020), although not the
+use of a weighted average. For the remainder of the paper, we will adopt
+Model 2 (\emph{lag11}) as our best model. In the following section we
+discuss the results of the analysis in more depth.
 
@@
 
-Summary of best model.
+\hypertarget{results-and-discussion}{%
+\section{Results and Discussion}\label{results-and-discussion}}
+
+Table \ref{tab:summary-best-model} presents a summary.
 
@@
 
-\caption{\label{tab:summary-best-model}\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)}
+\caption{\label{tab:summary-best-model}\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)}
 \centering
@@
 \rowcolor{gray!6}  log(Older) & -0.737 & -0.737 & -0.737 & 0 & 0 & 1 & Constrained\\
-log(Density) & -0.220 & -0.097 & 0.153 & 19 & 1 & 10 & Non-constrained\\
-\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 10 & 9 & 11 & Non-constrained\\
+log(Density) & -0.220 & -0.097 & 0.153 & 19 & 0 & 11 & Non-constrained\\
+\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 8 & 8 & 14 & Non-constrained\\
 \addlinespace
@@
 log(Sunshine) & -0.258 & 0.097 & 0.206 & 7 & 2 & 21 & Non-constrained\\
+\rowcolor{gray!6}  Spatially lagged y (rho) & 0.015 & 0.142 & 0.528 & 12 & 3 & 15 & Non-constrained\\
 \bottomrule
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Figure \ref{fig:delta-time} shows the temporal evolution of the spatial
-autocorrelation coefficient (\(\lambda\)).
+Figure \ref{fig:results-time-1} shows the temporal evolution of the
+spatial autocorrelation coefficient (\(\rho\)) and the intercept of the
+model.
 
@@
 \centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/delta-time-1.pdf}
-\caption{\label{fig:delta-time}Temporal evolution of the spatial
-autocorrelation coefficient; dots are the point estimates and vertical
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/results-time-1-1.pdf}
+\caption{\label{fig:results-time-1}Temporal evolution of the spatial
+autocorrelation coefficient (rho) and the intercept of the model; dots
+are the point estimates and vertical lines are 95\% confidence
+intervals. In yellow is the period after the declaration of the state of
+emergency, and in orange is the period when only essential activities
+were allowed.}
+\end{figure}
+
+Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
+coefficients for the two control variables that were not fixed over
+time, i.e., \(\log(Density)\) and \(Transit\).
+
+\begin{figure}
+\centering
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-controls-time-1.pdf}
+\caption{\label{fig:beta-controls-time}Temporal evolution of coefficient
+for the control variables; dots are the point estimates and vertical
 lines are 95\% confidence intervals. In yellow is the period after the
@@
 
-Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the
-intercept.
+Figure \ref{fig:beta-environmental-time} shows the temporal evolution of
+the coefficient for the three environmental variables.
 
@@
 \centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-intercept-time-1.pdf}
-\caption{\label{fig:beta-intercept-time}Temporal evolution of intercept;
-dots are the point estimates and vertical lines are 95\% confidence
-intervals. In yellow is the period after the declaration of the state of
-emergency, and in orange is the period when only essential activities
-were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the
-coefficient for \(\log(Density)\).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-density-time-1.pdf}
-\caption{\label{fig:beta-density-time}Temporal evolution of coefficient
-for log(Density); dots are the point estimates and vertical lines are
-95\% confidence intervals. In yellow is the period after the declaration
-of the state of emergency, and in orange is the period when only
-essential activities were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-transit-time} shows the temporal evolution of the
-coefficient for \(Transit\).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-transit-time-1.pdf}
-\caption{\label{fig:beta-transit-time}Temporal evolution of coefficient
-for Transit; dots are the point estimates and vertical lines are 95\%
-confidence intervals. In yellow is the period after the declaration of
-the state of emergency, and in orange is the period when only essential
-activities were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the
-coefficient for \(\log(Humidity)\).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-humidity-time-1.pdf}
-\caption{\label{fig:beta-humidity-time}Temporal evolution of coefficient
-for log(Humidity); dots are the point estimates and vertical lines are
-95\% confidence intervals. In yellow is the period after the declaration
-of the state of emergency, and in orange is the period when only
-essential activities were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-temperature-time} shows the temporal evolution of
-the coefficient for \(\log(Temperature)\).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-temperature-time-1.pdf}
-\caption{\label{fig:beta-temperature-time}Temporal evolution of
-coefficient for log(Temperature); dots are the point estimates and
-vertical lines are 95\% confidence intervals. In yellow is the period
-after the declaration of the state of emergency, and in orange is the
-period when only essential activities were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the
-coefficient for \(\log(Sunshine + 0.1)\).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-sunshine-time-1.pdf}
-\caption{\label{fig:beta-sunshine-time}Temporal evolution of coefficient
-for log(Sunshine + 0.1); dots are the point estimates and vertical lines
-are 95\% confidence intervals. In yellow is the period after the
-declaration of the state of emergency, and in orange is the period when
-only essential activities were allowed.}
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmenta-time-1.pdf}
+\caption{\label{fig:beta-environmental-time}Temporal evolution of
+coefficient for the environmental variables; dots are the point
+estimates and vertical lines are 95\% confidence intervals. In yellow is
+the period after the declaration of the state of emergency, and in
+orange is the period when only essential activities were allowed.}
 \end{figure}
@@
 \item
-  HAY QUE INCLUIR INTERVALOS DE CONFIANZA EN LOS GRÁFICOS QUE HACEMOS DE
-  LOS COEFICIENTES DEL MODELO
-\item
   ¿hasta que punto las variables de control no recojen también factores
@@
   ha tenido menos incidencia.
-\item
-  TITULO: A spatial analysis of the environmental correlates ofCOVID-19
-  incidence in Spain ¿during the lookdown?
 \item

### 2020-04-20 README.Rmd
Finished editing Analysis, improved the presentation of some figures, added alternate models for reference, for sensitivity analysis. Next step: Results and discussion, and conclusions.

@@
 
-# A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
+# A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain
 
@@
 
->https://github.com/paezha/covid19-environmental-correlates
-
-Context, Data, and Methods
+https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
+
+Context and Data
 ============
@@
 
-Explain the rationale for selecting the variables.
-
-For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world.
-Controlling for socioeconomic characteristics the objective of this paper is observe the effect of clime on COVID-19  proliferation.
-
-Some questions are important to point out.
-Is not evident present a diary association between the temprature (resp.
-humidity) with declarate case.
-There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate
+The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
+In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
+Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
+As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
+These variables have been implicated in the viability and ease of transmission of similar viruses.
+In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.
+
+The first variable that we consider is GDP per capita.
+Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
+A18)}.
+The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated].
+In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect.
+In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic.
+Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional].
+Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown.
+On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.
+
+The percentaje of older adults (over 65) is the second variable that we consider as a control.
+Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g.
+@Novel2020epidemiological].
+However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection.
+The tool of choice in containing the spread of the disease has been social distancing.
+In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
+In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation.
+Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age].
+Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].
+
+The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
+The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g.
+cumulative incidence in urban areas like NYC).
+For this reason, we anticipate a positive relationship between population density and the incidence of the disease.
+
+The last control variable is the presence of mass transit systems in a province.
+Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g.
+metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
+Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
+After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.
 
@@
 
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
+\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t$ is a $N \times k_t$ matrix of the $k_t$ independent variables, with associated vector of coefficients $\beta_t$,; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
 -->
@@
 
-Equation \ref{eq:sur-sim}) can be rewriten in compact form:
+Equation (\ref{eq:sur-sim}) can be rewriten in compact form:
 
@@
 \begin{equation}
-y = X \beta + \epsilon
+\bf{Y} = \bf{X} \beta + \epsilon
+\label{eq:sur-sim-block}
+\end{equation}
+
+\noindent where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
+(with $K = \sum_t{k_t}$) and $\epsilon$ is an $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:
+
+\begin{equation}
+E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
 \end{equation}
@@
 
-As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
+\noindent where ![$\bf{Y}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BY%7D%24) is now a vector of dimension ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24), ![$\bf{X}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BX%7D%24) is a block-diagonal matrix ![$NT \times K$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%20K%24)
+(with ![$K = \sum_t{k_t}$](https://render.githubusercontent.com/render/math?math=%24K%20%3D%20%5Csum_t%7Bk_t%7D%24)) and ![$\epsilon$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon%24) is an ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24) vector.
+Using the Kronecker product notation (![$\otimes$](https://render.githubusercontent.com/render/math?math=%24%5Cotimes%24)), the error matrix structure is expressed concisely as:
+
+![\begin{equation} E\[\epsilon \epsilon'\]=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'}) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5CSigma%20%5Cotimes%20I_N%20%3B%20%5C%20%5CSigma%3D(%5Csigma_%7Btt'%7D)%20%5Cend%7Bequation%7D)
+
+As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
 When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
@@
 \label{eq:sur-slm}
-\bf{A}y = X \beta + \epsilon \\
-\epsilon =N(0,\Omega)
+\begin{aligned}
+\bf{AY} = \bf{X} \beta + \epsilon \\
+\epsilon =N(0,\bf{\Sigma})
+\end{aligned}
 \end{equation}
 
-\noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
-
-This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$).
+\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ is the spatially lagged dependent variable, and $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.
+
+This specification assumes that outcome ($y_{st}$) at location $s$ and time $t$ is partially determined by the weighted average ($Wy_{st}$) of the outcome in neighboring provinces, with neighborhood defined by matrix $W$ of spatial weights.
+In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way.
+The coefficients of the spatially lagged variable are estimated for each time period $\rho_t$ and identify the intensity and the sign of the contagion effect.
+It is possible test the null hypothesis of identical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
 The correspond Wald test is available in the `R` package `spsur`.
@@
 
-![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
-
-\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
-
-This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period ![$\lambda_t$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence (![$\lambda_i=\lambda_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_i%3D%5Clambda_j%2C%20%5Cforall%20i%2Cj%24)).
+![\begin{equation} \label{eq:sur-slm} \begin{aligned} \bf{AY} = \bf{X} \beta + \epsilon \\ \epsilon =N(0,\bf{\Sigma}) \end{aligned} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbegin%7Baligned%7D%20%5Cbf%7BAY%7D%20%3D%20%5Cbf%7BX%7D%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5Cbf%7B%5CSigma%7D)%20%5Cend%7Baligned%7D%20%5Cend%7Bequation%7D)
+
+\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
+
+This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights.
+In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way.
+The coefficients of the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identify the intensity and the sign of the contagion effect.
+It is possible test the null hypothesis of identical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)).
 The correspond Wald test is available in the `R` package `spsur`.
@@
 The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
-Another alternative methodologies could be use.
-By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g.
-Elhorst 2014, Cap.
-4), but those models are more dont take account correlation between errors.
-Therefore, a spatial SUR approach is more reasonable for our purpose.
-
-Analysis and Results
+
+Analysis
 ===================
@@
   geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 4.9, 
+               x = madrid_xy[1] - 9, 
                y = madrid_xy[2] + 0.6,
                family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
+  geom_segment(x = madrid_xy[1] - 6, 
                y = madrid_xy[2] + 0.5, 
@@
         plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 4.9, y = madrid_xy[2] + 0.6) +
-  geom_segment(x = madrid_xy[1] - 3.4, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
+  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 9, y = madrid_xy[2] + 0.6) +
+  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
 
 week13.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  #filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
   geom_sf_text(label = "Madrid",
-               x = madrid_xy[1] - 4.9, 
+               x = madrid_xy[1] - 9, 
                y = madrid_xy[2] + 0.6,
                family = "serif") +
-  geom_segment(x = madrid_xy[1] - 3.4, 
+  geom_segment(x = madrid_xy[1] - 6, 
                y = madrid_xy[2] + 0.5, 
@@
   ylab("Value") +
-  facet_wrap(~ Variable, nrow = 2, 
+  facet_wrap(~ Variable, nrow = 3, 
              scales = "free_y")
@@
 
-```{r control-map, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
+Figure \ref{fig:map-controls} includes three maps that display the spatial variation of our control variables, namely GDP per capita, percentage of older adults in province, population density, and presence of mass transit systems.
+As seen there, GDP per capita is higher in Madrid and the northeast part of the country, mainly in Pais Vasco and Cataluña.
+Percentage of older adults is somewhat more checkered, with high values in Madrid and other provinces in the center-west part of the country, but also in some provinces in the north.
+Outside of provinces with major cities (e.g., Madrid; Bizkaia and Gipuzkoa in Pais Vasco; Pontevedra in Galicia), population density tends to be higher in provinces along the Mediterranean coast.
+The final panel in the figure shows the five provinces in the country that have higher order mass transit systems (e.g., metro).
+
+```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province", warning=FALSE, cache = TRUE}
+# GDP per capita:
+q <- quantile(provinces_spain$GDPpc)
 gdppc.plot <- provinces_spain %>%
-  filter(CCAA != "Canarias") %>%
+  mutate(GDPpc = GDPpc,
+         GDPpc = case_when(GDPpc < q[2] ~ "1st",
+                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
+                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
+                             GDPpc > q[4] ~ "4th")) %>%
+  #filter(CCAA != "Canarias") %>%
   ggplot() +
-  geom_sf(aes(fill = GDPpc)) +
-  scale_fill_distiller(name = "GDP per capita", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
+  geom_sf(aes(fill = GDPpc), color = "gray69") +
+  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  ggtitle("GDP per capita") +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
-        legend.position = "bottom")
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 9, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 # Older adults:
+q <- quantile(provinces_spain$Older)
 older.plot <- provinces_spain %>% 
-  filter(CCAA != "Canarias") %>%
+  mutate(Older = Older,
+         Older = case_when(Older < q[2] ~ "1st",
+                             Older > q[2] & Older <= q[3]  ~ "2nd",
+                             Older > q[3] & Older <= q[4] ~ "3rd",
+                             Older > q[4] ~ "4th")) %>%
+  #filter(CCAA != "Canarias") %>%
   ggplot() +
-  geom_sf(aes(fill = as.numeric(Older))) +
-  scale_fill_distiller(name = "Percentage of older adults", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
+  geom_sf(aes(fill = Older), color = "gray69") +
+  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  ggtitle("Percentage of older adults") +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
-        legend.position = "bottom")
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 9, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 # Population density:
+q <- as.numeric(quantile(provinces_spain$Density))
 density.plot <- provinces_spain %>% 
-  filter(CCAA != "Canarias") %>%
+  mutate(Density = as.numeric(Density),
+         Density = case_when(Density < q[2] ~ "1st",
+                             Density > q[2] & Density <= q[3]  ~ "2nd",
+                             Density > q[3] & Density <= q[4] ~ "3rd",
+                             Density > q[4] ~ "4th")) %>%
+  #filter(CCAA != "Canarias") %>%
   ggplot() +
-  geom_sf(aes(fill = as.numeric(Density))) +
-  scale_fill_distiller(name = "Population Density", 
-                       palette = "Blues", 
-                       direction = 1) +
-  theme_tufte() +
+  geom_sf(aes(fill = Density), color = "gray69") +
+  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
+  ggtitle("Population density") +
+  theme_bw(base_size = 9) +
   theme(axis.text = element_blank(),
-        legend.position = "bottom")
-
-grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
-```
-
-Figure \ref{fig:daily-correlations} shows the distribution of daily correlations of the independent variables with incidence of COVID-19, after log-transforming all variables.
-It can be seen there that the correlation of GDPpc and temperature (in its three definitions) have the strongest positive and negative correlations with incidence, respectively.
-Percentage of older adults displays somewhat weaker negative correlations with incidence, as does density.
-It can be seen that the humidity variable, in its three forms, tends to be possitively correlated with incidence of COVID-19.
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 9, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
+
+# Transit systems:
+transit.plot <- provinces_spain %>% 
+  mutate(Transit = factor(Transit, 
+                          levels = c(0, 1),
+                          labels = c("No",
+                                     "Yes"))) %>%
+  #filter(CCAA != "Canarias") %>%
+  ggplot() +
+  geom_sf(aes(fill = Transit), color = "gray69") +
+  scale_fill_manual(name = "Transit", values=c("#FFFEDE", "#D5610D")) +
+  ggtitle("Mass transit systems") +
+  theme_bw(base_size = 9) +
+  theme(axis.text = element_blank(),
+        legend.position = "bottom") +
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 9, 
+               y = madrid_xy[2] + 0.6,
+               family = "serif") +
+  geom_segment(x = madrid_xy[1] - 6, 
+               y = madrid_xy[2] + 0.5, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
+
+# Define layout for plot
+lay <- rbind(c(1, 2),
+             c(3, 4))
+
+grid.arrange(gdppc.plot, older.plot, density.plot, transit.plot, layout_matrix = lay)
+```
+
+Figure \ref{fig:daily-correlations} shows the distribution of daily simple correlations of incidence of COVID-19 with the independent variables (with the exception of Transit, which is a categorical variable).
+These correlations are calculated after log-transforming all variables.
+As previously discussed, the environmental variables have a temporal lag and were calculated different time windows.
+
+It can be seen in the figure that temperature (in its three forms) has the highest simple correlation with incidence of COVID-19.
+After temperature, GDP per capita has the highest positive correlation with the dependent variable.
+The distribution of these correlations is also quite tight over the 30-day period of the study.
+Hours of sunshine tends to have a moderately high correlation with incidence of COVID-19, however the distribution of these correlations is more spread, and in some cases strays into negative values.
+A similar thing happens with humidity, which also tends to display more day to day variation in the correlation with the dependent variable.
+The percentage of older adults shows a relatively tight distribution of day-to-day correlations, and is negative.
+Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive.
 
@@
 ## SUR Models
+
+Correlation analysis in the preceding section provides some insights about the potential associations between incidence of COVID-19 and the various environmental and control variables.
+In this section we estimate three spatial SUR models to test the differences between the various temporal lags and weighting schemes for the environmental variables.
+Accordingly, we define three models: Model 1, which is estimated using the lagged 8-day averages of the environmental variables (*lag8*); Model 2, which is estimated using the lagged 11-day averages of the environmental variables (*lag11*); and finally, Model 3, which is estimated using the lagged 11-day *weighted* averages of the environmental variables (*lag11w*).
+
+To implement the SUR approach, we must define a matrix of spatial weights $W$.
+In this case, we consider neighborhoods based on adjacency, based on the well-known rook criterion (two provinces are adjacent if they share a boundary, but not if they touch only at a vertex).
+We modify this criterion in two cases.
+First, we make an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are adjacent between each other.
+And secondly, we assume that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello).
+After matrix $W$ has been specified, estimation of the model can proceed as usual.
+
+For estimation, we log-transform the dependent and quantitative independent variables.
+The only variable that is not transformed is the categorical variable for transity systems.
+A log-log transformation is appropriate to capture non-linear relationships between variables and provides a straightforward interpretation of the coefficients as percentage change.
+Furthermore, we introduce restrictions so that the coefficients of two of our control variables are constant over time, namely GDP per capita and percentage of older adults.
+We do not see an  *a priori* reason to let those two variables vary across equations, and the correlation analysis in Figure \ref{fig:daily-correlations} also suggest little temporal variation.
+In contrast, we let the spatial autocorrelation parameter, as well as the parameters of the rest of the independent variables (including the constant) to vary over time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious.
+These results are available in the source R markdown document.}.
 
@@
         col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
-        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
+        caption = "\\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)") %>%
   #kable_styling(latex_options = c("striped", "scale_down")) %>%
@@
 
-Discussion
-===================
-
-Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\lambda$).
-
-```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
+```{r summary-best-model-all, include=FALSE, eval=FALSE}
+best_model_all <- sur.slm_lag11_all
+
+data.frame(Coefficient = c(best_model_all$coefficients, 
+                           # Copy best model
+                           best_model_all$deltas), 
+           sd = c(best_model_all$rest.se, 
+                  best_model_all$deltas.se)) %>%
+  rownames_to_column(var = "Variable") %>%
+  mutate(zscore = Coefficient/sd,
+         pval = 2 * pnorm(-abs(zscore)),
+         Variable = c(rep(c("Intercept", 
+                            "log(GDPpc)", 
+                            "log(Older)",
+                            "log(Density)",
+                            "Transit",
+                            "log(Humidity)",
+                            "log(Temperature)",
+                            "log(Sunshine)"),
+                          30),
+                      rep("Spatially lagged y (rho)",
+                          30)),
+         Variable = factor(Variable, 
+                           levels = c("Intercept", 
+                                      "log(GDPpc)", 
+                                      "log(Older)", 
+                                      "log(Density)",
+                                      "Transit", 
+                                      "log(Humidity)",
+                                      "log(Temperature)",
+                                      "log(Sunshine)",
+                                      "Spatially lagged y (rho)"))) %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Coefficient), 
+            Mean = mean(Coefficient), 
+            Max = max(Coefficient),
+            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
+            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
+            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Non-constrained", 
+                  "Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained")) %>%
+  kable(#"latex",
+    "html",
+    booktabs = TRUE,
+    digits = 3,
+    col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
+    caption = "\\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)") %>%
+  #kable_styling(latex_options = c("striped", "scale_down")) %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
+  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
+                       "Non-constrained: coefficient was allowed to vary across equations",
+                       "Constrained: coefficient as constant across equations"))
+```
+
+```{r function-for-plotting-coefficients, include=FALSE}
+# Define a function for plotting the coefficients over time including their confidence intervals and significance
+plot_coeff <- function(model = model, n = n){
   data.frame(Date = seq(ymd("2020-03-13"), 
@@
                         by = "days"),
-           Deltas = sur.slm_lag11$deltas,
-           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
+             Beta = matrix(model$coefficients[-c(2,3)],ncol = T)[n,],
+             sd = matrix(model$rest.se[-c(2,3)], ncol = T)[n,]) %>%
+    mutate(zscore = Beta/sd,
+           pval = 2 * pnorm(-abs(zscore)),
+           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
+           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
+           pval = case_when(pval <= 0.05 ~ "p <= 0.05",
+                            pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
                             TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Deltas, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.
-
-```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
-#Intercept
-n = 1
-data.frame(Date = seq(ymd("2020-03-13"), 
+    ggplot(aes(x = Date, y = Beta, color = pval)) +
+    geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
+    geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+    geom_point(size = 1) +
+    geom_errorbar(aes(x = Date,
+                      ymin = CI90low,
+                      ymax = CI90up), 
+                  size = 0.6,
+                  width = 0.8) +
+    geom_linerange(aes(x = Date,
+                       ymin = CI95low,
+                       ymax = CI95up),
+                   size = 0.4) +
+    scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
+    geom_hline(yintercept = 0) +
+    theme_bw()
+  }
+```
+
+```{r function-extract-legend, include=FALSE}
+# Define a function to extract a legend from a ggplot
+#extract legend
+#https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs
+g_legend<-function(a.gplot){
+  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)}
+```
+
+Figure \ref{fig:results-time-1} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$) and the intercept of the model.
+
+```{r results-time-1, cache = TRUE, fig.height= 4, echo=FALSE, fig.cap="\\label{fig:results-time-1}Temporal evolution of the spatial autocorrelation coefficient (rho) and the intercept of the model; dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
+rho.plot <- data.frame(Date = seq(ymd("2020-03-13"), 
                       ymd("2020-04-11"), 
                       by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
+           Beta = best_model$deltas,
+           sd = best_model$deltas.se) %>%
+  mutate(zscore = Beta/sd,
+         pval = 2 * pnorm(-abs(zscore)),
+           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
+           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
+           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
+         pval = case_when(pval <= 0.05 ~ "p <= 0.05",
+                          pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
                          TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Density)$.
-
-```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)", cache = TRUE}
+  ggplot(aes(x = Date, y = Beta, color = pval)) +
+  geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
+  geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
+            ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+  geom_point(size = 1) +
+  geom_errorbar(aes(x = Date,
+                     ymin = CI90low,
+                    ymax = CI90up),
+                 size = 0.6,
+                width = 0.8) +
+  geom_linerange(aes(x = Date,
+                     ymin = CI95low,
+                    ymax = CI95up),
+                 size = 0.4) +
+  scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
+  geom_hline(yintercept = 0) +
+  theme_bw() +
+  ylab("rho") +
+  theme(legend.position = "bottom")
+
+common_legend <- g_legend(rho.plot)
+
+# Plot without legend
+rho.plot <- rho.plot +
+   theme(legend.position = "none")
+
+#Intercept
+intercept.plot <- plot_coeff(model = best_model, n = 1) + 
+  ylab("Beta (Intercept)") +
+  theme(legend.position = "none")
+
+# Define layout for plot
+lay <- rbind(c(1, 2),
+             c(3, 3))
+
+grid.arrange(rho.plot, intercept.plot, common_legend, 
+             layout_matrix = lay,
+             heights = c(5, 1))
+```
+
+Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ and $Transit$.
+
+```{r beta-controls-time, echo=FALSE,  fig.height= 4, fig.cap="\\label{fig:beta-controls-time}Temporal evolution of coefficient for the control variables; dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
 #Density:
-n = 2 # Although this is the fourth coefficient, after intercept, log(GDPpc), and log(Older), we start at 2 here because the two constrained coefficients are removed from the matrix of coefficients 
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-transit-time} shows the temporal evolution of the coefficient for $Transit$.
-
-```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit", cache = TRUE}
-n = 3
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the coefficient for $\log(Humidity)$.
-
-```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)", cache = TRUE}
-n = 4
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-temperature-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.
-
-```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)", cache = TRUE}
-n = 5
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the coefficient for $\log(Sunshine + 0.1)$.
-
-```{r beta-sunshine-time, echo=FALSE, fig.cap="\\label{fig:beta-sunshine-time}Temporal evolution of coefficient for log(Sunshine + 0.1)", cache = TRUE}
-n = 6
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
+density.plot <- plot_coeff(model = best_model, n = 2) + 
+  ylab("Beta (log[Density]))") +
+  theme(legend.position = "none")
+
+# Transit
+transit.plot <- plot_coeff(model = best_model, n = 3) + 
+  ylab("Beta (Transit)") +
+  theme(legend.position = "none")
+
+# Define layout for plot
+lay <- rbind(c(1, 2),
+             c(3, 3))
+
+grid.arrange(density.plot, transit.plot, common_legend, 
+             layout_matrix = lay,
+             heights = c(5, 1))
+```
+
+Figure \ref{fig:beta-environmental-time} shows the temporal evolution of the coefficient for the three environmental variables.
+
+```{r beta-environmenta-time, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:beta-environmental-time}Temporal evolution of coefficient for the environmental variables; dots are the point estimates and vertical lines are 95% confidence intervals.
+In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
+# Humidity
+humidity.plot <- plot_coeff(model = best_model, n = 4) + 
+  ylab("Beta (log[Humidity])") +
+  theme(legend.position = "none")
+
+# Temperature
+temperature.plot <- plot_coeff(model = best_model, n = 5) + 
+  ylab("Beta (log[Temperature])") +
+  theme(legend.position = "none")
+
+# Sunshine
+sunshine.plot <- plot_coeff(model = best_model, n = 6) + 
+  ylab("Beta (log[Sunshine + 0.1])") +
+  theme(legend.position = "none")
+
+# Define layout for plot
+lay <- rbind(c(1, 1, 2, 2),
+             c(1, 1, 2, 2),
+             c(NA, 3, 3, NA),
+             c(NA, 3, 3, NA),
+             c(NA, 4, 4, NA))
+
+grid.arrange(humidity.plot, temperature.plot, sunshine.plot, common_legend, 
+             layout_matrix = lay,
+             heights = c(4, 4, 4, 4, 1))
+
+```
+
+```{r impacts, echo=FALSE, include=FALSE}
+Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
+trsur <- spatialreg::trW(Wsur, type = "MC")
+impacts.best_model <- impactspsur(best_model, tr = trsur,R = 1000)
+
+impacts.best_model[[1]]
+impacts.best_model[[2]]
+impacts.best_model[[3]]
+
+myimpact <- matrix(0,ncol = 5, nrow = 29)
+for (i in 2:30){
+  myimpact[i-1,] <- impacts.best_model[[i]]$res$direct
+}
+
+matplot(myimpact, type='l')
+
 ```
@@
 
+Limitaciones
+
+* La temperatura puedes ser un factor pero cabría esperar que no tuviera un impacto lineal.
+Por el contrario deberiamos esperar un 'punto de corte': Una temperatua mantenida superior a X grados durante siete días sea la mejor forma de incorporarla al modelo.
+En nuestro modelo (log-log) en incremento en un 1% de la temperatura se asocia con un incremento beta% en la incidencia.
+Esto es lo mismo si la temperatura es baja que si es alta.
+
+* Idem para la humedad y horas de sol
+
+* Los datos son 'provisionales'.
+Hay gran confusión sobre la incidencia real.
+La ausencia de test de diagnóstico PCRs al inicio de la pandemia (también ahora) puede desvirtuar el número de casos dianosticados.
+
+* Los datos oficiales (que tampoco son fiables) son reportados a nivel de Comunidades Autónomas.
+La recopilación de datos a nivel de provincial son el resultado de un esfuerzo colaborativo de recopilacion entre distintas fuentes (principalmente gobiernos locales).
+Nuevamente puede haber sesgos importanes.
+
+* En carácter insular de las Islas Canarias y de las Islas baleares no se ha tenido en cuenta.
+
+* Baleares se ha linkado artificialmente con 3/4 provincias y debería de haberse dejado aislada para respetar su carácter insular al definir la matriz W.
+
+* La incidencia depende de un estado inicial.
+Al inicio del estudio había provincias en las que la epidemia estaba muy desarrollada (Madrid/Alava) mientras que en otras apenas habías casos.
+Este hecho no ha sido considerado en el modelo.
+Decretar el confinamiento debe tener distintos impactos entre provincias.
+QUIZAR METER UNA VARAIBLE DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
+
+* No se ha controlado por el sistema sanitario de cada provincia.
+Uno de los principales focos de contagio han sido los hospitales y los centros de salud.
+En aquellas provincias donde se ha promovido el mensaje "NO IR AL MEDICO" han presentado menor incidencia.
+
+* ¿hasta que punto las variables de control no recojen también factores climáticos? por ejemplo, la gente jóven vive en el sur de España que ha tenido menos incidencia.
+
+* Las estaciones meteorológicas para la obtencion de datos climáticos has sido elegidas aleatoriamente (una para cada provincia).
+otra seleccion puede dar otros resultados.
+AQUI SE PODRÍA HACER EL ESFUERZO DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR PROVINCIA
+
+*IDEM para la humedad. idem para sunshine
+
+* HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA DARLE RELEVANCIA A LA ESTIMACION SUR
+
 Acknowledgments {#acknowledgments .unnumbered}

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
eso

--- 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 
+**Direct, Indirect and Total effect**
+
+For classical regression model, the parameter interpretation is easy since parameters can be interpreted as the partial derivatives of the dependent variable relative to the explanatory variable.
+For spatial regression models, parameter interpretation becomes richer but more complicated  (@golgher2016interpret).
+Changes in the value of an explanatory variable in a province can have influence in other provinces.
+In essence, the idea is that spatial dependence expands the information set to include information from neighbouring regions.
+The results of @LeSage2009 can be extended to the spatial SUR framework.
+Note that,
+
+According to @elhors2014, these partial derivatives have the following interpretation: if an explanatory variable ($X_k$) in a particular province changes, not only the incidence rate in that province change, also incidence rates in other provinces change.
+Therefore, a change in $X_k$ in a particular province has a *direct effect* on that province, but also an *indirect effect* on neighbouring provinces.
+Note that the ith diagonal element of the matrix of partial derivatives represents the direct effect on the ith province and non-diagonal elements of ith row of the matrix of partial derivatives represent the indirect effects on other provinces.
+In order to obtain a global indicator, the direct effect is calculated as the mean of the diagonal elements and captures the average change in incidence ratio caused by the change in $X_k$.
+Likewise, a global indicator of the indirect effects is calculated as the mean of the non-diagonal elements.
+The total effect is the sum of direct and indirect effects.
+Finally, note that if $\lambda_k = 0$, the indirect effects are 0 and the direct effects are equal to $\beta_kt$.
+
 More words go here.

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Started working on the results and discussion

--- 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 #linenumbers: true
-#numbersections: true
+numbersections: true
 csl: elsevier-harvard.csl
@@
 
-Introduction
+Introduction {#introduction}
 ==========================
@@
 
-Context and Data
+Context and Data {#context-and-data}
 ============
 
-## Covid-19 in Spain
+## Covid-19 in Spain {#covid-19-in-spain}
 
@@
 
-## Selection of Variables
+## Selection of Variables {#variables}
 
@@
 
-## Data
+## Data {#data}
 
@@
 
-Methods: the Spatial SUR Model
+Methods: the Spatial SUR Model {#methods}
 ===================
@@
 
-Analysis
+A note regarding the interpretation of the model is in order.
+It is well-known that coefficients in linear regression models are partial derivatives of the dependent variable with respect to the independent variables, and therefore directly give the marginal effects or rates of change.
+Spatially lagged models, however, are no longer linear.
+The intuition behind the non-linearity is that the spatial lag expands the information set to include information from neighbouring regions: in other words, the value of an explanatory variable in a spatial unit can have influence in other spatial units via the spatial lag.
+This makes interpretation of the coefficients less straightforward but also richer [@Golgher2016interpret].
+The results of LeSage [-@LeSage2009introduction] for cross-sectional spatial lag models can be extended to the spatial SUR framework.
+Note that, according to Elhorst [-@Elhorst2014spatial], the partial derivatives have the following interpretation: if an explanatory variable ($X_k$) in a particular province changes, not only the incidence rate in that province changes, also incidence rates in other provinces change via the contagion effect.
+Therefore, a change in $X_k$ in a particular province has a *direct effect* on that province, but also an *indirect effect* on neighbouring provinces.
+In this way, the $i$th diagonal element of the matrix of partial derivatives represents the direct effect on the $i$th province, whereas the non-diagonal elements of $i$th row of the matrix of partial derivatives represent the indirect effects on other provinces.
+In order to obtain a global indicator, the direct effect is calculated as the mean of the diagonal elements and captures the average change in incidence ratio caused by the change in $X_k$.
+Likewise, a global indicator of the indirect effects is calculated as the mean of the non-diagonal elements.
+The total effect is the sum of direct and indirect effects.
+Finally, note that if $\rho_k = 0$, the indirect effects are 0 and the direct effects are equal to $\beta_kt$.
+
+Analysis {#analysis}
 ===================
 
-## Exploratory Data Analysis
+## Exploratory Data Analysis {#eda}
 
@@
   geom_boxplot() +
-  theme_tufte() +
+  theme_bw() +
   theme(axis.text.x = element_text(angle = 90)) +
@@
   geom_boxplot() +
-  theme_tufte() +
+  theme_bw() +
   theme(axis.text.x = element_text(angle = 90))
@@
 
-## SUR Models
+## SUR Models {#sur-models}
 
@@
 These results are available in the source R markdown document.}.
+This will be useful to detect whether there are behavioral adaptations at the population level over the course of the period examined.
+As an example of behavioral adaptations, the effect of density might weaken over time, in the measure that the effects of the lockdown are felt: at full compliance with the lockdown, with people practicing social avoidance, density might matter less than other factors.
 
@@
 
-Results and Discussion
+Results and Discussion {#results-and-discussion}
 ===================
 
-Table \ref{tab:summary-best-model} presents a summary.
+Table \ref{tab:summary-best-model} presents a summary of the results of Model 2 (*lag11*).
+Recall that two coefficients were constrained and are estimated only for the first equation of the system, and thus are assumed to be constant over time.
+These are the coefficients corresponding to GDP per capita and percentage of older adults.
+Both coefficients are significant at $p<=0.05$.
+The sign of the coefficient for GDP per capita is positive, which indicates that wealthier regions tend to have a higher incidence of COVID-19.
+This is in line with the idea that the epidemic started earlier in wealthier places due to their connections to a globalized world.
+The sign of the coefficient for percentage of older adults, on the other hand, is negative.
+As previously discussed, the level of social contact of older adults even under normal circumstances tends to be lower than for younger people.
+As a consequence, places with larger populations of older adults appear to have a natural level of social distancing in place.
+It is important to note that this does not detract from evidence that older adults are more vulnerable individually and in institutional settings, where their case mortality rates are perhaps the highest of all age groups.
+Instead, this result indicates that their presence in the community at large tends to depress transmission of the virus.
+
+Of the two other control variables, the coefficient of population density is significant at $p<=0.05$ in 11 equations, and not significant in 19.
+The coefficient for transit is significant at $p<=0.10$ in 22 equations, and of those, significant at $p<=0.05$ in 14 equations.
+The next four variables are environmental factors.
+The coefficient for humidity is significant at $p<=0.10$ in 20 equations, and of those, significant at $p<=0.05$ in 19 equations.
+Of the environmental variables, temperature is the only variable that has significant coefficients in every equation at $p<=0.05$.
+Finally, sunshine has significant coefficients at $p<=0.10$ in 18 equations, including 15 coefficients that are significant at $p<=0.05$.
 
@@
 
-Figure \ref{fig:results-time-1} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$) and the intercept of the model.
+To better understand the results, we proceed to plot the coefficients in their temporal sequence.
+We begin with the evolution of the spatial autocorrelation coefficient ($\rho$) and the intercept of the model, both of which are shown in Figure \ref{fig:results-time-1}.
 
@@
 Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ and $Transit$.
+In Section \ref{variables} we had anticipated a positive sign for the coefficient of this variable.
+The negative sign in the results is therefore somewhat surprising.
+Upon reflection, however, it seems reasonable that the sign would be negative during the period under analysis.
 
@@
 
-```{r beta-environmenta-time, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:beta-environmental-time}Temporal evolution of coefficient for the environmental variables; dots are the point estimates and vertical lines are 95% confidence intervals.
+```{r beta-environmental-time, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:beta-environmental-time}Temporal evolution of coefficient for the environmental variables; dots are the point estimates and vertical lines are 95% confidence intervals.
 In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
@@
 sunshine.plot <- plot_coeff(model = best_model, n = 6) + 
-  ylab("Beta (log[Sunshine + 0.1])") +
+  ylab("Beta (log[Sunshine])") +
   theme(legend.position = "none")
@@
 
-```{r impacts, echo=FALSE, include=FALSE}
+```{r impacts, evalcho=FALSE, include=FALSE}
 Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
@@
 
-Concluding Remarks
+```{r calculate-impacts-best-model, echo=FALSE, cache = TRUE}
+# Calculate impacts
+Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
+trsur <- spatialreg::trW(Wsur, type = "MC")
+best_model.impacts <- impactspsur(best_model, tr = trsur,R = 1000)
+
+# Organize in dataframe
+best_model.impacts <- data.frame(Impact = unlist(best_model.impacts)) %>%
+  # Identify point estimates of effects for each variable and simulated values 
+  mutate(Variable = c(c(rep(c("log(GDPpc)", 
+                              "log(Older)",
+                              "log(Density)",
+                              "Transit",
+                              "log(Humidity)",
+                              "log(Temperature)",
+                              "log(Sunshine)"),
+                            3),
+                        rep("junk", 21000)),
+                      rep(c(rep(c("log(Density)",
+                                  "Transit",
+                                  "log(Humidity)",
+                                  "log(Temperature)",
+                                  "log(Sunshine)"),
+                                3),
+                            rep("junk", 15000)), # These are the simulated values
+                          29))) %>%
+  filter(Variable != "junk") %>% # Remove all simulations, keep only point estimates
+  # Add type of effect and date
+  mutate(Type = c(rep(c("Direct", 
+                        "Indirect", 
+                        "Total"),
+                      each = 7),
+                  rep(rep(c("Direct", 
+                            "Indirect", 
+                            "Total"),
+                          each = 5),
+                      29)),
+         Date = c(rep(ymd("2020-03-13"),
+                      21),
+                  rep(seq(ymd("2020-03-14"),
+                          ymd("2020-04-11"), 
+                          by = "days"), 
+                      each = 15))) %>%
+  # Convert to factors
+  mutate(Variable = factor(Variable,
+                           levels = c("log(GDPpc)", 
+                              "log(Older)",
+                              "log(Density)",
+                              "Transit",
+                              "log(Humidity)",
+                              "log(Temperature)",
+                              "log(Sunshine)")),
+         Type = factor(Type,
+                       levels = c("Direct", 
+                            "Indirect", 
+                            "Total")))
+```
+
+The effects of the different variables are summarized in Table \ref{tab:summary-impacts-best-model}.
+
+```{r summary-impacts-best-model, echo=FALSE}
+rbind(best_model.impacts %>%
+  filter(Type == "Direct") %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Impact), 
+            Mean = mean(Impact), 
+            Max = max(Impact)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained")),
+  best_model.impacts %>%
+  filter(Type == "Indirect") %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Impact), 
+            Mean = mean(Impact), 
+            Max = max(Impact)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained")),
+  best_model.impacts %>%
+  filter(Type == "Total") %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Impact), 
+            Mean = mean(Impact), 
+            Max = max(Impact)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained"))) %>%
+  kable("latex",
+        #"html",
+        booktabs = TRUE,
+        digits = 3,
+        col.names = c("Variable", "Min", "Mean", "Max", "Note"),
+        caption = "\\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)") %>%
+  kable_styling(latex_options = c("striped")) %>%
+  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  pack_rows("Direct Effects", 1, 7) %>%
+  pack_rows("Indirect Effects", 8, 14) %>%
+  pack_rows("Total Effects", 15, 21) %>%
+  footnote(general = c("Non-constrained: coefficient was allowed to vary across equations",
+                       "Constrained: coefficient as constant across equations"))
+```
+
+The temporal variation of the effects is shown in Figure \ref{fig:plot-impacts}.
+
+```{r plot-impacts, echo=FALSE, fig.cap="\\label{fig:plot-impacts}Temporal evolution of effects by date; the impact is percentage change in incidence as a result of a one percent change in value of variable."}
+best_model.impacts %>%
+  filter(Variable != "log(GDPpc)", 
+         Variable != "log(Older)") %>%
+  ggplot(aes(x = Date, 
+             y = Impact, 
+             color = Variable)) +
+  geom_line(size = 0.8) +
+  scale_color_brewer(palette = "Set1") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90),
+        legend.position = "bottom") +
+  facet_wrap(~ Type, ncol = 3)
+
+```
+
+Concluding Remarks {#conclusion}
 ===================

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Started working on the results and discussion

@@
 
-\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{5}
 % Pandoc toggle for numbering sections (defaults to be off)
-\setcounter{secnumdepth}{0}
 
@@
 
-\hypertarget{selection-of-variables}{%
-\subsection{Selection of Variables}\label{selection-of-variables}}
+\hypertarget{variables}{%
+\subsection{Selection of Variables}\label{variables}}
 
@@
 
-\hypertarget{methods-the-spatial-sur-model}{%
-\section{Methods: the Spatial SUR
-Model}\label{methods-the-spatial-sur-model}}
+\hypertarget{methods}{%
+\section{Methods: the Spatial SUR Model}\label{methods}}
 
@@
 
+A note regarding the interpretation of the model is in order. It is
+well-known that coefficients in linear regression models are partial
+derivatives of the dependent variable with respect to the independent
+variables, and therefore directly give the marginal effects or rates of
+change. Spatially lagged models, however, are no longer linear. The
+intuition behind the non-linearity is that the spatial lag expands the
+information set to include information from neighbouring regions: in
+other words, the value of an explanatory variable in a spatial unit can
+have influence in other spatial units via the spatial lag. This makes
+interpretation of the coefficients less straightforward but also richer
+(Golgher and Voss, 2016). The results of LeSage (2009) for
+cross-sectional spatial lag models can be extended to the spatial SUR
+framework. Note that, according to Elhorst (2014), the partial
+derivatives have the following interpretation: if an explanatory
+variable (\(X_k\)) in a particular province changes, not only the
+incidence rate in that province changes, also incidence rates in other
+provinces change via the contagion effect. Therefore, a change in
+\(X_k\) in a particular province has a \emph{direct effect} on that
+province, but also an \emph{indirect effect} on neighbouring provinces.
+In this way, the \(i\)th diagonal element of the matrix of partial
+derivatives represents the direct effect on the \(i\)th province,
+whereas the non-diagonal elements of \(i\)th row of the matrix of
+partial derivatives represent the indirect effects on other provinces.
+In order to obtain a global indicator, the direct effect is calculated
+as the mean of the diagonal elements and captures the average change in
+incidence ratio caused by the change in \(X_k\). Likewise, a global
+indicator of the indirect effects is calculated as the mean of the
+non-diagonal elements. The total effect is the sum of direct and
+indirect effects. Finally, note that if \(\rho_k = 0\), the indirect
+effects are 0 and the direct effects are equal to \(\beta_kt\).
+
 \hypertarget{analysis}{%
@@
 
-\hypertarget{exploratory-data-analysis}{%
-\subsection{Exploratory Data Analysis}\label{exploratory-data-analysis}}
+\hypertarget{eda}{%
+\subsection{Exploratory Data Analysis}\label{eda}}
 
@@
 These results are available in the source R markdown document.}.
+This will be useful to detect whether there are behavioral adaptations
+at the population level over the course of the period examined. As an
+example of behavioral adaptations, the effect of density might weaken
+over time, in the measure that the effects of the lockdown are felt: at
+full compliance with the lockdown, with people practicing social
+avoidance, density might matter less than other factors.
 
@@
 perform well towards the end of the study period. The most balanced
-model in terms of equation-level goodness-of-fit appears to be Model 1
-(\emph{lag11}), and this impression is further supported with a slightly
-higher value of the \(\text{pooled}-R^2\). This analysis is in line with
-the incubation period reported by Lauer et al.~(2020), although not the
-use of a weighted average. For the remainder of the paper, we will adopt
-Model 2 (\emph{lag11}) as our best model. In the following section we
-discuss the results of the analysis in more depth.
+model in terms of equation-level goodness-of-fit appears to be Model 2
+(\emph{lag11}), and this impression is further supported by a slightly
+higher value of the \(\text{pooled}-R^2\). The analysis using a lagged
+moving average of the environmental variables is in line with the
+incubation period reported by Lauer et al.~(2020), although the results
+do not support the use of a weighted average. For the remainder of the
+paper, we will adopt Model 2 (\emph{lag11}) as our best model. In the
+following section we discuss the results of the analysis in more depth.
 
@@
 
-Table \ref{tab:summary-best-model} presents a summary.
+Table \ref{tab:summary-best-model} presents a summary of the results of
+Model 2 (\emph{lag11}). Recall that two coefficients were constrained
+and are estimated only for the first equation of the system, and thus
+are assumed to be constant over time. These are the coefficients
+corresponding to GDP per capita and percentage of older adults. Both
+coefficients are significant at \(p<=0.05\). The sign of the coefficient
+for GDP per capita is positive, which indicates that wealthier regions
+tend to have a higher incidence of COVID-19. This is in line with the
+idea that the epidemic started earlier in wealthier places due to their
+connections to a globalized world. The sign of the coefficient for
+percentage of older adults, on the other hand, is negative. As
+previously discussed, the level of social contact of older adults even
+under normal circumstances tends to be lower than for younger people. As
+a consequence, places with larger populations of older adults appear to
+have a natural level of social distancing in place. It is important to
+note that this does not detract from evidence that older adults are more
+vulnerable individually and in institutional settings, where their case
+mortality rates are perhaps the highest of all age groups. Instead, this
+result indicates that their presence in the community at large tends to
+depress transmission of the virus.
+
+Of the two other control variables, the coefficient of population
+density is significant at \(p<=0.05\) in 11 equations, and not
+significant in 19. The coefficient for transit is significant at
+\(p<=0.10\) in 22 equations, and of those, significant at \(p<=0.05\) in
+14 equations. The next four variables are environmental factors. The
+coefficient for humidity is significant at \(p<=0.10\) in 20 equations,
+and of those, significant at \(p<=0.05\) in 19 equations. Of the
+environmental variables, temperature is the only variable that has
+significant coefficients in every equation at \(p<=0.05\). Finally,
+sunshine has significant coefficients at \(p<=0.10\) in 18 equations,
+including 15 coefficients that are significant at \(p<=0.05\).
 
@@
 
-Figure \ref{fig:results-time-1} shows the temporal evolution of the
-spatial autocorrelation coefficient (\(\rho\)) and the intercept of the
-model.
+To better understand the results, we proceed to plot the coefficients in
+their temporal sequence. We begin with the evolution of of the spatial
+autocorrelation coefficient (\(\rho\)) and the intercept of the model,
+both of which are shown in Figure \ref{fig:results-time-1}.
 
@@
 coefficients for the two control variables that were not fixed over
-time, i.e., \(\log(Density)\) and \(Transit\).
+time, i.e., \(\log(Density)\) and \(Transit\). In Section
+\ref{variables} we had anticipated a positive sign for the coefficient
+of this variable. The negative sign in the results is therefore somewhat
+surprising. Upon reflection, however, it seems reasonable that the sign
+would be negative during the period under analysis.
 
@@
 \centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmenta-time-1.pdf}
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmental-time-1.pdf}
 \caption{\label{fig:beta-environmental-time}Temporal evolution of
@@
 
-\hypertarget{concluding-remarks}{%
-\section{Concluding Remarks}\label{concluding-remarks}}
+The effects of the different variables are summarized in Table
+\ref{tab:summary-impacts-best-model}.
+
+\begin{table}
+
+\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
+\centering
+\begin{tabular}[t]{lrrrl}
+\toprule
+Variable & Min & Mean & Max & Note\\
+\midrule
+\rowcolor{gray!6}  \addlinespace[0.3em]
+\multicolumn{5}{l}{\textbf{Direct Effects}}\\
+\hspace{1em}log(GDPpc) & 0.681 & 0.681 & 0.681 & Constrained\\
+\hspace{1em}log(Older) & -0.809 & -0.809 & -0.809 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.168 & Non-constrained\\
+\hspace{1em}Transit & 0.345 & 0.517 & 0.584 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.575 & -0.545 & -0.031 & Non-constrained\\
+\hspace{1em}log(Temperature) & -2.019 & -1.422 & -0.929 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.284 & 0.095 & 0.206 & Non-constrained\\
+\addlinespace[0.3em]
+\multicolumn{5}{l}{\textbf{Indirect Effects}}\\
+\hspace{1em}log(GDPpc) & 0.632 & 0.632 & 0.632 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.751 & -0.751 & -0.751 & Constrained\\
+\hspace{1em}log(Density) & -0.032 & -0.004 & 0.156 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.088 & 0.338 & Non-constrained\\
+\hspace{1em}log(Humidity) & -1.462 & -0.156 & -0.001 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.629 & -0.273 & -0.014 & Non-constrained\\
+\hspace{1em}log(Sunshine) & -0.263 & -0.006 & 0.028 & Non-constrained\\
+\rowcolor{gray!6}  \addlinespace[0.3em]
+\multicolumn{5}{l}{\textbf{Total Effects}}\\
+\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
+\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
+\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
+\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
+\bottomrule
+\multicolumn{5}{l}{\textit{Note: }}\\
+\multicolumn{5}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
+\multicolumn{5}{l}{Constrained: coefficient as constant across equations}\\
+\end{tabular}
+\end{table}
+
+The temporal variation of the effects is shown in Figure
+\ref{fig:plot-impacts}.
+
+\begin{figure}
+\centering
+\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plot-impacts-1.pdf}
+\caption{\label{fig:plot-impacts}Temporal evolution of effects by date;
+the impact is percentage change in incidence as a result of a one
+percent change in value of variable.}
+\end{figure}
+
+\hypertarget{conclusion}{%
+\section{Concluding Remarks}\label{conclusion}}
 
@@
 
+\leavevmode\hypertarget{ref-Elhorst2014spatial}{}%
+Elhorst, J.P., 2014. Spatial econometrics: From cross-sectional data to
+spatial panels. Springer.
+
 \leavevmode\hypertarget{ref-Fernandes2020economic}{}%
@@
 on the world economy. Available at SSRN 3557504.
+
+\leavevmode\hypertarget{ref-Golgher2016interpret}{}%
+Golgher, A.B., Voss, P.R., 2016. How to interpret the coefficients of
+spatial models: Spillovers, direct and indirect effects. Spatial
+Demography 4, 175--205.
 
@@
 of Regional Science 46, 269--288.
+
+\leavevmode\hypertarget{ref-LeSage2009introduction}{}%
+LeSage, J., Pace, R.K., 2009. Introduction to spatial econometrics.
+Chapman; Hall/CRC.
 

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.Rmd
Wrapped up the conclusions.

--- 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
@@
 ---
-
-<!-- load cache -->
-<!-- whoppeR::load_knitr_cache('/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain_cache/latex/') -->
-
-```{r load-packages, include=FALSE}
-# The data package needs to be installed from github repository:
-# devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
-
-# This version of spsur may be needed:
-# Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install
-library(covid19env)
-library(ggthemes)
-library(gridExtra)
-library(kableExtra)
-library(lubridate)
-#library(meteoland) # Ask Fernando if this package was used in the end
-library(plm)
-library(sf)
-library(spdep)
-library(spsur)
-library(tidyverse)
-library(units)
-```
-
-```{r load-data, include=FALSE, cache=TRUE}
-# Load data from package `covid19env`
-data("covid19_spain")
-data("provinces_spain")
-```
-
-```{r data-preparation, include=FALSE, cache=TRUE}
-# Convert GDP per capita to thousands of euros:
-provinces_spain <- provinces_spain %>%
-  mutate(GDPpc = GDPpc/1000)
-
-# Join provincial data to incidence data and convert to simple features:
-covid19_spain <- covid19_spain %>% 
-  left_join(provinces_spain %>% st_drop_geometry(),
-            by = c("province", "CCAA", "ID_INE"))
-```
-
-```{r sort-communities-by-latitude, include=FALSE, cache=TRUE}
-# To visualize the distribution of temperature by CCAA, we sort the communities by latitude, from north to south.
-We do this by community instead of province because the large number of provinces clutters the plots.
-
-# Dissolve internal provincial boundaries to obtain a simple features object with the autonomous communities: 
-ccaa.sf <-  provinces_spain %>% 
-  group_by(CCAA) %>% 
-  summarize(provinces = n())
-
-# Extract coordinates of autonomous communities
-ccaa.coords <- ccaa.sf %>%
-  st_centroid() %>%
-  st_coordinates() %>%
-  as.data.frame()
-
-# Join Y coordinate to ccaa.sf
-ccaa.sf <- ccaa.sf %>% 
-  mutate(long = ccaa.coords$Y)
-
-# Sort autonomous communities from north to south
-ccaa.levels <- ccaa.sf %>% 
-  arrange(desc(long)) %>% select(CCAA)
-
-ccaa.levels <- as.character(ccaa.levels$CCAA)
-
-# Relevel autonomous communities
-covid19_spain <- covid19_spain %>%
-  mutate(CCAA = factor(CCAA, levels = ccaa.levels, ordered = TRUE))
-```
-
-Introduction
-==========================
-
-From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus has threatened to overrun health systems the world over.
-In efforts to contain the spread, numerous governments in many nations and regions have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place.
-There are encouraging signs that these measures have arrested the spread of the virus where they have been implemented, and have thus helped to keep a bad situation from becoming even worse [e.g., -@Lancastle2020impact].
-However, this has come at a high cost, and the consequences for all spheres of the economy, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how].
-As a result, there is a sense of urgency to anticipate the progression of the pandemic, in order to plan for progressive lifting of restrictions to movement and social contact [e.g., @Kissler2020projecting].
-Needless to say, this has become a delicate, and politically charged, balancing act between public health and the economy [@Gong2020balance].
-
-A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations.
-Earlier, diverse studies have shown the effect of temperature and humidity on the incidence of influenza [e.g., @Makinen2009cold; @Jaakkola2014decline; @Kudo2019low].
-Jaakkola et al.
-[-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low.
-After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV.
-In this way, Casanova et al.
-[-@Casanova2010effects] used two surrogates, namely the gastroenteritis (TGEV) and mouse hepatitis viruses (MHV), to find that virus inactivation was more rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
-In a similar vein, but working directly with SARS-CoV, Chan et al.
-[-@Chan2011effects] found that viability of the virus was lost at temperatures higher than 38C and relative humidity superior to 95%.
-
-While existing research on similar pathogens suggests that SARS-CoV is more stable and potentially easier to transmit in conditions of low temperature and low humidity, it is far from certain that this will also be the case with the novel SARS-CoV2.
-If such is the case, there is the possibility of easing restrictions to social contact as the weather warms; however, weeks or possibly months of costly measures could become undone if not, and the restrictions are lifted prematurely.
-Not surprisingly, given the stakes involved, this issue has already triggered a lively debate.
-
-Some of what is thought about the possible seasonality of COVID-19 is based on analogies to the patterns of other known respiratory viruses.
-However, de Ángel Solá et al.
-[-@deangel2020weathering] note that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" (section 4).
-This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive.
-At a global scale, de Ángel Solá et al.
-[-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al.
-[@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate.
-Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al.
-[-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States.
-Yao et al.
-[-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19.
-This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al.
-2016; Dąbrowska-Leonik et al.
-2018; Dinlen et al.
-2016; Mathyssen et al.
-2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).
-To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread).
-In the United States, the National Academy of Sciences, Engineering, and Medicine was engaged by the Office of the Executive for guidance on this matter [see @National2020rapid].
-Their conclusion summarizes the situation well (see p.
-6): "Some limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV-2 pandemic unfolds could shed more light on the effects of climate on transmission."
-
-With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature, humidity, and sunshine.
-We adopt a population health approach, and report results from a spatial model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
-We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
-We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables.
-These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a spatial contagion process.
-
-The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
-The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
-Our control variables also provide some intriguing insights.
-Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
-The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
-
-Please note that this paper is prepared as a reproducible research document.
-The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following repository:
-
-https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
-
-Context, Data, and Methods
-============
-
-## Covid-19 in Spain
-
-The first reported case of COVID-19 in Spain was on January 31th, when a German tourist in the Canary Islands tested positive for the virus.
-However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
-In a short period of time, after this relatively slow start, the outbreak flared.
-By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic.
-This declaration marked a turning point for the public in Spain too.
-As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe, after Italy.
-In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th.
-As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g.
-food, health) and some economic subsectors of industry and construction.
-A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents.
-The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed.
-During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
-
-## Selection of Variables
-
-The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
-In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
-Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
-As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
-These variables have been implicated in the viability and ease of transmission of similar viruses.
-In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.
-
-The first variable that we consider is GDP per capita.
-Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
-A18)}.
-The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when its cause may still be unclear [@Zhou2016accelerated].
-In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] remarks how Toronto's status as a global city was a vulnerability.
-In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and their potential to be further ahead in the trajectory of the pandemic.
-Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional].
-Therefore, it is possible that wealthier regions remained relatively more active even during the lockdown.
-On the other hand, it is also possible the less wealthy regions have a higher proportion of workers in manual occupations that cannot telework.
-
-The percentaje of older adults (over 65) is the second variable that we consider as a control.
-Early evidence suggest that the case rate mortality is higher at older ages [e.g.
-@Novel2020epidemiological].
-It is not clear that presence of older adults necessarily translates into higher transmission in the population.
-The public health tool of choice in containing the spread of the disease has been social distancing.
-In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
-In other words, older adults are, by preference or necessity, already in a form of social isolation.
-Social distancing during the pandemic may actually reinforce that condition for older adults, as the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age] suggests.
-Accordingly, our expectation is that in the provinces with higher proportions of older adults will tend to have similarl levels of lower social contact, particularly since age-structured social contact matrix in Spain is similar to Italy [@Prem2017projecting].
-
-The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
-The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g.
-cumulative incidence in urban areas like NYC).
-For this reason, we anticipate a positive relationship between population density and the incidence of the disease.
-
-The last control variable is the presence of mass transit systems in a province.
-Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
-Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches [@Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
-After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.
-
-## Data
-
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
-For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**].
-This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments.
-In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food).
-The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute).
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.
-
-The spatial and temporal coverage of the data is as follows.
-Our dataset begins on March 13, which is the first date when every province had reported at least one case of COVID-19, and continues until April 11, for a period of 30 days.
-The unit of analysis is the province.
-Provinces are the equivalent of states, and are embedded in Autonomous Communities.
-As an example, Cataluña is an Autonomous Community and consists of four provinces, namely Barcelona, Gerona, Lerida, and Tarragona.
-The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}.
-The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the US) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the US).
-While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is inconsistent at smaller geographies, or cases are not reported at that level at all.
-The analysis must therefore be considered ecological.
-
-```{r descriptive-statistics, echo=FALSE, cache=TRUE}
-data.frame(Variable = c("COVID-19 Incidence",
-                        "Area", 
-                        "GDPpc", 
-                        "Older", 
-                        "Population Density",
-                        "Mean Temperature",
-                        "Humidity"),
-           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
-                    "Area of province in sq.km",
-                    "GDP per capita in €1,000s",
-                    "Percentage of people aged 65 and older in the province",
-                    "Population density in the province in people per sq.km",
-                    "Mean temperature in province by date, in C",
-                    "Relative humidity in province by date"),
-           Min = c(min(covid19_spain$Incidence),
-                   min(provinces_spain %>% 
-                         st_area() %>%
-                         set_units(km^2)),
-                   min(provinces_spain$GDPpc),
-                   min(provinces_spain$Older),
-                   min(provinces_spain$Density),
-                   min(covid19_spain$Mean_Temp),
-                   min(covid19_spain$Humidity)),
-           Mean = c(mean(covid19_spain$Incidence),
-                    mean(provinces_spain %>% 
-                         st_area() %>%
-                         set_units(km^2)),
-                    mean(provinces_spain$GDPpc),
-                    mean(provinces_spain$Older),
-                    mean(provinces_spain$Density),
-                    mean(covid19_spain$Mean_Temp),
-                    mean(covid19_spain$Humidity)),
-           Max = c(max(covid19_spain$Incidence),
-                   max(provinces_spain %>% 
-                         st_area() %>%
-                         set_units(km^2)),
-                   max(provinces_spain$GDPpc),
-                   max(provinces_spain$Older),
-                   max(provinces_spain$Density),
-                   max(covid19_spain$Mean_Temp),
-                   max(covid19_spain$Humidity)),
-           SD = c(sd(covid19_spain$Incidence),
-                  sd(provinces_spain$GDPpc),
-                  sd(provinces_spain %>% 
-                         st_area() %>%
-                         set_units(km^2)),
-                  sd(provinces_spain$Older),
-                  sd(provinces_spain$Density),
-                  sd(covid19_spain$Mean_Temp),
-                  sd(covid19_spain$Humidity)),
-           Source = c("Montera34",
-                      "INE",
-                      "INE",
-                      "INE",
-                      "INE",
-                      "AEMET",
-                      "MAPA")) %>%
-  kable("latex",
-        #"html",
-        booktabs = TRUE,
-        digits = 2,
-        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
-  column_spec(2, width = "15em") %>%
-  column_spec(7, width = "5em") %>%
-  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
-                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
-                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
-                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
-```
-
-An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease.
-Lauer et al.
-[-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days).
-The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days).
-For this reason, we judge it best to use lagged values of the environmental variables.
-We test different time lags as follows.
-We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*).
-Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*).
-Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*).
-The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al.
-[-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418.
-With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.
-
-## Methods: Spatial SUR
-
-The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consisits of cross-sections for different time periods.
-The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient], and has become a popular methodology included in several econometrics textbook [e.g., @Greene2003econometric].
-To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective.
-In his landmark text, Anselin discussed a model made of "an equation for each time period, which is estimated for a cross section of spatial units" (p.
-141).
-From this milestone, a large body of research has developed to extend the classical SUR into a spatial framework [e.g., @Rey1999us; @Lauridsen2010spatiotemporal; @Legallo2006evaluating; @Lopez2017spatial].
-
-The classical SUR model without spatial effects (from here, SUR-SIM) is a stack of equations as follows:
-
-\begin{equation}
-\label{eq:sur-sim}
-\begin{bmatrix}
-y_1 \\ y_2 \\ \vdots \\ y_T
-\end{bmatrix}
-=
-  \begin{bmatrix}
-X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
-\end{bmatrix}
-\
-\begin{bmatrix}
-\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
-\end{bmatrix}
-+
-\begin{bmatrix}
-\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
-\end{bmatrix}
-\end{equation}
-
-\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t$ is a $N \times k_t$ matrix of the $k_t$ independent variables, with associated vector of coefficients $\beta_t$,; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
-
-<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
-  <!--
-  ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
-
-\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in the province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) the day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the vector of residuals.
--->
-
-A key feature of the SUR model is the dependence structure among the vectors of residuals, namely:
-
-\begin{equation}
-\label{eq:sur-err}
-E[\epsilon_t \epsilon'_{t'}]=\sigma_{tt'}
-\end{equation}
-
-Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$.
-This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations.
-This is equivalent to specifying some effects to be invariant over time.
-
-<!--
-![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
-
-Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
-This flexibility can be reduced and it is posible to impose restrictions to consider some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
-In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations.
-This is equivalent to specifying some effects to be invariant over time.
--->
-
-Equation (\ref{eq:sur-sim}) can be rewriten in compact form:
-
-\begin{equation}
-\bf{Y} = \bf{X} \beta + \epsilon
-\label{eq:sur-sim-block}
-\end{equation}
-
-where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
-(with $K = \sum_t{k_t}$) and $\epsilon$ is a $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:
-
-\begin{equation}
-E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
-\end{equation}
-
-<!--
-  ![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
--->
-
-As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
-When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
-In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor.
-Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio].
-While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid].
-For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.
-
-The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:
-
-\begin{equation}
-\label{eq:sur-slm}
-\begin{aligned}
-\bf{AY} = \bf{X} \beta + \epsilon \\
-\epsilon =N(0,\bf{\Sigma})
-\end{aligned}
-\end{equation}
-
-\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ with $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.
-
-<!--
-  ![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
-
-\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
--->
-
-This specification assumes that incidence in a province ($y_{st}$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypothesis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$).
-The correspond Wald test is available in the `R` package `spsur`.
-
-<!--
-This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces.
-Coefficients for the spatially lagged variable are estimated for each time period ![$\lambda_t$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_t%24) and identifies the intensity and the sign of the impacts of neighbourhood.
-It is possible test the null hypotheis of identtical levels of spatial dependence (![$\lambda_i=\lambda_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_i%3D%5Clambda_j%2C%20%5Cforall%20i%2Cj%24)).
-The correspond Wald test is available in the `R` package `spsur`.
--->
-
-The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).
-Another alternative methodologies could be use.
-By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g.
-Elhorst 2014, Cap.
-4), but those models don't take account correlation between errors.
-Therefore, a spatial SUR approach is more reasonable for our purpose.
-
-Analysis and Results
-===================
-
-## Exploratory Data Analysis
-
-We begin with the exploratory analysis of the data.
-
-Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages.
-Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19.
-It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain.
-While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces.
-Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao.
-The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava.
-The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.
-
-```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people", warning=FALSE, cache = TRUE}
-# Retrieve the coordinates of Madrid
-madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()
-
-week11.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 11) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("March 13 - March 15") +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-week12.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 12) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds",
-                       direction = 1) +
-  ggtitle("March 16 - March 22") +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 8, y = madrid_xy[2] + 0.6) +
-  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
-
-week13.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 13) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("March 23 - March 29") +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid",
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-week14.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 14) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("March 30 - April 5") +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-week15.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
-  group_by(province, week = isoweek(Date)) %>% 
-  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
-  filter(week == 15) %>%
-  left_join(provinces_spain,
-            by = c("ID_INE")) %>%
-  st_as_sf() %>%
-  ggplot() +
-  geom_sf(aes(fill = mean_weekly_incidence)) +
-  scale_fill_distiller(name = "Mean Weekly Incidence", 
-                       palette = "Reds", 
-                       direction = 1) +
-  ggtitle("April 6 - April 11") +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom",
-        plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, nrow = 3)
-```
-
-To visualize the distribution of temperature and humidity we aggregate the provinces by Autonomous Community.
-In Figure \ref{fig:descriptives-temperature} the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost.
-There is a relatively wide range of values both within and between provinces over the 30 day period examined.
-The top panel of the figure shows the distribution of mean temperature.
-The lowest mean temperature for a community on any given day is 2.8C, and the highest is 22.4C for a range of approximately 20 degrees.
-Likewise, there is a great deal of variability in humidity, as seen in the bottom panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
-The actual values for the provinces display somewhat more variability even.
-
-```{r descriptives-temperature, echo = FALSE, fig.height=7, fig.cap="\\label{fig:descriptives-temperature} Distribution of mean temperatures and humidities in the Autonomous Communities in Spain between March 12, 2020 and April 11, 2020.
-The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost", cache = TRUE}
-# Summary of temperature by Autonomous Community
-temp.summary <- covid19_spain %>% 
-  group_by(CCAA, Date) %>%
-  summarize(Value = mean(Mean_Temp)) %>%
-  mutate(Variable = "Mean Temperature in Community")
-
-# Summary of humidity by Autonomous Community
-humidity.summary <- covid19_spain %>% 
-  group_by(CCAA, Date) %>%
-  summarize(Value = mean(Humidity)) %>%
-  mutate(Variable = "Mean Humidity in Community")
-
-# Bind tables
-env.summary <- rbind(temp.summary, humidity.summary) %>%
-  mutate(Variable = factor(Variable, 
-                           levels = c("Mean Temperature in Community",
-                                      "Mean Humidity in Community")))
-
-# Plot
-ggplot(data = env.summary, 
-       aes(x = CCAA,
-           y = Value)) +
-  geom_boxplot() +
-  theme_tufte() +
-  theme(axis.text.x = element_text(angle = 90)) +
-  xlab("Autonomous Community (sorted from north to south)") +
-  ylab("Value") +
-  facet_wrap(~ Variable, nrow = 2, 
-             scales = "free_y")
-```
-
-```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province", warning=FALSE, cache = TRUE}
-# GDP per capita:
-q <- quantile(provinces_spain$GDPpc)
-gdppc.plot <- provinces_spain %>%
-  mutate(GDPpc = GDPpc,
-         GDPpc = case_when(GDPpc < q[2] ~ "1st",
-                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
-                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
-                             GDPpc > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
-  ggplot() +
-  geom_sf(aes(fill = GDPpc)) +
-  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom") +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-# Older adults:
-q <- quantile(provinces_spain$Older)
-older.plot <- provinces_spain %>% 
-  mutate(Older = Older,
-         Older = case_when(Older < q[2] ~ "1st",
-                             Older > q[2] & Older <= q[3]  ~ "2nd",
-                             Older > q[3] & Older <= q[4] ~ "3rd",
-                             Older > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
-  ggplot() +
-  geom_sf(aes(fill = Older)) +
-  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom") +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-# Population density:
-q <- as.numeric(quantile(provinces_spain$Density))
-density.plot <- provinces_spain %>% 
-  mutate(Density = as.numeric(Density),
-         Density = case_when(Density < q[2] ~ "1st",
-                             Density > q[2] & Density <= q[3]  ~ "2nd",
-                             Density > q[3] & Density <= q[4] ~ "3rd",
-                             Density > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
-  ggplot() +
-  geom_sf(aes(fill = Density)) +
-  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
-  theme_bw(base_size = 9) +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom") +
-  geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 8, 
-               y = madrid_xy[2] + 0.6,
-               family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
-               xend = madrid_xy[1], 
-               yend = madrid_xy[2])
-
-grid.arrange(gdppc.plot, older.plot, density.plot, nrow = 3)
-```
-
-Figure \ref{fig:daily-correlations} shows the distribution of daily correlations of the independent variables with incidence of COVID-19, after log-transforming all variables.
-It can be seen there that the correlation of GDPpc and temperature (in its three definitions) have the strongest positive and negative correlations with incidence, respectively.
-Percentage of older adults displays somewhat weaker negative correlations with incidence, as does density.
-It can be seen that the humidity variable, in its three forms, tends to be possitively correlated with incidence of COVID-19.
-
-```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)", cache = TRUE}
-covid19_spain %>% 
-  group_by(Date) %>%
-  summarize(Older = cor(log(Older), log(Incidence)),
-            GDPpc = cor(log(GDPpc), log(Incidence)),
-            Density = cor(log(Density), log(Incidence)),
-            Humidity_lag8 = cor(Humidity_lag8, Incidence),
-            Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
-            Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
-            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
-            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
-            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
-  select(-Date) %>%
-  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
-  mutate(Variable = factor(Variable,
-                           levels = c("GDPpc", 
-                                      "Older", 
-                                      "Density", 
-                                      "Humidity_lag8",
-                                      "Humidity_lag11",
-                                      "Humidity_lag11w",
-                                      "Mean_Temp_lag8",
-                                      "Mean_Temp_lag11",
-                                      "Mean_Temp_lag11w"))) %>%
-  ggplot(aes(x = Variable, y = Correlation)) +
-  geom_boxplot() +
-  theme_tufte() +
-  theme(axis.text.x = element_text(angle = 90))
-```
-
-## SUR Models
-
-```{r data-preparation-for-modelling, include=FALSE, cache = TRUE}
-# Organize data for SUR modelling 
-GPanel <- plm::pdata.frame(covid19_spain %>%
-                             select(province, 
-                                    Date,
-                                    Incidence, 
-                                    Median_Age,
-                                    Male2Female,
-                                    Older,
-                                    GDPpc,
-                                    Density,
-                                    Transit,
-                                    Mean_Temp_lag8,
-                                    Humidity_lag8,
-                                    Sunshine_Hours_lag8,
-                                    Mean_Temp_lag11,
-                                    Humidity_lag11,
-                                    Sunshine_Hours_lag11,
-                                    Mean_Temp_lag11w,
-                                    Humidity_lag11w,
-                                    Sunshine_Hours_lag11w),
-                           c("province","Date"))
-
-```
-
-```{r spatial-weights, include=FALSE, cache = TRUE}
-# Create spatial weights matrix: 
-Wmat <- provinces_spain %>%
-  #drop_na() %>%
-  as("Spatial") %>%
-  poly2nb(queen = FALSE) %>%
-  nb2mat(zero.policy = TRUE)
-
-Wmat <- (Wmat > 0) * 1
-
-# Join the two provinces in Canarias
-Wmat[which(provinces_spain$province == "Palmas(Las)"), 
-     which(provinces_spain$province == "Santa Cruz de Tenerife")] <- 1
-Wmat[which(provinces_spain$province == "Santa Cruz de Tenerife"), 
-     which(provinces_spain$province == "Palmas(Las)")] <- 1
-
-# 'Paises Catalans'
-#n = 8
-Wmat[which(provinces_spain$province == "Barcelona"), 
-     which(provinces_spain$province == "Baleares")] <- 1
-Wmat[which(provinces_spain$province == "Baleares"), 
-     which(provinces_spain$province == "Barcelona")] <- 1
-Wmat[which(provinces_spain$province == "Baleares"), 
-     which(provinces_spain$province == "Castellon/Castello")] <- 1 
-Wmat[which(provinces_spain$province == "Castellon/Castello"), 
-     which(provinces_spain$province == "Baleares")] <- 1
-Wmat[which(provinces_spain$province == "Baleares"), 
-     which(provinces_spain$province == "Tarragona")] <- 1 
-Wmat[which(provinces_spain$province == "Tarragona"), 
-     which(provinces_spain$province == "Baleares")] <- 1
-miW <- Wmat/rowSums(Wmat)
-
-# Convert to listw
-listw <- mat2listw(Wmat,style = "W")
-```
-
-```{r formulas, include=FALSE, cache = TRUE}
-# Define formulas with three different lagged variables:
-formula_lag8 <- log(Incidence) ~ 
-  log(GDPpc) +
-  log(Older) +
-  log(Density) +
-  Transit +
-  log(Humidity_lag8) +
-  log(Mean_Temp_lag8) +
-  log(Sunshine_Hours_lag8 + 0.1)
-
-formula_lag11 <- log(Incidence) ~ 
-  log(GDPpc) +
-  log(Older) +
-  log(Density) +
-  Transit +
-  log(Humidity_lag11) +
-  log(Mean_Temp_lag11) +
-  log(Sunshine_Hours_lag11 + 0.1)
-
-formula_lag11w <- log(Incidence) ~ 
-  log(GDPpc) +
-  log(Older) +
-  log(Density) +
-  Transit +
-  log(Humidity_lag11w) +
-  log(Mean_Temp_lag11w) +
-  log(Sunshine_Hours_lag11w + 0.1)
-```
-
-```{r model-restrictions, include=FALSE, cache = TRUE}
-T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # Recall that T is the number of days, i.e., time periods, i.e., equations; add 1 to include the starting day
-k <- 8 # Number of independent variables, including the constant
-coef_rest <- 2 # Number of restrictions
-
-# nrow is number of equations (time periods) minus 1, times the number of restrictions
-# ncol is number of variables times number of equations
-R2 <- matrix(0, nrow = (T - 1) * coef_rest, ncol = k * T)
-
-for (i in 1:(T-1)){
-  R2[i, 2] <- 1
-  R2[i, (2 + i * k)] <- -1
-  R2[(i + T - 1), 3] <- 1
-  R2[(i + T - 1), (3 + i * k)] <- -1
-  # Use if more restrictions are needed
-  #R2[(i + T - 1) * 2, 4] <- 1
-  #R2[(i + T - 1) * 2, (4 + i * k)] <- -1
-}
-b2 <- matrix(0, ncol = (T - 1) * coef_rest)
-```
-
-```{r model1-8-day-moving-average, include=FALSE, cache = TRUE}
-# Model with a lagged 8-day moving average of climatic variables:
-sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
-                                 data=GPanel, 
-                                 time = GPanel$Date, 
-                                 type = "slm", 
-                                 fit_method = "3sls", 
-                                 listw=  listw,
-                                 R = R2,
-                                 b = b2)
-#summary(sur.slm_lag8)
-#print(paste("Pooled R^2 = ", sur.slm_lag8$R2[1]))
-```
-
-```{r model2-11-day-moving-average, include=FALSE, cache = TRUE}
-# Model with 11-day moving average of climatic variables:
-sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
-                                  data=GPanel, 
-                                  time = GPanel$Date, 
-                                  type = "slm", 
-                                  fit_method = "3sls", 
-                                  listw=  listw,
-                                  R = R2,
-                                  b = b2)
-#summary(sur.slm_lag11)
-#print(paste("Pooled R^2 = ", sur.slm_lag11$R2[1]))
-```
-
-```{r model3-11-day-weighted-moving-average, include=FALSE, cache = TRUE}
-# Model with 11-day weighted moving average of climatic variables:
-sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
-                                   data=GPanel, 
-                                   time = GPanel$Date, 
-                                   type = "slm", 
-                                   fit_method = "3sls", 
-                                   listw=  listw,
-                                   R = R2,
-                                   b = b2)
-#summary(sur.slm_lag11w)
-#print(paste("Pooled R^2 = ", sur.slm_lag11w$R2[1]))
-```
-
-The goodness of fit of the three systems of equations is shown in Figure \ref{fig:goodness-of-fit}.
-
-```{r goodness-of-fit, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
-data.frame(Model1 = sur.slm_lag8$R2, 
-           Model2 = sur.slm_lag11$R2,
-           Model3 = sur.slm_lag11w$R2) %>%
-  slice(2:n()) %>%
-  rownames_to_column(var = "Equation") %>%
-  mutate(Date = seq(ymd("2020-03-13"), 
-                    ymd("2020-04-11"), 
-                    by = "days")) %>%
-  pivot_longer(cols = starts_with("M"), names_to = "Model", values_to = "R2") %>%
-  mutate(Model = factor(Model, 
-                        levels = c("Model1", "Model2", "Model3"),
-                        labels = c("Model 1: lag8", "Model 2: lag11", "Model 3: lag11w"))) %>%
-  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("Model 1: lag8" = "blue", "Model 2: lag11" = "orange", "Model 3: lag11w" = "green") ) + 
-  annotate(geom = "text", 
-           label = c(paste("pooled-R2 (Model 1: lag8)=", 
-                           round(sur.slm_lag8$R2[1], 4)),
-                     paste("pooled-R2 (Model 2: lag11)=", 
-                           round(sur.slm_lag11$R2[1], 4)),
-                     paste("pooled-R2 (Model 3: lag11w)=", 
-                           round(sur.slm_lag11w$R2[1], 4))),
-           x = as_date("2020-04-01"),
-           y = c(0.3, 0.27, 0.24),
-           family = "serif",
-           size = 3.5) +
-  theme_tufte()
-
-```
-
-Summary of best model.
-```{r summary-best-model, echo=FALSE, cache = TRUE}
-data.frame(Coefficient = sur.slm_lag11$coefficients, 
-           pval = dnorm(abs(sur.slm_lag11$coefficients/sur.slm_lag11$rest.se))) %>%
-  rownames_to_column(var = "Variable") %>%
-  mutate(Variable = c("Intercept", 
-                      "log(GDPpc)", 
-                      "log(Older)",
-                      "log(Density)",
-                      "Transit",
-                      "log(Humidity)",
-                      "log(Temperature)",
-                      "log(Sunshine)",
-                      rep(c("Intercept",
-                            "log(Density)", 
-                            "Transit", 
-                            "log(Humidity)", 
-                            "log(Temperature)",
-                            "log(Sunshine)"),
-                          29)),
-         Variable = factor(Variable, 
-                           levels = c("Intercept", 
-                                      "log(GDPpc)", 
-                                      "log(Older)", 
-                                      "log(Density)",
-                                      "Transit", 
-                                      "log(Humidity)",
-                                      "log(Temperature)",
-                                      "log(Sunshine)"))) %>%
-  group_by(Variable) %>%
-  summarize(Min = min(Coefficient), 
-            Mean = mean(Coefficient), 
-            Max = max(Coefficient),
-            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
-            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
-            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
-  mutate(Note = c("Non-constrained", 
-                  "Constrained", 
-                  "Constrained", 
-                  "Non-constrained", 
-                  "Non-constrained",
-                  "Non-constrained",
-                  "Non-constrained",
-                  "Non-constrained")) %>%
-  kable("latex",
-        #"html",
-        booktabs = TRUE,
-        digits = 3,
-        col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
-        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
-  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
-  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
-                       "Non-constrained: coefficient was allowed to vary across equations",
-                       "Constrained: coefficient as constant across equations"))
-```
-
-Discussion
-===================
-
-Figure \ref{fig:delta-time} shows the temporal evolution of the spatial autocorrelation coefficient ($\lambda$).
-
-```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Deltas = sur.slm_lag11$deltas,
-           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Deltas, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the intercept.
-
-```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
-#Intercept
-n = 1
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-density-time} shows the temporal evolution of the coefficient for $\log(Density)$.
-
-```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)", cache = TRUE}
-#Density:
-n = 2 # Although this is the fourth coefficient, after intercept, log(GDPpc), and log(Older), we start at 2 here because the two constrained coefficients are removed from the matrix of coefficients 
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-transit-time} shows the temporal evolution of the coefficient for $Transit$.
-
-```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit", cache = TRUE}
-n = 3
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the coefficient for $\log(Humidity)$.
-
-```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)", cache = TRUE}
-n = 4
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-temperature-time} shows the temporal evolution of the coefficient for $\log(Temperature)$.
-
-```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)", cache = TRUE}
-n = 5
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the coefficient for $\log(Sunshine + 0.1)$.
-
-```{r beta-sunshine-time, echo=FALSE, fig.cap="\\label{fig:beta-sunshine-time}Temporal evolution of coefficient for log(Sunshine + 0.1)", cache = TRUE}
-n = 6
-data.frame(Date = seq(ymd("2020-03-13"), 
-                      ymd("2020-04-11"), 
-                      by = "days"),
-           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
-           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
-  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
-                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
-                         TRUE ~ "p > 0.10")) %>%
-  ggplot(aes(x = Date, y = Beta, color = sig)) +
-  geom_point(size = 2) +
-  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
-  theme_tufte()
-```
-
-```{r impacts, echo=FALSE}
-Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
-trsur <- spatialreg::trW(Wsur, type = "MC")
-impacts.sur.slm_lag11 <- impactspsur(sur.slm_lag11, tr = trsur,R = 1000)
-
-impacts.sur.slm_lag11[[1]]
-impacts.sur.slm_lag11[[2]]
-impacts.sur.slm_lag11[[3]]
-
-myimpact <- matrix(0,ncol = 5, nrow = 29)
-for (i in 2:30){
-  myimpact[i-1,] <- impacts.sur.slm_lag11[[i]]$res$direct
-}
-matplot(myimpact, type='l')
-
-```
-
-Concluding Remarks
-===================
-
-**Direct, Indirect and Total effect**
-
-For classical regression model, the parameter interpretation is easy since parameters can be interpreted as the partial derivatives of the dependent variable relative to the explanatory variable.
-For spatial regression models, parameter interpretation becomes richer but more complicated  (@golgher2016interpret).
-Changes in the value of an explanatory variable in a province can have influence in other provinces.
-In essence, the idea is that spatial dependence expands the information set to include information from neighbouring regions.
-The results of @LeSage2009 can be extended to the spatial SUR framework.
-Note that,
-
-According to @elhors2014, these partial derivatives have the following interpretation: if an explanatory variable ($X_k$) in a particular province changes, not only the incidence rate in that province change, also incidence rates in other provinces change.
-Therefore, a change in $X_k$ in a particular province has a *direct effect* on that province, but also an *indirect effect* on neighbouring provinces.
-Note that the ith diagonal element of the matrix of partial derivatives represents the direct effect on the ith province and non-diagonal elements of ith row of the matrix of partial derivatives represent the indirect effects on other provinces.
-In order to obtain a global indicator, the direct effect is calculated as the mean of the diagonal elements and captures the average change in incidence ratio caused by the change in $X_k$.
-Likewise, a global indicator of the indirect effects is calculated as the mean of the non-diagonal elements.
-The total effect is the sum of direct and indirect effects.
-Finally, note that if $\lambda_k = 0$, the indirect effects are 0 and the direct effects are equal to $\beta_kt$.
-
-More words go here.
-
-Limitaciones
-
-* La temperatura puedes ser un factor pero cabría esperar que no tuviera un impacto lineal.
-Por el contrario deberiamos esperar un 'punto de corte': Una temperatua mantenida superior a X grados durante siete días sea la mejor forma de incorporarla al modelo.
-En nuestro modelo (log-log) en incremento en un 1% de la temperatura se asocia con un incremento beta% en la incidencia.
-Esto es lo mismo si la temperatura es baja que si es alta.
-
-* Idem para la humedad y horas de sol
-
-* Los datos son 'provisionales'.
-Hay gran confusión sobre la incidencia real.
-La ausencia de test de diagnóstico PCRs al inicio de la pandemia (también ahora) puede desvirtuar el número de casos dianosticados.
-
-* Los datos oficiales (que tampoco son fiables) son reportados a nivel de Comunidades Autónomas.
-La recopilación de datos a nivel de provincial son el resultado de un esfuerzo colaborativo de recopilacion entre distintas fuentes (principalmente gobiernos locales).
-Nuevamente puede haber sesgos importanes.
-
-* En carácter insular de las Islas Canarias y de las Islas baleares no se ha tenido en cuenta.
-
-* Baleares se ha linkado artificialmente con 3/4 provincias y debería de haberse dejado aislada para respetar su carácter insular al definir la matriz W.
-
-* La incidencia depende de un estado inicial.
-Al inicio del estudio había provincias en las que la epidemia estaba muy desarrollada (Madrid/Alava) mientras que en otras apenas habías casos.
-Este hecho no ha sido considerado en el modelo.
-Decretar el confinamiento debe tener distintos impactos entre provincias.
-QUIZAR METER UNA VARAIBLE DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
-
-* No se ha controlado por el sistema sanitario de cada provincia.
-Uno de los principales focos de contagio han sido los hospitales y los centros de salud.
-En aquellas provincias donde se ha promovido el mensaje "NO IR AL MEDICO" han presentado menor incidencia.
-
-* HAY QUE INCLUIR INTERVALOS DE CONFIANZA EN LOS GRÁFICOS QUE HACEMOS DE LOS COEFICIENTES DEL MODELO
-
-* ¿hasta que punto las variables de control no recojen también factores climáticos? por ejemplo, la gente jóven vive en el sur de España que ha tenido menos incidencia.
-
-* TITULO: A spatial analysis of the environmental correlates ofCOVID-19 incidence in Spain ¿during the lookdown?
-
-* Las estaciones meteorológicas para la obtencion de datos climáticos has sido elegidas aleatoriamente (una para cada provincia).
-otra seleccion puede dar otros resultados.
-AQUI SE PODRÍA HACER EL ESFUERZO DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR PROVINCIA
-
-*IDEM para la humedad. idem para sunshine
-
-* HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA DARLE RELEVANCIA A LA ESTIMACION SUR
-
-Acknowledgments {#acknowledgments .unnumbered}
-==========
-
-Add acknowledgments as appropriate in final draft.
-
-The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
-
-References {#references .unnumbered}
-==========

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Wrapped up the conclusions.

@@
 author:
-  - name: Antonio Paez
-    email: paezha@mcmaster.ca
-    affiliation: McMaster University
+  - name: Author A
+    email: author.a@example.com
+    affiliation: Some School
     footnote: 1
-  - name: Fernando A. Lopez
-    email: fernando.lopez@upct.es
-    affiliation: Universidad Politecnica de Cartagena
-  - name: Tatiane Menezes
-    email: tatiane.menezes@ufpe.br
-    affiliation: Departamento de Economia
-  - name: Renata Cavalcanti
-    email: renata.vcsantos@gmail.com
-    affiliation: Nucleo de Pesquisa
-  - name: Maira Galdino da Rocha Pitta
-    email: mgrpitta@ufpe.br
-    affiliation: Nucleo de Pesquisa
+  - name: Author B
+    email: author.b@example.com
+    affiliation: Some Institute
 address:
-  - code: McMaster University
-    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
-  - code: Universidad Politecnica de Cartagena
-    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
+  - code: Some School
+    address: School of the Things
+  - code: Some Institute
+    address: Institute of Everything
   - code: Departamento de Economia
-    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
-  - code: Nucleo de Pesquisa
-    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
 footnote:
@@
 
-<!-- load cache -->
-<!-- whoppeR::load_knitr_cache('/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain_cache/latex/') -->
-
 ```{r load-packages, include=FALSE}
-# The data package needs to be installed from github repository:
-# devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
-
-# This version of spsur may be needed:
-# Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install
+# Download and install these two packages:
+# covid19env
+# spsur
+
 library(covid19env)
@@
 library(lubridate)
-#library(meteoland) # Ask Fernando if this package was used in the end
 library(plm)
@@
 Please note that this paper is prepared as a reproducible research document.
-The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following repository:
-
-https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
+The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following anonymous Drive folder:
+
+https://drive.google.com/open?id=1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_
 
@@
 These are the coefficients corresponding to GDP per capita and percentage of older adults.
-Both coefficients are significant at $p<=0.05$.
+Both coefficients are significant at $p\leq0.05$.
 The sign of the coefficient for GDP per capita is positive, which indicates that wealthier regions tend to have a higher incidence of COVID-19.
@@
 
-Of the two other control variables, the coefficient of population density is significant at $p<=0.05$ in 11 equations, and not significant in 19.
-The coefficient for transit is significant at $p<=0.10$ in 22 equations, and of those, significant at $p<=0.05$ in 14 equations.
+Of the two other control variables, the coefficient of population density is significant at $p\leq0.05$ in 11 equations, and not significant in 19.
+The coefficient for transit is significant at $p\leq0.10$ in 22 equations, and of those, significant at $p\leq0.05$ in 14 equations.
 The next four variables are environmental factors.
-The coefficient for humidity is significant at $p<=0.10$ in 20 equations, and of those, significant at $p<=0.05$ in 19 equations.
-Of the environmental variables, temperature is the only variable that has significant coefficients in every equation at $p<=0.05$.
-Finally, sunshine has significant coefficients at $p<=0.10$ in 18 equations, including 15 coefficients that are significant at $p<=0.05$.
+The coefficient for humidity is significant at $p\leq0.10$ in 20 equations, and of those, significant at $p\leq0.05$ in 19 equations.
+Of the environmental variables, temperature is the only variable that has significant coefficients in every equation at $p\leq0.05$.
+Finally, sunshine has significant coefficients at $p\leq0.10$ in 18 equations, including 15 coefficients that are significant at $p\leq0.05$.
 
@@
               ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+    geom_rect(xmin = as_date("2020-04-12"), xmax = as_date("2020-04-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
   geom_point(size = 1) +
@@
 To better understand the results, we proceed to plot the coefficients in their temporal sequence.
-We begin with the evolution of the spatial autocorrelation coefficient ($\rho$) and the intercept of the model, both of which are shown in Figure \ref{fig:results-time-1}.
+At this point it is worth recalling that the state of emergency went into effect on March 14.
+In the following figures, the periods of time indicated in yellow starting on March 14 correspond to the state of emergency, with only essential travel and selected industrial activities allowed; the period of time in orange was the stricter lockdown when only essential travel was allowed.
+
+We begin our discussion with the evolution of the spatial autocorrelation coefficient ($\rho$) in Figure \ref{fig:results-time-1} (left panel).
+It is interesting to note that the magnitude of the spatial autocorrelation coefficient $\rho$ declines over the period under analysis, and is not significant for some days.
+This is an interesting result: immediately prior to the declaration of the state of emergency, there appears to have been a strong inter-provincial contagion effect.
+This situation persisted for a few days after the lockdown, and became less strong over time.
+At the end of the period under examination, the strength of this effect was much diminished.
+We would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero.
+The fact that it is not is evidence of some weak contagion effect due to restrictions to the movement between provinces.
+
+The intercept (right panel in Figure \ref{fig:results-time-1}) is indicative of the variation of the incidence of COVID-19, other things being equal.
+Here we see that at the incidence declines somewhat immediately after the state of emergency, only to begin increasing again over time.
+Then, the incidence declines again after the stricter lockdown and rebounds to a lower level by April 11.
 
@@
             ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
+  geom_rect(xmin = as_date("2020-04-12"), xmax = as_date("2020-04-30"),
+              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
   geom_point(size = 1) +
@@
 
-Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ and $Transit$.
-In Section \ref{variables} we had anticipated a positive sign for the coefficient of this variable.
-The negative sign in the results is therefore somewhat surprising.
-Upon reflection, however, it seems reasonable that the sign would be negative during the period under analysis.
+Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ (left panel) and $Transit$ (right panel).
+
+In Section \ref{variables} we had anticipated a positive sign for the coefficient of this variable, and indeed, at the beginning of the period the coefficient is positive, albeit not significant, and then remains non-significant for the earlier part of the lockdown.
+We are somewhat surprised by the way this coefficient turns significant _and_ negative in the later part of the lockdown, after April 1st.
+This effect, we surmise, is the result of risk compensation, a situation where people adapt their behavior according to the *perceived* level of risk, becoming more careful when the perceived risk is higher and viceversa [e.g., @Noland1995perceived; @Richens2000risk; @Phillips2011risk].
+Consequently, residents in high density regions may perceive the risk of infection as being higher, and adapt their behavior accordingly - while the opposite may be true of residents in low density regions.
+Anecdotally, this seems to have been the case with some governors in the US, who resisted stay-at-home orders citing their sparse \footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions} and rural populations\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}.
+The coefficient for Transit is positive, as expected, but with very wide confidence intervals, and in fact not significant in the earlier part of the period.
 
@@
 
-Figure \ref{fig:beta-environmental-time} shows the temporal evolution of the coefficient for the three environmental variables.
+The evolution of the coefficients for the three environmental variables are shown in Figure \ref{fig:beta-environmental-time}.
+Despite a mostly positive simple correlation with incidence (see Figure \ref{fig:daily-correlations}) once that we control for other factors, humidity has a negative association with incidence of COVID-19 in Spain (top-right panel).
+This is in line with the literature that describes the viability and transmission of different viruses at higher levels of humidity.
+The coefficients for temperature (top-right panel) are consistently negative and this variable is, besides the intercept, the only one with significant coefficients in all equations.
+The range of variation of this coefficient during the period examined is approximately between -1 and -2 - although it is important to recall that these values should not be interpreted directly as effects; more on this below.
+Finally the plot for the coefficients associated with hours of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the coefficient was negative, but significant only at $p\leq10$.
+At the beginning of the lockdown, the coefficient is no longer significative, before becoming *positive* and significant on March 23rd.
+This result stands in contrast to previous findings regarding influenza, where more hours of sunlight reduced the strength and duration of epidemic durations [@Yu2013characterization].
+A difference with previous studies is the temporal scale of the analysis: where Yu et al.
+[-@Yu2013characterization] use monthly averages, we use daily data for a much shorter period of time.
+The positive sign of sunshine may well be another instance of behavioral adaptations, whereby compliance with lockdown orders weakens on sunny days.
 
@@
 
-The effects of the different variables are summarized in Table \ref{tab:summary-impacts-best-model}.
+The preceding discussion helps to establish the inferential contributions of the analysis.
+The remaining question is, what are the implications.
+As discussed in Section \ref{methods} the question of the magnitude of the effects is not clear from the coefficients alone, since a change to a variable in a province influences, via the contagion effect, its neighbors.
+For this reason, the appropriate way to estimate the effects is to calculate both the own and contagion effects (in other words the direct and indirect effects, respectively), as well as their total.
+A summary of the different variable effects appears in Table \ref{tab:summary-impacts-best-model}.
+All effects in the table are interpreted as percentage change in the incidence of COVID-19 as a consequence of a one percent change in the variable.
+The exception is Transit (which was not log-transformed), and this variable represents the percentage change in incidence between provinces without and with mass transit systems.
+
+Two variables had temporally constrained coefficients.
+The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.675% for each percentual increase of this variable (in €1,000s).
+In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemia.
+Percentage of older adults has a negative effect, and each percentual increase in this variable is associated with a relatively small reduction of the incidence of less than 1%.
+
+The temporal variation of the effects for the rest of the variables is shown in Figure \ref{fig:plot-impacts}.
+The largest positive direct effect is Transit, and the largest direct negative effects are temperature and humidity.
+The direct effect of these variables is as follows: for each percentual point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
+This effect is amplified via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 1.5%.
+The effect of humidity is also to reduce the incidence: each percentual point of increase in humidity is associated with a reduction of up to 1.5% in incidence.
+With the addition of the indirect effect, the total effect of a 1% increase in humidity is to reduce incidence by up to 3%.
+As seen in the figure, the indirect (i.e., contagion) effects are almost as strong as the direct effects immediately before and at the beginning of the lockdown period.
+However, by the end of the lockdown period, the indirect effects have weakened considerably.
+The effect of density goes from positive (+0.334) to negative (-0.244).
+It is crucial to keep in mind here that this is the effect of density over a period of *lockdown* and possibly subject to risk compensation, as discussed above.
+In other words, the lockdown seems to be more effective in places with higher density.
+Accordingly, cities should not rush to ease lockdowns, but rather efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
+
+What do these effects mean?
+Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the effect of temperature is estimated to reduce the incidence of COVID-19 by between one and two percent with each percentual increase of this variable.
+Without a lockdown, or a relaxed lockdown, the effect is even greater, but only because the disease is transmitted inter-regionally!
+Imagine a region that is closed to its neighbors: increases in temperature are associated with smaller reductions in incidence - however, it does not import or export cases.
+A province that is not closed and not under lockdown will see a greater effect of temperature, but at the cost of importing/exporting cases.
 
@@
 
-The temporal variation of the effects is shown in Figure \ref{fig:plot-impacts}.
-
-```{r plot-impacts, echo=FALSE, fig.cap="\\label{fig:plot-impacts}Temporal evolution of effects by date; the impact is percentage change in incidence as a result of a one percent change in value of variable."}
+```{r plot-impacts, echo=FALSE, fig.cap="\\label{fig:plot-impacts}Temporal evolution of direct, indirect, and total effects by date."}
 best_model.impacts %>%
@@
 
-More words go here.
-
-Limitaciones
-
-* La temperatura puedes ser un factor pero cabría esperar que no tuviera un impacto lineal.
-Por el contrario deberiamos esperar un 'punto de corte': Una temperatua mantenida superior a X grados durante siete días sea la mejor forma de incorporarla al modelo.
-En nuestro modelo (log-log) en incremento en un 1% de la temperatura se asocia con un incremento beta% en la incidencia.
-Esto es lo mismo si la temperatura es baja que si es alta.
-
-* Idem para la humedad y horas de sol
-
-* Los datos son 'provisionales'.
-Hay gran confusión sobre la incidencia real.
-La ausencia de test de diagnóstico PCRs al inicio de la pandemia (también ahora) puede desvirtuar el número de casos dianosticados.
-
-* Los datos oficiales (que tampoco son fiables) son reportados a nivel de Comunidades Autónomas.
-La recopilación de datos a nivel de provincial son el resultado de un esfuerzo colaborativo de recopilacion entre distintas fuentes (principalmente gobiernos locales).
-Nuevamente puede haber sesgos importanes.
-
-* En carácter insular de las Islas Canarias y de las Islas baleares no se ha tenido en cuenta.
-
-* Baleares se ha linkado artificialmente con 3/4 provincias y debería de haberse dejado aislada para respetar su carácter insular al definir la matriz W.
-
-* La incidencia depende de un estado inicial.
-Al inicio del estudio había provincias en las que la epidemia estaba muy desarrollada (Madrid/Alava) mientras que en otras apenas habías casos.
-Este hecho no ha sido considerado en el modelo.
-Decretar el confinamiento debe tener distintos impactos entre provincias.
-QUIZAR METER UNA VARAIBLE DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
-
-* No se ha controlado por el sistema sanitario de cada provincia.
-Uno de los principales focos de contagio han sido los hospitales y los centros de salud.
-En aquellas provincias donde se ha promovido el mensaje "NO IR AL MEDICO" han presentado menor incidencia.
-
-* ¿hasta que punto las variables de control no recojen también factores climáticos? por ejemplo, la gente jóven vive en el sur de España que ha tenido menos incidencia.
-
-* Las estaciones meteorológicas para la obtencion de datos climáticos has sido elegidas aleatoriamente (una para cada provincia).
-otra seleccion puede dar otros resultados.
-AQUI SE PODRÍA HACER EL ESFUERZO DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR PROVINCIA
-
-*IDEM para la humedad. idem para sunshine
-
-* HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA DARLE RELEVANCIA A LA ESTIMACION SUR
+In this paper we have presented a spatio-temporal analysis of incidence of COVID-19 in Spain.
+The analysis covers a 30-day period that begins immediately before the state of emergency was declared in the country.
+The focus of the research has been on the environmental correlates of incidence of the disease.
+There is a need for more empirical evidence, as policy makers, public health practitioners, and the public begin planning for the months ahead at this early stage of the pandemic.
+
+Our results offer strong support for the hypothesis that incidence of COVID-19 at the population level is lower at higher temperatures and levels of humidity: the estiamated effect is a reduction in the neighborhood of 3% percent in incidence per each 1% increase in temperature, and a 3% reduction in incidence per 1% increase in humidity *under conditions of contagion*.
+These reductions are estimated to be lower when lockdown conditions are in place.
+The question here seems to be whether these environmental variables can yield a bigger reduction of *more* cases, or a smaller reduction of *fewer* cases.
+Our control variables also offer some interesting insights.
+In particular, there is evidence of behavioral adaptations in the form of risk compensation (density) and compliance with the lockdown (sunshine).
+These results offer a cautionary tale, with regards to the effectiveness of the lockdown in more dense areas, and also the implications for compliance with stay-at-home orders as the northern hemisphere moves towards summer and more hours of sunshine during the day.
+
+A key aspect of the analysis using spatial SUR models is that we were able to model incidence of COVID-19 as an interregional contagion process.
+Here, we find that the strength of the contagion effect was dramatically reduced by the lockdown.
+
+Needless to say, the analysis presented here is ecological, and at the level of population health.
+For this reason, the analysis does not make any claims with respect to the effect of ultraviolet light on the virus, but rather about transmission of the virus in the population.
+In this way, the analysis does not say that the virus moves less effectively where more people live in close proximity to each other, but rather that humans are more contagious when they feel safe in less dense regions.
+Similarly, more sunshine does not mean that the virus thrives, but rather that humans are more contagious to each other when their behavior adapts to this environmental condition.
+
+It is important, in this respect, to highlight some limitations of this study.
+
+First, the analysis was conducted under a situation of lockdown, and therefore, besides the results are the beginning of the period examined, one must exercise caution when trying to extrapolate the findings to other situations.
+Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
+However, we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results - even if we are aware that the number of true cases is likely higher.
+Thirdly, all environmental data are based on a single station in a province.
+Here, we made an effort to select the most representative station based on the distribution of the population.
+An alternative approach would be to calculate the environmental variables based on the average of multiple stations within each province: this would likely smooth to some extent these variables.
+Finally, we defined spatial neighborhoods based on adjacency.
+It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services.
+We flag this as a matter for future research.
 

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Wrapped up the conclusions.

@@
 incidence in Spain}
-    \author[McMaster University]{Antonio Paez\corref{1}}
-   \ead{paezha@mcmaster.ca} 
-    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
-   \ead{fernando.lopez@upct.es} 
-    \author[Departamento de Economia]{Tatiane Menezes}
-   \ead{tatiane.menezes@ufpe.br} 
-    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
-   \ead{renata.vcsantos@gmail.com} 
-    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
-   \ead{mgrpitta@ufpe.br} 
-      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
-St W, Hamilton, ON, L8S 4K1, Canada}
-    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
-Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
-30201, Cartagena, Murcia, Spain}
-    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
-Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
-    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
-Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
-50670-901, Brazil}
+    \author[Some School]{Author A\corref{1}}
+   \ead{author.a@example.com} 
+    \author[Some Institute]{Author B}
+   \ead{author.b@example.com} 
+      \address[Some School]{School of the Things}
+    \address[Some Institute]{Institute of Everything}
+    \address[Departamento de Economia]{}
       \cortext[1]{Corresponding Author}
@@
 needed to reproduce/review/extend the analysis can be obtained from the
-following repository:
-
-https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
+following anonymous Drive folder:
+
+https://drive.google.com/open?id=1d\_40N\_QXo2Fl14r3T3CHs84IED-B1rV\_
 
@@
 corresponding to GDP per capita and percentage of older adults. Both
-coefficients are significant at \(p<=0.05\). The sign of the coefficient
-for GDP per capita is positive, which indicates that wealthier regions
-tend to have a higher incidence of COVID-19. This is in line with the
-idea that the epidemic started earlier in wealthier places due to their
-connections to a globalized world. The sign of the coefficient for
-percentage of older adults, on the other hand, is negative. As
-previously discussed, the level of social contact of older adults even
-under normal circumstances tends to be lower than for younger people. As
-a consequence, places with larger populations of older adults appear to
-have a natural level of social distancing in place. It is important to
-note that this does not detract from evidence that older adults are more
-vulnerable individually and in institutional settings, where their case
-mortality rates are perhaps the highest of all age groups. Instead, this
-result indicates that their presence in the community at large tends to
-depress transmission of the virus.
+coefficients are significant at \(p\leq0.05\). The sign of the
+coefficient for GDP per capita is positive, which indicates that
+wealthier regions tend to have a higher incidence of COVID-19. This is
+in line with the idea that the epidemic started earlier in wealthier
+places due to their connections to a globalized world. The sign of the
+coefficient for percentage of older adults, on the other hand, is
+negative. As previously discussed, the level of social contact of older
+adults even under normal circumstances tends to be lower than for
+younger people. As a consequence, places with larger populations of
+older adults appear to have a natural level of social distancing in
+place. It is important to note that this does not detract from evidence
+that older adults are more vulnerable individually and in institutional
+settings, where their case mortality rates are perhaps the highest of
+all age groups. Instead, this result indicates that their presence in
+the community at large tends to depress transmission of the virus.
 
 Of the two other control variables, the coefficient of population
-density is significant at \(p<=0.05\) in 11 equations, and not
+density is significant at \(p\leq0.05\) in 11 equations, and not
 significant in 19. The coefficient for transit is significant at
-\(p<=0.10\) in 22 equations, and of those, significant at \(p<=0.05\) in
-14 equations. The next four variables are environmental factors. The
-coefficient for humidity is significant at \(p<=0.10\) in 20 equations,
-and of those, significant at \(p<=0.05\) in 19 equations. Of the
-environmental variables, temperature is the only variable that has
-significant coefficients in every equation at \(p<=0.05\). Finally,
-sunshine has significant coefficients at \(p<=0.10\) in 18 equations,
-including 15 coefficients that are significant at \(p<=0.05\).
+\(p\leq0.10\) in 22 equations, and of those, significant at
+\(p\leq0.05\) in 14 equations. The next four variables are environmental
+factors. The coefficient for humidity is significant at \(p\leq0.10\) in
+20 equations, and of those, significant at \(p\leq0.05\) in 19
+equations. Of the environmental variables, temperature is the only
+variable that has significant coefficients in every equation at
+\(p\leq0.05\). Finally, sunshine has significant coefficients at
+\(p\leq0.10\) in 18 equations, including 15 coefficients that are
+significant at \(p\leq0.05\).
 
@@
 To better understand the results, we proceed to plot the coefficients in
-their temporal sequence. We begin with the evolution of of the spatial
-autocorrelation coefficient (\(\rho\)) and the intercept of the model,
-both of which are shown in Figure \ref{fig:results-time-1}.
+their temporal sequence. At this point it is worth recalling that the
+state of emergency went into effect on March 14. In the following
+figures, the periods of time indicated in yellow starting on March 14
+correspond to the state of emergency, with only essential travel and
+selected industrial activities allowed; the period of time in orange was
+the stricter lockdown when only essential travel was allowed.
+
+We begin our discussion with the evolution of the spatial
+autocorrelation coefficient (\(\rho\)) in Figure
+\ref{fig:results-time-1} (left panel). It is interesting to note that
+the magnitude of the spatial autocorrelation coefficient \(\rho\)
+declines over the period under analysis, and is not significant for some
+days. This is an interesting result: immediately prior to the
+declaration of the state of emergency, there appears to have been a
+strong inter-provincial contagion effect. This situation persisted for a
+few days after the lockdown, and became less strong over time. At the
+end of the period under examination, the strength of this effect was
+much diminished. We would expect that under full compliance with strict
+lockdown conditions (meaning no inter-provincial mobility) the spatial
+contagion effect would be zero. The fact that it is not is evidence of
+some weak contagion effect due to restrictions to the movement between
+provinces.
+
+The intercept (right panel in Figure \ref{fig:results-time-1}) is
+indicative of the variation of the incidence of COVID-19, other things
+being equal. Here we see that at the incidence declines somewhat
+immediately after the state of emergency, only to begin increasing again
+over time. Then, the incidence declines again after the stricter
+lockdown and rebounds to a lower level by April 11.
 
@@
 coefficients for the two control variables that were not fixed over
-time, i.e., \(\log(Density)\) and \(Transit\). In Section
-\ref{variables} we had anticipated a positive sign for the coefficient
-of this variable. The negative sign in the results is therefore somewhat
-surprising. Upon reflection, however, it seems reasonable that the sign
-would be negative during the period under analysis.
+time, i.e., \(\log(Density)\) (left panel) and \(Transit\) (right
+panel).
+
+In Section \ref{variables} we had anticipated a positive sign for the
+coefficient of this variable, and indeed, at the beginning of the period
+the coefficient is positive, albeit not significant, and then remains
+non-significant for the earlier part of the lockdown. We are somewhat
+surprised by the way this coefficient turns significant \emph{and}
+negative in the later part of the lockdown, after April 1st. This
+effect, we surmise, is the result of risk compensation, a situation
+where people adapt their behavior according to the \emph{perceived}
+level of risk, becoming more careful when the perceived risk is higher
+and viceversa (e.g., Noland, 1995; Phillips et al., 2011; Richens et
+al., 2000). Consequently, residents in high density regions may perceive
+the risk of infection as being higher, and adapt their behavior
+accordingly - while the opposite may be true of residents in low density
+regions. Anecdotally, this seems to have been the case with some
+governors in the US, who resisted stay-at-home orders citing their
+sparse
+\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}
+and rural
+populations\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}.
+The coefficient for Transit is positive, as expected, but with very wide
+confidence intervals, and in fact not significant in the earlier part of
+the period.
 
@@
 
-Figure \ref{fig:beta-environmental-time} shows the temporal evolution of
-the coefficient for the three environmental variables.
+The evolution of the coefficients for the three environmental variables
+are shown in Figure \ref{fig:beta-environmental-time}. Despite a mostly
+positive simple correlation with incidence (see Figure
+\ref{fig:daily-correlations}) once that we control for other factors,
+humidity has a negative association with incidence of COVID-19 in Spain
+(top-right panel). This is in line with the literature that describes
+the viability and transmission of different viruses at higher levels of
+humidity. The coefficients for temperature (top-right panel) are
+consistently negative and this variable is, besides the intercept, the
+only one with significant coefficients in all equations. The range of
+variation of this coefficient during the period examined is
+approximately between -1 and -2 - although it is important to recall
+that these values should not be interpreted directly as effects; more on
+this below. Finally the plot for the coefficients associated with hours
+of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the
+coefficient was negative, but significant only at \(p\leq10\). At the
+beginning of the lockdown, the coefficient is no longer significative,
+before becoming \emph{positive} and significant on March 23rd. This
+result stands in contrast to previous findings regarding influenza,
+where more hours of sunlight reduced the strength and duration of
+epidemic durations (Yu et al., 2013). A difference with previous studies
+is the temporal scale of the analysis: where Yu et al.~(2013) use
+monthly averages, we use daily data for a much shorter period of time.
+The positive sign of sunshine may well be another instance of behavioral
+adaptations, whereby compliance with lockdown orders weakens on sunny
+days.
 
@@
 
-The effects of the different variables are summarized in Table
-\ref{tab:summary-impacts-best-model}.
+The preceding discussion helps to establish the inferential
+contributions of the analysis. The remaining question is, what are the
+implications. As discussed in Section \ref{methods} the question of the
+magnitude of the effects is not clear from the coefficients alone, since
+a change to a variable in a province influences, via the contagion
+effect, its neighbors. For this reason, the appropriate way to estimate
+the effects is to calculate both the own and contagion effects (in other
+words the direct and indirect effects, respectively), as well as their
+total. A summary of the different variable effects appears in Table
+\ref{tab:summary-impacts-best-model}. All effects in the table are
+interpreted as percentage change in the incidence of COVID-19 as a
+consequence of a one percent change in the variable. The exception is
+Transit (which was not log-transformed), and this variable represents
+the percentage change in incidence between provinces without and with
+mass transit systems.
+
+Two variables had temporally constrained coefficients. The estimated
+effect of GDP per capita is to increase the incidence of COVID-19 by
+0.675\% for each percentual increase of this variable (in €1,000s). In
+our view, this is a measure of inertia, as provinces with higher GDP per
+capita where among the first to see exponential growth in the pandemia.
+Percentage of older adults has a negative effect, and each percentual
+increase in this variable is associated with a relatively small
+reduction of the incidence of less than 1\%.
+
+The temporal variation of the effects for the rest of the variables is
+shown in Figure \ref{fig:plot-impacts}. The largest positive direct
+effect is Transit, and the largest direct negative effects are
+temperature and humidity. The direct effect of these variables is as
+follows: for each percentual point increase in temperature, there is
+between a 1\% and 2\% reduction in the incidence of the disease. This
+effect is amplified via contagion, as seen in the central panel in the
+figure, and the indirect effect can further reduce the incidence by up
+to 1.5\%. The effect of humidity is also to reduce the incidence: each
+percentual point of increase in humidity is associated with a reduction
+of up to 1.5\% in incidence. With the addition of the indirect effect,
+the total effect of a 1\% increase in humidity is to reduce incidence by
+up to 3\%. As seen in the figure, the indirect (i.e., contagion) effects
+are almost as strong as the direct effects immediately before and at the
+beginning of the lockdown period. However, by the end of the lockdown
+period, the indirect effects have weakened considerably. The effect of
+density goes from positive (+0.334) to negative (-0.244). It is crucial
+to keep in mind here that this is the effect of density over a period of
+\emph{lockdown} and possibly subject to risk compensation, as discussed
+above. In other words, the lockdown seems to be more effective in places
+with higher density. Accordingly, cities should not rush to ease
+lockdowns, but rather efforts should be made to reduce or eliminate risk
+compensation in less densely populated regions.
+
+What do these effects mean? Under a situation of lockdown,
+inter-regional contagion is reduced, as expected, and the effect of
+temperature is estimated to reduce the incidence of COVID-19 by between
+one and two percent with each percentual increase of this variable.
+Without a lockdown, or a relaxed lockdown, the effect is even greater,
+but only because the disease is transmitted inter-regionally! Imagine a
+region that is closed to its neighbors: increases in temperature are
+associated with smaller reductions in incidence - however, it does not
+import or export cases. A province that is not closed and not under
+lockdown will see a greater effect of temperature, but at the cost of
+importing/exporting cases.
 
@@
 \multicolumn{5}{l}{\textbf{Direct Effects}}\\
-\hspace{1em}log(GDPpc) & 0.681 & 0.681 & 0.681 & Constrained\\
-\hspace{1em}log(Older) & -0.809 & -0.809 & -0.809 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.168 & Non-constrained\\
-\hspace{1em}Transit & 0.345 & 0.517 & 0.584 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.575 & -0.545 & -0.031 & Non-constrained\\
-\hspace{1em}log(Temperature) & -2.019 & -1.422 & -0.929 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.284 & 0.095 & 0.206 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.675 & 0.675 & 0.675 & Constrained\\
+\hspace{1em}log(Older) & -0.802 & -0.802 & -0.802 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.167 & Non-constrained\\
+\hspace{1em}Transit & 0.342 & 0.517 & 0.584 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.562 & -0.544 & -0.031 & Non-constrained\\
+\hspace{1em}log(Temperature) & -2.019 & -1.421 & -0.929 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.281 & 0.095 & 0.206 & Non-constrained\\
 \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Indirect Effects}}\\
-\hspace{1em}log(GDPpc) & 0.632 & 0.632 & 0.632 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.751 & -0.751 & -0.751 & Constrained\\
-\hspace{1em}log(Density) & -0.032 & -0.004 & 0.156 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.088 & 0.338 & Non-constrained\\
-\hspace{1em}log(Humidity) & -1.462 & -0.156 & -0.001 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.629 & -0.273 & -0.014 & Non-constrained\\
-\hspace{1em}log(Sunshine) & -0.263 & -0.006 & 0.028 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.638 & 0.638 & 0.638 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.758 & -0.758 & -0.758 & Constrained\\
+\hspace{1em}log(Density) & -0.032 & -0.004 & 0.157 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.341 & Non-constrained\\
+\hspace{1em}log(Humidity) & -1.475 & -0.157 & -0.001 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.644 & -0.274 & -0.014 & Non-constrained\\
+\hspace{1em}log(Sunshine) & -0.266 & -0.006 & 0.028 & Non-constrained\\
 \rowcolor{gray!6}  \addlinespace[0.3em]
@@
 
-The temporal variation of the effects is shown in Figure
-\ref{fig:plot-impacts}.
-
 \begin{figure}
@@
 \includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plot-impacts-1.pdf}
-\caption{\label{fig:plot-impacts}Temporal evolution of effects by date;
-the impact is percentage change in incidence as a result of a one
-percent change in value of variable.}
+\caption{\label{fig:plot-impacts}Temporal evolution of direct, indirect,
+and total effects by date.}
 \end{figure}
@@
 
-More words go here.
-
-Limitaciones
-
-\begin{itemize}
-\item
-  La temperatura puedes ser un factor pero cabría esperar que no tuviera
-  un impacto lineal. Por el contrario deberiamos esperar un `punto de
-  corte': Una temperatua mantenida superior a X grados durante siete
-  días sea la mejor forma de incorporarla al modelo. En nuestro modelo
-  (log-log) en incremento en un 1\% de la temperatura se asocia con un
-  incremento beta\% en la incidencia. Esto es lo mismo si la temperatura
-  es baja que si es alta.
-\item
-  Idem para la humedad y horas de sol
-\item
-  Los datos son `provisionales'. Hay gran confusión sobre la incidencia
-  real. La ausencia de test de diagnóstico PCRs al inicio de la pandemia
-  (también ahora) puede desvirtuar el número de casos dianosticados.
-\item
-  Los datos oficiales (que tampoco son fiables) son reportados a nivel
-  de Comunidades Autónomas. La recopilación de datos a nivel de
-  provincial son el resultado de un esfuerzo colaborativo de
-  recopilacion entre distintas fuentes (principalmente gobiernos
-  locales). Nuevamente puede haber sesgos importanes.
-\item
-  En carácter insular de las Islas Canarias y de las Islas baleares no
-  se ha tenido en cuenta.
-\item
-  Baleares se ha linkado artificialmente con 3/4 provincias y debería de
-  haberse dejado aislada para respetar su carácter insular al definir la
-  matriz W.
-\item
-  La incidencia depende de un estado inicial. Al inicio del estudio
-  había provincias en las que la epidemia estaba muy desarrollada
-  (Madrid/Alava) mientras que en otras apenas habías casos. Este hecho
-  no ha sido considerado en el modelo. Decretar el confinamiento debe
-  tener distintos impactos entre provincias. QUIZAR METER UNA VARAIBLE
-  DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON
-  MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
-\item
-  No se ha controlado por el sistema sanitario de cada provincia. Uno de
-  los principales focos de contagio han sido los hospitales y los
-  centros de salud. En aquellas provincias donde se ha promovido el
-  mensaje ``NO IR AL MEDICO'' han presentado menor incidencia.
-\item
-  ¿hasta que punto las variables de control no recojen también factores
-  climáticos? por ejemplo, la gente jóven vive en el sur de España que
-  ha tenido menos incidencia.
-\item
-  Las estaciones meteorológicas para la obtencion de datos climáticos
-  has sido elegidas aleatoriamente (una para cada provincia). otra
-  seleccion puede dar otros resultados. AQUI SE PODRÍA HACER EL ESFUERZO
-  DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR
-  PROVINCIA
-\end{itemize}
-
-*IDEM para la humedad. idem para sunshine
-
-\begin{itemize}
-\tightlist
-\item
-  HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA
-  DARLE RELEVANCIA A LA ESTIMACION SUR
-\end{itemize}
+In this paper we have presented a spatio-temporal analysis of incidence
+of COVID-19 in Spain. The analysis covers a 30-day period that begins
+immediately before the state of emergency was declared in the country.
+The focus of the research has been on the environmental correlates of
+incidence of the disease. There is a need for more empirical evidence,
+as policy makers, public health practitioners, and the public begin
+planning for the months ahead at this early stage of the pandemic.
+
+Our results offer strong support for the hypothesis that incidence of
+COVID-19 at the population level is lower at higher temperatures and
+levels of humidity: the estiamated effect is a reduction in the
+neighborhood of 3\% percent in incidence per each 1\% increase in
+temperature, and a 3\% reduction in incidence per 1\% increase in
+humidity \emph{under conditions of contagion}. These reductions are
+estimated to be lower when lockdown conditions are in place. The
+question here seems to be whether these environmental variables can
+yield a bigger reduction of \emph{more} cases, or a smaller reduction of
+\emph{fewer} cases. Our control variables also offer some interesting
+insights. In particular, there is evidence of behavioral adaptations in
+the form of risk compensation (density) and compliance with the lockdown
+(sunshine). These results offer a cautionary tale, with regards to the
+effectiveness of the lockdown in more dense areas, and also the
+implications for compliance with stay-at-home orders as the northern
+hemisphere moves towards summer and more hours of sunshine during the
+day.
+
+A key aspect of the analysis using spatial SUR models is that we were
+able to model incidence of COVID-19 as an interregional contagion
+process. Here, we find that the strength of the contagion effect was
+dramatically reduced by the lockdown.
+
+Needless to say, the analysis presented here is ecological, and at the
+level of population health. For this reason, the analysis does not make
+any claims with respect to the effect of ultraviolet light on the virus,
+but rather about transmission of the virus in the population. In this
+way, the analysis does not say that the virus moves less effectively
+where more people live in close proximity to each other, but rather that
+humans are more contagious when they feel safe in less dense regions.
+Similarly, more sunshine does not mean that the virus thrives, but
+rather that humans are more contagious to each other when their behavior
+adapts to this environmental condition.
+
+It is important, in this respect, to highlight some limitations of this
+study.
+
+First, the analysis was conducted under a situation of lockdown, and
+therefore, besides the results are the beginning of the period examined,
+one must exercise caution when trying to extrapolate the findings to
+other situations. Secondly, it is well known that there is in many
+countries substantial underreporting of cases of COVID-19 due to limited
+testing. However, we do not suspect systematic provincial bias in
+reporting, and as long as the underreporting is consistent across units
+of analysis, we do not expect biased results - even if we are aware that
+the number of true cases is likely higher. Thirdly, all environmental
+data are based on a single station in a province. Here, we made an
+effort to select the most representative station based on the
+distribution of the population. An alternative approach would be to
+calculate the environmental variables based on the average of multiple
+stations within each province: this would likely smooth to some extent
+these variables. Finally, we defined spatial neighborhoods based on
+adjacency. It would be interesting to compare other connectivity
+criteria, for instance based on domestic transportation instrastructure
+and services. We flag this as a matter for future research.
 
@@
 
+\leavevmode\hypertarget{ref-Noland1995perceived}{}%
+Noland, R.B., 1995. Perceived risk and modal choice - risk compensation
+in transportation system. Accident Analysis and Prevention 27, 503--521.
+doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}
+
 \leavevmode\hypertarget{ref-Pebesma2018}{}%
@@
 
+\leavevmode\hypertarget{ref-Phillips2011risk}{}%
+Phillips, R.O., Fyhri, A., Sagberg, F., 2011. Risk compensation and
+bicycle helmets. Risk Analysis 31, 1187--1195.
+doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}
+
 \leavevmode\hypertarget{ref-Prem2017projecting}{}%
@@
 spatial econometric perspective. Regional studies 33, 143--156.
+
+\leavevmode\hypertarget{ref-Richens2000risk}{}%
+Richens, J., Imrie, J., Copas, A., 2000. Condoms and seat belts: The
+parallels and the lessons. Lancet 355, 400--403.
+doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}
 
@@
 
+\leavevmode\hypertarget{ref-Yu2013characterization}{}%
+Yu, H.J., Alonso, W.J., Feng, L.Z., Tan, Y., Shu, Y.L., Yang, W.Z.,
+Viboud, C., 2013. Characterization of regional influenza seasonality
+patterns in china and implications for vaccination strategies:
+Spatio-temporal modeling of surveillance data. Plos Medicine 10, 16.
+doi:\href{https://doi.org/10.1371/journal.pmed.1001552}{10.1371/journal.pmed.1001552}
+
 \leavevmode\hypertarget{ref-Zellner1962efficient}{}%

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.Rmd
Revised abstract before submission

@@
 abstract: |
-  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress.
-While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees.
-How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine.
-In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain.
-Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems.
-An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process.
-Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations.
+  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health systems and the economy.
+Public health interventions have slowed the spread of the virus, albeit at a high cost.
+How and when to ease restrictions to movement hinges in part on whether SARS-CoV2 will display seasonality due to variations in temperature, humidity, and hours of sunshine.
+Here, we address this by means of a spatio-temporal analysis of the incidence of COVID-19 in Spain.
+Use of spatial Seemingly Unrelated Regressions (SUR) allows us to model the incidence of reported cases of the disease per 100,000 population as an interregional contagion process, in addition to a function of temperature, humidity, and sunshine.
+In the analysis we also control for GDP per capita, percentage of older adults in the population, population density, and presence of mass transit systems.
+The results support the hypothesis that incidence of the disease is lower at higher temperatures and higher levels of humidity.
 Sunshine, in contrast, displays a positive association with incidence of the disease.

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.tex
Revised abstract before submission

--- 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
@@
 \bibliographystyle{elsarticle-harv}
-\usepackage{graphicx}
-% We will generate all images so they have a width \maxwidth. This means
-% that they will get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-\else\Gin@nat@width\fi}
-\makeatother
-\let\Oldincludegraphics\includegraphics
-\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 \ifxetex
@@
 
-\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{5}
 % Pandoc toggle for numbering sections (defaults to be off)
-\setcounter{secnumdepth}{0}
 
 % Pandoc header
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 
@@
   \begin{abstract}
-  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
-  globe, causing enormous stress to health systems and prompting social
-  distance guidelines and mandates to arrest its progress. While there is
-  encouraging evidence that early public health interventions have slowed
-  the spread of the virus, this has come at a high cost as the global
-  economy is brought to its knees. How and when to ease restrictions to
-  movement hinges in part on the question whether SARS-CoV2 will display
-  seasonality associated with variations in temperature, humidity, and
-  hours of sunshine. In this research, we address this question by means
-  of a spatial analysis of the incidence of COVID-19 in the provinces in
-  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
-  allows us to model the incidence of reported cases of the disease per
-  100,000 population, as a function of temperature and humidity, while
-  controlling for GDP per capita, population density, percentage of older
-  adults in the population, and presence of mass transit systems. An
-  interesting aspect of the spatial SUR approach is that it models
-  incidence as a contagion process. Our results indicate that incidence of
-  the disease is lower at higher temperatures and higher levels of
-  humidity, although coefficients for this variable are significant only
-  in some equations. Sunshine, in contrast, displays a positive
+  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
+  systems and the economy. Public health interventions have slowed the
+  spread of the virus, albeit at a high cost. How and when to ease
+  restrictions to movement hinges in part on whether SARS-CoV2 will
+  display seasonality due to variations in temperature, humidity, and
+  hours of sunshine. Here, we address this by means of a spatio-temporal
+  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
+  Unrelated Regressions (SUR) allows us to model the incidence of reported
+  cases of the disease per 100,000 population as an interregional
+  contagion process, in addition to a function of temperature, humidity,
+  and sunshine. In the analysis we also control for GDP per capita,
+  percentage of older adults in the population, population density, and
+  presence of mass transit systems. The results support the hypothesis
+  that incidence of the disease is lower at higher temperatures and higher
+  levels of humidity. Sunshine, in contrast, displays a positive
   association with incidence of the disease. Our control variables also
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-From a small outbreak linked to a live animal market at the end of 2019
-to a global pandemic in a matter of weeks, the SARS-CoV2 virus has
-threatened to overrun health systems the world over. In efforts to
-contain the spread, numerous governments in many nations and regions
-have either recommended or mandated social distancing measures, and as
-of this writing, millions of people in five continents shelter in place.
-There are encouraging signs that these measures have arrested the spread
-of the virus where they have been implemented, and have thus helped to
-keep a bad situation from becoming even worse (e.g., 2020). However,
-this has come at a high cost, and the consequences for all spheres of
-the economy, social, and cultural life have been dire (e.g., Fernandes,
-2020; Luo and Tsang, 2020). As a result, there is a sense of urgency to
-anticipate the progression of the pandemic, in order to plan for
-progressive lifting of restrictions to movement and social contact
-(e.g., Kissler et al., 2020). Needless to say, this has become a
-delicate, and politically charged, balancing act between public health
-and the economy (Gong et al., 2020).
-
-A salient question in the debate on how and when to ease social
-distancing measures is whether the virus will display seasonal
-variations. Earlier, diverse studies have shown the effect of
-temperature and humidity on the incidence of influenza (e.g., Mäkainen
-et al., 2009; Jaakkola et al., 2014; Kudo et al., 2019). Jaakkola et
-al.~(2014), for example, found that a decrease of temperature and
-absolute humidity precedes the onset of symptoms of influenza A and B
-viruses by 3 days in places where the temperature is low. After the
-2002-2004 outbreak of SARS, researchers also began to investigate the
-relationship between these factors and SARS-CoV. In this way, Casanova
-et al.~(2010) used two surrogates, namely the gastroenteritis (TGEV) and
-mouse hepatitis viruses (MHV), to find that virus inactivation was more
-rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of
-humidity, these researchers reported that survival of the virus was
-lower at moderate relative humidity levels. In a similar vein, but
-working directly with SARS-CoV, Chan et al.~(2011) found that viability
-of the virus was lost at temperatures higher than 38C and relative
-humidity superior to 95\%.
-
-While existing research on similar pathogens suggests that SARS-CoV is
-more stable and potentially easier to transmit in conditions of low
-temperature and low humidity, it is far from certain that this will also
-be the case with the novel SARS-CoV2. If such is the case, there is the
-possibility of easing restrictions to social contact as the weather
-warms; however, weeks or possibly months of costly measures could become
-undone if not, and the restrictions are lifted prematurely. Not
-surprisingly, given the stakes involved, this issue has already
-triggered a lively debate.
-
-Some of what is thought about the possible seasonality of COVID-19 is
-based on analogies to the patterns of other known respiratory viruses.
-However, de Ángel Solá et al.~(2020) note that ``not all seasonal
-respiratory viruses experience the same spatiotemporal patterns''
-(section 4). This urges caution when extrapolating from known viruses,
-and indeed, the evidence in this respect is inconclusive. At a global
-scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
-however, Coelho et al.~(Coelho et al., 2020) warn that at least during
-the exponential phase, expansion of the virus is not driven by climate.
-Similarly, whereas Araujo and Naimi (2020) argue that spread of
-SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
-remain unconvinced that spatial modelling can currently discriminate the
-distribution of the disease on the basis of climate, at least in the
-United States. Yao et al.~(2020), examined data from China and came to
-the conclusion that neither temperature nor ultraviolet indices had an
-association with transmission of COVID-19. This is despite previous
-research that has linked less exposure to UVB radiation to higher
-prevalence and severity of acute respiratory tract infections
-(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
-al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
-Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
-much of the relevant work has yet to be peer-reviewed and therefore is
-open to change (see for example the challenge of Harbert et al. (2020)
-to Araujo and Naimi (2020)). In the United States, the National Academy
-of Sciences, Engineering, and Medicine was engaged by the Office of the
-Executive for guidance on this matter (see National Academies of
-Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
-the situation well (see p.~6): ``Some limited data support a potential
-waning of cases in warmer and more humid seasons, yet none are without
-major limitations\ldots Additional studies as the SARS-CoV-2 pandemic
-unfolds could shed more light on the effects of climate on
-transmission.''
-
-With the above considerations in mind, our objective with this paper is
-to contribute to the knowledge basis regarding the spread of COVID-19
-and the influence of environmental factors, particularly temperature,
-humidity, and sunshine. We adopt a population health approach, and
-report results from a spatial model of the incidence of COVID-19 in
-fifty provinces in Spain, one of the countries hardest hit by the
-pandemic. We combine data on reported cases of the disease with
-metereological information, to create a spatio-temporal dataset covering
-a period of 30 days. We then join this dataset with provincial-level
-economic and demographic information to act as controls to our key
-environmental variables. These data are analyzed using a spatial
-Seemingly Unrelated Regressions (SUR) approach, which allows us to model
-incidence of COVID-19 as a spatial contagion process.
-
-The results provide evidence of the effect of temperature, humidity, and
-sunshine on the incidence of COVID-19. The clearest result with respect
-to these variables is a lower incidence of COVID-19 at higher
-temperatures and levels of humidity, while the opposite happens with
-respect to hours of sunshine. Our control variables also provide some
-intriguing insights. Higher incidence is associated with higher GDP per
-capita and presence of mass transit systems in the province; in
-contrast, population density and percentage of older adults display
-negative associations with incidence of COVID-19. The results of this
-analysis provide support to the hypothesis of seasonality of the novel
-SARS-CoV2, and should be of interest to public health officials and
-policy makers grappling with the question of the trajectory of the
-pandemic.
-
-Please note that this paper is prepared as a reproducible research
-document. The source R Markdown document, as well as all data and code
-needed to reproduce/review/extend the analysis can be obtained from the
-following repository:
-
-https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
-
-\hypertarget{context-and-data}{%
-\section{Context and Data}\label{context-and-data}}
-
-\hypertarget{covid-19-in-spain}{%
-\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}
-
-The first reported case of COVID-19 in Spain was on January 31th, when a
-German tourist in the Canary Islands tested positive for the virus.
-However, it was still a few weeks before the first domestic case was
-reported, on February 27th in Sevilla province (Andalusia). In a short
-period of time, after this relatively slow start, the outbreak flared.
-By March 11th the World Health Organization (WHO) declared COVID-19
-officially a pandemic. This declaration marked a turning point for the
-public in Spain too. As of March 13th, the number of cases of COVID-19
-reported in Spain was 4,473, with a majority of cases (1,990)
-concentrated in Madrid: these numbers were at the time the worst
-outbreak in Europe, after Italy. In response to the situation, on March
-13th the Spanish National Government declared a state of emergency, to
-go into effect on Saturday March 14th. As part of the state of emergency
-restrictions to most activities were imposed, with the exception of
-essential services (e.g.~food, health) and some economic subsectors of
-industry and construction. A few days later, on March 17th, Spain closed
-its lands borders to allow entry only to returnee nationals and
-permanent residents. The lockdown was further hardened between March
-30th and April 12th (including the Easter weekend of April 10th-12th)
-and during this period only essential activities were allowed. During
-this period, there was a dramatic reduction in overall mobility, both
-within provinces as between
-\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
-
-\hypertarget{selection-of-variables}{%
-\subsection{Selection of Variables}\label{selection-of-variables}}
-
-The global emergence of infectious diseases is mostly driven by
-environmental, ecological, and socio-economic factors (Jones et al.,
-2008). In the case of SARS-CoV2, the ecological factors include the
-interaction between humans and wildlife. Once transmission of a disease
-begins to happen between humans, socio-economic and environmental
-factors become increasingly important. As noted in the introduction, the
-focus of the paper is on environmental variables, namely temperature,
-humidity, and sunshine. These variables have been implicated in the
-viability and ease of transmission of similar viruses. In addition to
-these variables, we also aim to use a set of controls, in the form of
-specific socio-economic and demographic characteristics of each
-province.
-
-The first variable that we consider is GDP per capita. Much has been
-said about globalization and the spread of infectious
-disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
-A18)}.
-The growth in global connections has presented a challenge to spatial
-approaches in the initial stages of disease management, when the cause
-of a disease may still be unclear, but the plane has already departed
-(Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
-van Wagner (Van Wagner, 2008) chronicles how Toronto's status as a
-global city turned out to be a vulnerability in this respect. In our
-case, we think of GDP per capita as a marker of a region's relative
-position in a network of global cities, and its potential to be further
-ahead in the trajectory of the pandemic. Furthermore, wealthier regions
-also tend to concentrate more activities that produce non-traded goods,
-including building and construction (Hallet, 2002). Therefore, it is
-possible that wealthier regions remain relatively more active even
-during a lockdown. On the other hand, it is also possible that less
-wealthy regions have a higher proportion of workers in manual
-occupations that cannot telework, and therefore have more difficulties
-complying with shelter-in-place orders.
-
-The percentaje of older adults (over 65) is the second variable that we
-consider as a control. Early evidence regarding COVID-19 suggests that
-the case rate mortality is higher at older ages (e.g.~The Novel
-Coronavirus Pneumonia Emergency Response Epidemiology Team, 2020).
-However, it is not clear that a relatively large population of older
-adults necessarily translates into higher transmission of the infection.
-The tool of choice in containing the spread of the disease has been
-social distancing. In this respect, the evidence from the field of
-transportation is that older adults tend to travel less frequently, for
-shorter distances, and have higher rates of immobility (e.g., Roorda et
-al., 2010; Morency et al., 2011; Sikder and Pinjari, 2012). In other
-words, many older adults are, whether by preference or otherwise,
-already in a form of social isolation. Social distancing during the
-pandemic may actually reinforce that condition for them, as suggested by
-the analysis of age-structured social contact in India, China, and Italy
-of Sing and Adhikari (2020). Accordingly, our expectation is that
-provinces with higher percentages of older adults will tend to have
-similarly low levels of social contact, particularly since the
-age-structured matrix of social contact in Spain is similar to Italy
-(Prem et al., 2017).
-
-The third population variable the we consider is population density,
-since it directly affects the contact patterns and contact rates between
-individuals in a population (Hu et al., 2013). The evidence available
-suggests a positive relationship between the transmission of COVID-19
-and population density (e.g.~cumulative incidence in urban areas like
-NYC). For this reason, we anticipate a positive relationship between
-population density and the incidence of the disease.
-
-The last control variable is the presence of mass transit systems in a
-province. Every province in Spain offers some form of public
-transportation, however only five provinces have higher order systems of
-mass mobility (e.g.~metro or subway), namely Barcelona, Madrid, Sevilla,
-Valencia, and Bizkaia. Public transportation has been hypothesized to
-relate to the spread of contagious disease by some researchers using
-agent-based approaches and simulation (e.g., Perez and Dragicevic, 2009;
-Wang et al., 2010), and while we find scant evidence of a link in the
-literature, the idea is intuitively appealing. After all, unlike the
-isolation that a car offers to travellers, most mass transit system are
-cauldrons of social contact.
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-Our dataset includes information about the daily number of cases of
-COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
-terminology) for the period between March 13th and April 11th,
-inclusive. For our purposes, we consider positive cases reported, but
-excluding symptomatic cases diagnosed by a doctor without a Polymerase
-Chain Reaction (PCR) test. The Spanish National Government publishes
-periodic updates at the regional level (NUTII) and the information is
-also released at the provincial level as part of a collaborative project
-by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
-ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
-and ESRI España. This information is compiled from several sources,
-mainly the official web pages of the Spanish regional goverments, as
-documented in Centro de Datos
-Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
-In addition, we consider two sets of explanatory variables. The first
-one, and the focus of this research, is set of two environmental
-variables, namely temperature and humidity, which are collected from
-official sources (i.g., \emph{AEMET}, the state meteorology agency, and
-\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
-second set provides some relevant controls for multivariate analysis,
-and refers to economic and demographic attributes of the province (also
-collected from official sources, i.e., INE, the national statistics
-institute). Table \ref{tab:descriptive-statistics} shows the descriptive
-statistics and the provenance of the data.
-
-The spatial and temporal coverage of the data is as follows. Our dataset
-begins on March 13, which is the first date when every province had
-reported at least one case of COVID-19, and continues until April 11,
-for a period of 30 days. The unit of analysis is the province. Provinces
-are the equivalent of states, and are embedded in Autonomous
-Communities. As an example, Cataluña is an Autonomous Community and
-consists of four provinces, namely Barcelona, Gerona, Lerida, and
-Tarragona. The size of the provinces is relatively large, as seen in
-Table \ref{tab:descriptive-statistics}. The smallest province is
-\(1,978.12km^2\) (this is smaller than Rhode Island in the US) and the
-largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
-in the US). While this is a relatively large degree of spatial
-aggregation, reporting on COVID-19 is inconsistent at smaller
-geographies, or cases are not reported at that level at all. The
-analysis must therefore be considered ecological.
-
-\begin{table}
-
-\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics} Descriptive statistics}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
-\toprule
-Variable & Note & Min & Mean & Max & SD & Source\\
-\midrule
-\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & ProvidencialData19\\
-Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
-\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4805.98 & INE\\
-Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
-\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
-\addlinespace
-Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
-\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
-\bottomrule
-\multicolumn{7}{l}{\textit{Note: }}\\
-\multicolumn{7}{l}{ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19}\\
-\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
-\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
-\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
-\end{tabular}}
-\end{table}
-
-An important aspect of working with environmental data such as
-temperature and humidity is the incubation period of the disease. Lauer
-et al.~(2020) report for the case of COVID-19 a median incubation period
-of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The
-vast majority of cases (95\%) develop symptoms between 2.6 days (CI, 2.1
-to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we
-judge it best to use lagged values of the environmental variables. We
-test different time lags as follows. We consider a lagged 8-day average,
-from date-minus-12 to date-minus-5 days (hereafter \emph{lag8}).
-Secondly, we consider a lagged 11-day average, from date-minus-12 to
-date-minus-2 days (hereafter \emph{lag11}). Finally, to account for the
-likely duration of incubation, we consider a lagged 11-day
-\emph{weighted} average, from date-minus-12 to date-minus-2 days
-(hereafter \emph{lag11w}). The weights for this variable are calculated
-using the parameters of the log-normal distribution reported by Lauer et
-al.~(2020), i.e., a log-mean of 1.621 and a log-standard deviation of
-0.418. With these weights, the environmental variables at date-minus-2
-and date-minus-12 days are weighted as 0.041 and 0.009 respectively,
-whereas the environmental variables at date-minus-5 days are weighted as
-0.194.
-
-\hypertarget{methods-the-spatial-sur-model}{%
-\section{Methods: the Spatial SUR
-Model}\label{methods-the-spatial-sur-model}}
-
-The Seemingly Unrelated Regression equations model (SUR hereafter) is a
-multivariate econometric model used in different fields when the
-structure of the data consisits of cross-sections for different time
-periods. The basis of this approach is well-known since the initial
-works of Zellner (1962), and has become a popular methodology included
-in several econometrics textbook (e.g., Greene, 2003). To our knowledge,
-Anselin (1988) was the first author to discuss SUR from a spatial
-perspective. In his landmark text, Anselin discussed a model made of
-``an equation for each time period, which is estimated for a cross
-section of spatial units'' (p.~141). From this milestone, a large body
-of research has developed to extend the classical SUR into a spatial
-framework (e.g., Rey and Montouri, 1999; Lauridsen et al., 2010; Le
-Gallo and Dall'Erba, 2006; López et al., 2017).
-
-The classical SUR model without spatial effects (from here, SUR-SIM) is
-a stack of equations as follows:
-
-\begin{equation}
-\label{eq:sur-sim}
-\begin{bmatrix}
-y_1 \\ y_2 \\ \vdots \\ y_T
-\end{bmatrix}
-=
-  \begin{bmatrix}
-X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
-\end{bmatrix}
-\
-\begin{bmatrix}
-\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
-\end{bmatrix}
-+
-\begin{bmatrix}
-\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
-\end{bmatrix}
-\end{equation}
-
-\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
-vector, and in our case \(y_{st}\) is the incidence ratio in the
-province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\); \(X_t\) is
-a \(N \times k_t\) matrix of the \(k_t\) independent variables, with
-associated vector of coefficients \(\beta_t\),;
-\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
-\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
-residuals.
-
-A key feature of the SUR model is the dependence structure among the
-vectors of residuals, namely:
-
-\begin{equation}
-\label{eq:sur-err}
-E[\epsilon_t \epsilon'_{t'}]=\sigma_{tt'}
-\end{equation}
-
-Note that this specification is very flexible, in that it allows changes
-in the coefficients \(\beta_{it}\) in order to modulate the effect of
-\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
-posible to impose restrinctions considering some \(\beta\) coefficients
-as being constant over time. In this case, we can reformulate the
-coefficients expression
-\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
-to restrict the first \(r\) coefficients to be constant across
-equations. This is equivalent to specifying some effects to be invariant
-over time.
-
-Equation (\ref{eq:sur-sim}) can be rewriten in compact form:
-
-\begin{equation}
-\bf{Y} = \bf{X} \beta + \epsilon
-\label{eq:sur-sim-block}
-\end{equation}
-
-\noindent where \(\bf{Y}\) is now a vector of dimension \(NT \times 1\),
-\(\bf{X}\) is a block-diagonal matrix \(NT \times K\) (with
-\(K = \sum_t{k_t}\)) and \(\epsilon\) is an \(NT \times 1\) vector.
-Using the Kronecker product notation (\(\otimes\)), the error matrix
-structure is expressed concisely as:
-
-\begin{equation}
-E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
-\end{equation}
-
-As is the case with cross-sectional data, it is possible to test the
-residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation,
-and several tests have been developed to test the null hypothesis of
-spatial independence (see López et al., 2014). When the null hypothesis
-is rejected, several alternative specifications have been proposed to
-include spatial effects (Anselin, 1988, see also 2016). In this paper we
-consider a spatial SUR model that incorporates a spatial lag of the
-dependent variable as an explanatory factor. Spatial analytical
-approaches were used to understand contagion-difussion processes in the
-case of infectious disease in general (e.g., Cliff et al., 1998) and the
-2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
-al., 2010). While we are mindful of the same caveat that the novel
-SARS-CoV2 may not follow the patterns of previous diseases, there is
-still evidence from the United States that COVID-19 displays spatial
-patterns that are consistent with a diffusion process (Desjardins et
-al., 2020). For this reason, the spatial lag model is appropriate to
-model incidence of COVID-19 geographically, since it accounts for
-potential spatial patterns that result from a process of contagion, as
-explained next.
-
-The stack expresion for the SUR model with a spatially lagged dependent
-variable (SUR-SLM) is as follows:
-
-\begin{equation}
-\label{eq:sur-slm}
-\begin{aligned}
-\bf{AY} = \bf{X} \beta + \epsilon \\
-\epsilon =N(0,\bf{\Sigma})
-\end{aligned}
-\end{equation}
-
-\noindent where \(\bf{A} =I_{TN}-\bf{\Gamma} \otimes W\) is the
-spatially lagged dependent variable, and
-\(\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)\).
-
-This specification assumes that outcome (\(y_{st}\)) at location \(s\)
-and time \(t\) is partially determined by the weighted average
-(\(Wy_{st}\)) of the outcome in neighboring provinces, with neighborhood
-defined by matrix \(W\) of spatial weights. In other words, the
-spatially lagged dependent variable represents a process of contagion,
-where the disease in neighboring provinces can spillover in a spatial
-way. The coefficients of the spatially lagged variable are estimated for
-each time period \(\rho_t\) and identify the intensity and the sign of
-the contagion effect. It is possible test the null hypothesis of
-identical levels of spatial dependence (\(\rho_i=\rho_j, \forall i,j\)).
-The correspond Wald test is available in the \texttt{R} package
-\texttt{spsur}.
-
-The SUR-SLM model can be estimated using maximum likelihood (López et
-al. (2014)) or instrumental variables (Mínguez et al. (2019)).
-
-\hypertarget{analysis}{%
-\section{Analysis}\label{analysis}}
-
-\hypertarget{exploratory-data-analysis}{%
-\subsection{Exploratory Data Analysis}\label{exploratory-data-analysis}}
-
-Figure \ref{fig:weekly-average-incidence-map} shows the geographical
-variation in the incidence of COVID-19 in Spain, as well as the temporal
-progression of the disease in weekly averages. Our analysis begins on
-March 13, which is the first date when every province had reported at
-least one case of COVID-19. It can be seen that the highest incidence at
-this early date was in the provice of Álava, in the North of Spain.
-While not the most populous province, with a population of only 331,549,
-Álava has the highest GDP per capita of all provinces. Vitoria, its main
-city, is the capital of the Basque Country and has been the focus of
-efforts, along with San Sebastian and Bilbao, to develop a ``Global
-Basque City'' (Meijers et al., 2008). The other early focus of the
-pandemic in Spain was Madrid, which is the most populous province in the
-country and has the second highest GDP per capita after Álava. The early
-outbreaks in these two provinces can be traced throughout the
-progression of the pandemic over time, although by the end of the period
-under consideration, other provinces had matched and even surpased their
-incidence rates, including Segovia and Soria to the north of Madrid, and
-Ciudad Real and Albacete to the south.
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/weekly-average-incidence-map-1.pdf}
-\caption{\label{fig:weekly-average-incidence-map}Mean weekly incidence
-of COVID-19 by province, in reported cases by 100,000 people}
-\end{figure}
-
-Figure \ref{fig:descriptives-temperature} shows the distribution of the
-environmental variables in Spain. For ease of visualization we aggregate
-the provinces by Autonomous Community. Each box-and-whisker in the
-figure represents the distribution of the variable for a community over
-the 30-day period. In the plot, the communities have been sorted by
-latitude, so that Principado de Asturias is the northernmost community,
-and Canarias the southernmost. As seen in the figure, there is a
-relatively wide range of values both within and between provinces over
-the 30-day period examined. The top panel of the figure shows the
-distribution of mean temperatures. The lowest mean temperature for a
-community on any given day was 2.8C, and the highest 22.4C, for a range
-of approximately 20 degrees. Likewise, there is a great deal of
-variability in humidity, as seen in the middle panel of the figure,
-where the lowest mean humidity for any community is 48.3 and the highest
-is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
-in the community. This variable displays the most variability within
-communities over time, but remains relatively constant across
-communities. It is important to note that these are summaries by
-community, and the actual values of these variables for the provinces
-display somewhat more variability.
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/descriptives-temperature-1.pdf}
-\caption{\label{fig:descriptives-temperature} Distribution of mean
-temperatures and humidities in the Autonomous Communities in Spain
-between March 12, 2020 and April 11, 2020. The Autonomous Communities
-have been sorted by latitude, with communities to the left being the
-northermost, and to the right the southernmost}
-\end{figure}
-
-Figure \ref{fig:map-controls} includes three maps that display the
-spatial variation of our control variables, namely GDP per capita,
-percentage of older adults in province, population density, and presence
-of mass transit systems. As seen there, GDP per capita is higher in
-Madrid and the northeast part of the country, mainly in Pais Vasco and
-Cataluña. Percentage of older adults is somewhat more checkered, with
-high values in Madrid and other provinces in the center-west part of the
-country, but also in some provinces in the north. Outside of provinces
-with major cities (e.g., Madrid; Bizkaia and Gipuzkoa in Pais Vasco;
-Pontevedra in Galicia), population density tends to be higher in
-provinces along the Mediterranean coast. The final panel in the figure
-shows the five provinces in the country that have higher order mass
-transit systems (e.g., metro).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/map-controls-1.pdf}
-\caption{\label{fig:map-controls}Spatial distribution of control
-variables by province}
-\end{figure}
-
-Figure \ref{fig:daily-correlations} shows the distribution of daily
-simple correlations of incidence of COVID-19 with the independent
-variables (with the exception of Transit, which is a categorical
-variable). These correlations are calculated after log-transforming all
-variables. As previously discussed, the environmental variables have a
-temporal lag and were calculated different time windows.
-
-It can be seen in the figure that temperature (in its three forms) has
-the highest simple correlation with incidence of COVID-19. After
-temperature, GDP per capita has the highest positive correlation with
-the dependent variable. The distribution of these correlations is also
-quite tight over the 30-day period of the study. Hours of sunshine tends
-to have a moderately high correlation with incidence of COVID-19,
-however the distribution of these correlations is more spread, and in
-some cases strays into negative values. A similar thing happens with
-humidity, which also tends to display more day to day variation in the
-correlation with the dependent variable. The percentage of older adults
-shows a relatively tight distribution of day-to-day correlations, and is
-negative. Population density, in contrast, tends to be negative, but is
-relatively spread, and on some days, the simple correlation between
-density and incidence of COVID-19 is weakly positive.
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/daily-correlations-1.pdf}
-\caption{\label{fig:daily-correlations}Distribution of daily
-correlations of the independent variables with daily incidence of
-COVID-19 (all variables have been log-transformed)}
-\end{figure}
-
-\hypertarget{sur-models}{%
-\subsection{SUR Models}\label{sur-models}}
-
-Correlation analysis in the preceding section provides some insights
-about the potential associations between incidence of COVID-19 and the
-various environmental and control variables. In this section we estimate
-three spatial SUR models to test the differences between the various
-temporal lags and weighting schemes for the environmental variables.
-Accordingly, we define three models: Model 1, which is estimated using
-the lagged 8-day averages of the environmental variables (\emph{lag8});
-Model 2, which is estimated using the lagged 11-day averages of the
-environmental variables (\emph{lag11}); and finally, Model 3, which is
-estimated using the lagged 11-day \emph{weighted} averages of the
-environmental variables (\emph{lag11w}).
-
-To implement the SUR approach, we must define a matrix of spatial
-weights \(W\). In this case, we consider neighborhoods based on
-adjacency, based on the well-known rook criterion (two provinces are
-adjacent if they share a boundary, but not if they touch only at a
-vertex). We modify this criterion in two cases. First, we make an
-allowance for adjacency between the two islands in the Autonomous
-Community of Canarias in the Pacific (Las Palmas and Santa Cruz de
-Tenerife), which we assume are adjacent between each other. And
-secondly, we assume that Islas Baleares in the Mediterranean are
-adjacent to three provinces in Pais Catalans (i.e., Barcelona,
-Tarragona, and Castello). After matrix \(W\) has been specified,
-estimation of the model can proceed as usual.
-
-For estimation, we log-transform the dependent and quantitative
-independent variables. The only variable that is not transformed is the
-categorical variable for transity systems. A log-log transformation is
-appropriate to capture non-linear relationships between variables and
-provides a straightforward interpretation of the coefficients as
-percentage change. Furthermore, we introduce restrictions so that the
-coefficients of two of our control variables are constant over time,
-namely GDP per capita and percentage of older adults. We do not see an
-\emph{a priori} reason to let those two variables vary across equations,
-and the correlation analysis in Figure \ref{fig:daily-correlations} also
-suggest little temporal variation. In contrast, we let the spatial
-autocorrelation parameter, as well as the parameters of the rest of the
-independent variables (including the constant) to vary over
-time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious.
-These results are available in the source R markdown document.}.
-
-After estimation, we compare the goodness of fit of the three SUR
-models. Figure \ref{fig:goodness-of-fit} shows the equation-level
-coefficient of determination \(R^2\), one for each time period/day. As
-well, the overall coefficient of determination for the system is
-reported for each model \(\text{pooled}-R^2\). The general trend is
-identical for the three models, with the goodness-of-fit improving over
-time and plateuing around a value of \(R^2\) of 0.6. Model 1
-(\emph{lag8}) performs somewhat better in the first few equations/days,
-when the goodness-of-fit is relatively poor, and then again in the last
-few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
-perform well towards the end of the study period. The most balanced
-model in terms of equation-level goodness-of-fit appears to be Model 1
-(\emph{lag11}), and this impression is further supported with a slightly
-higher value of the \(\text{pooled}-R^2\). This analysis is in line with
-the incubation period reported by Lauer et al.~(2020), although not the
-use of a weighted average. For the remainder of the paper, we will adopt
-Model 2 (\emph{lag11}) as our best model. In the following section we
-discuss the results of the analysis in more depth.
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/goodness-of-fit-1.pdf}
-\caption{\label{fig:goodness-of-fit} Goodness of fit of the SUR systems:
-by date and pooled}
-\end{figure}
-
-\hypertarget{results-and-discussion}{%
-\section{Results and Discussion}\label{results-and-discussion}}
-
-Table \ref{tab:summary-best-model} presents a summary.
-
-\begin{table}
-
-\caption{\label{tab:summary-best-model}\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrrrrrrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Estimates} & \multicolumn{3}{c}{Significance} & \multicolumn{1}{c}{ } \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
-Variable & Min & Mean & Max & p > 0.10 & 0.10 <= p < 0.05 & p <= 0.05 & Note\\
-\midrule
-\rowcolor{gray!6}  Intercept & 7.370 & 10.037 & 12.968 & 0 & 0 & 30 & Non-constrained\\
-log(GDPpc) & 0.620 & 0.620 & 0.620 & 0 & 0 & 1 & Constrained\\
-\rowcolor{gray!6}  log(Older) & -0.737 & -0.737 & -0.737 & 0 & 0 & 1 & Constrained\\
-log(Density) & -0.220 & -0.097 & 0.153 & 19 & 0 & 11 & Non-constrained\\
-\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 8 & 8 & 14 & Non-constrained\\
-\addlinespace
-log(Humidity) & -1.434 & -0.534 & -0.031 & 10 & 1 & 19 & Non-constrained\\
-\rowcolor{gray!6}  log(Temperature) & -2.014 & -1.406 & -0.929 & 0 & 0 & 30 & Non-constrained\\
-log(Sunshine) & -0.258 & 0.097 & 0.206 & 7 & 2 & 21 & Non-constrained\\
-\rowcolor{gray!6}  Spatially lagged y (rho) & 0.015 & 0.142 & 0.528 & 12 & 3 & 15 & Non-constrained\\
-\bottomrule
-\multicolumn{8}{l}{\textit{Note: }}\\
-\multicolumn{8}{l}{Significance: This is the number of coefficients with p-values as indicated}\\
-\multicolumn{8}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
-\multicolumn{8}{l}{Constrained: coefficient as constant across equations}\\
-\end{tabular}}
-\end{table}
-
-Figure \ref{fig:results-time-1} shows the temporal evolution of the
-spatial autocorrelation coefficient (\(\rho\)) and the intercept of the
-model.
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/results-time-1-1.pdf}
-\caption{\label{fig:results-time-1}Temporal evolution of the spatial
-autocorrelation coefficient (rho) and the intercept of the model; dots
-are the point estimates and vertical lines are 95\% confidence
-intervals. In yellow is the period after the declaration of the state of
-emergency, and in orange is the period when only essential activities
-were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
-coefficients for the two control variables that were not fixed over
-time, i.e., \(\log(Density)\) and \(Transit\).
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-controls-time-1.pdf}
-\caption{\label{fig:beta-controls-time}Temporal evolution of coefficient
-for the control variables; dots are the point estimates and vertical
-lines are 95\% confidence intervals. In yellow is the period after the
-declaration of the state of emergency, and in orange is the period when
-only essential activities were allowed.}
-\end{figure}
-
-Figure \ref{fig:beta-environmental-time} shows the temporal evolution of
-the coefficient for the three environmental variables.
-
-\begin{figure}
-\centering
-\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmenta-time-1.pdf}
-\caption{\label{fig:beta-environmental-time}Temporal evolution of
-coefficient for the environmental variables; dots are the point
-estimates and vertical lines are 95\% confidence intervals. In yellow is
-the period after the declaration of the state of emergency, and in
-orange is the period when only essential activities were allowed.}
-\end{figure}
-
-\hypertarget{concluding-remarks}{%
-\section{Concluding Remarks}\label{concluding-remarks}}
-
-More words go here.
-
-Limitaciones
-
-\begin{itemize}
-\item
-  La temperatura puedes ser un factor pero cabría esperar que no tuviera
-  un impacto lineal. Por el contrario deberiamos esperar un `punto de
-  corte': Una temperatua mantenida superior a X grados durante siete
-  días sea la mejor forma de incorporarla al modelo. En nuestro modelo
-  (log-log) en incremento en un 1\% de la temperatura se asocia con un
-  incremento beta\% en la incidencia. Esto es lo mismo si la temperatura
-  es baja que si es alta.
-\item
-  Idem para la humedad y horas de sol
-\item
-  Los datos son `provisionales'. Hay gran confusión sobre la incidencia
-  real. La ausencia de test de diagnóstico PCRs al inicio de la pandemia
-  (también ahora) puede desvirtuar el número de casos dianosticados.
-\item
-  Los datos oficiales (que tampoco son fiables) son reportados a nivel
-  de Comunidades Autónomas. La recopilación de datos a nivel de
-  provincial son el resultado de un esfuerzo colaborativo de
-  recopilacion entre distintas fuentes (principalmente gobiernos
-  locales). Nuevamente puede haber sesgos importanes.
-\item
-  En carácter insular de las Islas Canarias y de las Islas baleares no
-  se ha tenido en cuenta.
-\item
-  Baleares se ha linkado artificialmente con 3/4 provincias y debería de
-  haberse dejado aislada para respetar su carácter insular al definir la
-  matriz W.
-\item
-  La incidencia depende de un estado inicial. Al inicio del estudio
-  había provincias en las que la epidemia estaba muy desarrollada
-  (Madrid/Alava) mientras que en otras apenas habías casos. Este hecho
-  no ha sido considerado en el modelo. Decretar el confinamiento debe
-  tener distintos impactos entre provincias. QUIZAR METER UNA VARAIBLE
-  DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON
-  MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
-\item
-  No se ha controlado por el sistema sanitario de cada provincia. Uno de
-  los principales focos de contagio han sido los hospitales y los
-  centros de salud. En aquellas provincias donde se ha promovido el
-  mensaje ``NO IR AL MEDICO'' han presentado menor incidencia.
-\item
-  ¿hasta que punto las variables de control no recojen también factores
-  climáticos? por ejemplo, la gente jóven vive en el sur de España que
-  ha tenido menos incidencia.
-\item
-  Las estaciones meteorológicas para la obtencion de datos climáticos
-  has sido elegidas aleatoriamente (una para cada provincia). otra
-  seleccion puede dar otros resultados. AQUI SE PODRÍA HACER EL ESFUERZO
-  DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR
-  PROVINCIA
-\end{itemize}
-
-*IDEM para la humedad. idem para sunshine
-
-\begin{itemize}
-\tightlist
-\item
-  HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA
-  DARLE RELEVANCIA A LA ESTIMACION SUR
-\end{itemize}
-
-\hypertarget{acknowledgments}{%
-\section*{Acknowledgments}\label{acknowledgments}}
-\addcontentsline{toc}{section}{Acknowledgments}
-
-Add acknowledgments as appropriate in final draft.
-
-The following \texttt{R} packages were used in the course of this
-investigation and the authors wish to acknowledge their developers:
-\texttt{aemet} {[}{]}, \texttt{ggthemes} (Arnold, 2019),
-\texttt{gridExtra} (Auguie, 2017), \texttt{kableExtra} (Zhu, 2019),
-\texttt{knitr} (Xie, 2015, 2014), \texttt{lubridate} (Grolemund and
-Wickham, 2011), \texttt{plm} (Millo, 2017), \texttt{rticles} (Allaire et
-al., 2020), \texttt{sf} (Pebesma, 2018), \texttt{spdep} (Bivand et al.,
-2013), spsur (Angulo et al., 2020) \texttt{tidyverse} (Wickham et al.,
-2019), \texttt{units} (Pebesma et al., 2016).
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Allaire2020}{}%
-Allaire, J., Xie, Y., R Foundation, Wickham, H., Journal of Statistical
-Software, Vaidyanathan, R., Association for Computing Machinery,
-Boettiger, C., Elsevier, Broman, K., Mueller, K., Quast, B., Pruim, R.,
-Marwick, B., Wickham, C., Keyes, O., Yu, M., Emaasit, D., Onkelinx, T.,
-Gasparini, A., Desautels, M.-A., Leutnant, D., MDPI, Taylor and Francis,
-Ögreden, O., Hance, D., Nüst, D., Uvesten, P., Campitelli, E.,
-Muschelli, J., Kamvar, Z.N., Ross, N., Cannoodt, R., Luguern, D.,
-Kaplan, D.M., 2020. Rticles: Article formats for r markdown.
-
-\leavevmode\hypertarget{ref-Angulo2020spsur}{}%
-Angulo, A., Lopez, F.A., Minguez, R., Mur, J., 2020. Spsur: Spatial
-seemingly unrelated regression models.
-
-\leavevmode\hypertarget{ref-Anselin1988spatial}{}%
-Anselin, L., 1988. Spatial econometrics: Methods and models, Studies in
-operational regional science. Kluwer Academic Publishers, Dordrecht.
-
-\leavevmode\hypertarget{ref-Anselin2016estimation}{}%
-Anselin, L., 2016. Estimation and testing in the spatial seemingly
-unrelated regression (sur). Geoda Center for Geospatial Analysis;
-Computation, Arizona State University. Working Paper 2016-01.
-
-\leavevmode\hypertarget{ref-Araujo2020spread}{}%
-Araujo, M.B., Naimi, B., 2020. Spread of sars-cov-2 coronavirus likely
-to be constrained by climate. medRxiv.
-
-\leavevmode\hypertarget{ref-Arnold2019}{}%
-Arnold, J.B., 2019. Ggthemes: Extra themes, scales and geoms for
-'ggplot2'.
-
-\leavevmode\hypertarget{ref-Auguie2017gridextra}{}%
-Auguie, B., 2017. GridExtra: Miscellaneous functions for "grid"
-graphics.
-
-\leavevmode\hypertarget{ref-deangel2020weathering}{}%
-Ángel Solá, D.E. de, Wang, L., Vázquez, M., Lázaro, P.A.M., 2020.
-Weathering the pandemic: How the caribbean basin can use viral and
-environmental patterns to predict, prepare and respond to covid‐19.
-Journal of Medical Virology.
-
-\leavevmode\hypertarget{ref-Bivand2013}{}%
-Bivand, R.S., Pebesma, E., Gomez-Rubio, V., 2013. Applied spatial data
-analysis with R, second edition. Springer, NY.
-
-\leavevmode\hypertarget{ref-Cao2010spatio}{}%
-Cao, Z., Zeng, D., Zheng, X., Wang, Q., Wang, F., Wang, J., Wang, X.,
-2010. Spatio-temporal evolution of beijing 2003 sars epidemic. Science
-China Earth Sciences 53, 1017--1028.
-doi:\href{https://doi.org/10.1007/s11430-010-0043-x}{10.1007/s11430-010-0043-x}
-
-\leavevmode\hypertarget{ref-Casanova2010effects}{}%
-Casanova, L.M., Jeon, S., Rutala, W.A., Weber, D.J., Sobsey, M.D., 2010.
-Effects of air temperature and relative humidity on coronavirus survival
-on surfaces. Appl. Environ. Microbiol. 76, 2712--2717.
-
-\leavevmode\hypertarget{ref-Chan2011effects}{}%
-Chan, K., Peiris, J., Lam, S., Poon, L., Yuen, K., Seto, W., 2011. The
-effects of temperature and relative humidity on the viability of the
-sars coronavirus. Advances in virology 2011.
-
-\leavevmode\hypertarget{ref-Cliff1998detecting}{}%
-Cliff, A., Haggett, P., Smallman-Raynor, M., 1998. Detecting
-space---time patterns in geocoded disease data. Cholera in london, 1854
-measles in the united states, 1962--95, in: Geomed'97. Springer, pp.
-13--42.
-
-\leavevmode\hypertarget{ref-Coelho2020exponential}{}%
-Coelho, M.T.P., Rodrigues, J.F.M., Medina, A.M., Scalco, P., Terribile,
-L.C., Vilela, B., Diniz-Filho, J.A.F., Dobrovolski, R., 2020.
-Exponential phase of covid19 expansion is not driven by climate at
-global scale. medRxiv.
-
-\leavevmode\hypertarget{ref-Desjardins2020rapid}{}%
-Desjardins, M., Hohl, A., Delmelle, E., 2020. Rapid surveillance of
-covid-19 in the united states using a prospective space-time scan
-statistic: Detecting and evaluating emerging clusters. Applied Geography
-102202.
-
-\leavevmode\hypertarget{ref-Fernandes2020economic}{}%
-Fernandes, N., 2020. Economic effects of coronavirus outbreak (covid-19)
-on the world economy. Available at SSRN 3557504.
-
-\leavevmode\hypertarget{ref-Gong2020balance}{}%
-Gong, B., Zhang, S., Yuan, L., Chen, K.Z., 2020. A balance act:
-Minimizing economic loss while controlling novel coronavirus pneumonia.
-Journal of Chinese Governance 1--20.
-
-\leavevmode\hypertarget{ref-Greene2003econometric}{}%
-Greene, W.H., 2003. Econometric analysis. Pearson Education India.
-
-\leavevmode\hypertarget{ref-Grolemund2011dates}{}%
-Grolemund, G., Wickham, H., 2011. Dates and times made easy with
-lubridate. Journal of Statistical Software 40, 1--25.
-
-\leavevmode\hypertarget{ref-Hallet2002regional}{}%
-Hallet, M., 2002. Regional specialisation and concentration in the eu,
-in: Regional Convergence in the European Union. Springer, pp. 53--76.
-
-\leavevmode\hypertarget{ref-Harbert2020spatial}{}%
-Harbert, R.S., Cunningham, S.W., Tessler, M., 2020. Spatial modeling
-cannot currently differentiate sars-cov-2 coronavirus and human
-distributions on the basis of climate in the united states. medRxiv.
-
-\leavevmode\hypertarget{ref-Hu2013scaling}{}%
-Hu, H., Nigmatulina, K., Eckhoff, P., 2013. The scaling of contact rates
-with population density for the infectious disease models. Mathematical
-Biosciences 244, 125--134.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.mbs.2013.04.013}{https://doi.org/10.1016/j.mbs.2013.04.013}
-
-\leavevmode\hypertarget{ref-Jaakkola2014decline}{}%
-Jaakkola, K., Saukkoriipi, A., Jokelainen, J., Juvonen, R., Kauppila,
-J., Vainio, O., Ziegler, T., Rönkkö, E., Jaakkola, J.J., Ikäheimo, T.M.,
-2014. Decline in temperature and humidity increases the occurrence of
-influenza in cold climate. Environmental Health 13, 22.
-
-\leavevmode\hypertarget{ref-Jones2008global}{}%
-Jones, K.E., Patel, N.G., Levy, M.A., Storeygard, A., Balk, D.,
-Gittleman, J.L., Daszak, P., 2008. Global trends in emerging infectious
-diseases. Nature 451, 990--993.
-doi:\href{https://doi.org/10.1038/nature06536}{10.1038/nature06536}
-
-\leavevmode\hypertarget{ref-Kissler2020projecting}{}%
-Kissler, S.M., Tedijanto, C., Goldstein, E., Grad, Y.H., Lipsitch, M.,
-2020. Projecting the transmission dynamics of sars-cov-2 through the
-postpandemic period. Science eabb5793.
-doi:\href{https://doi.org/10.1126/science.abb5793}{10.1126/science.abb5793}
-
-\leavevmode\hypertarget{ref-Kudo2019low}{}%
-Kudo, E., Song, E., Yockey, L.J., Rakib, T., Wong, P.W., Homer, R.J.,
-Iwasaki, A., 2019. Low ambient humidity impairs barrier function and
-innate resistance against influenza infection. Proceedings of the
-National Academy of Sciences 116, 10905--10910.
-
-\leavevmode\hypertarget{ref-Lancastle2020impact}{}%
-Lancastle, N.M., 2020. Is the impact of social distancing on coronavirus
-growth rates effective across different settings? A non-parametric and
-local regression approach to test and compare the growth rate. medRxiv.
-
-\leavevmode\hypertarget{ref-Lauer2020incubation}{}%
-Lauer, S.A., Grantz, K.H., Bi, Q., Jones, F.K., Zheng, Q., Meredith,
-H.R., Azman, A.S., Reich, N.G., Lessler, J., 2020. The incubation period
-of coronavirus disease 2019 (covid-19) from publicly reported confirmed
-cases: Estimation and application. Annals of Internal Medicine.
-doi:\href{https://doi.org/10.7326/m20-0504}{10.7326/m20-0504}
-
-\leavevmode\hypertarget{ref-Lauridsen2010spatiotemporal}{}%
-Lauridsen, J., Bech, M., López, F., Maté, M., 2010. A spatiotemporal
-analysis of public pharmaceutical expenditure. The Annals of Regional
-Science 44, 299--314.
-
-\leavevmode\hypertarget{ref-Legallo2006evaluating}{}%
-Le Gallo, J., Dall'Erba, S., 2006. Evaluating the temporal and spatial
-heterogeneity of the european convergence process, 1980--1999. Journal
-of Regional Science 46, 269--288.
-
-\leavevmode\hypertarget{ref-Lopez2017spatial}{}%
-López, F.A., Martínez-Ortiz, P.J., Cegarra-Navarro, J.-G., 2017. Spatial
-spillovers in public expenditure on a municipal level in spain. The
-Annals of Regional Science 58, 39--65.
-
-\leavevmode\hypertarget{ref-Lopez2014spatial}{}%
-López, F.A., Mur, J., Angulo, A., 2014. Spatial model selection
-strategies in a sur framework. The case of regional productivity in eu.
-The Annals of Regional Science 53, 197--220.
-
-\leavevmode\hypertarget{ref-Luo2020how}{}%
-Luo, S., Tsang, K.P., 2020. How much of china and world gdp has the
-coronavirus reduced? Available at SSRN 3543760.
-
-\leavevmode\hypertarget{ref-Makinen2009cold}{}%
-Mäkainen, T.M., Juvonen, R., Jokelainen, J., Harju, T.H., Peitso, A.,
-Bloigu, A., Silvennoinen-Kassinen, S., Leinonen, M., Hassi, J., 2009.
-Cold temperature and low humidity are associated with increased
-occurrence of respiratory tract infections. Respiratory medicine 103,
-456--462.
-
-\leavevmode\hypertarget{ref-Meijers2008strategic}{}%
-Meijers, E., Hoekstra, J., Aguado, R., 2008. Strategic planning for city
-networks: The emergence of a basque global city? International Planning
-Studies 13, 239--259.
-doi:\href{https://doi.org/10.1080/13563470802521440}{10.1080/13563470802521440}
-
-\leavevmode\hypertarget{ref-Meng2005understanding}{}%
-Meng, B., Wang, J., Liu, J., Wu, J., Zhong, E., 2005. Understanding the
-spatial diffusion process of severe acute respiratory syndrome in
-beijing. Public Health 119, 1080--1087.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.puhe.2005.02.003}{https://doi.org/10.1016/j.puhe.2005.02.003}
-
-\leavevmode\hypertarget{ref-Millo2017robust}{}%
-Millo, G., 2017. Robust standard error estimators for panel models: A
-unifying approach. Journal of Statistical Software 82, 1--27.
-doi:\href{https://doi.org/10.18637/jss.v082.i03}{10.18637/jss.v082.i03}
-
-\leavevmode\hypertarget{ref-Minguez2019}{}%
-Mínguez, R., López, F., Mur, J., 2019. ML versus iv estimates of spatial
-sur models: Evidence from the case of airbnb in madrid urban area. The
-Annals of Regional Science 1--35.
-
-\leavevmode\hypertarget{ref-Morency2011distance}{}%
-Morency, C., Páez, A., Roorda, M.J., Mercado, R.G., Farber, S., 2011.
-Distance traveled in three canadian cities: Spatial analysis from the
-perspective of vulnerable population segments. Journal of Transport
-Geography 19, 39--50.
-
-\leavevmode\hypertarget{ref-National2020rapid}{}%
-National Academies of Sciences, Engineering and Medicine, 2020. Rapid
-expert consultation on sars-cov-2 survival in relation to temperature
-and humidity and potential for seasonality for the covid-19 pandemic
-(april 7, 2020). The National Academies Press, Washington, DC.
-doi:\href{https://doi.org/doi:10.17226/25771}{doi:10.17226/25771}
-
-\leavevmode\hypertarget{ref-Pebesma2018}{}%
-Pebesma, E., 2018. Simple Features for R: Standardized Support for
-Spatial Vector Data. The R Journal 10, 439--446.
-doi:\href{https://doi.org/10.32614/RJ-2018-009}{10.32614/RJ-2018-009}
-
-\leavevmode\hypertarget{ref-Pebesma2016}{}%
-Pebesma, E., Mailund, T., Hiebert, J., 2016. Measurement units in R. R
-Journal 8, 486--494.
-doi:\href{https://doi.org/10.32614/RJ-2016-061}{10.32614/RJ-2016-061}
-
-\leavevmode\hypertarget{ref-Perez2009agent}{}%
-Perez, L., Dragicevic, S., 2009. An agent-based approach for modeling
-dynamics of contagious disease spread. International journal of health
-geographics 8, 50.
-
-\leavevmode\hypertarget{ref-Prem2017projecting}{}%
-Prem, K., Cook, A.R., Jit, M., 2017. Projecting social contact matrices
-in 152 countries using contact surveys and demographic data. PLoS
-computational biology 13, e1005697.
-
-\leavevmode\hypertarget{ref-Rey1999us}{}%
-Rey, S.J., Montouri, B.D., 1999. US regional income convergence: A
-spatial econometric perspective. Regional studies 33, 143--156.
-
-\leavevmode\hypertarget{ref-Roorda2010trip}{}%
-Roorda, M.J., Paez, A., Morency, C., Mercado, R., Farber, S., 2010. Trip
-generation of vulnerable populations in three canadian cities: A spatial
-ordered probit approach. Transportation 37, 525--548.
-doi:\href{https://doi.org/10.1007/s11116-010-9263-3}{10.1007/s11116-010-9263-3}
-
-\leavevmode\hypertarget{ref-Sikder2012immobility}{}%
-Sikder, S., Pinjari, A.R., 2012. Immobility levels and mobility
-preferences of the elderly in the united states evidence from 2009
-national household travel survey. Transportation Research Record
-137--147. doi:\href{https://doi.org/10.3141/2318-16}{10.3141/2318-16}
-
-\leavevmode\hypertarget{ref-Singh2020age}{}%
-Singh, R., Adhikari, R., 2020. Age-structured impact of social
-distancing on the covid-19 epidemic in india. arXiv preprint
-arXiv:2003.12055.
-
-\leavevmode\hypertarget{ref-Novel2020epidemiological}{}%
-The Novel Coronavirus Pneumonia Emergency Response Epidemiology Team,
-2020. The epidemiological characteristics of an outbreak of 2019 novel
-coronavirus diseases (covid-19)---china, 2020. China CDC Weekly 2,
-113--122.
-
-\leavevmode\hypertarget{ref-vanWagner2008toward}{}%
-Van Wagner, E., 2008. Toward a dialectical understanding of networked
-disease in the global city: Vulnerability, connectivity, topologies.
-Networked disease: Emerging infections in the global city 13--26.
-
-\leavevmode\hypertarget{ref-Wang2010gis}{}%
-Wang, J., Xiong, J., Yang, K., Peng, S., Xu, Q., 2010. Use of gis and
-agent-based modeling to simulate the spread of influenza, in: 2010 18th
-International Conference on Geoinformatics. IEEE, pp. 1--6.
-
-\leavevmode\hypertarget{ref-Wickham2019}{}%
-Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L.D., François,
-R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen,
-T.L., Miller, E., Bache, S.M., Müller, K., Ooms, J., Robinson, D.,
-Seidel, D.P., Spinu, V., Takahashi, K., Vaughan, D., Wilke, C., Woo, K.,
-Yutani, H., 2019. Welcome to the tidyverse. Journal of Open Source
-Software 4, 1686.
-doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}
-
-\leavevmode\hypertarget{ref-Xie2014}{}%
-Xie, Y., 2014. Knitr: A comprehensive tool for reproducible research in
-R, in: Stodden, V., Leisch, F., Peng, R.D. (Eds.), Implementing
-Reproducible Computational Research. Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-Xie2015}{}%
-Xie, Y., 2015. Dynamic documents with R and knitr, 2nd ed. Chapman;
-Hall/CRC, Boca Raton, Florida.
-
-\leavevmode\hypertarget{ref-Yao2020association}{}%
-Yao, Y., Pan, J., Liu, Z., Meng, X., Wang, W., Kan, H., Wang, W., 2020.
-No association of covid-19 transmission with temperature or uv radiation
-in chinese cities. European Respiratory Journal 2000517.
-doi:\href{https://doi.org/10.1183/13993003.00517-2020}{10.1183/13993003.00517-2020}
-
-\leavevmode\hypertarget{ref-Zellner1962efficient}{}%
-Zellner, A., 1962. An efficient method of estimating seemingly unrelated
-regressions and tests for aggregation bias. Journal of the American
-statistical Association 57, 348--368.
-
-\leavevmode\hypertarget{ref-Zhou2016accelerated}{}%
-Zhou, Y.R., Coleman, W.D., 2016. Accelerated contagion and response:
-Understanding the relationships among globalization, time, and disease.
-Globalizations 13, 285--299.
-
-\leavevmode\hypertarget{ref-Zhu2019}{}%
-Zhu, H., 2019. KableExtra: Construct complex table with 'kable' and pipe
-syntax.
-
 \end{document}

### 2020-04-20 README.Rmd
Revised abstract before submission

@@
 
-  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress.
-While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees.
-How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine.
-In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain.
-Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems.
-An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process.
-Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations.
+Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health systems and the economy.
+Public health interventions have slowed the spread of the virus, albeit at a high cost.
+How and when to ease restrictions to movement hinges in part on whether SARS-CoV2 will display seasonality due to variations in temperature, humidity, and hours of sunshine.
+Here, we address this by means of a spatio-temporal analysis of the incidence of COVID-19 in Spain.
+Use of spatial Seemingly Unrelated Regressions (SUR) allows us to model the incidence of reported cases of the disease per 100,000 population as an interregional contagion process, in addition to a function of temperature, humidity, and sunshine.
+In the analysis we also control for GDP per capita, percentage of older adults in the population, population density, and presence of mass transit systems.
+The results support the hypothesis that incidence of the disease is lower at higher temperatures and higher levels of humidity.
 Sunshine, in contrast, displays a positive association with incidence of the disease.

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Made some revisions based on Fernando's comments about the lag between the lockdown and its effect.

@@
 This is an interesting result: immediately prior to the declaration of the state of emergency, there appears to have been a strong inter-provincial contagion effect.
-This situation persisted for a few days after the lockdown, and became less strong over time.
-At the end of the period under examination, the strength of this effect was much diminished.
+Keeping in mind that the incubation period ranges between 2 and 11 days with a median of 5, it is reasonable to expect that the effect of the lockdown will be observed with some delay.
+Indeed, as seen in the figure, the autocorrelation coefficient remains high for three days, then declines on March 16, and continues to weaken over time.
+At the end of the period under examination, the strength of this effect is much diminished.
 We would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero.

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Cover-Letter.Rmd
Prepared cover letter

---
title: "Cover-Letter"
author: "Author"
date: "4/20/2020"
output: pdf_document
---

Dear Prof. Franklin,

We are pleased to submit the manuscript enclosed for publication consideration in Geographical Analysis. The title of the paper is *A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain*. In this research, we analyze the incidence of COVID-19 (in reported cases by 100,000 people) at the level of the provinces in Spain. Spain, as you are surely aware, is one the countries hardest hit by the pandemic.

Our analysis covers a 30-day period that includes the totality of the lockdown to the date. In terms of methods, we adopt a spatial Seemingly Unrelated Regressions (SUR) approach. The inclusion of a lagged dependent variable allows us to model the incidence of COVID-19 as an interregional contagion effect. The focus of the paper is on some key environmental variables: temperature, humidity, and sunshine. These variables are the subject of considerable debate at the moment, because the potential seasonality of the pandemic will likely influence the decisions of many governments whether and when to ease restrictions to social distancing measures. In addition, our control variables produce some interesting insights as well. We find evidence of behavioral adaptations at the population level, regarding risk compensation and compliance with lockdown measures.

In summary, the results presented here are novel, and can potentially have an impact both in terms of informing the science around the novel coronavirus, as well as policy towards the pandemic.

Hoping that the paper will meet your approval, and that of your expert reviewers, we remain sincerely yours.

The authors

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd

@@
 
-Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$.
+Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of independent varaibles on $y_t$.
 This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
@@
 \begin{equation}
-E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
+E[\epsilon \epsilon']=\bf{\Sigma} \otimes \bf{I_N} ; \ \bf{\Sigma}=(\sigma_{tt'})
 \end{equation}

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Revised README.md after submission

@@
 
-<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/ -->
   <!--
@@
 
-Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of independent varaibles on $y_t$.
-This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
+Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of the independent variables on $y_t$.
+This flexibility can be reduced and it is posible to impose restrictions considering some $\beta$ coefficients as being constant over time.
 In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations.
@@
 
-Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
-This flexibility can be reduced and it is posible to impose restrictions to consider some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
+Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of the independent variables on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24).
+This flexibility can be reduced and it is posible to impose restrictions considering some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time.
 In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations.
@@
 <!--
-![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
+![\begin{equation} y = X \beta + \epsilon ![E\[\epsilon \epsilon'\]=\bf{\Sigma} \otimes \bf{I_N} ; \ \bf{\Sigma}=(\sigma_{tt'})](https://render.githubusercontent.com/render/math?math=E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5Cbf%7B%5CSigma%7D%20%5Cotimes%20%5Cbf%7BI_N%7D%20%3B%20%5C%20%5Cbf%7B%5CSigma%7D%3D(%5Csigma_%7Btt'%7D))
 

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Revised README.md after submission

@@
 in the coefficients \(\beta_{it}\) in order to modulate the effect of
-\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
-posible to impose restrinctions considering some \(\beta\) coefficients
-as being constant over time. In this case, we can reformulate the
-coefficients expression
+independent varaibles on \(y_t\). This flexibility can be reduced and it
+is posible to impose restrinctions considering some \(\beta\)
+coefficients as being constant over time. In this case, we can
+reformulate the coefficients expression
 \(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
@@
 declaration of the state of emergency, there appears to have been a
-strong inter-provincial contagion effect. This situation persisted for a
-few days after the lockdown, and became less strong over time. At the
-end of the period under examination, the strength of this effect was
-much diminished. We would expect that under full compliance with strict
-lockdown conditions (meaning no inter-provincial mobility) the spatial
-contagion effect would be zero. The fact that it is not is evidence of
-some weak contagion effect due to restrictions to the movement between
-provinces.
+strong inter-provincial contagion effect. Keeping in mind that the
+incubation period ranges between 2 and 11 days with a median of 5, it is
+reasonable to expect that the effect of the lockdown will be observed
+with some delay. Indeed, as seen in the figure, the autocorrelation
+coefficient remains high for three days, then declines on March 16, and
+continues to weaken over time. At the end of the period under
+examination, the strength of this effect is much diminished. We would
+expect that under full compliance with strict lockdown conditions
+(meaning no inter-provincial mobility) the spatial contagion effect
+would be zero. The fact that it is not is evidence of some weak
+contagion effect due to restrictions to the movement between provinces.
 

### 2020-04-20 README.Rmd
Revised README.md after submission

@@
 ```{r load-packages, include=FALSE}
-# The data package needs to be installed from github repository:
-# devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
-
-# This version of spsur may be needed:
-# Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install
+# Download and install these two packages:
+# covid19env
+# spsur
+
 library(covid19env)
@@
 Please note that this paper is prepared as a reproducible research document.
-The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following repository:
-
-https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
-
-Context and Data
+The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following anonymous Drive folder:
+
+https://drive.google.com/open?id=1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_
+
+Context and Data {#context-and-data}
 ============
 
-## Covid-19 in Spain
+## Covid-19 in Spain {#covid-19-in-spain}
 
@@
 
-<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
+<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/ -->
 ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
@@
 
-Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$.
-This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time.
+Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of the independent variables on $y_t$.
+This flexibility can be reduced and it is posible to impose restrictions considering some $\beta$ coefficients as being constant over time.
 In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations.
@@
 
-![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
+![\begin{equation} y = X \beta + \epsilon ![E\[\epsilon \epsilon'\]=\bf{\Sigma} \otimes \bf{I_N} ; \ \bf{\Sigma}=(\sigma_{tt'})](https://render.githubusercontent.com/render/math?math=E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5Cbf%7B%5CSigma%7D%20%5Cotimes%20%5Cbf%7BI_N%7D%20%3B%20%5C%20%5Cbf%7B%5CSigma%7D%3D(%5Csigma_%7Btt'%7D))
 
@@
 
-Results and Discussion
+Results and Discussion {#results-and-discussion}
 ===================
 
-Table \ref{tab:summary-best-model} presents a summary.
+Table \ref{tab:summary-best-model} presents a summary of the results of Model 2 (*lag11*).
+Recall that two coefficients were constrained and are estimated only for the first equation of the system, and thus are assumed to be constant over time.
+These are the coefficients corresponding to GDP per capita and percentage of older adults.
+Both coefficients are significant at $p\leq0.05$.
+The sign of the coefficient for GDP per capita is positive, which indicates that wealthier regions tend to have a higher incidence of COVID-19.
+This is in line with the idea that the epidemic started earlier in wealthier places due to their connections to a globalized world.
+The sign of the coefficient for percentage of older adults, on the other hand, is negative.
+As previously discussed, the level of social contact of older adults even under normal circumstances tends to be lower than for younger people.
+As a consequence, places with larger populations of older adults appear to have a natural level of social distancing in place.
+It is important to note that this does not detract from evidence that older adults are more vulnerable individually and in institutional settings, where their case mortality rates are perhaps the highest of all age groups.
+Instead, this result indicates that their presence in the community at large tends to depress transmission of the virus.
+
+Of the two other control variables, the coefficient of population density is significant at $p\leq0.05$ in 11 equations, and not significant in 19.
+The coefficient for transit is significant at $p\leq0.10$ in 22 equations, and of those, significant at $p\leq0.05$ in 14 equations.
+The next four variables are environmental factors.
+The coefficient for humidity is significant at $p\leq0.10$ in 20 equations, and of those, significant at $p\leq0.05$ in 19 equations.
+Of the environmental variables, temperature is the only variable that has significant coefficients in every equation at $p\leq0.05$.
+Finally, sunshine has significant coefficients at $p\leq0.10$ in 18 equations, including 15 coefficients that are significant at $p\leq0.05$.
 
@@
 
-Figure \ref{fig:results-time-1} shows the temporal evolution of the spatial autocorrelation coefficient ($\rho$) and the intercept of the model.
+To better understand the results, we proceed to plot the coefficients in their temporal sequence.
+At this point it is worth recalling that the state of emergency went into effect on March 14.
+In the following figures, the periods of time indicated in yellow starting on March 14 correspond to the state of emergency, with only essential travel and selected industrial activities allowed; the period of time in orange was the stricter lockdown when only essential travel was allowed.
+
+We begin our discussion with the evolution of the spatial autocorrelation coefficient ($\rho$) in Figure \ref{fig:results-time-1} (left panel).
+It is interesting to note that the magnitude of the spatial autocorrelation coefficient $\rho$ declines over the period under analysis, and is not significant for some days.
+This is an interesting result: immediately prior to the declaration of the state of emergency, there appears to have been a strong inter-provincial contagion effect.
+Keeping in mind that the incubation period ranges between 2 and 11 days with a median of 5, it is reasonable to expect that the effect of the lockdown will be observed with some delay.
+Indeed, as seen in the figure, the autocorrelation coefficient remains high for three days, then declines on March 16, and continues to weaken over time.
+At the end of the period under examination, the strength of this effect is much diminished.
+We would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero.
+The fact that it is not is evidence of some weak contagion effect due to restrictions to the movement between provinces.
+
+The intercept (right panel in Figure \ref{fig:results-time-1}) is indicative of the variation of the incidence of COVID-19, other things being equal.
+Here we see that at the incidence declines somewhat immediately after the state of emergency, only to begin increasing again over time.
+Then, the incidence declines again after the stricter lockdown and rebounds to a lower level by April 11.
 
@@
 
-Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ and $Transit$.
+Figure \ref{fig:beta-controls-time} shows the temporal evolution of the coefficients for the two control variables that were not fixed over time, i.e., $\log(Density)$ (left panel) and $Transit$ (right panel).
+
+In Section \ref{variables} we had anticipated a positive sign for the coefficient of this variable, and indeed, at the beginning of the period the coefficient is positive, albeit not significant, and then remains non-significant for the earlier part of the lockdown.
+We are somewhat surprised by the way this coefficient turns significant _and_ negative in the later part of the lockdown, after April 1st.
+This effect, we surmise, is the result of risk compensation, a situation where people adapt their behavior according to the *perceived* level of risk, becoming more careful when the perceived risk is higher and viceversa [e.g., @Noland1995perceived; @Richens2000risk; @Phillips2011risk].
+Consequently, residents in high density regions may perceive the risk of infection as being higher, and adapt their behavior accordingly - while the opposite may be true of residents in low density regions.
+Anecdotally, this seems to have been the case with some governors in the US, who resisted stay-at-home orders citing their sparse \footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions} and rural populations\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}.
+The coefficient for Transit is positive, as expected, but with very wide confidence intervals, and in fact not significant in the earlier part of the period.
 
@@
 
-Figure \ref{fig:beta-environmental-time} shows the temporal evolution of the coefficient for the three environmental variables.
-
-```{r beta-environmenta-time, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:beta-environmental-time}Temporal evolution of coefficient for the environmental variables; dots are the point estimates and vertical lines are 95% confidence intervals.
+The evolution of the coefficients for the three environmental variables are shown in Figure \ref{fig:beta-environmental-time}.
+Despite a mostly positive simple correlation with incidence (see Figure \ref{fig:daily-correlations}) once that we control for other factors, humidity has a negative association with incidence of COVID-19 in Spain (top-right panel).
+This is in line with the literature that describes the viability and transmission of different viruses at higher levels of humidity.
+The coefficients for temperature (top-right panel) are consistently negative and this variable is, besides the intercept, the only one with significant coefficients in all equations.
+The range of variation of this coefficient during the period examined is approximately between -1 and -2 - although it is important to recall that these values should not be interpreted directly as effects; more on this below.
+Finally the plot for the coefficients associated with hours of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the coefficient was negative, but significant only at $p\leq10$.
+At the beginning of the lockdown, the coefficient is no longer significative, before becoming *positive* and significant on March 23rd.
+This result stands in contrast to previous findings regarding influenza, where more hours of sunlight reduced the strength and duration of epidemic durations [@Yu2013characterization].
+A difference with previous studies is the temporal scale of the analysis: where Yu et al.
+[-@Yu2013characterization] use monthly averages, we use daily data for a much shorter period of time.
+The positive sign of sunshine may well be another instance of behavioral adaptations, whereby compliance with lockdown orders weakens on sunny days.
+
+```{r beta-environmental-time, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:beta-environmental-time}Temporal evolution of coefficient for the environmental variables; dots are the point estimates and vertical lines are 95% confidence intervals.
 In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
@@
 
-Concluding Remarks
+```{r calculate-impacts-best-model, echo=FALSE, cache = TRUE}
+# Calculate impacts
+Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
+trsur <- spatialreg::trW(Wsur, type = "MC")
+best_model.impacts <- impactspsur(best_model, tr = trsur,R = 1000)
+
+# Organize in dataframe
+best_model.impacts <- data.frame(Impact = unlist(best_model.impacts)) %>%
+  # Identify point estimates of effects for each variable and simulated values 
+  mutate(Variable = c(c(rep(c("log(GDPpc)", 
+                              "log(Older)",
+                              "log(Density)",
+                              "Transit",
+                              "log(Humidity)",
+                              "log(Temperature)",
+                              "log(Sunshine)"),
+                            3),
+                        rep("junk", 21000)),
+                      rep(c(rep(c("log(Density)",
+                                  "Transit",
+                                  "log(Humidity)",
+                                  "log(Temperature)",
+                                  "log(Sunshine)"),
+                                3),
+                            rep("junk", 15000)), # These are the simulated values
+                          29))) %>%
+  filter(Variable != "junk") %>% # Remove all simulations, keep only point estimates
+  # Add type of effect and date
+  mutate(Type = c(rep(c("Direct", 
+                        "Indirect", 
+                        "Total"),
+                      each = 7),
+                  rep(rep(c("Direct", 
+                            "Indirect", 
+                            "Total"),
+                          each = 5),
+                      29)),
+         Date = c(rep(ymd("2020-03-13"),
+                      21),
+                  rep(seq(ymd("2020-03-14"),
+                          ymd("2020-04-11"), 
+                          by = "days"), 
+                      each = 15))) %>%
+  # Convert to factors
+  mutate(Variable = factor(Variable,
+                           levels = c("log(GDPpc)", 
+                              "log(Older)",
+                              "log(Density)",
+                              "Transit",
+                              "log(Humidity)",
+                              "log(Temperature)",
+                              "log(Sunshine)")),
+         Type = factor(Type,
+                       levels = c("Direct", 
+                            "Indirect", 
+                            "Total")))
+```
+
+The preceding discussion helps to establish the inferential contributions of the analysis.
+The remaining question is, what are the implications.
+As discussed in Section \ref{methods} the question of the magnitude of the effects is not clear from the coefficients alone, since a change to a variable in a province influences, via the contagion effect, its neighbors.
+For this reason, the appropriate way to estimate the effects is to calculate both the own and contagion effects (in other words the direct and indirect effects, respectively), as well as their total.
+A summary of the different variable effects appears in Table \ref{tab:summary-impacts-best-model}.
+All effects in the table are interpreted as percentage change in the incidence of COVID-19 as a consequence of a one percent change in the variable.
+The exception is Transit (which was not log-transformed), and this variable represents the percentage change in incidence between provinces without and with mass transit systems.
+
+Two variables had temporally constrained coefficients.
+The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.675% for each percentual increase of this variable (in €1,000s).
+In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemia.
+Percentage of older adults has a negative effect, and each percentual increase in this variable is associated with a relatively small reduction of the incidence of less than 1%.
+
+The temporal variation of the effects for the rest of the variables is shown in Figure \ref{fig:plot-impacts}.
+The largest positive direct effect is Transit, and the largest direct negative effects are temperature and humidity.
+The direct effect of these variables is as follows: for each percentual point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
+This effect is amplified via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 1.5%.
+The effect of humidity is also to reduce the incidence: each percentual point of increase in humidity is associated with a reduction of up to 1.5% in incidence.
+With the addition of the indirect effect, the total effect of a 1% increase in humidity is to reduce incidence by up to 3%.
+As seen in the figure, the indirect (i.e., contagion) effects are almost as strong as the direct effects immediately before and at the beginning of the lockdown period.
+However, by the end of the lockdown period, the indirect effects have weakened considerably.
+The effect of density goes from positive (+0.334) to negative (-0.244).
+It is crucial to keep in mind here that this is the effect of density over a period of *lockdown* and possibly subject to risk compensation, as discussed above.
+In other words, the lockdown seems to be more effective in places with higher density.
+Accordingly, cities should not rush to ease lockdowns, but rather efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
+
+What do these effects mean?
+Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the effect of temperature is estimated to reduce the incidence of COVID-19 by between one and two percent with each percentual increase of this variable.
+Without a lockdown, or a relaxed lockdown, the effect is even greater, but only because the disease is transmitted inter-regionally!
+Imagine a region that is closed to its neighbors: increases in temperature are associated with smaller reductions in incidence - however, it does not import or export cases.
+A province that is not closed and not under lockdown will see a greater effect of temperature, but at the cost of importing/exporting cases.
+
+```{r summary-impacts-best-model, echo=FALSE}
+rbind(best_model.impacts %>%
+  filter(Type == "Direct") %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Impact), 
+            Mean = mean(Impact), 
+            Max = max(Impact)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained")),
+  best_model.impacts %>%
+  filter(Type == "Indirect") %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Impact), 
+            Mean = mean(Impact), 
+            Max = max(Impact)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained")),
+  best_model.impacts %>%
+  filter(Type == "Total") %>%
+  group_by(Variable) %>%
+  summarize(Min = min(Impact), 
+            Mean = mean(Impact), 
+            Max = max(Impact)) %>% # Number of coefficients with pval <=5 
+  mutate(Note = c("Constrained", 
+                  "Constrained", 
+                  "Non-constrained", 
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained",
+                  "Non-constrained"))) %>%
+  kable(#"latex",
+        "html",
+        booktabs = TRUE,
+        digits = 3,
+        col.names = c("Variable", "Min", "Mean", "Max", "Note"),
+        caption = "\\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)") %>%
+  #kable_styling(latex_options = c("striped")) %>%
+  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
+  pack_rows("Direct Effects", 1, 7) %>%
+  pack_rows("Indirect Effects", 8, 14) %>%
+  pack_rows("Total Effects", 15, 21) %>%
+  footnote(general = c("Non-constrained: coefficient was allowed to vary across equations",
+                       "Constrained: coefficient as constant across equations"))
+```
+
+```{r plot-impacts, echo=FALSE, fig.cap="\\label{fig:plot-impacts}Temporal evolution of direct, indirect, and total effects by date."}
+best_model.impacts %>%
+  filter(Variable != "log(GDPpc)", 
+         Variable != "log(Older)") %>%
+  ggplot(aes(x = Date, 
+             y = Impact, 
+             color = Variable)) +
+  geom_line(size = 0.8) +
+  scale_color_brewer(palette = "Set1") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90),
+        legend.position = "bottom") +
+  facet_wrap(~ Type, ncol = 3)
+
+```
+
+Concluding Remarks {#conclusion}
 ===================
 
-More words go here.
-
-Limitaciones
-
-* La temperatura puedes ser un factor pero cabría esperar que no tuviera un impacto lineal.
-Por el contrario deberiamos esperar un 'punto de corte': Una temperatua mantenida superior a X grados durante siete días sea la mejor forma de incorporarla al modelo.
-En nuestro modelo (log-log) en incremento en un 1% de la temperatura se asocia con un incremento beta% en la incidencia.
-Esto es lo mismo si la temperatura es baja que si es alta.
-
-* Idem para la humedad y horas de sol
-
-* Los datos son 'provisionales'.
-Hay gran confusión sobre la incidencia real.
-La ausencia de test de diagnóstico PCRs al inicio de la pandemia (también ahora) puede desvirtuar el número de casos dianosticados.
-
-* Los datos oficiales (que tampoco son fiables) son reportados a nivel de Comunidades Autónomas.
-La recopilación de datos a nivel de provincial son el resultado de un esfuerzo colaborativo de recopilacion entre distintas fuentes (principalmente gobiernos locales).
-Nuevamente puede haber sesgos importanes.
-
-* En carácter insular de las Islas Canarias y de las Islas baleares no se ha tenido en cuenta.
-
-* Baleares se ha linkado artificialmente con 3/4 provincias y debería de haberse dejado aislada para respetar su carácter insular al definir la matriz W.
-
-* La incidencia depende de un estado inicial.
-Al inicio del estudio había provincias en las que la epidemia estaba muy desarrollada (Madrid/Alava) mientras que en otras apenas habías casos.
-Este hecho no ha sido considerado en el modelo.
-Decretar el confinamiento debe tener distintos impactos entre provincias.
-QUIZAR METER UNA VARAIBLE DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
-
-* No se ha controlado por el sistema sanitario de cada provincia.
-Uno de los principales focos de contagio han sido los hospitales y los centros de salud.
-En aquellas provincias donde se ha promovido el mensaje "NO IR AL MEDICO" han presentado menor incidencia.
-
-* ¿hasta que punto las variables de control no recojen también factores climáticos? por ejemplo, la gente jóven vive en el sur de España que ha tenido menos incidencia.
-
-* Las estaciones meteorológicas para la obtencion de datos climáticos has sido elegidas aleatoriamente (una para cada provincia).
-otra seleccion puede dar otros resultados.
-AQUI SE PODRÍA HACER EL ESFUERZO DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR PROVINCIA
-
-*IDEM para la humedad. idem para sunshine
-
-* HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA DARLE RELEVANCIA A LA ESTIMACION SUR
+In this paper we have presented a spatio-temporal analysis of incidence of COVID-19 in Spain.
+The analysis covers a 30-day period that begins immediately before the state of emergency was declared in the country.
+The focus of the research has been on the environmental correlates of incidence of the disease.
+There is a need for more empirical evidence, as policy makers, public health practitioners, and the public begin planning for the months ahead at this early stage of the pandemic.
+
+Our results offer strong support for the hypothesis that incidence of COVID-19 at the population level is lower at higher temperatures and levels of humidity: the estiamated effect is a reduction in the neighborhood of 3% percent in incidence per each 1% increase in temperature, and a 3% reduction in incidence per 1% increase in humidity *under conditions of contagion*.
+These reductions are estimated to be lower when lockdown conditions are in place.
+The question here seems to be whether these environmental variables can yield a bigger reduction of *more* cases, or a smaller reduction of *fewer* cases.
+Our control variables also offer some interesting insights.
+In particular, there is evidence of behavioral adaptations in the form of risk compensation (density) and compliance with the lockdown (sunshine).
+These results offer a cautionary tale, with regards to the effectiveness of the lockdown in more dense areas, and also the implications for compliance with stay-at-home orders as the northern hemisphere moves towards summer and more hours of sunshine during the day.
+
+A key aspect of the analysis using spatial SUR models is that we were able to model incidence of COVID-19 as an interregional contagion process.
+Here, we find that the strength of the contagion effect was dramatically reduced by the lockdown.
+
+Needless to say, the analysis presented here is ecological, and at the level of population health.
+For this reason, the analysis does not make any claims with respect to the effect of ultraviolet light on the virus, but rather about transmission of the virus in the population.
+In this way, the analysis does not say that the virus moves less effectively where more people live in close proximity to each other, but rather that humans are more contagious when they feel safe in less dense regions.
+Similarly, more sunshine does not mean that the virus thrives, but rather that humans are more contagious to each other when their behavior adapts to this environmental condition.
+
+It is important, in this respect, to highlight some limitations of this study.
+
+First, the analysis was conducted under a situation of lockdown, and therefore, besides the results are the beginning of the period examined, one must exercise caution when trying to extrapolate the findings to other situations.
+Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
+However, we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results - even if we are aware that the number of true cases is likely higher.
+Thirdly, all environmental data are based on a single station in a province.
+Here, we made an effort to select the most representative station based on the distribution of the population.
+An alternative approach would be to calculate the environmental variables based on the average of multiple stations within each province: this would likely smooth to some extent these variables.
+Finally, we defined spatial neighborhoods based on adjacency.
+It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services.
+We flag this as a matter for future research.
 

### 2020-04-20 README.Rmd
Fixed the link to the repository in README.md

@@
 Please note that this paper is prepared as a reproducible research document.
-The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following anonymous Drive folder:
-
-https://drive.google.com/open?id=1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_
+The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following link (this repository):
+
+https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain
 

### 2020-04-25 Visualization-Contagion/Visualizing-Contagion.Rmd
Created a new folder Visualization-Contagion with some preliminary experiments in visualizing the contagion term of the SUR models.

---
title: "Visualize contagion"
output:
  html_document:
    df_print: paged
---

In this notebook I'd like to visualize the pattern of inter-provincial contagion. My idea is to get the fitted values of the model minus the effect of the covariates, to isolate the lagged dependent variable, and then do an animation with that.

## Preliminaries

Load packages:
```{r load-packages, message=FALSE}
# Download and install these two packages:
# covid19env
# spsur
library(covid19env)
library(gganimate)
#library(ggthemes)
#library(gridExtra)
#library(kableExtra)
library(lubridate)
library(plm)
#library(plotly)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
```

Load datasets:
```{r load-data, cache=TRUE}
# Load data from package `covid19env`
data("covid19_spain")
data("provinces_spain")
```

Preprocess the data:
```{r data-preparation, cache=TRUE}
# Convert GDP per capita to thousands of euros:
provinces_spain <- provinces_spain %>%
  mutate(GDPpc = GDPpc/1000)

# Join provincial data to incidence data and convert to simple features:
covid19_spain <- covid19_spain %>% 
  left_join(provinces_spain %>% st_drop_geometry(),
            by = c("province", "CCAA", "ID_INE"))
```

## The contagion effect

Consider an estimated model of the following form:
$$
(I -  \hat{\rho}W)\hat{Y} = X\hat{\beta}
$$

Expand:
$$
\hat{Y} - \hat{\rho}W\hat{Y} = X\hat{\beta}
$$

Rearrange:
$$
\hat{\rho}W\hat{Y} = \hat{Y} - X\hat{\beta}
$$

\noindent where $\hat{Y}$ is the prediction of incidence in the provinces, and $X\hat{\beta}$ is the estimated impact on incidence due to all $X$ factors. The right-hand side of the equation is the estimated incidence discounted by its own factors.

The term on the right hand side is the contribution to $\hat{Y}$ due to incidence in the neighboring provinces. In other words, this is the contagion effect.

When the estimated incidence $\hat{Y}$ is greater than $X\hat{\beta}$, this means that the province has a greater incidence of the disease than what its own conditions explain. In this case $\hat{\rho}W\hat{Y}$ is positive, and contagion contributes to increase the incidence (the province tends to be on the receiving end of contagion).

On the other hand, when the estimated incidence $\hat{Y}$ is less than $X\hat{\beta}$, the province has lower incidence than what its own conditions explain. Now, $\hat{\rho}W\hat{Y}$ is negative, which means that the incidence is reduced by lower rates of incidence in their neighborhood.

Finally, when the estimated incidence $\hat{Y}$ is equal to $X\hat{\beta}$, the incidence in the province is explained completely by its own conditions.

By mapping this term, we can explore how vulnerable to contagion, or contagious, the provinces were over time. Remember, positive means on the receiving end (or vulnerable) of contagion, red means contagious.

## Model

Organize data for SUR modelling: 
```{r data-preparation-for-modelling, cache = TRUE}
GPanel <- plm::pdata.frame(covid19_spain %>%
                             select(province, 
                                    Date,
                                    Incidence, 
                                    Median_Age,
                                    Male2Female,
                                    Older,
                                    GDPpc,
                                    Density,
                                    Transit,
                                    Mean_Temp_lag8,
                                    Humidity_lag8,
                                    Sunshine_Hours_lag8,
                                    Mean_Temp_lag11,
                                    Humidity_lag11,
                                    Sunshine_Hours_lag11,
                                    Mean_Temp_lag11w,
                                    Humidity_lag11w,
                                    Sunshine_Hours_lag11w),
                           c("province","Date"))

```

Create spatial weights matrix: 
```{r spatial-weights, cache = TRUE}

Wmat <- provinces_spain %>%
  #drop_na() %>%
  as("Spatial") %>%
  poly2nb(queen = FALSE) %>%
  nb2mat(zero.policy = TRUE)

Wmat <- (Wmat > 0) * 1

# Join the two provinces in Canarias
Wmat[which(provinces_spain$province == "Palmas(Las)"), 
     which(provinces_spain$province == "Santa Cruz de Tenerife")] <- 1
Wmat[which(provinces_spain$province == "Santa Cruz de Tenerife"), 
     which(provinces_spain$province == "Palmas(Las)")] <- 1

# 'Paises Catalans'
#n = 8
Wmat[which(provinces_spain$province == "Barcelona"), 
     which(provinces_spain$province == "Baleares")] <- 1
Wmat[which(provinces_spain$province == "Baleares"), 
     which(provinces_spain$province == "Barcelona")] <- 1
Wmat[which(provinces_spain$province == "Baleares"), 
     which(provinces_spain$province == "Castellon/Castello")] <- 1 
Wmat[which(provinces_spain$province == "Castellon/Castello"), 
     which(provinces_spain$province == "Baleares")] <- 1
Wmat[which(provinces_spain$province == "Baleares"), 
     which(provinces_spain$province == "Tarragona")] <- 1 
Wmat[which(provinces_spain$province == "Tarragona"), 
     which(provinces_spain$province == "Baleares")] <- 1
miW <- Wmat/rowSums(Wmat)

# Convert to listw
listw <- mat2listw(Wmat,style = "W")
```

Define formula for model:
```{r formulas, cache = TRUE}
formula_lag11 <- log(Incidence) ~ 
  log(GDPpc) +
  log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag11) +
  log(Mean_Temp_lag11) +
  log(Sunshine_Hours_lag11 + 0.1)
```

Add restrictions to model for temporally fixed coefficients:
```{r model-restrictions, cache = TRUE}
T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # Recall that T is the number of days, i.e., time periods, i.e., equations; add 1 to include the starting day
k <- 8 # Number of independent variables, including the constant
coef_rest <- 2 # Number of restrictions

# nrow is number of equations (time periods) minus 1, times the number of restrictions
# ncol is number of variables times number of equations
R2 <- matrix(0, nrow = (T - 1) * coef_rest, ncol = k * T)

for (i in 1:(T-1)){
  R2[i, 2] <- 1
  R2[i, (2 + i * k)] <- -1
  R2[(i + T - 1), 3] <- 1
  R2[(i + T - 1), (3 + i * k)] <- -1
  # Use if more restrictions are needed
  #R2[(i + T - 1) * 2, 4] <- 1
  #R2[(i + T - 1) * 2, (4 + i * k)] <- -1
}
b2 <- matrix(0, ncol = (T - 1) * coef_rest)
```

Estimate model:
```{r model2-11-day-moving-average, cache = TRUE}
# Model with 11-day moving average of climatic variables:
sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
                                  data=GPanel, 
                                  time = GPanel$Date, 
                                  type = "slm", 
                                  fit_method = "3sls", 
                                  listw=  listw,
                                  R = R2,
                                  b = b2)
#summary(sur.slm_lag11)
#print(paste("Pooled R^2 = ", sur.slm_lag11$R2[1]))
```

## Prepare results for analysis

To analyze the contagion effect we need to calculate it. This can be done in two different ways. Directly, by computing this term:

$$
\hat{\rho}W\hat{Y}
$$

The estimated autocorrelation coefficient $\hat{\rho}$ (by date) and the fitted values $\hat{Y}$ are part of the output of the estimation procedure.

As an alternative, the contagion effect can be calculated indirectly, as:
$$
\hat{Y} - X\hat{\beta}
$$

The fitted values $\hat{Y}$ and the coefficients $\hat{\beta}$ (by date) are part of the output of the estimation procedure. Here, I take the second approach, which avoid having to work with matrix $W$ in the calculations.

Extract and organize coefficients:
```{r}
Coefficients <- data.frame(Coefficients = sur.slm_lag11$coefficients,
                           Variable = c(c("Intercept",
                                        "b_GDPpc",
                                        "b_Older",
                                        "b_Density",
                                        "b_Transit",
                                        "b_Humidity",
                                        "b_Temperature",
                                        "b_Sunshine"),
                                        rep(c("Intercept",
                                              "b_Density",
                                              "b_Transit",
                                              "b_Humidity",
                                              "b_Temperature",
                                              "b_Sunshine"),
                                            29)),
                           Date = c(rep(ymd("2020-03-13"), 8),
                                    rep(seq(ymd("2020-03-14"),
                                        ymd("2020-04-11"),
                                        by = "days"),
                                        each = 6)))  %>%
  pivot_wider(names_from = "Variable", 
              values_from = "Coefficients")  %>%
  mutate(b_GDPpc = max(sur.slm_lag11$coefficients[2]),
         b_Older = max(sur.slm_lag11$coefficients[3]))
```

Join coefficients with data table:
```{r}
covid19_spain <- covid19_spain %>% left_join(Coefficients, by = "Date")
```

Calculate $\hat{Y}$ minus $X\hat{\beta}$:
```{r}
Contagion <- covid19_spain %>%
  mutate(fitted.values = sur.slm_lag11$fitted.values,
         X_beta = Intercept + 
           b_GDPpc * log(GDPpc) + 
           b_Older * log(Older) + 
           b_Density * log(as.numeric(Density)) +
           b_Transit * Transit +
           b_Humidity * log(Humidity_lag11) +
           b_Temperature * log(Mean_Temp_lag11) + 
           b_Sunshine * log(Sunshine_Hours_lag11 + 0.01), 
         Contagion = fitted.values - X_beta)
```

Select variables from `Contagion` dataframe:
```{r}
Contagion <- Contagion %>%
  select(ID_INE, province, CCAA, Contagion, Date) %>%
  mutate(Day = factor(rep(1:50, 30)))
```

Next, join geometry and convert to sf:
```{r}
Contagion <- Contagion %>%
  left_join(provinces_spain %>% 
              select(ID_INE, geometry),
            by = "ID_INE") %>%
  st_as_sf()
```

Create a choropleth plot of the provinces with the contagion effect ($\hat{\rho}W\hat{Y}$):
```{r}
p1 <- ggplot() + 
  geom_sf(data = Contagion %>% 
            filter(CCAA != "Canarias"), 
          aes(fill = Contagion, 
              group = interaction(factor(ID_INE), 
                                  Date))) +
  scale_fill_gradient2() + 
  theme_bw()
```

Create an animation using `gganimate`:
```{r}
anim1 <-  p1 +
  labs(title = "Date: {frame_time}") +
  transition_time(Date)
```

Render animation:
```{r cache=TRUE}
animate(anim1, fps = 2, end_pause = 5)
```

What general trends to we see?

Now plot the trajectory of the contagion term by province:
```{r}
ggplot(data = Contagion, aes(x = Date, y = Contagion, group = province)) +
  geom_line(color = "gray") +
  geom_line(data = Contagion %>% 
              filter(province == "Araba/alava"), 
            color = "orange",
            size = 1) +
  geom_line(data = Contagion %>% 
              filter(province == "Madrid"), 
            color = "purple",
            size = 1) +
  labs(title = "Contagion effect by province by date",
              #subtitle = "Plot by date",
              caption = "Alava is the orange line; Madrid is the purple line") +
  theme_bw()
```

Where could this analysis be taken? LISA analysis by date? Spatial Markov Chains? Cluster membership?

### 2020-05-04 Donwload_climatic_data_from_AEMET.Rmd
COVID-19 Municipios v0

---
title: "Download climatic data from AEMET"
author: "Fernando"
date: "4/23/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse = TRUE,message = FALSE)
```

```{r libraries}
library(meteoland)
library(aemet)
library(sf)
library(kableExtra)
library(ggplot2)
```

## First step: Get API

Get the API key from AEMET open data service (is free). https://opendata.aemet.es/centrodedescargas/inicio

## API key for AEMET
```{r key}
apikey <- "eyJhbGciOiJIUzI1NiJ9.eyJzdWIiOiJmOGw1aDlAZ21haWwuY29tIiwianRpIjoiNzQ4NzU1ZTgtZDBjYi00MzE3LThmMTUtNzcwZDE0ZGNiNDg0IiwiaXNzIjoiQUVNRVQiLCJpYXQiOjE1ODYxODI2MzIsInVzZXJJZCI6Ijc0ODc1NWU4LWQwY2ItNDMxNy04ZjE1LTc3MGQxNGRjYjQ4NCIsInJvbGUiOiIifQ.Ez8SgcBusQ-514k9vDtEXL29wbRi-cGIYkxnXl5AVv0"
```

## List of dates
Select the set of dates 
```{r}
mydates <- seq(as.Date("2020/3/01"), as.Date("2020/4/20"), "day")
```

## Donwload Meteo Stations

En  la AEMET se puede descargar shp de las estaciones meteorológicas. Están clasificadas en 4 tipos (automáticas, completas, termoétricas, plubiometricas). Aún no se la diferencia entre ellas.

Hasta donde yo se ninguna ofrece datos de humedad.

He utilizado el paquete **aemet** para descargarme información de las estaciones meteorológicas

Para el papel seleccione aquella que me pareció mas correcta y que tuviera datos (hay algunas con muchos datos perdidos)

```{r}
stations <- aemet::aemet_stations(apikey)
stations.sf <- st_as_sf(stations, coords = c("longitud", "latitud"), crs = 4326)
stations.sf$Lon <- stations$longitud
stations.sf$Lat <- stations$latitud
```

## Geometry of provinces

```{r}
load("/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/covid19env/data/provinces_spain.RData")
```

## Geometry of Meteo Stations

```{r plot-stations}
ggplot(data = stations.sf) +
  geom_sf(data = provinces_spain, aes()) +
  geom_sf(data = stations.sf, aes(color = provincia), show.legend = FALSE) +
  theme_bw(base_size=6)
```

## Domwload climatic data from each Meteo Station

En el shp "provinces_spain".shp se incluye el campo "Meteo_Station" especificando el ID de la estación seleccionada en cada provincia.

Hay que uncluir un lag temporal. Hay limitación de descargas por minuto.

```{r}
num_prov <- 5
Tem <- list()
for (i in 1:num_prov){
  Tem[[i]] <- downloadAEMEThistorical(apikey, mydates, station_id=provinces_spain$Meteo_Station[i])[[1]]
  print(i)
  Sys.sleep(2)
}
```

## Example: List information
En la lista Tem se obtienen los valores climático

```{r}
kableExtra::kable(head(Tem[[1]]))
```

### 2020-05-04 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.tex
COVID-19 Municipios v0

@@
   contrast, population density and percentage of older adults display
-  negative associations with incidence of COVID-19.
+  negative associations with incidence of COVID-19.\\
   \end{abstract}

### 2020-05-04 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
COVID-19 Municipios v0

@@
 waning of cases in warmer and more humid seasons, yet none are without
-major limitations\ldots Additional studies as the SARS-CoV-2 pandemic
+major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
 unfolds could shed more light on the effects of climate on
@@
 in the coefficients \(\beta_{it}\) in order to modulate the effect of
-independent varaibles on \(y_t\). This flexibility can be reduced and it
-is posible to impose restrinctions considering some \(\beta\)
+the independent variables on \(y_t\). This flexibility can be reduced
+and it is posible to impose restrictions considering some \(\beta\)
 coefficients as being constant over time. In this case, we can
@@
 \multicolumn{5}{l}{\textbf{Direct Effects}}\\
-\hspace{1em}log(GDPpc) & 0.675 & 0.675 & 0.675 & Constrained\\
-\hspace{1em}log(Older) & -0.802 & -0.802 & -0.802 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.167 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.674 & 0.674 & 0.674 & Constrained\\
+\hspace{1em}log(Older) & -0.801 & -0.801 & -0.801 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.166 & Non-constrained\\
 \hspace{1em}Transit & 0.342 & 0.517 & 0.584 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.562 & -0.544 & -0.031 & Non-constrained\\
-\hspace{1em}log(Temperature) & -2.019 & -1.421 & -0.929 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.559 & -0.544 & -0.031 & Non-constrained\\
+\hspace{1em}log(Temperature) & -2.019 & -1.420 & -0.929 & Non-constrained\\
 \rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.281 & 0.095 & 0.206 & Non-constrained\\
@@
 \multicolumn{5}{l}{\textbf{Indirect Effects}}\\
-\hspace{1em}log(GDPpc) & 0.638 & 0.638 & 0.638 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.758 & -0.758 & -0.758 & Constrained\\
-\hspace{1em}log(Density) & -0.032 & -0.004 & 0.157 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.341 & Non-constrained\\
-\hspace{1em}log(Humidity) & -1.475 & -0.157 & -0.001 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.644 & -0.274 & -0.014 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.639 & 0.639 & 0.639 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.759 & -0.759 & -0.759 & Constrained\\
+\hspace{1em}log(Density) & -0.032 & -0.004 & 0.158 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.342 & Non-constrained\\
+\hspace{1em}log(Humidity) & -1.478 & -0.157 & -0.001 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.647 & -0.275 & -0.014 & Non-constrained\\
 \hspace{1em}log(Sunshine) & -0.266 & -0.006 & 0.028 & Non-constrained\\

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Response-to-Reviewers-1.Rmd
First round of revisions after invitation to revise and resubmit.

---
title: "Response to Reviewers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor:

Thank you for your decision letter of May 5th, concerning our jointly authored submission *A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain*.

We would like to express our gratitude to the three anonymous reviewers for their thoughtful comments, which we addressed in this revision.

In what follows, we describe the actions taken in response to the feedback received. We trust that you, and the reviewers, will find that we have addressed all recommendations in a thorough and complete manner.

We look forward to your opinion regarding the revisions of this paper.

Sincerely,

The authors

\newpage

## Reviewer 1

Narrative

This timely paper examines the association between environmental variables and COVID-19 incidence in Spain.  It is well-written and contains a robust statistical analysis. The SUR method is very robust and appropriate to address the authors’ research questions; while it is very well articulated in the text.  The results and discussion are well presented and the exhibited decrease in Rho (spatial autocorrelation) over time is particularly very insightful.

I believe that the paper is suitable for publication after some minor revisions.

> *Thank you for your kind comments and positive assessment of the paper.*

Please see my specific suggestions and comments below:

Introduction and Purpose

Page 1, Lines 42-43: “the SARS-CoV2 virus has threatened to overrun health systems the world over” – sentence is awkward and should be rephrased to “around the world” or “global health systems”.

Page 1, Line 47: “arrested the spread” should be “mitigated the spread of the virus”.

Page 1, Line 48: “helped to keep a bad situation from becoming even worse” is awkward and should be rephrased.  Do you have any citations regarding the efficacy of social distancing?

> *These items were edited as per your suggestion, and we have included references as requested.*

Page 2, Lines 24-28: You need to add the degree symbol ° in front of C.  Please do so throughout the paper.

> *Done.*

Page 3, Lines 32-42: This paragraph does not really belong in the introduction.  You are discussing the key findings and results here, which belong in the discussion and conclusion.  Instead, I suggest adding some hypotheses and research questions.

> *The introduction was reworked and most of this material was moved to the Background. This helped to make the introduction shorter and more engaging.*

Page 3, Line 47: The link should be a footnote, rather than in the main body of the text.

> *Changed to a footnote.*

Context and Data Methods

Page 3, Line 51: “January 31th” should be “January 31st, 2020”.

> *Corrected.*

Page 3, Lines 9-10: Is the relatively slow start due to testing lags? You may want to add something to clarify.  In the United States, many people who died or were infected with the virus in January and February were considered influenza cases and not COVID-19 cases.

> *This is true. We have edited the text to note the relationship with testing capabilities.*

Page 4, Lines 13-14: Was Madrid the original hotspot of COVID-19 because of testing resources and population?

> *The original hotspot was actually Álava in Pais Vasco; after reorganizing the sections, we hope that it is clearer that our hypothesis is that population and GDP are related to the early flare of the outbreak in Álava and Madrid (the first and second wealthiest Autonomous Communities in Spain, in GDP per capita).*

Section 2.2: This is all great information, but it almost reads like a literature review and could be condensed and discussed in the introduction.  I think the paper can be rearranged to Introduction and Context; then Data and Methods.

> *The sections have been reorganized following this suggestion. The introduction is now a little bit shorter and more in-depth discussion of the state of knowledge is now in the Background.*

Page 6, Lines 29-30: US should be abbreviated as U.S.  Please fix throughout paper.

> *Done.*

Table 1: Is the sunshine variable missing from the table?

> *It was. We have corrected this oversight.*

Methods

Page 8, Line 15. “varaibles” is misspelled. Should be “variables”.

> *Fixed.*

Is the model actually a space-time model, rather than spatial? I don’t think it is explicitly stated.

> *Indeed, the model is space-time, with a lagged dependent variable for space, and temporal correlations in the residuals. We have emphasized this in the methods.*

The methods section is well written and model structure is clearly articulated.  Although spatial SUR is not the only approach to answer your research questions, it is a very robust statistical approach.  I am curious how the results from the SUR approach compares to Geographic and Temporally Weighted Regression and Space-Time Conditional Autoregressive Models (MCMC or INLA).

> *We now comment on other potential modelling approaches in the conclusions.*

Analysis

Section 4.1 is well-written.

Figure 1 needs to be improved.  The class breaks in the legends are inconsistent, making it difficult to compare incidence rates.  The projection seems to be a bit odd.  Mainland Spain should be much larger, while the Canary Islands should be in separate insets, or remove altogether.  Finally, the black line for the Madrid label masks many provinces.

> *We changed the scale for the colors to be consistent for all maps. We tested the insets for the Canary Islands, but they crowded the figures and added relatively little information. We have decided instead to exclude them from the maps. Once this is done, the line for Madrid is less obtrusive, and overall the figures are easier to read.*

Suggest caption for Figure 2 should be “Autonomous Communities in Spain (sorted from north to south)”

> *Changed.*

I have the same suggestions for Figure 3 as I did for Figure 1.  The resolution is poor, and the mainland should be much larger and the center of the map frames.

> *Done.*

Figure 4: Are the correlations statistically significant? Please provide p-values.

>*The purpose of the daily correlation analysis in Figure 4 is to see how much day to day variability there is in the statistical associations with incidence. The p-values of of these correlation coefficients are somewhat misleading, because of the smaller samples used in their calculation. The information from this Figure helped us decide which variables to constrain in the SUR systems. At that point, the significance of the correlation is given by the p-values of the coefficients of the model. Instead of reporting the p-values for the daily correlations, we added a sentence to the effect that overall for the period  under examination, the pairwise correlations  between the independent variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables. More details concerning the correlation tests can be found in the R markdown document in the chunk `correlation-tests`.*

What is the justification of using Rook instead of Queen?  Queen would strengthen the examination of spatial interaction and dependence.  I think you are losing important information with a Rook weight matrix.

> *For this revision, we used the rook criterion, and the results do not change in any major ways.*

Is it justifiable to assume that the islands are adjacent? I think that introduces a degree of uncertainty, especially after the lockdown order.

> *This is a good point. We repeated the analysis, and removing the three islands in the sample (Baleares, Las Palmas, and Tenerife) does not change the results in any major way. The results appear to be relatively robust to these changes in W. Conceptually, we agree that it makes more sense to exclude the islands given the lockdown implications. New footnotes comment on this, and the alternative analyses can be consulted in the source R markdown document.*

Figure 5: Please increase the font size of the pooled-R2 labels.

> *Done.*

Results and Discussion

I suggest that this section should just be “Discussion”, while “Analysis” should be renamed to “Results”.

> *Done.*

Page 20, Line 32: “pandemia” is misspelled.  Should be “pandemic”.

> *Typo fixed.*

Conclusion

Page 24, Lines 43-44: “Thirdly, all environmental data are based on a single station in a province”.  This is a bit concerning since a province is essentially similar to a state.  I suppose the justification that you selected a single station based on population distribution is fair, but an interpolation could have improved the accuracy and precision of the environmental variables.

> *Many thanks for this comment. We were also curious about using a summary of several stations for the environmental variables. To test this, we interpolated the environmental variables using a k-points algorithm. When using these variables, instead of seeing improvements on accuracy and precision, the goodness-of-fit of the models declines, the intervals of confidence become wider, and consequently several coefficients lose significance. It seems to us that the interpolated variables are more noisy than the manually selected stations. It is possible that a different interpolation algorithm might improve this, but that would involve experimenting with several interpolation methods to derive several environmental variables for a period of 51 days. For verification purposes, we have recreated the data package to include both datasets (single station and average of stations). In the paper we discuss only the results of the first.*

I also think that you should not finish your paper with limitations.  I suggest moving the limitations to the discussion, then finish with a strong conclusion regarding the main findings and public health impacts.

> *Thank you for this suggestion, which we have adopted.*

\newpage

## Reviewer 2

Comments to the Author
This is a well written paper exploring environmental impacts on incidence of CV19 with spatial models, controlling for a number of socioeconomic variables. There is value in this paper, but a number of issues need to be considered:

> *Thank you for your positive comments and feedback*

It would be helpful for the authors choose a name for the virus (e.g. SARS-COV2, COVID19) and use it consistently or explain if there are differences in when/how these names are use. Additionally, make sure presentation is consistent: e.g. sars-cov2 vs. sars-cov-2, COVID/Covid, etc.

> *SARS-CoV2 is the virus, and COVID-19 the disease it causes. We have clarified this, and revised the text to make consistent use of these terms.*

pg 5, ln 13 - percentage*

> *Typo was corrected.*

pg 6, ln 46 - I'm not sure I fully understand the weighting ... given the results indicate this approach doesn't produce a good fit, perhaps it can left out

> *We feel that this is, although a minor point, of interest for modellers and would prefer to retain it. We have revised the discussion of the weights and hopefully they are more clear now.*

pg 8 ln 15 "variable" misspelled, ln 16 "restrictions" misspelled.

> *Typos were corrected*

use of mean temp/humidity - does daily variation matter? For example do some regions have lows near 2 and highs near 20 c, where others have lows near 5 and highs near 6?

> *Indeed, which is why there are experiments with the moving averages of the environmental variables: 1) to smooth the daily variation in accordance to the incubation period of the disease. The narrowest window is 8 days, and the two 11-day windows weight the daily variations in different ways: uniformly for the $lagw$ variables, and giving greater weight to days closer to the median incubation period of 4-5 days in the $lag11w$ variables.*

pg 10 ln 50 - sentence grammar

could W be defined using commuter flow to gain a better sense of interaction? Are there past examples in the literature that define W for provinces across Spain? Perhaps there's more interaction between Madrid and Barcelona, etc.

> *In principle, it is possible, but unfortunately commuter flow data are not readily available. The Ministry of Transportation is working on flow data for Spain based on mobile data, to be released at a future (and yet undefined) date. In our view, once in lockdown, hierarchical flows are perhaps less relevant than cross-boundary flows, and even those should become relatively less important as the lockdown persists: this is, we believe, what the autocorrelation coefficient is telling us. See this paragraph at the end of the discussion:*

>>*Finally, we defined neighborhoods based on adjacency. It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services. We flag this as a matter for future research.*

pg 15 ln 22 - 'transit' systems

> *Typo corrected.*

given multiple equations would it be appropriate to account for multiple testing and adjust accordingly?

> *Not with SUR, because the multiple equations are estimated jointly, and the calculations account for the size of the sample.*

pg 18 ln 32-33, do the authors mean 'due to *violations* of the restrictions of movement'?

> *This sentence was revised to conform to the slight changes in the results after the models were reestimated without the islands.*

pg 19, not sure the anecdote is appropriate - suggest this be deleted

> *Deleted.*

pg 21, ln 52 - can this recommendation be backed up with further evidence about lockdown effectiveness?

> *We think that this is a reference to pg 20, ln 52. This paragraph was too wordy, and we have edited it for clarity. Instead of effectiveness of the lockdown, we tie this discussion to risk compensation, which seems to a clearer case.*

pg 22 ln 36 - I'm not sure I follow this statement ("Without a lockdown..."). Also, please remove "!". I've tried to work through this logic, but can't quite seem to figure out how you have come to this conclusion. Wouldn't cases generated elsewhere also be affected by temperatures in those locations? Would this not make the process more complex/less clear?

> *You are right that the process becomes more complicated, and explaining it is also not simple. We were not as clear as we had hoped, but made an effort with this new paragraph:*

>>*What do these effects mean? Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the total effects of the variables tend towards their direct effects. In those first few days covered by our analysis the total effect of all variables is greater due to the spatial contagion effect. This also makes analysis and intervention more complex: the contagion effect essentially acts like a multiplier, whereby developments in each province spill over to their neighbors. Once the contagion effect has been tamed, each province can be "treated" independently from its neighbors.*

please double check spelling/grammar throughout.

> *We have proof-read the paper and hopefully fixed all remaining spelling/grammar mistakes.*

\newpage

## Reviewer 3

Comments to the Author

The paper is well written and very timely. I truly enjoyed reading it and I think it should be granted publication only with some very minor revisions. Although I understand and I use econometric and statistical models, I am not an expert on the technical part and therefore I will not comment on that specifically. My comments are more "general".

> *Many thanks for your positive assessment and suggestions for revisions.*

There is a clear need of looking at environmental factors such as those you are considering. However, there has been a huge debate on pollution as a possible environmental factor affecting the spread of the disease. I would suggest that you clarify somewhere that you look only at weather-related factor and not pollution and explain why (bad data I assume?).

> *Indeed, working with meterological stations was already challenging, and we are not aware of pollution data at the scale of the analysis. We now discuss the role of pollution as an environmental factor in need of further attention.*

Working on the topic myself I often wonder how the results would change if the disease had hit somewhere else first? Clearly that would have affected the spread...Could this be a problem in your case?

>*This is an interesting question and a problem insofar as we lack a counterfactual, This is where the use of simulation techniques are useful, and the work of Muller et al. with MATSim is relevant in this respect; see:*

>>*Muller, S. A., Balmer, M., Neumann, A., & Nagel, K. (2020). Mobility traces and spreading of COVID-19. medRxiv, 2020.2003.2027.20045302. doi:10.1101/2020.03.27.20045302*

Very minor:

I would write NUTS2, NUTS3...I have almost never seen NUTII and NUTIII...

>*We made this change.*

Page 6: when you say you use a rather large geographical unit of analysis. I get that, to be quick, you have to use whatever data are available in the short run. However, I do not get what you mean by 'the analysis must therefore be considered "ecological"'. Either you explain it better or you simply remove the sentence acknowledging your results are preliminary, but could be easily applied to a finer scale once, and if, better data become available.

>*We agree, and removed the sentence to avoid overcomplicating the discussion. Earlier on in the paper we already had explained that this is a population health approach.*

Here and there I would remove some "however".

>*Done.*

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
First round of revisions after invitation to revise and resubmit.

@@
 # Load data from package `covid19env`
-data("covid19_spain")
+data("covid19_spain_1") # Use data sourced from one representative station for the province
+#data("covid19_spain_2") # Use data with the average of all stations in provinces
 data("provinces_spain")
@@
 # Join provincial data to incidence data and convert to simple features:
-covid19_spain <- covid19_spain %>% 
+covid19_spain <- covid19_spain_1 %>% 
+#covid19_spain <- covid19_spain_2 %>% # data with the average of all stations in provinces   
   left_join(provinces_spain %>% st_drop_geometry(),
@@
 
-From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus has threatened to overrun health systems the world over.
-In efforts to contain the spread, numerous governments in many nations and regions have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place.
-There are encouraging signs that these measures have arrested the spread of the virus where they have been implemented, and have thus helped to keep a bad situation from becoming even worse [e.g., -@Lancastle2020impact].
-However, this has come at a high cost, and the consequences for all spheres of the economy, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how].
+From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and COVID-19, the disease it causes), has threatened to overrun health systems around the world.
+In efforts to contain the spread of the disease, numerous governments in many regions and nations have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place.
+There are encouraging signs that these measures have mitigated the spread of the virus [e.g., @Lancastle2020impact; @Lewnard2020scientific; @Wilder2020isolation].
+Even so, this has come at a high cost, and the consequences for all spheres of economic, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how].
 As a result, there is a sense of urgency to anticipate the progression of the pandemic, in order to plan for progressive lifting of restrictions to movement and social contact [e.g., @Kissler2020projecting].
@@
 A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations.
+Existing research on similar pathogens suggests that the virus could be more stable and potentially easier to transmit in conditions of low temperature and low humidity.
+While encouraging this is encouraging, it is important to keep in mind that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" [@deangel2020weathering, section 4].
+This urges caution when extrapolating from known viruses.
+The evidence in this respect is as yet inconclusive, and although easing restrictions as the weather warms may appear tempting, doing so prematurely could well undo weeks or possibly months of costly measures.
+
+It is not surprising, given the stakes involved, that this issue has already triggered a lively debate.
+The current state of knowledge is was well-summarized by the National Academy of Sciences, Engineering, and Medicine in the U.S.
+[see @National2020rapid].
+Engaged by the Office of the Executive for guidance on this matter, this organization concludes that: "[some] limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV2 pandemic unfolds could shed more light on the effects of climate on transmission" (p.
+6).
+To further complicate matters, much of the relevant work has yet to be peer-reviewed (see for instance the challenge of @Harbert2020spatial to @Araujo2020spread).
+
+With the above considerations in mind, our objective with this paper is to investigate the influence of environmental factors, concretely temperature, humidity, and sunshine, on the progression of the pandemic.
+We adopt a population health approach, and report results from a spatio-temporal model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
+We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
+We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables.
+These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a contagion process.
+
+The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
+The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
+Our control variables also provide some intriguing insights.
+Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
+The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
+
+Please note that this paper is prepared as a reproducible research document.
+The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.
+
+Background {#background}
+============
+
+The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
+In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
+Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
+As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
+
+Much of what is known about the potential seasonality of SARS-CoV2 is a result of research on other pathogens.
 Earlier, diverse studies have shown the effect of temperature and humidity on the incidence of influenza [e.g., @Makinen2009cold; @Jaakkola2014decline; @Kudo2019low].
@@
 [-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low.
-After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV.
-In this way, Casanova et al.
-[-@Casanova2010effects] used two surrogates, namely the gastroenteritis (TGEV) and mouse hepatitis viruses (MHV), to find that virus inactivation was more rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
-In a similar vein, but working directly with SARS-CoV, Chan et al.
-[-@Chan2011effects] found that viability of the virus was lost at temperatures higher than 38C and relative humidity superior to 95%.
-
-While existing research on similar pathogens suggests that SARS-CoV is more stable and potentially easier to transmit in conditions of low temperature and low humidity, it is far from certain that this will also be the case with the novel SARS-CoV2.
-If such is the case, there is the possibility of easing restrictions to social contact as the weather warms; however, weeks or possibly months of costly measures could become undone if not, and the restrictions are lifted prematurely.
-Not surprisingly, given the stakes involved, this issue has already triggered a lively debate.
-
-Some of what is thought about the possible seasonality of COVID-19 is based on analogies to the patterns of other known respiratory viruses.
-However, de Ángel Solá et al.
-[-@deangel2020weathering] note that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" (section 4).
-This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive.
+After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV2 [@Casanova2010effects; @Chan2011effects].
+Casanova et al.
+[-@Casanova2010effects], for instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse hepatitis MHV viruses), to find that virus inactivation was more rapid at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
+Chan et al.
+[-@Chan2011effects] also found that viability of the virus that causes SARS is also lost at higher temperatures (>38°C) and relative humidity superior to 95%.
+
+Whether these results from laboratory experiments will hold in general remains uncertain.
 At a global scale, de Ángel Solá et al.
@@
 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).
-To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread).
-In the United States, the National Academy of Sciences, Engineering, and Medicine was engaged by the Office of the Executive for guidance on this matter [see @National2020rapid].
-Their conclusion summarizes the situation well (see p.
-6): "Some limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV-2 pandemic unfolds could shed more light on the effects of climate on transmission."
-
-With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature, humidity, and sunshine.
-We adopt a population health approach, and report results from a spatial model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
-We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
-We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables.
-These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a spatial contagion process.
-
-The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
-The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
-Our control variables also provide some intriguing insights.
-Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
-The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
-
-Please note that this paper is prepared as a reproducible research document.
-The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following anonymous Drive folder:
-
-https://drive.google.com/open?id=1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_
-
-Context and Data {#context-and-data}
-============
-
-## Covid-19 in Spain {#covid-19-in-spain}
-
-The first reported case of COVID-19 in Spain was on January 31th, when a German tourist in the Canary Islands tested positive for the virus.
-However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
-In a short period of time, after this relatively slow start, the outbreak flared.
-By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic.
-This declaration marked a turning point for the public in Spain too.
-As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe, after Italy.
-In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th.
-As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g.
-food, health) and some economic subsectors of industry and construction.
-A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents.
-The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed.
-During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
-
-## Selection of Variables {#variables}
-
-The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
-In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
-Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
-As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
-These variables have been implicated in the viability and ease of transmission of similar viruses.
-In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.
+
+In addition to the environmental variables above, we also consider a set of potential socio-economic and demographic confounders.
 
@@
 
-The percentaje of older adults (over 65) is the second variable that we consider as a control.
+Secondly, we consider percentage of older adults (over 65) in a region.
 Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g.
@@
 Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age].
-Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].
-
-The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
+Accordingly, our expectation is that higher percentages of older adults will tend to have similarly low levels of social contact [the age-structured matrix of social contact in Spain is similar to Italy; see @Prem2017projecting].
+
+Population density is also relevant since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
 The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g.
@@
 
-The last control variable is the presence of mass transit systems in a province.
+The last variable that we consider as a control is the presence of mass transit systems in a province.
 Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g.
@@
 
+Context and Data {#context-and-data}
+===================
+
+## Covid-19 in Spain {#covid-19-in-spain}
+
+The first reported case of COVID-19 in Spain was on January 31st, 2020, when a German tourist in the Canary Islands tested positive for the virus.
+However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
+In a short period of time, as testing started to ramp up, it became clear that an outbreak was flaring.
+By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic.
+This declaration marked a turning point for the public in Spain too.
+As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe after Italy.
+In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th.
+As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g.
+food, health) and some economic subsectors of industry and construction.
+A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents.
+The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed.
+During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
+
 ## Data {#data}
 
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
+Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
 For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España.
+The Spanish National Government publishes periodic updates at the regional level (NUTS2) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España.
 This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
@@
 The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}.
-The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the US) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the US).
-While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is inconsistent at smaller geographies, or cases are not reported at that level at all.
-The analysis must therefore be considered ecological.
+The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the U.S.) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the U.S.).
+While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is not yet consistent at smaller geographies, or cases are not reported at that level at all.
 
@@
                         "Mean Temperature",
-                        "Humidity"),
+                        "Humidity",
+                        "Sunshine"),
            Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
@@
                     "Population density in the province in people per sq.km",
-                    "Mean temperature in province by date, in C",
-                    "Relative humidity in province by date"),
+                    "Mean temperature in province by date, in °C",
+                    "Relative humidity in province by date",
+                    "Daily hours of sunshine in province by date"),
            Min = c(min(covid19_spain$Incidence),
@@
                    min(covid19_spain$Mean_Temp),
-                   min(covid19_spain$Humidity)),
+                   min(covid19_spain$Humidity),
+                   min(covid19_spain$Sunshine_Hours)),
            Mean = c(mean(covid19_spain$Incidence),
@@
                     mean(covid19_spain$Mean_Temp),
-                    mean(covid19_spain$Humidity)),
+                    mean(covid19_spain$Humidity),
+                    mean(covid19_spain$Sunshine_Hours)),
            Max = c(max(covid19_spain$Incidence),
@@
                    max(covid19_spain$Mean_Temp),
-                   max(covid19_spain$Humidity)),
+                   max(covid19_spain$Humidity),
+                   max(covid19_spain$Sunshine_Hours)),
            SD = c(sd(covid19_spain$Incidence),
@@
                   sd(covid19_spain$Mean_Temp),
-                  sd(covid19_spain$Humidity)),
+                  sd(covid19_spain$Humidity),
+                  sd(covid19_spain$Sunshine_Hours)),
            Source = c("ProvidencialData19",
@@
                       "AEMET",
-                      "MAPA")) %>%
+                      "MAPA",
+                      "AEMET")) %>%
   kable("latex",
@@
 
-The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consisits of cross-sections for different time periods.
+The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consists of cross-sections for different time periods.
 The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient], and has become a popular methodology included in several econometrics textbook [e.g., @Greene2003econometric].
-To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective.
+To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective, in the context of spatio-temporal analysis.
 In his landmark text, Anselin discussed a model made of "an equation for each time period, which is estimated for a cross section of spatial units" (p.
@@
 
-A key feature of the SUR model is the dependence structure among the vectors of residuals, namely:
+A key feature of the SUR model is the temporal dependence structure among the vectors of residuals, namely:
 
@@
 
-Analysis {#analysis}
+Results {#results}
 ===================
@@
 week11.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
                        palette = "Reds", 
-                       direction = 1) +
+                       direction = 1,
+                       limits = c(0.755, 988.71)) +
   ggtitle("March 13 - March 15") +
@@
   geom_sf_text(label = "Madrid", 
-               x = madrid_xy[1] - 9, 
+               x = madrid_xy[1] - 4.7, 
                y = madrid_xy[2] + 0.6,
                family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
+  geom_segment(x = madrid_xy[1] - 3.4, 
+               y = madrid_xy[2] + 0.6, 
                xend = madrid_xy[1], 
@@
 week12.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
                        palette = "Reds",
-                       direction = 1) +
+                       direction = 1,
+                       limits = c(0.755, 988.71)) +
   ggtitle("March 16 - March 22") +
@@
         plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 9, y = madrid_xy[2] + 0.6) +
-  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 4.7, 
+               y = madrid_xy[2] + 0.6) +
+  geom_segment(x = madrid_xy[1] - 3.4,
+               y = madrid_xy[2] + 0.6, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 week13.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
                        palette = "Reds", 
-                       direction = 1) +
+                       direction = 1,
+                       limits = c(0.755, 988.71)) +
   ggtitle("March 23 - March 29") +
@@
   geom_sf_text(label = "Madrid",
-               x = madrid_xy[1] - 9, 
+               x = madrid_xy[1] - 4.7, 
                y = madrid_xy[2] + 0.6,
                family = "serif") +
-  geom_segment(x = madrid_xy[1] - 6, 
-               y = madrid_xy[2] + 0.5, 
+  geom_segment(x = madrid_xy[1] - 3.4, 
+               y = madrid_xy[2] + 0.6, 
                xend = madrid_xy[1], 
@@
 week14.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
                        palette = "Reds", 
-                       direction = 1) +
+                       direction = 1,
+                       limits = c(0.755, 988.71)) +
   ggtitle("March 30 - April 5") +
@@
 week15.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
                        palette = "Reds", 
-                       direction = 1) +
+                       direction = 1,
+                       limits = c(0.755, 988.71)) +
   ggtitle("April 6 - April 11") +
@@
 The top panel of the figure shows the distribution of mean temperatures.
-The lowest mean temperature for a community on any given day was 2.8C, and the highest 22.4C, for a range of approximately 20 degrees.
+The lowest mean temperature for a community on any given day was 2.8°C, and the highest 22.4°C, for a range of approximately 20 degrees.
 Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
@@
   theme(axis.text.x = element_text(angle = 90)) +
-  xlab("Autonomous Community (sorted from north to south)") +
+  xlab("Autonomous Communities in Spain (sorted from north to south)") +
   ylab("Value") +
@@
                              GDPpc > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   ggplot() +
@@
                              Older > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   ggplot() +
@@
                              Density > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   ggplot() +
@@
                                      "Yes"))) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias") %>%
   ggplot() +
@@
 Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive.
+Overall for the period under examination, the pairwise correlations  between these variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables.
 
@@
   group_by(Date) %>%
-  summarize(Older = cor(log(Older), log(Incidence)),
-            GDPpc = cor(log(GDPpc), log(Incidence)),
+  summarize(GDPpc = cor(log(GDPpc), log(Incidence)),
+            Older = cor(log(Older), log(Incidence)),
             Density = cor(log(Density), log(Incidence)),
@@
 
+```{r correlation-tests, include=FALSE}
+cor.test(log(covid19_spain$GDPpc), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Older), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Density), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Humidity_lag8), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Humidity_lag11), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Humidity_lag11w), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Mean_Temp_lag8), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Mean_Temp_lag11), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Mean_Temp_lag11w), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Sunshine_Hours_lag8 + 0.1), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Sunshine_Hours_lag11 + 0.1), log(covid19_spain$Incidence))
+cor.test(log(covid19_spain$Sunshine_Hours_lag11w + 0.1), log(covid19_spain$Incidence))
+```
+
 ## SUR Models {#sur-models}
@@
 To implement the SUR approach, we must define a matrix of spatial weights $W$.
-In this case, we consider neighborhoods based on adjacency, based on the well-known rook criterion (two provinces are adjacent if they share a boundary, but not if they touch only at a vertex).
+In this case, we consider neighborhoods based on adjacency, based on the well-known queen criterion (two provinces are adjacent if they share a boundary or touch at a vertex).
 We modify this criterion in two cases.
 First, we make an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are adjacent between each other.
-And secondly, we assume that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello).
-After matrix $W$ has been specified, estimation of the model can proceed as usual.
+And secondly, we assume that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello)\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and removed the three islands in the sample.
+The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
 
 For estimation, we log-transform the dependent and quantitative independent variables.
-The only variable that is not transformed is the categorical variable for transity systems.
+The only variable that is not transformed is the categorical variable for transit systems.
 A log-log transformation is appropriate to capture non-linear relationships between variables and provides a straightforward interpretation of the coefficients as percentage change.
@@
 
-```{r data-preparation-for-modelling, include=FALSE, cache = TRUE}
+```{r data-preparation-for-modelling, eval = FALSE, include=FALSE, cache = TRUE}
 # Organize data for SUR modelling 
@@
 
-```{r spatial-weights, include=FALSE, cache = TRUE}
+```{r spatial-weights, eval = FALSE, include=FALSE, cache = TRUE}
 # Create spatial weights matrix: 
@@
   as("Spatial") %>%
-  poly2nb(queen = FALSE) %>%
+  #poly2nb(queen = FALSE) %>%
+  poly2nb(queen = TRUE) %>%
   nb2mat(zero.policy = TRUE)
@@
 
+```{r data-preparation-for-modelling-minus-islands, eval = TRUE, include=FALSE, cache = TRUE}
+# Organize data for SUR modelling 
+GPanel <- plm::pdata.frame(covid19_spain %>%
+                             filter(province != "Baleares", CCAA != "Canarias") %>%
+                             select(province, 
+                                    Date,
+                                    Incidence, 
+                                    Median_Age,
+                                    Male2Female,
+                                    Older,
+                                    GDPpc,
+                                    Density,
+                                    Transit,
+                                    Mean_Temp_lag8,
+                                    Humidity_lag8,
+                                    Sunshine_Hours_lag8,
+                                    Mean_Temp_lag11,
+                                    Humidity_lag11,
+                                    Sunshine_Hours_lag11,
+                                    Mean_Temp_lag11w,
+                                    Humidity_lag11w,
+                                    Sunshine_Hours_lag11w),
+                           c("province","Date"))
+
+```
+
+```{r spatial-weights-minus-islands, eval=TRUE, include=FALSE, cache = TRUE}
+# Create spatial weights matrix: 
+Wmat <- provinces_spain %>%
+  filter(province != "Baleares", CCAA != "Canarias") %>%
+  #drop_na() %>%
+  as("Spatial") %>%
+  poly2nb(queen = TRUE) %>%
+  nb2mat(zero.policy = TRUE)
+
+#Wmat <- (Wmat > 0) * 1
+
+#miW <- Wmat/rowSums(Wmat)
+
+# Convert to listw
+listw <- mat2listw(Wmat,style = "W")
+```
+
 ```{r formulas, include=FALSE, cache = TRUE}
@@
 ```
-
 ```{r model-restrictions, include=FALSE, cache = TRUE}
@@
            x = as_date("2020-04-01"),
-           y = c(0.3, 0.27, 0.24),
+           y = c(0.3, 0.25, 0.20),
            family = "serif",
-           size = 3.5) +
+           size = 5) +
   theme_bw() +
@@
 
-Results and Discussion {#results-and-discussion}
+Discussion {#discussion}
 ===================
@@
 ```{r summary-best-model-all, include=FALSE, eval=FALSE}
-best_model_all <- sur.slm_lag11_all
+best_model_all <- sur.slm_lag11
 
@@
 We begin our discussion with the evolution of the spatial autocorrelation coefficient ($\rho$) in Figure \ref{fig:results-time-1} (left panel).
-It is interesting to note that the magnitude of the spatial autocorrelation coefficient $\rho$ declines over the period under analysis, and is not significant for some days.
+We notice that the magnitude of the spatial autocorrelation coefficient $\rho$ declines over the period under analysis, and is not significant for some days.
 This is an interesting result: immediately prior to the declaration of the state of emergency, there appears to have been a strong inter-provincial contagion effect.
@@
 Indeed, as seen in the figure, the autocorrelation coefficient remains high for three days, then declines on March 16, and continues to weaken over time.
-At the end of the period under examination, the strength of this effect is much diminished.
-We would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero.
-The fact that it is not is evidence of some weak contagion effect due to restrictions to the movement between provinces.
+At the end of the period under examination, the strength of this effect is much diminished and we would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero - as seems to be the case.
 
@@
 Consequently, residents in high density regions may perceive the risk of infection as being higher, and adapt their behavior accordingly - while the opposite may be true of residents in low density regions.
-Anecdotally, this seems to have been the case with some governors in the US, who resisted stay-at-home orders citing their sparse \footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions} and rural populations\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}.
 The coefficient for Transit is positive, as expected, but with very wide confidence intervals, and in fact not significant in the earlier part of the period.
@@
 
-```{r calculate-impacts-best-model, echo=FALSE, cache = TRUE}
+```{r calculate-impacts-best-model, echo=FALSE}
 # Calculate impacts
@@
 Two variables had temporally constrained coefficients.
-The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.675% for each percentual increase of this variable (in €1,000s).
-In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemia.
-Percentage of older adults has a negative effect, and each percentual increase in this variable is associated with a relatively small reduction of the incidence of less than 1%.
+The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.675% for each percent increase of this variable (in €1,000s).
+In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemic.
+Percentage of older adults has a negative effect, and each percent increase in this variable is associated with a relatively small reduction of the incidence of less than 1%.
 
@@
 The largest positive direct effect is Transit, and the largest direct negative effects are temperature and humidity.
-The direct effect of these variables is as follows: for each percentual point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
+The direct effects of these variables are as follows: for each percent point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
 This effect is amplified via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 1.5%.
-The effect of humidity is also to reduce the incidence: each percentual point of increase in humidity is associated with a reduction of up to 1.5% in incidence.
+The effect of humidity is also to reduce the incidence: each percent point of increase in humidity is associated with a reduction of up to 1.5% in incidence.
 With the addition of the indirect effect, the total effect of a 1% increase in humidity is to reduce incidence by up to 3%.
 As seen in the figure, the indirect (i.e., contagion) effects are almost as strong as the direct effects immediately before and at the beginning of the lockdown period.
-However, by the end of the lockdown period, the indirect effects have weakened considerably.
+Nonetheless, by the end of the lockdown period, the indirect effects have weakened considerably.
 The effect of density goes from positive (+0.334) to negative (-0.244).
-It is crucial to keep in mind here that this is the effect of density over a period of *lockdown* and possibly subject to risk compensation, as discussed above.
-In other words, the lockdown seems to be more effective in places with higher density.
-Accordingly, cities should not rush to ease lockdowns, but rather efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
+It is crucial to keep in mind here that this is the effect of density over a period of *lockdown* and possibly subject to risk compensation, as discussed above which suggests, in turn, that efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
 
 What do these effects mean?
-Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the effect of temperature is estimated to reduce the incidence of COVID-19 by between one and two percent with each percentual increase of this variable.
-Without a lockdown, or a relaxed lockdown, the effect is even greater, but only because the disease is transmitted inter-regionally!
-Imagine a region that is closed to its neighbors: increases in temperature are associated with smaller reductions in incidence - however, it does not import or export cases.
-A province that is not closed and not under lockdown will see a greater effect of temperature, but at the cost of importing/exporting cases.
+Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the total effects of the variables tend towards their direct effects.
+In those first few days covered by our analysis the total effect of all variables is greater due to the spatial contagion effect.
+This also makes analysis and intervention more complex: the contagion effect essentially acts like a multiplier, whereby developments in each province spill over to their neighbors.
+Once the contagion effect has been tamed, each province can be "treated" independently from its neighbors.
+
+It is important, before concluding our discussion, to highlight some limitations of this study.
+
+First, the analysis was conducted under a situation of lockdown, and therefore, besides the results at the beginning of the period examined, one must exercise caution when trying to extrapolate the findings to other situations.
+Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
+In this case we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results - even if we are aware that the number of true cases is likely higher.
+Finally, we defined neighborhoods based on adjacency.
+It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services.
+We flag this as a matter for future research.
 
@@
 
-It is important, in this respect, to highlight some limitations of this study.
-
-First, the analysis was conducted under a situation of lockdown, and therefore, besides the results are the beginning of the period examined, one must exercise caution when trying to extrapolate the findings to other situations.
-Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
-However, we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results - even if we are aware that the number of true cases is likely higher.
-Thirdly, all environmental data are based on a single station in a province.
-Here, we made an effort to select the most representative station based on the distribution of the population.
-An alternative approach would be to calculate the environmental variables based on the average of multiple stations within each province: this would likely smooth to some extent these variables.
-Finally, we defined spatial neighborhoods based on adjacency.
-It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services.
-We flag this as a matter for future research.
+Some directions for future research include investigating other modelling frameworks, such as geographically and temporally weighted regression and/or space-time conditional autoregressive models.
+In addition, the environmental variables examined here relate to meteorological conditions, and did not include other environmental factors that may incide in the transmission of the virus, such as pollution.
+Other factors could be incorporated in future studies.
 

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
First round of revisions after invitation to revise and resubmit.

@@
 From a small outbreak linked to a live animal market at the end of 2019
-to a global pandemic in a matter of weeks, the SARS-CoV2 virus has
-threatened to overrun health systems the world over. In efforts to
-contain the spread, numerous governments in many nations and regions
-have either recommended or mandated social distancing measures, and as
-of this writing, millions of people in five continents shelter in place.
-There are encouraging signs that these measures have arrested the spread
-of the virus where they have been implemented, and have thus helped to
-keep a bad situation from becoming even worse (e.g., 2020). However,
-this has come at a high cost, and the consequences for all spheres of
-the economy, social, and cultural life have been dire (e.g., Fernandes,
-2020; Luo and Tsang, 2020). As a result, there is a sense of urgency to
-anticipate the progression of the pandemic, in order to plan for
-progressive lifting of restrictions to movement and social contact
-(e.g., Kissler et al., 2020). Needless to say, this has become a
-delicate, and politically charged, balancing act between public health
-and the economy (Gong et al., 2020).
+to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and
+COVID-19, the disease it causes), has threatened to overrun health
+systems around the world. In efforts to contain the spread of the
+disease, numerous governments in many regions and nations have either
+recommended or mandated social distancing measures, and as of this
+writing, millions of people in five continents shelter in place. There
+are encouraging signs that these measures have mitigated the spread of
+the virus (e.g., Lancastle, 2020; Lewnard and Lo, 2020; Wilder-Smith and
+Freedman, 2020). Even so, this has come at a high cost, and the
+consequences for all spheres of economic, social, and cultural life have
+been dire (e.g., Fernandes, 2020; Luo and Tsang, 2020). As a result,
+there is a sense of urgency to anticipate the progression of the
+pandemic, in order to plan for progressive lifting of restrictions to
+movement and social contact (e.g., Kissler et al., 2020). Needless to
+say, this has become a delicate, and politically charged, balancing act
+between public health and the economy (Gong et al., 2020).
 
@@
 distancing measures is whether the virus will display seasonal
-variations. Earlier, diverse studies have shown the effect of
-temperature and humidity on the incidence of influenza (e.g., Mäkainen
-et al., 2009; Jaakkola et al., 2014; Kudo et al., 2019). Jaakkola et
-al.~(2014), for example, found that a decrease of temperature and
-absolute humidity precedes the onset of symptoms of influenza A and B
-viruses by 3 days in places where the temperature is low. After the
-2002-2004 outbreak of SARS, researchers also began to investigate the
-relationship between these factors and SARS-CoV. In this way, Casanova
-et al.~(2010) used two surrogates, namely the gastroenteritis (TGEV) and
-mouse hepatitis viruses (MHV), to find that virus inactivation was more
-rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of
-humidity, these researchers reported that survival of the virus was
-lower at moderate relative humidity levels. In a similar vein, but
-working directly with SARS-CoV, Chan et al.~(2011) found that viability
-of the virus was lost at temperatures higher than 38C and relative
-humidity superior to 95\%.
-
-While existing research on similar pathogens suggests that SARS-CoV is
-more stable and potentially easier to transmit in conditions of low
-temperature and low humidity, it is far from certain that this will also
-be the case with the novel SARS-CoV2. If such is the case, there is the
-possibility of easing restrictions to social contact as the weather
-warms; however, weeks or possibly months of costly measures could become
-undone if not, and the restrictions are lifted prematurely. Not
-surprisingly, given the stakes involved, this issue has already
-triggered a lively debate.
-
-Some of what is thought about the possible seasonality of COVID-19 is
-based on analogies to the patterns of other known respiratory viruses.
-However, de Ángel Solá et al.~(2020) note that ``not all seasonal
-respiratory viruses experience the same spatiotemporal patterns''
-(section 4). This urges caution when extrapolating from known viruses,
-and indeed, the evidence in this respect is inconclusive. At a global
-scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
-however, Coelho et al.~(Coelho et al., 2020) warn that at least during
-the exponential phase, expansion of the virus is not driven by climate.
-Similarly, whereas Araujo and Naimi (2020) argue that spread of
-SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
-remain unconvinced that spatial modelling can currently discriminate the
-distribution of the disease on the basis of climate, at least in the
-United States. Yao et al.~(2020), examined data from China and came to
-the conclusion that neither temperature nor ultraviolet indices had an
-association with transmission of COVID-19. This is despite previous
-research that has linked less exposure to UVB radiation to higher
-prevalence and severity of acute respiratory tract infections
-(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
-al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
-Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
-much of the relevant work has yet to be peer-reviewed and therefore is
-open to change (see for example the challenge of Harbert et al. (2020)
-to Araujo and Naimi (2020)). In the United States, the National Academy
-of Sciences, Engineering, and Medicine was engaged by the Office of the
-Executive for guidance on this matter (see National Academies of
-Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
-the situation well (see p.~6): ``Some limited data support a potential
-waning of cases in warmer and more humid seasons, yet none are without
-major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
-unfolds could shed more light on the effects of climate on
-transmission.''
+variations. Existing research on similar pathogens suggests that the
+virus could be more stable and potentially easier to transmit in
+conditions of low temperature and low humidity. While encouraging this
+is encouraging, it is important to keep in mind that ``not all seasonal
+respiratory viruses experience the same spatiotemporal patterns'' (Ángel
+Solá et al., 2020, sec. 4). This urges caution when extrapolating from
+known viruses. The evidence in this respect is as yet inconclusive, and
+although easing restrictions as the weather warms may appear tempting,
+doing so prematurely could well undo weeks or possibly months of costly
+measures.
+
+It is not surprising, given the stakes involved, that this issue has
+already triggered a lively debate. The current state of knowledge is was
+well-summarized by the National Academy of Sciences, Engineering, and
+Medicine in the U.S. (see National Academies of Sciences, Engineering
+and Medicine, 2020). Engaged by the Office of the Executive for guidance
+on this matter, this organization concludes that: ``{[}some{]} limited
+data support a potential waning of cases in warmer and more humid
+seasons, yet none are without major limitations\ldots Additional studies
+as the SARS-CoV2 pandemic unfolds could shed more light on the effects
+of climate on transmission'' (p.~6). To further complicate matters, much
+of the relevant work has yet to be peer-reviewed (see for instance the
+challenge of Harbert et al. (2020) to Araujo and Naimi (2020)).
 
 With the above considerations in mind, our objective with this paper is
-to contribute to the knowledge basis regarding the spread of COVID-19
-and the influence of environmental factors, particularly temperature,
-humidity, and sunshine. We adopt a population health approach, and
-report results from a spatial model of the incidence of COVID-19 in
-fifty provinces in Spain, one of the countries hardest hit by the
-pandemic. We combine data on reported cases of the disease with
-metereological information, to create a spatio-temporal dataset covering
-a period of 30 days. We then join this dataset with provincial-level
-economic and demographic information to act as controls to our key
-environmental variables. These data are analyzed using a spatial
-Seemingly Unrelated Regressions (SUR) approach, which allows us to model
-incidence of COVID-19 as a spatial contagion process.
+to investigate the influence of environmental factors, concretely
+temperature, humidity, and sunshine, on the progression of the pandemic.
+We adopt a population health approach, and report results from a
+spatio-temporal model of the incidence of COVID-19 in fifty provinces in
+Spain, one of the countries hardest hit by the pandemic. We combine data
+on reported cases of the disease with metereological information, to
+create a spatio-temporal dataset covering a period of 30 days. We then
+join this dataset with provincial-level economic and demographic
+information to act as controls to our key environmental variables. These
+data are analyzed using a spatial Seemingly Unrelated Regressions (SUR)
+approach, which allows us to model incidence of COVID-19 as a contagion
+process.
 
@@
 Please note that this paper is prepared as a reproducible research
-document. The source R Markdown document, as well as all data and code
-needed to reproduce/review/extend the analysis can be obtained from the
-following anonymous Drive folder:
-
-https://drive.google.com/open?id=1d\_40N\_QXo2Fl14r3T3CHs84IED-B1rV\_
-
-\hypertarget{context-and-data}{%
-\section{Context and Data}\label{context-and-data}}
-
-\hypertarget{covid-19-in-spain}{%
-\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}
-
-The first reported case of COVID-19 in Spain was on January 31th, when a
-German tourist in the Canary Islands tested positive for the virus.
-However, it was still a few weeks before the first domestic case was
-reported, on February 27th in Sevilla province (Andalusia). In a short
-period of time, after this relatively slow start, the outbreak flared.
-By March 11th the World Health Organization (WHO) declared COVID-19
-officially a pandemic. This declaration marked a turning point for the
-public in Spain too. As of March 13th, the number of cases of COVID-19
-reported in Spain was 4,473, with a majority of cases (1,990)
-concentrated in Madrid: these numbers were at the time the worst
-outbreak in Europe, after Italy. In response to the situation, on March
-13th the Spanish National Government declared a state of emergency, to
-go into effect on Saturday March 14th. As part of the state of emergency
-restrictions to most activities were imposed, with the exception of
-essential services (e.g.~food, health) and some economic subsectors of
-industry and construction. A few days later, on March 17th, Spain closed
-its lands borders to allow entry only to returnee nationals and
-permanent residents. The lockdown was further hardened between March
-30th and April 12th (including the Easter weekend of April 10th-12th)
-and during this period only essential activities were allowed. During
-this period, there was a dramatic reduction in overall mobility, both
-within provinces as between
-\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
-
-\hypertarget{variables}{%
-\subsection{Selection of Variables}\label{variables}}
+document. The source \texttt{R} markdown document, as well as all data
+and code needed to reproduce/review/extend the analysis can be obtained
+from a public
+repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.
+
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
@@
 focus of the paper is on environmental variables, namely temperature,
-humidity, and sunshine. These variables have been implicated in the
-viability and ease of transmission of similar viruses. In addition to
-these variables, we also aim to use a set of controls, in the form of
-specific socio-economic and demographic characteristics of each
-province.
+humidity, and sunshine.
+
+Much of what is known about the potential seasonality of SARS-CoV2 is a
+result of research on other pathogens. Earlier, diverse studies have
+shown the effect of temperature and humidity on the incidence of
+influenza (e.g., Mäkainen et al., 2009; Jaakkola et al., 2014; Kudo et
+al., 2019). Jaakkola et al.~(2014), for example, found that a decrease
+of temperature and absolute humidity precedes the onset of symptoms of
+influenza A and B viruses by 3 days in places where the temperature is
+low. After the 2002-2004 outbreak of SARS, researchers also began to
+investigate the relationship between these factors and SARS-CoV2
+(Casanova et al., 2010; Chan et al., 2011). Casanova et al.~(2010), for
+instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse
+hepatitis MHV viruses), to find that virus inactivation was more rapid
+at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of
+humidity, these researchers reported that survival of the virus was
+lower at moderate relative humidity levels. Chan et al.~(2011) also
+found that viability of the virus that causes SARS is also lost at
+higher temperatures (\textgreater38°C) and relative humidity superior to
+95\%.
+
+Whether these results from laboratory experiments will hold in general
+remains uncertain. At a global scale, de Ángel Solá et al.~(2020) see
+less risk in the Caribean Basin; however, Coelho et al.~(Coelho et al.,
+2020) warn that at least during the exponential phase, expansion of the
+virus is not driven by climate. Similarly, whereas Araujo and Naimi
+(2020) argue that spread of SARS-CoV2 will likely be constrained by
+climate, Harbert et al.~(2020) remain unconvinced that spatial modelling
+can currently discriminate the distribution of the disease on the basis
+of climate, at least in the United States. Yao et al.~(2020), examined
+data from China and came to the conclusion that neither temperature nor
+ultraviolet indices had an association with transmission of COVID-19.
+This is despite previous research that has linked less exposure to UVB
+radiation to higher prevalence and severity of acute respiratory tract
+infections (Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen
+et al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
+Moriyama, Hugentobler, and Iwasaki 2020).
+
+In addition to the environmental variables above, we also consider a set
+of potential socio-economic and demographic confounders.
 
@@
 
-The percentaje of older adults (over 65) is the second variable that we
-consider as a control. Early evidence regarding COVID-19 suggests that
-the case rate mortality is higher at older ages (e.g.~The Novel
-Coronavirus Pneumonia Emergency Response Epidemiology Team, 2020).
-However, it is not clear that a relatively large population of older
-adults necessarily translates into higher transmission of the infection.
-The tool of choice in containing the spread of the disease has been
-social distancing. In this respect, the evidence from the field of
-transportation is that older adults tend to travel less frequently, for
-shorter distances, and have higher rates of immobility (e.g., Roorda et
-al., 2010; Morency et al., 2011; Sikder and Pinjari, 2012). In other
-words, many older adults are, whether by preference or otherwise,
-already in a form of social isolation. Social distancing during the
-pandemic may actually reinforce that condition for them, as suggested by
-the analysis of age-structured social contact in India, China, and Italy
-of Sing and Adhikari (2020). Accordingly, our expectation is that
-provinces with higher percentages of older adults will tend to have
-similarly low levels of social contact, particularly since the
-age-structured matrix of social contact in Spain is similar to Italy
-(Prem et al., 2017).
-
-The third population variable the we consider is population density,
-since it directly affects the contact patterns and contact rates between
-individuals in a population (Hu et al., 2013). The evidence available
-suggests a positive relationship between the transmission of COVID-19
-and population density (e.g.~cumulative incidence in urban areas like
-NYC). For this reason, we anticipate a positive relationship between
-population density and the incidence of the disease.
-
-The last control variable is the presence of mass transit systems in a
-province. Every province in Spain offers some form of public
-transportation, however only five provinces have higher order systems of
-mass mobility (e.g.~metro or subway), namely Barcelona, Madrid, Sevilla,
-Valencia, and Bizkaia. Public transportation has been hypothesized to
-relate to the spread of contagious disease by some researchers using
-agent-based approaches and simulation (e.g., Perez and Dragicevic, 2009;
-Wang et al., 2010), and while we find scant evidence of a link in the
-literature, the idea is intuitively appealing. After all, unlike the
-isolation that a car offers to travellers, most mass transit system are
-cauldrons of social contact.
+Secondly, we consider percentage of older adults (over 65) in a region.
+Early evidence regarding COVID-19 suggests that the case rate mortality
+is higher at older ages (e.g.~The Novel Coronavirus Pneumonia Emergency
+Response Epidemiology Team, 2020). However, it is not clear that a
+relatively large population of older adults necessarily translates into
+higher transmission of the infection. The tool of choice in containing
+the spread of the disease has been social distancing. In this respect,
+the evidence from the field of transportation is that older adults tend
+to travel less frequently, for shorter distances, and have higher rates
+of immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder
+and Pinjari, 2012). In other words, many older adults are, whether by
+preference or otherwise, already in a form of social isolation. Social
+distancing during the pandemic may actually reinforce that condition for
+them, as suggested by the analysis of age-structured social contact in
+India, China, and Italy of Sing and Adhikari (2020). Accordingly, our
+expectation is that higher percentages of older adults will tend to have
+similarly low levels of social contact (the age-structured matrix of
+social contact in Spain is similar to Italy; see Prem et al., 2017).
+
+Population density is also relevant since it directly affects the
+contact patterns and contact rates between individuals in a population
+(Hu et al., 2013). The evidence available suggests a positive
+relationship between the transmission of COVID-19 and population density
+(e.g.~cumulative incidence in urban areas like NYC). For this reason, we
+anticipate a positive relationship between population density and the
+incidence of the disease.
+
+The last variable that we consider as a control is the presence of mass
+transit systems in a province. Every province in Spain offers some form
+of public transportation, however only five provinces have higher order
+systems of mass mobility (e.g.~metro or subway), namely Barcelona,
+Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been
+hypothesized to relate to the spread of contagious disease by some
+researchers using agent-based approaches and simulation (e.g., Perez and
+Dragicevic, 2009; Wang et al., 2010), and while we find scant evidence
+of a link in the literature, the idea is intuitively appealing. After
+all, unlike the isolation that a car offers to travellers, most mass
+transit system are cauldrons of social contact.
+
+\hypertarget{context-and-data}{%
+\section{Context and Data}\label{context-and-data}}
+
+\hypertarget{covid-19-in-spain}{%
+\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}
+
+The first reported case of COVID-19 in Spain was on January 31st, 2020,
+when a German tourist in the Canary Islands tested positive for the
+virus. However, it was still a few weeks before the first domestic case
+was reported, on February 27th in Sevilla province (Andalusia). In a
+short period of time, as testing started to ramp up, it became clear
+that an outbreak was flaring. By March 11th the World Health
+Organization (WHO) declared COVID-19 officially a pandemic. This
+declaration marked a turning point for the public in Spain too. As of
+March 13th, the number of cases of COVID-19 reported in Spain was 4,473,
+with a majority of cases (1,990) concentrated in Madrid: these numbers
+were at the time the worst outbreak in Europe after Italy. In response
+to the situation, on March 13th the Spanish National Government declared
+a state of emergency, to go into effect on Saturday March 14th. As part
+of the state of emergency restrictions to most activities were imposed,
+with the exception of essential services (e.g.~food, health) and some
+economic subsectors of industry and construction. A few days later, on
+March 17th, Spain closed its lands borders to allow entry only to
+returnee nationals and permanent residents. The lockdown was further
+hardened between March 30th and April 12th (including the Easter weekend
+of April 10th-12th) and during this period only essential activities
+were allowed. During this period, there was a dramatic reduction in
+overall mobility, both within provinces as between
+\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
 
@@
 Our dataset includes information about the daily number of cases of
-COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
+COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat
 terminology) for the period between March 13th and April 11th,
@@
 Chain Reaction (PCR) test. The Spanish National Government publishes
-periodic updates at the regional level (NUTII) and the information is
+periodic updates at the regional level (NUTS2) and the information is
 also released at the provincial level as part of a collaborative project
@@
 Table \ref{tab:descriptive-statistics}. The smallest province is
-\(1,978.12km^2\) (this is smaller than Rhode Island in the US) and the
+\(1,978.12km^2\) (this is smaller than Rhode Island in the U.S.) and the
 largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
-in the US). While this is a relatively large degree of spatial
-aggregation, reporting on COVID-19 is inconsistent at smaller
-geographies, or cases are not reported at that level at all. The
-analysis must therefore be considered ecological.
+in the U.S.). While this is a relatively large degree of spatial
+aggregation, reporting on COVID-19 is not yet consistent at smaller
+geographies, or cases are not reported at that level at all.
 
@@
 \addlinespace
-Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
+Mean Temperature & Mean temperature in province by date, in °C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
 \rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
+Sunshine & Daily hours of sunshine in province by date & 0.00 & 5.74 & 12.40 & 3.96 & AEMET\\
 \bottomrule
@@
 multivariate econometric model used in different fields when the
-structure of the data consisits of cross-sections for different time
+structure of the data consists of cross-sections for different time
 periods. The basis of this approach is well-known since the initial
@@
 Anselin (1988) was the first author to discuss SUR from a spatial
-perspective. In his landmark text, Anselin discussed a model made of
-``an equation for each time period, which is estimated for a cross
-section of spatial units'' (p.~141). From this milestone, a large body
-of research has developed to extend the classical SUR into a spatial
-framework (e.g., Rey and Montouri, 1999; Lauridsen et al., 2010; Le
-Gallo and Dall'Erba, 2006; López et al., 2017).
+perspective, in the context of spatio-temporal analysis. In his landmark
+text, Anselin discussed a model made of ``an equation for each time
+period, which is estimated for a cross section of spatial units''
+(p.~141). From this milestone, a large body of research has developed to
+extend the classical SUR into a spatial framework (e.g., Rey and
+Montouri, 1999; Lauridsen et al., 2010; Le Gallo and Dall'Erba, 2006;
+López et al., 2017).
 
@@
 
-A key feature of the SUR model is the dependence structure among the
-vectors of residuals, namely:
+A key feature of the SUR model is the temporal dependence structure
+among the vectors of residuals, namely:
 
@@
 
-\hypertarget{analysis}{%
-\section{Analysis}\label{analysis}}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 distribution of mean temperatures. The lowest mean temperature for a
-community on any given day was 2.8C, and the highest 22.4C, for a range
-of approximately 20 degrees. Likewise, there is a great deal of
+community on any given day was 2.8°C, and the highest 22.4°C, for a
+range of approximately 20 degrees. Likewise, there is a great deal of
 variability in humidity, as seen in the middle panel of the figure,
@@
 relatively spread, and on some days, the simple correlation between
-density and incidence of COVID-19 is weakly positive.
+density and incidence of COVID-19 is weakly positive. Overall for the
+period under examination, the pairwise correlations between these
+variables and incidence are significant at \(p<0.05\), with the
+exception of the three \(Sunshine_hours\) variables.
 
@@
 weights \(W\). In this case, we consider neighborhoods based on
-adjacency, based on the well-known rook criterion (two provinces are
-adjacent if they share a boundary, but not if they touch only at a
-vertex). We modify this criterion in two cases. First, we make an
-allowance for adjacency between the two islands in the Autonomous
-Community of Canarias in the Pacific (Las Palmas and Santa Cruz de
-Tenerife), which we assume are adjacent between each other. And
-secondly, we assume that Islas Baleares in the Mediterranean are
-adjacent to three provinces in Pais Catalans (i.e., Barcelona,
-Tarragona, and Castello). After matrix \(W\) has been specified,
-estimation of the model can proceed as usual.
+adjacency, based on the well-known queen criterion (two provinces are
+adjacent if they share a boundary or touch at a vertex). We modify this
+criterion in two cases. First, we make an allowance for adjacency
+between the two islands in the Autonomous Community of Canarias in the
+Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are
+adjacent between each other. And secondly, we assume that Islas Baleares
+in the Mediterranean are adjacent to three provinces in Pais Catalans
+(i.e., Barcelona, Tarragona, and
+Castello)\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and removed the three islands in the sample.
+The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
 
@@
 independent variables. The only variable that is not transformed is the
-categorical variable for transity systems. A log-log transformation is
+categorical variable for transit systems. A log-log transformation is
 appropriate to capture non-linear relationships between variables and
@@
 
-\hypertarget{results-and-discussion}{%
-\section{Results and Discussion}\label{results-and-discussion}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 \midrule
-\rowcolor{gray!6}  Intercept & 7.370 & 10.037 & 12.968 & 0 & 0 & 30 & Non-constrained\\
-log(GDPpc) & 0.620 & 0.620 & 0.620 & 0 & 0 & 1 & Constrained\\
-\rowcolor{gray!6}  log(Older) & -0.737 & -0.737 & -0.737 & 0 & 0 & 1 & Constrained\\
-log(Density) & -0.220 & -0.097 & 0.153 & 19 & 0 & 11 & Non-constrained\\
-\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 8 & 8 & 14 & Non-constrained\\
+\rowcolor{gray!6}  Intercept & 6.878 & 9.889 & 15.294 & 0 & 0 & 30 & Non-constrained\\
+log(GDPpc) & 0.637 & 0.637 & 0.637 & 0 & 0 & 1 & Constrained\\
+\rowcolor{gray!6}  log(Older) & -0.854 & -0.854 & -0.854 & 0 & 0 & 1 & Constrained\\
+log(Density) & -0.213 & -0.101 & 0.101 & 18 & 3 & 9 & Non-constrained\\
+\rowcolor{gray!6}  Transit & 0.326 & 0.487 & 0.593 & 10 & 12 & 8 & Non-constrained\\
 \addlinespace
-log(Humidity) & -1.434 & -0.534 & -0.031 & 10 & 1 & 19 & Non-constrained\\
-\rowcolor{gray!6}  log(Temperature) & -2.014 & -1.406 & -0.929 & 0 & 0 & 30 & Non-constrained\\
-log(Sunshine) & -0.258 & 0.097 & 0.206 & 7 & 2 & 21 & Non-constrained\\
-\rowcolor{gray!6}  Spatially lagged y (rho) & 0.015 & 0.142 & 0.528 & 12 & 3 & 15 & Non-constrained\\
+log(Humidity) & -2.159 & -0.456 & 0.193 & 15 & 2 & 13 & Non-constrained\\
+\rowcolor{gray!6}  log(Temperature) & -2.006 & -1.496 & -0.988 & 0 & 0 & 30 & Non-constrained\\
+log(Sunshine) & -0.252 & 0.078 & 0.177 & 9 & 1 & 20 & Non-constrained\\
+\rowcolor{gray!6}  Spatially lagged y (rho) & 0.056 & 0.236 & 0.548 & 2 & 5 & 23 & Non-constrained\\
 \bottomrule
@@
 autocorrelation coefficient (\(\rho\)) in Figure
-\ref{fig:results-time-1} (left panel). It is interesting to note that
-the magnitude of the spatial autocorrelation coefficient \(\rho\)
-declines over the period under analysis, and is not significant for some
-days. This is an interesting result: immediately prior to the
-declaration of the state of emergency, there appears to have been a
-strong inter-provincial contagion effect. Keeping in mind that the
-incubation period ranges between 2 and 11 days with a median of 5, it is
-reasonable to expect that the effect of the lockdown will be observed
-with some delay. Indeed, as seen in the figure, the autocorrelation
-coefficient remains high for three days, then declines on March 16, and
-continues to weaken over time. At the end of the period under
-examination, the strength of this effect is much diminished. We would
-expect that under full compliance with strict lockdown conditions
-(meaning no inter-provincial mobility) the spatial contagion effect
-would be zero. The fact that it is not is evidence of some weak
-contagion effect due to restrictions to the movement between provinces.
+\ref{fig:results-time-1} (left panel). We notice that the magnitude of
+the spatial autocorrelation coefficient \(\rho\) declines over the
+period under analysis, and is not significant for some days. This is an
+interesting result: immediately prior to the declaration of the state of
+emergency, there appears to have been a strong inter-provincial
+contagion effect. Keeping in mind that the incubation period ranges
+between 2 and 11 days with a median of 5, it is reasonable to expect
+that the effect of the lockdown will be observed with some delay.
+Indeed, as seen in the figure, the autocorrelation coefficient remains
+high for three days, then declines on March 16, and continues to weaken
+over time. At the end of the period under examination, the strength of
+this effect is much diminished and we would expect that under full
+compliance with strict lockdown conditions (meaning no inter-provincial
+mobility) the spatial contagion effect would be zero - as seems to be
+the case.
 
@@
 accordingly - while the opposite may be true of residents in low density
-regions. Anecdotally, this seems to have been the case with some
-governors in the US, who resisted stay-at-home orders citing their
-sparse
-\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}
-and rural
-populations\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}.
-The coefficient for Transit is positive, as expected, but with very wide
-confidence intervals, and in fact not significant in the earlier part of
-the period.
+regions. The coefficient for Transit is positive, as expected, but with
+very wide confidence intervals, and in fact not significant in the
+earlier part of the period.
 
@@
 effect of GDP per capita is to increase the incidence of COVID-19 by
-0.675\% for each percentual increase of this variable (in €1,000s). In
-our view, this is a measure of inertia, as provinces with higher GDP per
-capita where among the first to see exponential growth in the pandemia.
-Percentage of older adults has a negative effect, and each percentual
+0.675\% for each percent increase of this variable (in €1,000s). In our
+view, this is a measure of inertia, as provinces with higher GDP per
+capita where among the first to see exponential growth in the pandemic.
+Percentage of older adults has a negative effect, and each percent
 increase in this variable is associated with a relatively small
@@
 effect is Transit, and the largest direct negative effects are
-temperature and humidity. The direct effect of these variables is as
-follows: for each percentual point increase in temperature, there is
+temperature and humidity. The direct effects of these variables are as
+follows: for each percent point increase in temperature, there is
 between a 1\% and 2\% reduction in the incidence of the disease. This
@@
 to 1.5\%. The effect of humidity is also to reduce the incidence: each
-percentual point of increase in humidity is associated with a reduction
-of up to 1.5\% in incidence. With the addition of the indirect effect,
-the total effect of a 1\% increase in humidity is to reduce incidence by
-up to 3\%. As seen in the figure, the indirect (i.e., contagion) effects
+percent point of increase in humidity is associated with a reduction of
+up to 1.5\% in incidence. With the addition of the indirect effect, the
+total effect of a 1\% increase in humidity is to reduce incidence by up
+to 3\%. As seen in the figure, the indirect (i.e., contagion) effects
 are almost as strong as the direct effects immediately before and at the
-beginning of the lockdown period. However, by the end of the lockdown
-period, the indirect effects have weakened considerably. The effect of
-density goes from positive (+0.334) to negative (-0.244). It is crucial
-to keep in mind here that this is the effect of density over a period of
-\emph{lockdown} and possibly subject to risk compensation, as discussed
-above. In other words, the lockdown seems to be more effective in places
-with higher density. Accordingly, cities should not rush to ease
-lockdowns, but rather efforts should be made to reduce or eliminate risk
-compensation in less densely populated regions.
+beginning of the lockdown period. Nonetheless, by the end of the
+lockdown period, the indirect effects have weakened considerably. The
+effect of density goes from positive (+0.334) to negative (-0.244). It
+is crucial to keep in mind here that this is the effect of density over
+a period of \emph{lockdown} and possibly subject to risk compensation,
+as discussed above which suggests, in turn, that efforts should be made
+to reduce or eliminate risk compensation in less densely populated
+regions.
 
 What do these effects mean? Under a situation of lockdown,
-inter-regional contagion is reduced, as expected, and the effect of
-temperature is estimated to reduce the incidence of COVID-19 by between
-one and two percent with each percentual increase of this variable.
-Without a lockdown, or a relaxed lockdown, the effect is even greater,
-but only because the disease is transmitted inter-regionally! Imagine a
-region that is closed to its neighbors: increases in temperature are
-associated with smaller reductions in incidence - however, it does not
-import or export cases. A province that is not closed and not under
-lockdown will see a greater effect of temperature, but at the cost of
-importing/exporting cases.
+inter-regional contagion is reduced, as expected, and the total effects
+of the variables tend towards their direct effects. In those first few
+days covered by our analysis the total effect of all variables is
+greater due to the spatial contagion effect. This also makes analysis
+and intervention more complex: the contagion effect essentially acts
+like a multiplier, whereby developments in each province spill over to
+their neighbors. Once the contagion effect has been tamed, each province
+can be ``treated'' independently from its neighbors.
+
+It is important, before concluding our discussion, to highlight some
+limitations of this study.
+
+First, the analysis was conducted under a situation of lockdown, and
+therefore, besides the results at the beginning of the period examined,
+one must exercise caution when trying to extrapolate the findings to
+other situations. Secondly, it is well known that there is in many
+countries substantial underreporting of cases of COVID-19 due to limited
+testing. In this case we do not suspect systematic provincial bias in
+reporting, and as long as the underreporting is consistent across units
+of analysis, we do not expect biased results - even if we are aware that
+the number of true cases is likely higher. Finally, we defined
+neighborhoods based on adjacency. It would be interesting to compare
+other connectivity criteria, for instance based on domestic
+transportation instrastructure and services. We flag this as a matter
+for future research.
 
@@
 \multicolumn{5}{l}{\textbf{Direct Effects}}\\
-\hspace{1em}log(GDPpc) & 0.674 & 0.674 & 0.674 & Constrained\\
-\hspace{1em}log(Older) & -0.801 & -0.801 & -0.801 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.166 & Non-constrained\\
-\hspace{1em}Transit & 0.342 & 0.517 & 0.584 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.559 & -0.544 & -0.031 & Non-constrained\\
-\hspace{1em}log(Temperature) & -2.019 & -1.420 & -0.929 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.281 & 0.095 & 0.206 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.694 & 0.694 & 0.694 & Constrained\\
+\hspace{1em}log(Older) & -0.930 & -0.930 & -0.930 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.213 & -0.102 & 0.110 & Non-constrained\\
+\hspace{1em}Transit & 0.351 & 0.496 & 0.602 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -2.351 & -0.475 & 0.194 & Non-constrained\\
+\hspace{1em}log(Temperature) & -2.184 & -1.528 & -0.990 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.275 & 0.077 & 0.180 & Non-constrained\\
 \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Indirect Effects}}\\
-\hspace{1em}log(GDPpc) & 0.639 & 0.639 & 0.639 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.759 & -0.759 & -0.759 & Constrained\\
-\hspace{1em}log(Density) & -0.032 & -0.004 & 0.158 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.342 & Non-constrained\\
-\hspace{1em}log(Humidity) & -1.478 & -0.157 & -0.001 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.647 & -0.275 & -0.014 & Non-constrained\\
-\hspace{1em}log(Sunshine) & -0.266 & -0.006 & 0.028 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.718 & 0.718 & 0.718 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.962 & -0.962 & -0.962 & Constrained\\
+\hspace{1em}log(Density) & -0.115 & -0.020 & 0.114 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}Transit & 0.035 & 0.158 & 0.447 & Non-constrained\\
+\hspace{1em}log(Humidity) & -2.431 & -0.279 & 0.023 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -2.259 & -0.538 & -0.095 & Non-constrained\\
+\hspace{1em}log(Sunshine) & -0.284 & 0.001 & 0.059 & Non-constrained\\
 \rowcolor{gray!6}  \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Total Effects}}\\
-\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
-\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
-\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
-\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 1.411 & 1.411 & 1.411 & Constrained\\
+\hspace{1em}log(Older) & -1.892 & -1.892 & -1.892 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.241 & -0.122 & 0.224 & Non-constrained\\
+\hspace{1em}Transit & 0.530 & 0.655 & 0.905 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -4.782 & -0.754 & 0.217 & Non-constrained\\
+\hspace{1em}log(Temperature) & -4.442 & -2.066 & -1.087 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.559 & 0.078 & 0.238 & Non-constrained\\
 \bottomrule
@@
 
-It is important, in this respect, to highlight some limitations of this
-study.
-
-First, the analysis was conducted under a situation of lockdown, and
-therefore, besides the results are the beginning of the period examined,
-one must exercise caution when trying to extrapolate the findings to
-other situations. Secondly, it is well known that there is in many
-countries substantial underreporting of cases of COVID-19 due to limited
-testing. However, we do not suspect systematic provincial bias in
-reporting, and as long as the underreporting is consistent across units
-of analysis, we do not expect biased results - even if we are aware that
-the number of true cases is likely higher. Thirdly, all environmental
-data are based on a single station in a province. Here, we made an
-effort to select the most representative station based on the
-distribution of the population. An alternative approach would be to
-calculate the environmental variables based on the average of multiple
-stations within each province: this would likely smooth to some extent
-these variables. Finally, we defined spatial neighborhoods based on
-adjacency. It would be interesting to compare other connectivity
-criteria, for instance based on domestic transportation instrastructure
-and services. We flag this as a matter for future research.
+Some directions for future research include investigating other
+modelling frameworks, such as geographically and temporally weighted
+regression and/or space-time conditional autoregressive models. In
+addition, the environmental variables examined here relate to
+meteorological conditions, and did not include other environmental
+factors that may incide in the transmission of the virus, such as
+pollution. Other factors could be incorporated in future studies.
 
@@
 
+\leavevmode\hypertarget{ref-Lewnard2020scientific}{}%
+Lewnard, J.A., Lo, N.C., 2020. Scientific and ethical basis for
+social-distancing interventions against covid-19. The Lancet. Infectious
+diseases S1473--3099(20)30190--0.
+doi:\href{https://doi.org/10.1016/S1473-3099(20)30190-0}{10.1016/S1473-3099(20)30190-0}
+
 \leavevmode\hypertarget{ref-Lopez2017spatial}{}%
@@
 
+\leavevmode\hypertarget{ref-Wilder2020isolation}{}%
+Wilder-Smith, A., Freedman, D.O., 2020. Isolation, quarantine, social
+distancing and community containment: Pivotal role for old-style public
+health measures in the novel coronavirus (2019-nCoV) outbreak. Journal
+of Travel Medicine 27.
+doi:\href{https://doi.org/10.1093/jtm/taaa020}{10.1093/jtm/taaa020}
+
 \leavevmode\hypertarget{ref-Xie2014}{}%

### 2020-05-06 Visualization-Contagion/Visualizing-Contagion.Rmd
First round of revisions after invitation to revise and resubmit.

@@
 Render animation:
-```{r cache=TRUE}
+```{r cache=TRUE, eval=FALSE}
 animate(anim1, fps = 2, end_pause = 5)
@@
 Where could this analysis be taken? LISA analysis by date? Spatial Markov Chains? Cluster membership?
+
+## LISA analysis of contagion
+
+Here I calculate LISA statistics for `Contagion` on each date:
+```{r}
+junk <- Contagion %>%
+  group_by(Date) %>%
+  summarize(Ii =, 
+            P_Ii = )
+localmoran(Contagion$Contagion[Contagion$Date == "2020-03-13"], listw)[,5]
+```
+
+LISA statistic calculated for each cross-section
+```{r}
+Ii <- Contagion %>% 
+  group_by(Date) %>%
+  group_modify(~ {(localmoran(.x$Contagion, listw)[,1]) %>%
+      enframe()}) %>%
+  ungroup()  %>%
+  transmute(Date,
+            ID_INE = rep(c(1:50), 30),
+            Ii = value)
+```
+
+LISA's p-value:
+```{r}
+P_Ii <- Contagion %>% 
+  group_by(Date) %>%
+  group_modify(~ {(localmoran(.x$Contagion, listw)[,5]) %>%
+      enframe()}) %>%
+  ungroup() %>%
+  transmute(Date,
+            ID_INE = rep(c(1:50), 30),
+            P_Ii = value)
+```
+
+Join the results of the LISA analysis to the dataframe:
+```{r}
+Contagion <- Contagion %>%
+  left_join(Ii, by = c("Date" = "Date", "ID_INE" = "ID_INE")) %>%
+  left_join(P_Ii, by = c("Date" = "Date", "ID_INE" = "ID_INE"))
+```
+
+Label provinces by whether they are positive or negative, and whether they are significant or not:
+```{r}
+Contagion <- Contagion %>%
+  mutate(LISA = case_when(Ii <= 0 & P_Ii <= 0.05 ~ "Hot Spot",
+                          Ii <= 0 & P_Ii > 0.05 ~ "NS (Hot)",
+                          Ii > 0 & P_Ii <= 0.05 ~ "Cold Spot",
+                          Ii > 0 & P_Ii > 0.05 ~ "NS (Cold)"),
+         LISA = factor(LISA, 
+                       levels = c("Hot Spot",
+                                  "NS (Hot)",
+                                  "NS (Cold)",
+                                  "Cold Spot"),
+                       ordered = TRUE))
+```
+
+Create a choropleth plot of the provinces based on the results of the LISA analysis:
+```{r}
+p2 <- ggplot() + 
+  geom_sf(data = Contagion %>% 
+            filter(CCAA != "Canarias"), 
+          aes(fill = LISA, 
+              group = interaction(factor(ID_INE), 
+                                  Date))) +
+  scale_fill_manual(values = c("Hot Spot" = "red", 
+                               "NS (Hot)" = "pink", 
+                               "NS (Cold)" = "royalblue", 
+                               "Cold Spot" = "royalblue4")) + 
+  theme_bw()
+```
+
+Create an animation using `gganimate`:
+```{r}
+anim2 <-  p2 +
+  labs(title = "Date: {frame_time}") +
+  transition_time(Date)
+```
+
+Render animation:
+```{r cache=TRUE, eval=FALSE}
+animate(anim2, fps = 2, end_pause = 5)
+```
+
+## Spatial Markov Chains
+
+I used to have Matlab code for spatial markov chains (SMC). It is gone. Here' I will try to reproduce it but now in `R`.
+
+The first step to implement SMC is to discretize the variable of interest.
+This can be done in two ways: within period quantiles, and pooled quantiles for the whole time series.
+According to [Rey and Kang](https://pysal.org/notebooks/explore/giddy/Markov_Based_Methods.html), these two forms of doing the analysis give different interpretations: pooled quantiles are absolute trends, whereas within period quantiles are relative dynamics.
+
+Here, we label each case by their pooled quintile (`q_p`):
+```{r}
+Contagion <- Contagion %>%
+  mutate(q_p = cut(Contagion, 
+                   quantile(Contagion, c(0, 0.2, 0.4, 0.6, 0.8, 1)), 
+                   labels = c("High Contagion", 
+                              "Medium-High Contagion", 
+                              "Medium Contagion", 
+                              "Medium-Low Contagion", 
+                              "Low Contagion"))) %>%
+  replace_na(replace = list(q_p = "High Contagion"))
+```
+
+To label each case by their within day quintile (`q_d`) we need to summarize the quintiles by date:
+```{r}
+quintiles_by_date <- Contagion %>%
+  st_drop_geometry() %>%
+  group_by(Date) %>%
+  summarize(x20 = quantile(Contagion, 0.2),
+            x40 = quantile(Contagion, 0.4),
+            x60 = quantile(Contagion, 0.6),
+            x80 = quantile(Contagion, 0.8))
+```
+
+Join the quintiles by date to the dataframe:
+```{r}
+Contagion <- Contagion %>%
+  left_join(quintiles_by_date, by = "Date")
+```
+
+Label the observations according to their values of `Contagion` and the quintiles by date:
+```{r}
+Contagion <- Contagion %>%
+  mutate(q_d = case_when(Contagion <= x20 ~ "High Contagion",
+                         Contagion > x20 & Contagion <= x40 ~ "Medium-High Contagion",
+                         Contagion > x40 & Contagion <= x60 ~ "Medium Contagion",
+                         Contagion > x60 & Contagion <= x80 ~ "Medium-Low Contagion",
+                         Contagion > x80 ~ "Low Contagion"),
+         q_d = factor(q_d, 
+                      levels = c("High Contagion", 
+                              "Medium-High Contagion", 
+                              "Medium Contagion", 
+                              "Medium-Low Contagion", 
+                              "Low Contagion"),
+                      ordered = TRUE))
+```

### 2020-05-06 covid19env/data_processing/00-Donwload-meteo-data-from-AEMET.Rmd
First round of revisions after invitation to revise and resubmit.

--- 2020-05-04 Donwload_climatic_data_from_AEMET.Rmd
@@
 library(meteoland)
+#devtools::install_github("SevillaR/aemet")
 library(aemet)
+library(lubridate)
+library(nngeo)
 library(sf)
 library(kableExtra)
-library(ggplot2)
+library(tidyverse)
 ```
@@
 ```{r}
-load("/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/covid19env/data/provinces_spain.RData")
+load("provinces_spain.RData")
 ```
@@
 ```{r}
-num_prov <- 5
-Tem <- list()
-for (i in 1:num_prov){
-  Tem[[i]] <- downloadAEMEThistorical(apikey, mydates, station_id=provinces_spain$Meteo_Station[i])[[1]]
-  print(i)
-  Sys.sleep(2)
-}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[1:40])
 ```
 
-## Example: List information
-En la lista Tem se obtienen los valores climático
+Extract meteorological data. First initialize the dataframe:
+```{r}
+df <- data.frame(indicativo = character(), 
+                 long = numeric(), 
+                 lat = numeric(), 
+                 date = ymd(), 
+                 Precipitation = numeric(), 
+                 MeanTemperature = numeric(), 
+                 MinTemperature = numeric(), 
+                 MaxTemperature = numeric(), 
+                 WindDirection = numeric(), 
+                 WindSpeed = numeric(), 
+                 SunshineHours = numeric()) 
+```
 
+Extract the data tables for these stations and bind to the dataframe:
 ```{r}
-kableExtra::kable(head(Tem[[1]]))
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
 ```
+
+Next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[41:80])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[81:120])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[121:160])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Retrieve the next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[161:200])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Retrieve the next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[201:240])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Retrieve the next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[241:280])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Retrieve the next set of stations:
+```{r}
+Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[281:291])
+```
+
+Extract the data tables for these stations and bind to the dataframe:
+```{r}
+coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
+for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
+                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
+                                                     long = rep(coords$long[1], 51),
+                                                     lat = rep(coords$lat[1], each = 51),
+                                                     date = Tem1@dates,
+                                                     Tem@data[[i]]))}
+```
+
+Rename dataframe:
+```{r}
+meteo_data <- df %>% left_join(stations.sf %>% st_drop_geometry() %>% transmute(indicativo, Lon, Lat)) %>%
+  select(-c(long, lat))
+```
+
+Save to RData file:
+```{r}
+save(meteo_data, file = "meteo_data.RData")
+```

### 2020-05-06 covid19env/data_processing/01-Preprocess-meteo-data.Rmd
First round of revisions after invitation to revise and resubmit.

---
title: "Download climatic data from AEMET"
author: "Fernando"
date: "4/23/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse = TRUE,message = FALSE)
```

```{r libraries}
library(meteoland)
#devtools::install_github("SevillaR/aemet")
library(aemet)
library(lubridate)
library(nngeo)
library(sf)
library(kableExtra)
library(tidyverse)
```

## Load data:

```{r}
load("provinces_spain.RData")
```

## Geometry of Meteo Stations

```{r plot-stations}
ggplot(data = provinces_spain) +
  geom_sf() +
  theme_bw(base_size=6)
```

Load RData file (this is the outcome of the file `00-Download-meteo-data-from-AEMET.Rmd`; check that file for details on how these data were downloaded):
```{r}
load("meteo_data.RData")
```

Convert `meteo_data` to simple features:
```{r}
meteo_data.sf <- meteo_data %>% st_as_sf(coords = c("Lon", "Lat"), crs = 4326)
```

Join the provincial identifier to calculate the meteorological variables as the average of stations in province:
```{r}
meteo_data_joined = st_join(meteo_data.sf, provinces_spain["ID_INE"])
```

Number of stations per day per province:
```{r}
meteo_data_joined %>% 
  group_by(ID_INE, date) %>% 
  summarize(n = n()) %>%
  summary()
```

Provincial averages will use between 1 and 11 stations.

Obtain provincial averages:
```{r}
meteo_data_province <- meteo_data_joined %>% 
  st_drop_geometry() %>%
  group_by(ID_INE, date) %>%
  summarize(Precipitation = mean(Precipitation, na.rm = TRUE),
            MeanTemperature = mean(MeanTemperature, na.rm = TRUE),
            MinTemperature = mean(MinTemperature, na.rm = TRUE),
            MaxTemperature = mean(MaxTemperature, na.rm = TRUE),
            SunshineHours = mean(SunshineHours, na.rm = TRUE))
```

Summarize the dataframe with the provincial daily averages of the meteo variables: 
```{r}
summary(meteo_data_province)
```
There are some NAs, due to some stations not having been read.

Load the dataset that Fernando produced originally:
```{r}
load("COVID-19-Datos_Climaticos_Spain-Extendido.RData")
```

Arrange the hours of sunshine:
```{r}
# Hours of sunshine 
Sun <- Sun %>% select(-ID, -provincia) %>%
  pivot_longer(starts_with("S"), 
               names_to = "Date",
               values_to = "Sunshine_Hours") %>%
  mutate(Date = rep(seq(dmy("01-03-2020"), 
                        dmy("11-04-2020"), 
                        by = "days"), 
                    50))
```

Join to interpolated meteo variables:
```{r}
meteo_data_province <- meteo_data_province %>% 
  left_join(Sun, by = c("ID_INE" = "ID_INE", "date" = "Date"))
```

Replace NAs with the values from Fernando's table:
```{r}
meteo_data_province <- meteo_data_province %>%
  mutate(SunshineHours = ifelse(is.na(SunshineHours), Sunshine_Hours, SunshineHours))
```

Drop meteo data not assigned to a province:
```{r}
meteo_data_province <- meteo_data_province %>%
  drop_na(ID_INE)
```

Summarize again after dropping stations that were not used for any province:
```{r}
summary(meteo_data_province)
```

Tidy up and save:
```{r}
meteo_data_province <- meteo_data_province %>%
  select(-Sunshine_Hours)
```

```{r}
summary(meteo_data_province)
```

What are the NAs in Precipitation?
```{r}
meteo_data_province %>%
  filter(is.na(Precipitation))
```

Replace the NAs by zeros:
```{r}
meteo_data_province <- meteo_data_province %>%
  mutate(Precipitation = ifelse(is.na(Precipitation), 0, Precipitation))
```

What are the NAs in Sunshine? These are dates that we are not using so we can ignore.
```{r}
meteo_data_province %>%
  filter(is.na(SunshineHours))
```

```{r}

save(meteo_data_province, file = "meteo_data_province.RData", compress = "xz")
```

### 2020-05-06 COVID-19 Municipalities v0.Rmd

---
title: "COVID-19 Municipalities"
author: "Fernando"
date: "5/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, collapse=TRUE, message=FALSE}
library(sf)
library(ggplot2)
library(gridExtra)
```
## leer datos y geometrias

### lectura geometrias
Geometría de España por municipios
```{r, message=FALSE}
dir <-"/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/"
municipios.sf <- st_read(paste0(dir,"Municipios_ETRS89_30N.shp"))
municipios.sf$CODIGO <- as.numeric(as.character(municipios.sf$CODIGO))
```

Extraer shp para regiones
```{r}
# Regional-Shp
cat.sf <- municipios.sf[municipios.sf$COD_CCAA=="09",]
pv.sf <- municipios.sf[municipios.sf$COD_CCAA=="16",]
```

Geometría de Madrid

El municipio de Madrid está dividido en distritos
```{r geometria_madrid}
madrid.sf <- st_read(paste0(dir,"municipios_y_distritos_madrid.shp"))
madrid.sf$cod <- as.numeric(as.character(madrid.sf$cod))
```

### leer datos. Casos diarios por municipios

Diferentes periodos para cada region

```{r}
cat <- read.csv(file=paste0(dir,"COVID-19-Municipios (Cataluña).csv"),header = TRUE,sep=";")
pv <- read.csv(file=paste0(dir,"COVID-19-Municipios (Pais Vasco).csv"),header = TRUE,sep=";")
mad <- read.csv(file=paste0(dir,"COVID-19-Municipios (Madrid).csv"),header = TRUE,sep=";")
names(mad)[1]<-"cod"
```

## Cataluña

```{r}
cat.sf <- merge(cat.sf,cat,id.x=CODIGO,id.y=CODIGO,all.x=TRUE)
Inc <- sf::st_drop_geometry(cat.sf[,14:78])*100000/cat.sf$Pob
names(Inc) <- gsub("D","I",names(Inc))
Inc$CODIGO <- cat.sf$CODIGO
cat.sf <- merge(cat.sf,Inc,id.x="CODIGO",id.y="CODIGO",id.all=TRUE)

q <- quantile(cat.sf$I30Mar)
cat.sf$Quantile<- as.factor((cat.sf$I30Mar > q[2]) + (cat.sf$I30Mar > q[3]) +(cat.sf$I30Mar >= q[4]) + 1)
p1 <- ggplot(data = cat.sf) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))

q <- quantile(cat.sf$I30Abr)
cat.sf$Quantile<- as.factor((cat.sf$I30Abr > q[2]) + (cat.sf$I30Abr > q[3]) +(cat.sf$I30Abr >= q[4]) + 1)
p2 <- ggplot(data = cat.sf) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))

# gridExtra::grid.arrange(p1,p2,nrow=1)
```
** Cataluña 30 de Marzo**
```{r cat-p1}
p1
```
** Cataluña 30 de Abril**
```{r cat-p2}
p2
```

## País Vasco

**Fechas 20 Marzo - 26 Abril**
```{r}
# lincar la geometría con la base de datos de casos
pv.sf <- merge(pv.sf,pv,id.x=CODIGO,id.y=CODIGO,all.x=TRUE)
# Calcular la incidencia
Inc <- sf::st_drop_geometry(pv.sf[,14:51])*100000/pv.sf$Pob
names(Inc) <- gsub("D","I",names(Inc))
Inc$CODIGO <- pv.sf$CODIGO
# lincar la incidencia con la geometría
pv.sf <- merge(pv.sf,Inc,id.x="CODIGO",id.y="CODIGO",id.all=TRUE)
```

```{r}
q <- quantile(pv.sf$I20Mar)
pv.sf$Quantile<- as.factor((pv.sf$I20Mar > q[2]) + (pv.sf$I20Mar > q[3]) +(pv.sf$I20Mar >= q[4]) + 1)
p1 <- ggplot(data = pv.sf) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))

q <- quantile(pv.sf$I20Abr)
pv.sf$Quantile<- as.factor((pv.sf$I20Abr > q[2]) + (pv.sf$I20Abr > q[3]) +(pv.sf$I20Abr >= q[4]) + 1)
p2 <- ggplot(data = pv.sf) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
# gridExtra::grid.arrange(p1,p2,nrow=1)
```

** País Vasco 20 de Marzo - 20 Abril**
```{r pv-p1}
p1
```
** País Vasco 26 de Abril**
```{r pv-p2}
p2
```

## Madrid

**Fechas 8 Abril - 04 Mayo**
```{r}
# lincar la geometría con la base de datos de casos
madrid.sf <- merge(madrid.sf,mad,id.x=CODIGO,id.y=CODIGO,all.x=TRUE)
# Calcular la incidencia
Inc <- sf::st_drop_geometry(madrid.sf[,12:38])*100000/pv.sf$Pob
names(Inc) <- gsub("D","I",names(Inc))
Inc$CODIGO <- madrid.sf$CODIGO
# lincar la incidencia con la geometría
madrid.sf <- merge(madrid.sf,Inc,id.x="CODIGO",id.y="CODIGO",id.all=TRUE)
```

```{r}
q <- quantile(madrid.sf$I08Abr)
madrid.sf$Quantile<- as.factor((madrid.sf$I08Abr > q[2]) + (madrid.sf$I08Abr > q[3]) +(madrid.sf$I08Abr >= q[4]) + 1)
p1 <- ggplot(data = madrid.sf) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))

q <- quantile(madrid.sf$I04May)
madrid.sf$Quantile<- as.factor((madrid.sf$I04May > q[2]) + (madrid.sf$I04May > q[3]) +(madrid.sf$I04May >= q[4]) + 1)
p2 <- ggplot(data = madrid.sf) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
```
**Madrid de 8 de Abril**
```{r mad-p1}
p1
```
**Madrid de 4 de Mayo**
```{r mad-p2}
p2
```

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Response-to-Reviewers-1.Rmd
Proof-reading the morning after.

@@
 >*The purpose of the daily correlation analysis in Figure 4 is to see how much day to day variability there is in the statistical associations with incidence.
-The p-values of of these correlation coefficients are somewhat misleading, because of the smaller samples used in their calculation.
+The p-values of these correlation coefficients are somewhat misleading because of the smaller samples used in their calculation.
 The information from this Figure helped us decide which variables to constrain in the SUR systems.
 At that point, the significance of the correlation is given by the p-values of the coefficients of the model.
-Instead of reporting the p-values for the daily correlations, we added a sentence to the effect that overall for the period  under examination, the pairwise correlations  between the independent variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables.
+Instead of reporting the p-values for the daily correlations, we added a sentence to the effect that overall for the period  under examination, the pairwise correlations  between the independent variables and incidence are significant at $p<0.05$, with the exception of the three sunshine variables.
 More details concerning the correlation tests can be found in the R markdown document in the chunk `correlation-tests`.*
@@
 When using these variables, instead of seeing improvements on accuracy and precision, the goodness-of-fit of the models declines, the intervals of confidence become wider, and consequently several coefficients lose significance.
-It seems to us that the interpolated variables are more noisy than the manually selected stations.
-It is possible that a different interpolation algorithm might improve this, but that would involve experimenting with several interpolation methods to derive several environmental variables for a period of 51 days.
-For verification purposes, we have recreated the data package to include both datasets (single station and average of stations).
+It seems to us that the interpolated variables introduce undesirable noise: on the one hand, they smooth the environmental variables over the province, but on the other hand they also include stations in places that are not necessarily informative (such as airports, up in Sierra Nevada, etc.) Having selected the stations manually seems to have kept this information "crisp".
+We recognize the possibility that a different interpolation algorithm might improve this, and recommend this as a topic for future research.
+A proper comparison of interpolation algorithms would involve experimenting with several methods to derive several environmental variables for a period of 51 days.
+For the purpose of this paper, we have recreated the data package to include both datasets (single station and average of stations).
 In the paper we discuss only the results of the first.*

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Proof-reading the morning after.

@@
 At a global scale, de Ángel Solá et al.
-[-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al.
+[-@deangel2020weathering] see less risk in the Caribean Basin; on the other hand, Coelho et al.
 [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate.
@@
 
-In addition to the environmental variables above, we also consider a set of potential socio-economic and demographic confounders.
-
-The first variable that we consider is GDP per capita.
+In addition to the environmental variables above, it is important to account for potential socio-economic and demographic confounders.
+
+In this respect, the first variable that we consider is GDP per capita.
 Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
@@
 The last variable that we consider as a control is the presence of mass transit systems in a province.
-Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g.
+Every province in Spain offers some form of public transportation, but only five provinces have higher order systems of mass mobility (e.g.
 metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
@@
 The first reported case of COVID-19 in Spain was on January 31st, 2020, when a German tourist in the Canary Islands tested positive for the virus.
-However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
+After this case, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
 In a short period of time, as testing started to ramp up, it became clear that an outbreak was flaring.
@@
 With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.
+These weights have the effect of changing the contribution of daily values to the lagged moving average.
+For instance, the temperature at date-minus-4-days is weighted more heavily than the temperature at date-minus-10-days, as a closer approximation of the conditions at the most likely time of contagion before the disease became manifest.
 
@@
 The distribution of these correlations is also quite tight over the 30-day period of the study.
-Hours of sunshine tends to have a moderately high correlation with incidence of COVID-19, however the distribution of these correlations is more spread, and in some cases strays into negative values.
+Hours of sunshine tends to have a moderately high correlation with incidence of COVID-19, but the distribution of these correlations is more spread, and in some cases strays into negative values.
 A similar thing happens with humidity, which also tends to display more day to day variation in the correlation with the dependent variable.

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.tex
Final edits after discussion with co-authors, then resubmission to Geographical Analysis.

--- 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.tex
@@
   \begin{abstract}
-  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
-  systems and the economy. Public health interventions have slowed the
-  spread of the virus, albeit at a high cost. How and when to ease
-  restrictions to movement hinges in part on whether SARS-CoV2 will
-  display seasonality due to variations in temperature, humidity, and
-  hours of sunshine. Here, we address this by means of a spatio-temporal
-  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
-  Unrelated Regressions (SUR) allows us to model the incidence of reported
-  cases of the disease per 100,000 population as an interregional
-  contagion process, in addition to a function of temperature, humidity,
-  and sunshine. In the analysis we also control for GDP per capita,
-  percentage of older adults in the population, population density, and
-  presence of mass transit systems. The results support the hypothesis
-  that incidence of the disease is lower at higher temperatures and higher
-  levels of humidity. Sunshine, in contrast, displays a positive
-  association with incidence of the disease. Our control variables also
-  yield interesting insights. Higher incidence is associated with higher
-  GDP per capita and presence of mass transit systems in the province; in
-  contrast, population density and percentage of older adults display
-  negative associations with incidence of COVID-19.
+  The novel SARS-CoV2 has disrupted health systems and the economy and
+  public health interventions to slow its spread have been costly. How and
+  when to ease restrictions to movement hinges in part on whether
+  SARS-CoV2 will display seasonality due to variations in temperature,
+  humidity, and hours of sunshine. Here, we address this question by means
+  of a spatio-temporal analysis in Spain of the incidence of COVID-19, the
+  disease caused by the virus. Use of spatial Seemingly Unrelated
+  Regressions (SUR) allows us to model the incidence of reported cases of
+  the disease per 100,000 population as an interregional contagion
+  process, in addition to a function of temperature, humidity, and
+  sunshine. In the analysis we also control for GDP per capita, percentage
+  of older adults in the population, population density, and presence of
+  mass transit systems. The results support the hypothesis that incidence
+  of the disease is lower at higher temperatures and higher levels of
+  humidity. Sunshine, in contrast, displays a positive association with
+  incidence of the disease. Our control variables also yield interesting
+  insights. Higher incidence is associated with higher GDP per capita and
+  presence of mass transit systems in the province; in contrast,
+  population density and percentage of older adults display negative
+  associations with incidence of COVID-19.
   \end{abstract}

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Final edits after discussion with co-authors, then resubmission to Geographical Analysis.

@@
 abstract: |
-  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health systems and the economy.
-Public health interventions have slowed the spread of the virus, albeit at a high cost.
+  The novel SARS-CoV2 has disrupted health systems and the economy and public health interventions to slow its spread have been costly.
 How and when to ease restrictions to movement hinges in part on whether SARS-CoV2 will display seasonality due to variations in temperature, humidity, and hours of sunshine.
-Here, we address this by means of a spatio-temporal analysis of the incidence of COVID-19 in Spain.
+Here, we address this question by means of a spatio-temporal analysis in Spain of the incidence of  COVID-19, the disease caused by the virus.
 Use of spatial Seemingly Unrelated Regressions (SUR) allows us to model the incidence of reported cases of the disease per 100,000 population as an interregional contagion process, in addition to a function of temperature, humidity, and sunshine.
@@
 
-From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and COVID-19, the disease it causes), has threatened to overrun health systems around the world.
+From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus and COVID-19, the disease it causes, have threatened to overrun health systems around the world.
 In efforts to contain the spread of the disease, numerous governments in many regions and nations have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place.
@@
 
-A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations.
+A salient question in the debate about how and when to ease social distancing measures is whether the virus will display seasonal variations.
 Existing research on similar pathogens suggests that the virus could be more stable and potentially easier to transmit in conditions of low temperature and low humidity.
-While encouraging this is encouraging, it is important to keep in mind that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" [@deangel2020weathering, section 4].
+While this is encouraging, it is important to keep in mind that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" [@deangel2020weathering, section 4].
 This urges caution when extrapolating from known viruses.
@@
 It is not surprising, given the stakes involved, that this issue has already triggered a lively debate.
-The current state of knowledge is was well-summarized by the National Academy of Sciences, Engineering, and Medicine in the U.S.
-[see @National2020rapid].
-Engaged by the Office of the Executive for guidance on this matter, this organization concludes that: "[some] limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV2 pandemic unfolds could shed more light on the effects of climate on transmission" (p.
+The current state of knowledge was well-summarized by the National Academy of Sciences, Engineering, and Medicine in the U.S.
+in a recent report [see @National2020rapid].
+Engaged by the Office of the Executive for guidance on this matter, this organization concluded that: "[some] limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV2 pandemic unfolds could shed more light on the effects of climate on transmission" (p.
 6).
-To further complicate matters, much of the relevant work has yet to be peer-reviewed (see for instance the challenge of @Harbert2020spatial to @Araujo2020spread).
+To further complicate matters, much of the relevant work has yet to be peer-reviewed [see for instance the challenge of @Harbert2020spatial; to @Araujo2020spread].
 
 With the above considerations in mind, our objective with this paper is to investigate the influence of environmental factors, concretely temperature, humidity, and sunshine, on the progression of the pandemic.
-We adopt a population health approach, and report results from a spatio-temporal model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
+We adopt a population health approach, and report results from a spatio-temporal model of the incidence of COVID-19 in the coterminous provinces in Spain, one of the countries hardest hit by the pandemic.
 We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
@@
 Please note that this paper is prepared as a reproducible research document.
-The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.
+The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{\url{https://drive.google.com/drive/u/1/folders/1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_}}.
 
@@
 Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
-As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
+As noted in the introduction, the focus of the paper is on environmental variables, concretely three related to meteorological conditions: temperature, humidity, and sunshine.
 
@@
 [-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low.
-After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV2 [@Casanova2010effects; @Chan2011effects].
+After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV [@Casanova2010effects; @Chan2011effects].
 Casanova et al.
-[-@Casanova2010effects], for instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse hepatitis MHV viruses), to find that virus inactivation was more rapid at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
+[-@Casanova2010effects], for instance, used surrogates to find that virus inactivation was likely more rapid at higher temperatures; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
 Chan et al.
@@
 
-Whether these results from laboratory experiments will hold in general remains uncertain.
+Whether results from laboratory experiments will hold when the virus circulates in the community remains uncertain.
 At a global scale, de Ángel Solá et al.
-[-@deangel2020weathering] see less risk in the Caribean Basin; on the other hand, Coelho et al.
+[-@deangel2020weathering] see less risk from SARS-CoV2 in the Caribean Basin; on the other hand, Coelho et al.
 [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate.
@@
 
-In addition to the environmental variables above, it is important to account for potential socio-economic and demographic confounders.
-
-In this respect, the first variable that we consider is GDP per capita.
-Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
+In addition to the environmental variables above, from a population health perspective it is also important to account for potential socio-economic and demographic confounders.
+
+To account for population-level factors, the first variable that we consider is GDP per capita.
+Much has been written about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
 A18)}.
-The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated].
-In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect.
+The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear but the plane has already departed [@Zhou2016accelerated].
+In reference to the earlier outbreak of SARS, van Wagner [-@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect.
 In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic.
@@
 Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown.
-On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.
+On the other hand, we cannot discount the possibility that less wealthy regions have a higher proportion of workers in manual occupations who cannot telework, and therefore have more difficulties complying with shelter-in-place orders.
 
@@
 @Novel2020epidemiological].
-However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection.
+However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission rates of the infection.
 The tool of choice in containing the spread of the disease has been social distancing.
-In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
+In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility than most everyone, except the youngest members of the public [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
 In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation.
 Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age].
-Accordingly, our expectation is that higher percentages of older adults will tend to have similarly low levels of social contact [the age-structured matrix of social contact in Spain is similar to Italy; see @Prem2017projecting].
+Since the age-structured matrix of social contact in Spain is similar to Italy [see @Prem2017projecting], our expectation is that populations with higher percentages of older adults will tend to have lower levels of social contact and hence of incidence.
 
@@
 Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
-For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
+For our purposes, we consider positive cases reported, but exclude symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
 The Spanish National Government publishes periodic updates at the regional level (NUTS2) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España.
 This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
-In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food).
-The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute).
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.
+We consider two sets of explanatory variables.
+The first one, and the focus of this research, are the three environmental variables, collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food).
+The second set provides some relevant controls as discussed above, and are also collected from official sources, i.e., INE, the national statistics institute.
+Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data used in this research.
 
@@
 
-An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease.
+An important aspect of working with environmental data such as temperature, humidity, and hours of sunshine, is the incubation period of the disease.
 Lauer et al.
@@
 Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of COVID-19 in Spain, as well as the temporal progression of the disease in weekly averages.
-Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19.
-It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain.
-While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces.
+Our analysis begins on March 13.
+Albeit still low, the highest incidence at this early date was in the provice of Álava, in the North of Spain.
+Álava is not the most populous province, with a population of only 331,549, but it has the highest GDP per capita of all provinces.
 Vitoria, its main city, is the capital of the Basque Country and has been the focus of efforts, along with San Sebastian and Bilbao, to develop a "Global Basque City" [@Meijers2008strategic].
 The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava.
-The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.
+The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under study, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.
 
@@
 week11.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week12.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week13.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week14.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week15.plot <- covid19_spain %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 Each box-and-whisker in the figure represents the distribution of the variable for a community over the 30-day period.
-In the plot, the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost.
+In the plot, the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Andalucia the southernmost.
 As seen in the figure, there is a relatively wide range of values both within and between provinces over the 30-day period examined.
 The top panel of the figure shows the distribution of mean temperatures.
-The lowest mean temperature for a community on any given day was 2.8°C, and the highest 22.4°C, for a range of approximately 20 degrees.
-Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
+The lowest mean temperature for a community on any given day was approximately 3°C, and the highest about 20°C, for a range of approximately 17 degrees.
+Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is approximately 48% and the highest is close to 100%.
 Finally, the bottom panel displays mean daily hours of sunshine in the community.
@@
 temp.summary <- covid19_spain %>% 
+  filter(CCAA != "Canarias" , ) %>%
   group_by(CCAA, Date) %>%
@@
 humidity.summary <- covid19_spain %>% 
+  filter(CCAA != "Canarias") %>%
   group_by(CCAA, Date) %>%
@@
 sunshine.summary <- covid19_spain %>% 
+  filter(CCAA != "Canarias") %>%
   group_by(CCAA, Date) %>%
@@
                              GDPpc > q[4] ~ "4th")) %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
                              Older > q[4] ~ "4th")) %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
                              Density > q[4] ~ "4th")) %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
                                      "Yes"))) %>%
-  filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
 These correlations are calculated after log-transforming all variables.
-As previously discussed, the environmental variables have a temporal lag and were calculated different time windows.
+As previously discussed, the environmental variables have a temporal lag and were calculated using different time windows.
 
@@
 Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive.
-Overall for the period under examination, the pairwise correlations  between these variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables.
+Overall for the period under examination, the pairwise correlations  between these variables and incidence are significant at $p<0.05$, with the exception of the three Sunshine_hours variables.
 
@@
 covid19_spain %>% 
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(Date) %>%
@@
 In this case, we consider neighborhoods based on adjacency, based on the well-known queen criterion (two provinces are adjacent if they share a boundary or touch at a vertex).
-We modify this criterion in two cases.
-First, we make an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are adjacent between each other.
-And secondly, we assume that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello)\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and removed the three islands in the sample.
-The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
+The analysis is of the coterminous provinces\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and included the three islands in the sample.
+In this case we made an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), and assumed that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello) The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
 
@@
 As well, the overall coefficient of determination for the system is reported for each model $\text{pooled}-R^2$.
-The general trend is identical for the three models, with the goodness-of-fit improving over time and plateuing around a value of $R^2$ of 0.6.
+The general trend is identical for the three models, with the goodness-of-fit improving over time and plateuing around a value of $R^2$ of 0.55.
 Model 1 (*lag8*) performs somewhat better in the first few equations/days, when the goodness-of-fit is relatively poor, and then again in the last few equations/days.
 Model 3 (*lag11w*), in contrast, does not perform well towards the end of the study period.
-The most balanced model in terms of equation-level goodness-of-fit appears to be Model 2 (*lag11*), and this impression is further supported by a slightly higher value of the $\text{pooled}-R^2$.
-The analysis using a lagged moving average of the environmental variables is in line with the incubation period reported by Lauer et al.
+The most balanced model in terms of equation-level goodness-of-fit appears to be Model 1 (*lag8*), and this impression is further supported by a slightly higher value of the $\text{pooled}-R^2$.
+The analysis using a lagged moving average of the environmental variables is generally in line with the incubation period reported by Lauer et al.
 [-@Lauer2020incubation], although the results do not support the use of a weighted average.
-For the remainder of the paper, we will adopt Model 2 (*lag11*) as our best model.
+For the remainder of the paper, we will adopt Model 1 (*lag8*) as our best model.
 In the following section we discuss the results of the analysis in more depth.
@@
 
-Table \ref{tab:summary-best-model} presents a summary of the results of Model 2 (*lag11*).
+Table \ref{tab:summary-best-model} presents a summary of the results of Model 1 (*lag8*).
 Recall that two coefficients were constrained and are estimated only for the first equation of the system, and thus are assumed to be constant over time.
-These are the coefficients corresponding to GDP per capita and percentage of older adults.
-Both coefficients are significant at $p\leq0.05$.
+These are the coefficients corresponding to GDP per capita ($p\leq0.10$) and percentage of older adults ($p\leq0.05$).
 The sign of the coefficient for GDP per capita is positive, which indicates that wealthier regions tend to have a higher incidence of COVID-19.
@@
 
-Of the two other control variables, the coefficient of population density is significant at $p\leq0.05$ in 11 equations, and not significant in 19.
-The coefficient for transit is significant at $p\leq0.10$ in 22 equations, and of those, significant at $p\leq0.05$ in 14 equations.
+Of the two other control variables, the coefficient of population density is significant at $p\leq0.05$ in 12, at $p\leq0.10$ in 3 equations, and not significant in 15.
+The coefficient for transit is significant at $p\leq0.10$ in 20 equations, and of those, significant at $p\leq0.05$ in 18 equations.
 The next four variables are environmental factors.
-The coefficient for humidity is significant at $p\leq0.10$ in 20 equations, and of those, significant at $p\leq0.05$ in 19 equations.
+The coefficient for humidity is significant at $p\leq0.19$ in 20 equations, and of those, significant at $p\leq0.05$ in 15 equations.
 Of the environmental variables, temperature is the only variable that has significant coefficients in every equation at $p\leq0.05$.
-Finally, sunshine has significant coefficients at $p\leq0.10$ in 18 equations, including 15 coefficients that are significant at $p\leq0.05$.
+Finally, sunshine has significant coefficients at $p\leq0.05$ in 24 equations.
 
@@
 # Copy best model
-best_model <- sur.slm_lag11
+best_model <- sur.slm_lag8
 
@@
 This is an interesting result: immediately prior to the declaration of the state of emergency, there appears to have been a strong inter-provincial contagion effect.
-Keeping in mind that the incubation period ranges between 2 and 11 days with a median of 5, it is reasonable to expect that the effect of the lockdown will be observed with some delay.
-Indeed, as seen in the figure, the autocorrelation coefficient remains high for three days, then declines on March 16, and continues to weaken over time.
+Keeping in mind that the incubation period ranges between 2 and 12 days with a median of 5, it is reasonable to expect that the effect of the lockdown will be observed with some delay.
+Indeed, as seen in the figure, the autocorrelation coefficient remains high for several days, then begins to decline around March 23, and continues to weaken over time.
 At the end of the period under examination, the strength of this effect is much diminished and we would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero - as seems to be the case.
@@
 
-In Section \ref{variables} we had anticipated a positive sign for the coefficient of this variable, and indeed, at the beginning of the period the coefficient is positive, albeit not significant, and then remains non-significant for the earlier part of the lockdown.
+In Section \ref{background} we had anticipated a positive sign for the coefficient of density, and indeed, at the beginning of the period the coefficient is positive, albeit not significant, and then remains mostly non-significant for the earlier part of the lockdown.
 We are somewhat surprised by the way this coefficient turns significant _and_ negative in the later part of the lockdown, after April 1st.
@@
 
-The evolution of the coefficients for the three environmental variables are shown in Figure \ref{fig:beta-environmental-time}.
+The evolution of the coefficients for the three environmental variables is shown in Figure \ref{fig:beta-environmental-time}.
 Despite a mostly positive simple correlation with incidence (see Figure \ref{fig:daily-correlations}) once that we control for other factors, humidity has a negative association with incidence of COVID-19 in Spain (top-right panel).
-This is in line with the literature that describes the viability and transmission of different viruses at higher levels of humidity.
+This is in line with the literature that describes the lower viability and transmission of different viruses at higher levels of humidity.
 The coefficients for temperature (top-right panel) are consistently negative and this variable is, besides the intercept, the only one with significant coefficients in all equations.
-The range of variation of this coefficient during the period examined is approximately between -1 and -2 - although it is important to recall that these values should not be interpreted directly as effects; more on this below.
-Finally the plot for the coefficients associated with hours of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the coefficient was negative, but significant only at $p\leq10$.
-At the beginning of the lockdown, the coefficient is no longer significative, before becoming *positive* and significant on March 23rd.
+The range of variation of this coefficient during the period examined is approximately between -1 and -2, although it is important to recall that these values should not be interpreted directly as effects; more on this below.
+Finally the plot for the coefficients associated with hours of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the coefficient was negative, but not significant.
+However, five days into the lockdown, the coefficient becomes significant and *positive*.
 This result stands in contrast to previous findings regarding influenza, where more hours of sunlight reduced the strength and duration of epidemic durations [@Yu2013characterization].
@@
 
-The preceding discussion helps to establish the inferential contributions of the analysis.
+The preceding discussion helps to establish the inferential contributions of the analysis, and indicate which variables display significant statistical associations with incidence of COVID-19.
 The remaining question is, what are the implications.
-As discussed in Section \ref{methods} the question of the magnitude of the effects is not clear from the coefficients alone, since a change to a variable in a province influences, via the contagion effect, its neighbors.
-For this reason, the appropriate way to estimate the effects is to calculate both the own and contagion effects (in other words the direct and indirect effects, respectively), as well as their total.
-A summary of the different variable effects appears in Table \ref{tab:summary-impacts-best-model}.
+
+As discussed in Section \ref{methods} the effect of a variable is not clear from its coefficient alone, since a change to a variable in a province influences, via the contagion effect, its neighbors.
+For this reason, the appropriate way to estimate the effects is to calculate both the own effect and the effect due to contagion, or in other words the direct and indirect effects, respectively.
+The total effect is the sum of the two.
+A summary of the effects appears in Table \ref{tab:summary-impacts-best-model}.
 All effects in the table are interpreted as percentage change in the incidence of COVID-19 as a consequence of a one percent change in the variable.
-The exception is Transit (which was not log-transformed), and this variable represents the percentage change in incidence between provinces without and with mass transit systems.
+The exception to this is Transit (which was not log-transformed).
+This variable instead represents the percentage change in incidence between provinces without and with mass transit systems.
 
 Two variables had temporally constrained coefficients.
-The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.675% for each percent increase of this variable (in €1,000s).
+The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.449% for each percent increase of this variable (in €1,000s).
 In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemic.
-Percentage of older adults has a negative effect, and each percent increase in this variable is associated with a relatively small reduction of the incidence of less than 1%.
+Percentage of older adults has a negative effect, and each percent increase in this variable is associated with a relatively small reduction of the incidence of approximately 0.67%.
 
@@
 The direct effects of these variables are as follows: for each percent point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
-This effect is amplified via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 1.5%.
-The effect of humidity is also to reduce the incidence: each percent point of increase in humidity is associated with a reduction of up to 1.5% in incidence.
+This effect is compounded via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 0.75%.
+The effect of humidity is also to reduce the incidence: each percent point of increase in humidity is associated with a reduction of up to 2% in incidence.
 With the addition of the indirect effect, the total effect of a 1% increase in humidity is to reduce incidence by up to 3%.
-As seen in the figure, the indirect (i.e., contagion) effects are almost as strong as the direct effects immediately before and at the beginning of the lockdown period.
-Nonetheless, by the end of the lockdown period, the indirect effects have weakened considerably.
-The effect of density goes from positive (+0.334) to negative (-0.244).
-It is crucial to keep in mind here that this is the effect of density over a period of *lockdown* and possibly subject to risk compensation, as discussed above which suggests, in turn, that efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
+As seen in the figure, the indirect (i.e., contagion) effects are stronger at the before and at the beginning of the lockdown period.
+Nonetheless, by the end of the period under study, the indirect effects have weakened considerably.
 
@@
 Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the total effects of the variables tend towards their direct effects.
-In those first few days covered by our analysis the total effect of all variables is greater due to the spatial contagion effect.
-This also makes analysis and intervention more complex: the contagion effect essentially acts like a multiplier, whereby developments in each province spill over to their neighbors.
+In the first few days covered by our analysis the total effect of all variables is greater due to the spatial contagion effect.
+Contagion makes analysis and intervention more complex: the contagion effect essentially acts like a multiplier, whereby developments in each province spill over to their neighbors.
 Once the contagion effect has been tamed, each province can be "treated" independently from its neighbors.
@@
 
-First, the analysis was conducted under a situation of lockdown, and therefore, besides the results at the beginning of the period examined, one must exercise caution when trying to extrapolate the findings to other situations.
+First, the analysis was conducted mostly under a situation of lockdown, and therefore, besides first days of the period examined, one must exercise caution when trying to extrapolate the findings to a situations without lockdown.
 Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
-In this case we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results - even if we are aware that the number of true cases is likely higher.
+In this case we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results; it is still important, however, to remain aware that the number of true cases is likely higher.
 Finally, we defined neighborhoods based on adjacency.
-It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services.
+It would be interesting to compare adjacency to other connectivity criteria, for instance based on domestic transportation instrastructure and services.
 We flag this as a matter for future research.
@@
         col.names = c("Variable", "Min", "Mean", "Max", "Note"),
-        caption = "\\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)") %>%
+        caption = "\\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag8: lagged 8-day moving average)") %>%
   kable_styling(latex_options = c("striped")) %>%
@@
 
-In this paper we have presented a spatio-temporal analysis of incidence of COVID-19 in Spain.
+In this paper we presented a spatio-temporal analysis of incidence of COVID-19 in Spain.
 The analysis covers a 30-day period that begins immediately before the state of emergency was declared in the country.
@@
 
-Our results offer strong support for the hypothesis that incidence of COVID-19 at the population level is lower at higher temperatures and levels of humidity: the estiamated effect is a reduction in the neighborhood of 3% percent in incidence per each 1% increase in temperature, and a 3% reduction in incidence per 1% increase in humidity *under conditions of contagion*.
-These reductions are estimated to be lower when lockdown conditions are in place.
+Our results offer strong support for the hypothesis that incidence of COVID-19 at the population level is lower at higher temperatures and levels of humidity: the estimated effect is a reduction in the neighborhood of 3% percent in incidence per each 1% increase in temperature, and a 3% reduction in incidence per 1% increase in humidity *under conditions of contagion*.
+These reductions are lower when contagion has ceased (i.e., due to lockdown conditions).
 The question here seems to be whether these environmental variables can yield a bigger reduction of *more* cases, or a smaller reduction of *fewer* cases.
+
 Our control variables also offer some interesting insights.
-In particular, there is evidence of behavioral adaptations in the form of risk compensation (density) and compliance with the lockdown (sunshine).
-These results offer a cautionary tale, with regards to the effectiveness of the lockdown in more dense areas, and also the implications for compliance with stay-at-home orders as the northern hemisphere moves towards summer and more hours of sunshine during the day.
+In particular, there is evidence of behavioral adaptations during lockdown in the form of risk compensation (density) and compliance with the lockdown (sunshine).
+These results offer a cautionary tale with regards to the effectiveness of the lockdown in more dense areas, and also the implications for compliance with stay-at-home orders as the northern hemisphere moves towards summer and more hours of sunshine during the day.
+If risk compensation is a factor, then efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
 
@@
 
-Needless to say, the analysis presented here is ecological, and at the level of population health.
+Needless to say, the analysis presented here is at the level of population health.
 For this reason, the analysis does not make any claims with respect to the effect of ultraviolet light on the virus, but rather about transmission of the virus in the population.
-In this way, the analysis does not say that the virus moves less effectively where more people live in close proximity to each other, but rather that humans are more contagious when they feel safe in less dense regions.
+For example, the analysis does not imply that the virus moves less effectively in places where more people live in close proximity to each other, but rather that humans are more contagious when they feel safe in less dense regions.
 Similarly, more sunshine does not mean that the virus thrives, but rather that humans are more contagious to each other when their behavior adapts to this environmental condition.
@@
 Some directions for future research include investigating other modelling frameworks, such as geographically and temporally weighted regression and/or space-time conditional autoregressive models.
-In addition, the environmental variables examined here relate to meteorological conditions, and did not include other environmental factors that may incide in the transmission of the virus, such as pollution.
-Other factors could be incorporated in future studies.
+In addition, the environmental variables examined here relate to meteorological conditions only, and did not include other environmental factors that may incide in the transmission of the virus, such as pollution.
+These other factors should be incorporated in future studies.
 

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Final edits after discussion with co-authors, then resubmission to Geographical Analysis.

@@
 From a small outbreak linked to a live animal market at the end of 2019
-to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and
-COVID-19, the disease it causes), has threatened to overrun health
+to a global pandemic in a matter of weeks, the SARS-CoV2 virus and
+COVID-19, the disease it causes, have threatened to overrun health
 systems around the world. In efforts to contain the spread of the
@@
 
-A salient question in the debate on how and when to ease social
+A salient question in the debate about how and when to ease social
 distancing measures is whether the virus will display seasonal
@@
 virus could be more stable and potentially easier to transmit in
-conditions of low temperature and low humidity. While encouraging this
-is encouraging, it is important to keep in mind that ``not all seasonal
+conditions of low temperature and low humidity. While this is
+encouraging, it is important to keep in mind that ``not all seasonal
 respiratory viruses experience the same spatiotemporal patterns'' (Ángel
@@
 It is not surprising, given the stakes involved, that this issue has
-already triggered a lively debate. The current state of knowledge is was
+already triggered a lively debate. The current state of knowledge was
 well-summarized by the National Academy of Sciences, Engineering, and
-Medicine in the U.S. (see National Academies of Sciences, Engineering
-and Medicine, 2020). Engaged by the Office of the Executive for guidance
-on this matter, this organization concludes that: ``{[}some{]} limited
-data support a potential waning of cases in warmer and more humid
-seasons, yet none are without major limitations\ldots Additional studies
-as the SARS-CoV2 pandemic unfolds could shed more light on the effects
-of climate on transmission'' (p.~6). To further complicate matters, much
-of the relevant work has yet to be peer-reviewed (see for instance the
-challenge of Harbert et al. (2020) to Araujo and Naimi (2020)).
+Medicine in the U.S. in a recent report (see National Academies of
+Sciences, Engineering and Medicine, 2020). Engaged by the Office of the
+Executive for guidance on this matter, this organization concluded that:
+``{[}some{]} limited data support a potential waning of cases in warmer
+and more humid seasons, yet none are without major
+limitations\ldots Additional studies as the SARS-CoV2 pandemic unfolds
+could shed more light on the effects of climate on transmission''
+(p.~6). To further complicate matters, much of the relevant work has yet
+to be peer-reviewed (see for instance the challenge of Harbert et al.,
+2020; to Araujo and Naimi, 2020).
 
@@
 We adopt a population health approach, and report results from a
-spatio-temporal model of the incidence of COVID-19 in fifty provinces in
-Spain, one of the countries hardest hit by the pandemic. We combine data
-on reported cases of the disease with metereological information, to
-create a spatio-temporal dataset covering a period of 30 days. We then
-join this dataset with provincial-level economic and demographic
-information to act as controls to our key environmental variables. These
-data are analyzed using a spatial Seemingly Unrelated Regressions (SUR)
-approach, which allows us to model incidence of COVID-19 as a contagion
-process.
+spatio-temporal model of the incidence of COVID-19 in the coterminous
+provinces in Spain, one of the countries hardest hit by the pandemic. We
+combine data on reported cases of the disease with metereological
+information, to create a spatio-temporal dataset covering a period of 30
+days. We then join this dataset with provincial-level economic and
+demographic information to act as controls to our key environmental
+variables. These data are analyzed using a spatial Seemingly Unrelated
+Regressions (SUR) approach, which allows us to model incidence of
+COVID-19 as a contagion process.
 
@@
 from a public
-repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.
+repository\footnote{\url{https://drive.google.com/drive/u/1/folders/1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_}}.
 
@@
 factors become increasingly important. As noted in the introduction, the
-focus of the paper is on environmental variables, namely temperature,
-humidity, and sunshine.
+focus of the paper is on environmental variables, concretely three
+related to meteorological conditions: temperature, humidity, and
+sunshine.
 
@@
 low. After the 2002-2004 outbreak of SARS, researchers also began to
-investigate the relationship between these factors and SARS-CoV2
+investigate the relationship between these factors and SARS-CoV
 (Casanova et al., 2010; Chan et al., 2011). Casanova et al.~(2010), for
-instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse
-hepatitis MHV viruses), to find that virus inactivation was more rapid
-at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of
-humidity, these researchers reported that survival of the virus was
-lower at moderate relative humidity levels. Chan et al.~(2011) also
-found that viability of the virus that causes SARS is also lost at
-higher temperatures (\textgreater38°C) and relative humidity superior to
-95\%.
-
-Whether these results from laboratory experiments will hold in general
-remains uncertain. At a global scale, de Ángel Solá et al.~(2020) see
-less risk in the Caribean Basin; however, Coelho et al.~(Coelho et al.,
-2020) warn that at least during the exponential phase, expansion of the
-virus is not driven by climate. Similarly, whereas Araujo and Naimi
-(2020) argue that spread of SARS-CoV2 will likely be constrained by
-climate, Harbert et al.~(2020) remain unconvinced that spatial modelling
-can currently discriminate the distribution of the disease on the basis
-of climate, at least in the United States. Yao et al.~(2020), examined
-data from China and came to the conclusion that neither temperature nor
-ultraviolet indices had an association with transmission of COVID-19.
-This is despite previous research that has linked less exposure to UVB
-radiation to higher prevalence and severity of acute respiratory tract
-infections (Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen
-et al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
+instance, used surrogates to find that virus inactivation was likely
+more rapid at higher temperatures; in terms of humidity, these
+researchers reported that survival of the virus was lower at moderate
+relative humidity levels. Chan et al.~(2011) also found that viability
+of the virus that causes SARS is also lost at higher temperatures
+(\textgreater38°C) and relative humidity superior to 95\%.
+
+Whether results from laboratory experiments will hold when the virus
+circulates in the community remains uncertain. At a global scale, de
+Ángel Solá et al.~(2020) see less risk from SARS-CoV2 in the Caribean
+Basin; on the other hand, Coelho et al.~(Coelho et al., 2020) warn that
+at least during the exponential phase, expansion of the virus is not
+driven by climate. Similarly, whereas Araujo and Naimi (2020) argue that
+spread of SARS-CoV2 will likely be constrained by climate, Harbert et
+al.~(2020) remain unconvinced that spatial modelling can currently
+discriminate the distribution of the disease on the basis of climate, at
+least in the United States. Yao et al.~(2020), examined data from China
+and came to the conclusion that neither temperature nor ultraviolet
+indices had an association with transmission of COVID-19. This is
+despite previous research that has linked less exposure to UVB radiation
+to higher prevalence and severity of acute respiratory tract infections
+(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
+al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
 Moriyama, Hugentobler, and Iwasaki 2020).
 
-In addition to the environmental variables above, we also consider a set
-of potential socio-economic and demographic confounders.
-
-The first variable that we consider is GDP per capita. Much has been
-said about globalization and the spread of infectious
+In addition to the environmental variables above, from a population
+health perspective it is also important to account for potential
+socio-economic and demographic confounders.
+
+To account for population-level factors, the first variable that we
+consider is GDP per capita. Much has been written about globalization
+and the spread of infectious
 disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
@@
 approaches in the initial stages of disease management, when the cause
-of a disease may still be unclear, but the plane has already departed
+of a disease may still be unclear but the plane has already departed
 (Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
-van Wagner (Van Wagner, 2008) chronicles how Toronto's status as a
-global city turned out to be a vulnerability in this respect. In our
-case, we think of GDP per capita as a marker of a region's relative
-position in a network of global cities, and its potential to be further
-ahead in the trajectory of the pandemic. Furthermore, wealthier regions
-also tend to concentrate more activities that produce non-traded goods,
-including building and construction (Hallet, 2002). Therefore, it is
-possible that wealthier regions remain relatively more active even
-during a lockdown. On the other hand, it is also possible that less
-wealthy regions have a higher proportion of workers in manual
-occupations that cannot telework, and therefore have more difficulties
-complying with shelter-in-place orders.
+van Wagner (2008) chronicles how Toronto's status as a global city
+turned out to be a vulnerability in this respect. In our case, we think
+of GDP per capita as a marker of a region's relative position in a
+network of global cities, and its potential to be further ahead in the
+trajectory of the pandemic. Furthermore, wealthier regions also tend to
+concentrate more activities that produce non-traded goods, including
+building and construction (Hallet, 2002). Therefore, it is possible that
+wealthier regions remain relatively more active even during a lockdown.
+On the other hand, we cannot discount the possibility that less wealthy
+regions have a higher proportion of workers in manual occupations who
+cannot telework, and therefore have more difficulties complying with
+shelter-in-place orders.
 
@@
 relatively large population of older adults necessarily translates into
-higher transmission of the infection. The tool of choice in containing
-the spread of the disease has been social distancing. In this respect,
-the evidence from the field of transportation is that older adults tend
-to travel less frequently, for shorter distances, and have higher rates
-of immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder
-and Pinjari, 2012). In other words, many older adults are, whether by
-preference or otherwise, already in a form of social isolation. Social
-distancing during the pandemic may actually reinforce that condition for
-them, as suggested by the analysis of age-structured social contact in
-India, China, and Italy of Sing and Adhikari (2020). Accordingly, our
-expectation is that higher percentages of older adults will tend to have
-similarly low levels of social contact (the age-structured matrix of
-social contact in Spain is similar to Italy; see Prem et al., 2017).
+higher transmission rates of the infection. The tool of choice in
+containing the spread of the disease has been social distancing. In this
+respect, the evidence from the field of transportation is that older
+adults tend to travel less frequently, for shorter distances, and have
+higher rates of immobility than most everyone, except the youngest
+members of the public (e.g., Roorda et al., 2010; Morency et al., 2011;
+Sikder and Pinjari, 2012). In other words, many older adults are,
+whether by preference or otherwise, already in a form of social
+isolation. Social distancing during the pandemic may actually reinforce
+that condition for them, as suggested by the analysis of age-structured
+social contact in India, China, and Italy of Sing and Adhikari (2020).
+Since the age-structured matrix of social contact in Spain is similar to
+Italy (see Prem et al., 2017), our expectation is that populations with
+higher percentages of older adults will tend to have lower levels of
+social contact and hence of incidence.
 
@@
 transit systems in a province. Every province in Spain offers some form
-of public transportation, however only five provinces have higher order
+of public transportation, but only five provinces have higher order
 systems of mass mobility (e.g.~metro or subway), namely Barcelona,
@@
 when a German tourist in the Canary Islands tested positive for the
-virus. However, it was still a few weeks before the first domestic case
-was reported, on February 27th in Sevilla province (Andalusia). In a
-short period of time, as testing started to ramp up, it became clear
-that an outbreak was flaring. By March 11th the World Health
-Organization (WHO) declared COVID-19 officially a pandemic. This
+virus. After this case, it was still a few weeks before the first
+domestic case was reported, on February 27th in Sevilla province
+(Andalusia). In a short period of time, as testing started to ramp up,
+it became clear that an outbreak was flaring. By March 11th the World
+Health Organization (WHO) declared COVID-19 officially a pandemic. This
 declaration marked a turning point for the public in Spain too. As of
@@
 inclusive. For our purposes, we consider positive cases reported, but
-excluding symptomatic cases diagnosed by a doctor without a Polymerase
+exclude symptomatic cases diagnosed by a doctor without a Polymerase
 Chain Reaction (PCR) test. The Spanish National Government publishes
@@
 Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
-In addition, we consider two sets of explanatory variables. The first
-one, and the focus of this research, is set of two environmental
-variables, namely temperature and humidity, which are collected from
-official sources (i.g., \emph{AEMET}, the state meteorology agency, and
-\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
-second set provides some relevant controls for multivariate analysis,
-and refers to economic and demographic attributes of the province (also
-collected from official sources, i.e., INE, the national statistics
-institute). Table \ref{tab:descriptive-statistics} shows the descriptive
-statistics and the provenance of the data.
+We consider two sets of explanatory variables. The first one, and the
+focus of this research, are the three environmental variables, collected
+from official sources (i.g., \emph{AEMET}, the state meteorology agency,
+and \emph{MAPA}, the ministry of agriculture, fisheries, and food). The
+second set provides some relevant controls as discussed above, and are
+also collected from official sources, i.e., INE, the national statistics
+institute. Table \ref{tab:descriptive-statistics} shows the descriptive
+statistics and the provenance of the data used in this research.
 
@@
 An important aspect of working with environmental data such as
-temperature and humidity is the incubation period of the disease. Lauer
-et al.~(2020) report for the case of COVID-19 a median incubation period
-of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The
-vast majority of cases (95\%) develop symptoms between 2.6 days (CI, 2.1
-to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we
-judge it best to use lagged values of the environmental variables. We
-test different time lags as follows. We consider a lagged 8-day average,
-from date-minus-12 to date-minus-5 days (hereafter \emph{lag8}).
-Secondly, we consider a lagged 11-day average, from date-minus-12 to
-date-minus-2 days (hereafter \emph{lag11}). Finally, to account for the
-likely duration of incubation, we consider a lagged 11-day
-\emph{weighted} average, from date-minus-12 to date-minus-2 days
-(hereafter \emph{lag11w}). The weights for this variable are calculated
-using the parameters of the log-normal distribution reported by Lauer et
-al.~(2020), i.e., a log-mean of 1.621 and a log-standard deviation of
-0.418. With these weights, the environmental variables at date-minus-2
-and date-minus-12 days are weighted as 0.041 and 0.009 respectively,
-whereas the environmental variables at date-minus-5 days are weighted as
-0.194.
+temperature, humidity, and hours of sunshine, is the incubation period
+of the disease. Lauer et al.~(2020) report for the case of COVID-19 a
+median incubation period of 5.7 days (with a confidence interval between
+4.9 to 7.8 days). The vast majority of cases (95\%) develop symptoms
+between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7
+days). For this reason, we judge it best to use lagged values of the
+environmental variables. We test different time lags as follows. We
+consider a lagged 8-day average, from date-minus-12 to date-minus-5 days
+(hereafter \emph{lag8}). Secondly, we consider a lagged 11-day average,
+from date-minus-12 to date-minus-2 days (hereafter \emph{lag11}).
+Finally, to account for the likely duration of incubation, we consider a
+lagged 11-day \emph{weighted} average, from date-minus-12 to
+date-minus-2 days (hereafter \emph{lag11w}). The weights for this
+variable are calculated using the parameters of the log-normal
+distribution reported by Lauer et al.~(2020), i.e., a log-mean of 1.621
+and a log-standard deviation of 0.418. With these weights, the
+environmental variables at date-minus-2 and date-minus-12 days are
+weighted as 0.041 and 0.009 respectively, whereas the environmental
+variables at date-minus-5 days are weighted as 0.194. These weights have
+the effect of changing the contribution of daily values to the lagged
+moving average. For instance, the temperature at date-minus-4-days is
+weighted more heavily than the temperature at date-minus-10-days, as a
+closer approximation of the conditions at the most likely time of
+contagion before the disease became manifest.
 
@@
 progression of the disease in weekly averages. Our analysis begins on
-March 13, which is the first date when every province had reported at
-least one case of COVID-19. It can be seen that the highest incidence at
-this early date was in the provice of Álava, in the North of Spain.
-While not the most populous province, with a population of only 331,549,
-Álava has the highest GDP per capita of all provinces. Vitoria, its main
-city, is the capital of the Basque Country and has been the focus of
-efforts, along with San Sebastian and Bilbao, to develop a ``Global
-Basque City'' (Meijers et al., 2008). The other early focus of the
-pandemic in Spain was Madrid, which is the most populous province in the
-country and has the second highest GDP per capita after Álava. The early
-outbreaks in these two provinces can be traced throughout the
-progression of the pandemic over time, although by the end of the period
-under consideration, other provinces had matched and even surpased their
-incidence rates, including Segovia and Soria to the north of Madrid, and
-Ciudad Real and Albacete to the south.
+March 13. Albeit still low, the highest incidence at this early date was
+in the provice of Álava, in the North of Spain. Álava is not the most
+populous province, with a population of only 331,549, but it has the
+highest GDP per capita of all provinces. Vitoria, its main city, is the
+capital of the Basque Country and has been the focus of efforts, along
+with San Sebastian and Bilbao, to develop a ``Global Basque City''
+(Meijers et al., 2008). The other early focus of the pandemic in Spain
+was Madrid, which is the most populous province in the country and has
+the second highest GDP per capita after Álava. The early outbreaks in
+these two provinces can be traced throughout the progression of the
+pandemic over time, although by the end of the period under study, other
+provinces had matched and even surpased their incidence rates, including
+Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
+to the south.
 
@@
 latitude, so that Principado de Asturias is the northernmost community,
-and Canarias the southernmost. As seen in the figure, there is a
+and Andalucia the southernmost. As seen in the figure, there is a
 relatively wide range of values both within and between provinces over
@@
 distribution of mean temperatures. The lowest mean temperature for a
-community on any given day was 2.8°C, and the highest 22.4°C, for a
-range of approximately 20 degrees. Likewise, there is a great deal of
-variability in humidity, as seen in the middle panel of the figure,
-where the lowest mean humidity for any community is 48.3 and the highest
-is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
-in the community. This variable displays the most variability within
-communities over time, but remains relatively constant across
-communities. It is important to note that these are summaries by
-community, and the actual values of these variables for the provinces
-display somewhat more variability.
+community on any given day was approximately 3°C, and the highest about
+20°C, for a range of approximately 17 degrees. Likewise, there is a
+great deal of variability in humidity, as seen in the middle panel of
+the figure, where the lowest mean humidity for any community is
+approximately 48\% and the highest is close to 100\%. Finally, the
+bottom panel displays mean daily hours of sunshine in the community.
+This variable displays the most variability within communities over
+time, but remains relatively constant across communities. It is
+important to note that these are summaries by community, and the actual
+values of these variables for the provinces display somewhat more
+variability.
 
@@
 variables. As previously discussed, the environmental variables have a
-temporal lag and were calculated different time windows.
+temporal lag and were calculated using different time windows.
 
@@
 quite tight over the 30-day period of the study. Hours of sunshine tends
-to have a moderately high correlation with incidence of COVID-19,
-however the distribution of these correlations is more spread, and in
-some cases strays into negative values. A similar thing happens with
-humidity, which also tends to display more day to day variation in the
-correlation with the dependent variable. The percentage of older adults
-shows a relatively tight distribution of day-to-day correlations, and is
+to have a moderately high correlation with incidence of COVID-19, but
+the distribution of these correlations is more spread, and in some cases
+strays into negative values. A similar thing happens with humidity,
+which also tends to display more day to day variation in the correlation
+with the dependent variable. The percentage of older adults shows a
+relatively tight distribution of day-to-day correlations, and is
 negative. Population density, in contrast, tends to be negative, but is
@@
 variables and incidence are significant at \(p<0.05\), with the
-exception of the three \(Sunshine_hours\) variables.
+exception of the three Sunshine\_hours variables.
 
@@
 adjacency, based on the well-known queen criterion (two provinces are
-adjacent if they share a boundary or touch at a vertex). We modify this
-criterion in two cases. First, we make an allowance for adjacency
-between the two islands in the Autonomous Community of Canarias in the
-Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are
-adjacent between each other. And secondly, we assume that Islas Baleares
-in the Mediterranean are adjacent to three provinces in Pais Catalans
-(i.e., Barcelona, Tarragona, and
-Castello)\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and removed the three islands in the sample.
-The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
+adjacent if they share a boundary or touch at a vertex). The analysis is
+of the coterminous
+provinces\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and included the three islands in the sample.
+In this case we made an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), and assumed that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello) The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
 
@@
 identical for the three models, with the goodness-of-fit improving over
-time and plateuing around a value of \(R^2\) of 0.6. Model 1
+time and plateuing around a value of \(R^2\) of 0.55. Model 1
 (\emph{lag8}) performs somewhat better in the first few equations/days,
@@
 perform well towards the end of the study period. The most balanced
-model in terms of equation-level goodness-of-fit appears to be Model 2
-(\emph{lag11}), and this impression is further supported by a slightly
+model in terms of equation-level goodness-of-fit appears to be Model 1
+(\emph{lag8}), and this impression is further supported by a slightly
 higher value of the \(\text{pooled}-R^2\). The analysis using a lagged
-moving average of the environmental variables is in line with the
-incubation period reported by Lauer et al.~(2020), although the results
-do not support the use of a weighted average. For the remainder of the
-paper, we will adopt Model 2 (\emph{lag11}) as our best model. In the
-following section we discuss the results of the analysis in more depth.
+moving average of the environmental variables is generally in line with
+the incubation period reported by Lauer et al.~(2020), although the
+results do not support the use of a weighted average. For the remainder
+of the paper, we will adopt Model 1 (\emph{lag8}) as our best model. In
+the following section we discuss the results of the analysis in more
+depth.
 
@@
 Table \ref{tab:summary-best-model} presents a summary of the results of
-Model 2 (\emph{lag11}). Recall that two coefficients were constrained
-and are estimated only for the first equation of the system, and thus
-are assumed to be constant over time. These are the coefficients
-corresponding to GDP per capita and percentage of older adults. Both
-coefficients are significant at \(p\leq0.05\). The sign of the
-coefficient for GDP per capita is positive, which indicates that
-wealthier regions tend to have a higher incidence of COVID-19. This is
-in line with the idea that the epidemic started earlier in wealthier
-places due to their connections to a globalized world. The sign of the
-coefficient for percentage of older adults, on the other hand, is
-negative. As previously discussed, the level of social contact of older
-adults even under normal circumstances tends to be lower than for
-younger people. As a consequence, places with larger populations of
-older adults appear to have a natural level of social distancing in
-place. It is important to note that this does not detract from evidence
-that older adults are more vulnerable individually and in institutional
-settings, where their case mortality rates are perhaps the highest of
-all age groups. Instead, this result indicates that their presence in
-the community at large tends to depress transmission of the virus.
+Model 1 (\emph{lag8}). Recall that two coefficients were constrained and
+are estimated only for the first equation of the system, and thus are
+assumed to be constant over time. These are the coefficients
+corresponding to GDP per capita (\(p\leq0.10\)) and percentage of older
+adults (\(p\leq0.05\)). The sign of the coefficient for GDP per capita
+is positive, which indicates that wealthier regions tend to have a
+higher incidence of COVID-19. This is in line with the idea that the
+epidemic started earlier in wealthier places due to their connections to
+a globalized world. The sign of the coefficient for percentage of older
+adults, on the other hand, is negative. As previously discussed, the
+level of social contact of older adults even under normal circumstances
+tends to be lower than for younger people. As a consequence, places with
+larger populations of older adults appear to have a natural level of
+social distancing in place. It is important to note that this does not
+detract from evidence that older adults are more vulnerable individually
+and in institutional settings, where their case mortality rates are
+perhaps the highest of all age groups. Instead, this result indicates
+that their presence in the community at large tends to depress
+transmission of the virus.
 
 Of the two other control variables, the coefficient of population
-density is significant at \(p\leq0.05\) in 11 equations, and not
-significant in 19. The coefficient for transit is significant at
-\(p\leq0.10\) in 22 equations, and of those, significant at
-\(p\leq0.05\) in 14 equations. The next four variables are environmental
-factors. The coefficient for humidity is significant at \(p\leq0.10\) in
-20 equations, and of those, significant at \(p\leq0.05\) in 19
-equations. Of the environmental variables, temperature is the only
-variable that has significant coefficients in every equation at
-\(p\leq0.05\). Finally, sunshine has significant coefficients at
-\(p\leq0.10\) in 18 equations, including 15 coefficients that are
-significant at \(p\leq0.05\).
+density is significant at \(p\leq0.05\) in 12, at \(p\leq0.10\) in 3
+equations, and not significant in 15. The coefficient for transit is
+significant at \(p\leq0.10\) in 20 equations, and of those, significant
+at \(p\leq0.05\) in 18 equations. The next four variables are
+environmental factors. The coefficient for humidity is significant at
+\(p\leq0.19\) in 20 equations, and of those, significant at
+\(p\leq0.05\) in 15 equations. Of the environmental variables,
+temperature is the only variable that has significant coefficients in
+every equation at \(p\leq0.05\). Finally, sunshine has significant
+coefficients at \(p\leq0.05\) in 24 equations.
 
@@
 \midrule
-\rowcolor{gray!6}  Intercept & 6.878 & 9.889 & 15.294 & 0 & 0 & 30 & Non-constrained\\
-log(GDPpc) & 0.637 & 0.637 & 0.637 & 0 & 0 & 1 & Constrained\\
-\rowcolor{gray!6}  log(Older) & -0.854 & -0.854 & -0.854 & 0 & 0 & 1 & Constrained\\
-log(Density) & -0.213 & -0.101 & 0.101 & 18 & 3 & 9 & Non-constrained\\
-\rowcolor{gray!6}  Transit & 0.326 & 0.487 & 0.593 & 10 & 12 & 8 & Non-constrained\\
+\rowcolor{gray!6}  Intercept & 6.172 & 9.441 & 14.071 & 0 & 0 & 30 & Non-constrained\\
+log(GDPpc) & 0.449 & 0.449 & 0.449 & 0 & 1 & 0 & Constrained\\
+\rowcolor{gray!6}  log(Older) & -0.676 & -0.676 & -0.676 & 0 & 0 & 1 & Constrained\\
+log(Density) & -0.212 & -0.105 & 0.143 & 15 & 3 & 12 & Non-constrained\\
+\rowcolor{gray!6}  Transit & 0.341 & 0.528 & 0.606 & 10 & 2 & 18 & Non-constrained\\
 \addlinespace
-log(Humidity) & -2.159 & -0.456 & 0.193 & 15 & 2 & 13 & Non-constrained\\
-\rowcolor{gray!6}  log(Temperature) & -2.006 & -1.496 & -0.988 & 0 & 0 & 30 & Non-constrained\\
-log(Sunshine) & -0.252 & 0.078 & 0.177 & 9 & 1 & 20 & Non-constrained\\
-\rowcolor{gray!6}  Spatially lagged y (rho) & 0.056 & 0.236 & 0.548 & 2 & 5 & 23 & Non-constrained\\
+log(Humidity) & -1.935 & -0.435 & 0.054 & 11 & 4 & 15 & Non-constrained\\
+\rowcolor{gray!6}  log(Temperature) & -1.904 & -1.236 & -0.817 & 0 & 0 & 30 & Non-constrained\\
+log(Sunshine) & -0.187 & 0.099 & 0.189 & 6 & 0 & 24 & Non-constrained\\
+\rowcolor{gray!6}  Spatially lagged y (rho) & 0.014 & 0.154 & 0.348 & 14 & 3 & 13 & Non-constrained\\
 \bottomrule
@@
 contagion effect. Keeping in mind that the incubation period ranges
-between 2 and 11 days with a median of 5, it is reasonable to expect
+between 2 and 12 days with a median of 5, it is reasonable to expect
 that the effect of the lockdown will be observed with some delay.
 Indeed, as seen in the figure, the autocorrelation coefficient remains
-high for three days, then declines on March 16, and continues to weaken
-over time. At the end of the period under examination, the strength of
-this effect is much diminished and we would expect that under full
-compliance with strict lockdown conditions (meaning no inter-provincial
-mobility) the spatial contagion effect would be zero - as seems to be
-the case.
+high for several days, then begins to decline around March 23, and
+continues to weaken over time. At the end of the period under
+examination, the strength of this effect is much diminished and we would
+expect that under full compliance with strict lockdown conditions
+(meaning no inter-provincial mobility) the spatial contagion effect
+would be zero - as seems to be the case.
 
@@
 
-In Section \ref{variables} we had anticipated a positive sign for the
-coefficient of this variable, and indeed, at the beginning of the period
-the coefficient is positive, albeit not significant, and then remains
+In Section \ref{background} we had anticipated a positive sign for the
+coefficient of density, and indeed, at the beginning of the period the
+coefficient is positive, albeit not significant, and then remains mostly
 non-significant for the earlier part of the lockdown. We are somewhat
@@
 The evolution of the coefficients for the three environmental variables
-are shown in Figure \ref{fig:beta-environmental-time}. Despite a mostly
+is shown in Figure \ref{fig:beta-environmental-time}. Despite a mostly
 positive simple correlation with incidence (see Figure
@@
 (top-right panel). This is in line with the literature that describes
-the viability and transmission of different viruses at higher levels of
-humidity. The coefficients for temperature (top-right panel) are
-consistently negative and this variable is, besides the intercept, the
-only one with significant coefficients in all equations. The range of
-variation of this coefficient during the period examined is
-approximately between -1 and -2 - although it is important to recall
-that these values should not be interpreted directly as effects; more on
-this below. Finally the plot for the coefficients associated with hours
-of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the
-coefficient was negative, but significant only at \(p\leq10\). At the
-beginning of the lockdown, the coefficient is no longer significative,
-before becoming \emph{positive} and significant on March 23rd. This
-result stands in contrast to previous findings regarding influenza,
+the lower viability and transmission of different viruses at higher
+levels of humidity. The coefficients for temperature (top-right panel)
+are consistently negative and this variable is, besides the intercept,
+the only one with significant coefficients in all equations. The range
+of variation of this coefficient during the period examined is
+approximately between -1 and -2, although it is important to recall that
+these values should not be interpreted directly as effects; more on this
+below. Finally the plot for the coefficients associated with hours of
+sunshine (bottom panel) is more ambiguous: prior to the lockdown, the
+coefficient was negative, but not significant. However, five days into
+the lockdown, the coefficient becomes significant and \emph{positive}.
+This result stands in contrast to previous findings regarding influenza,
 where more hours of sunlight reduced the strength and duration of
@@
 The preceding discussion helps to establish the inferential
-contributions of the analysis. The remaining question is, what are the
-implications. As discussed in Section \ref{methods} the question of the
-magnitude of the effects is not clear from the coefficients alone, since
-a change to a variable in a province influences, via the contagion
-effect, its neighbors. For this reason, the appropriate way to estimate
-the effects is to calculate both the own and contagion effects (in other
-words the direct and indirect effects, respectively), as well as their
-total. A summary of the different variable effects appears in Table
+contributions of the analysis, and indicate which variables display
+significant statistical associations with incidence of COVID-19. The
+remaining question is, what are the implications.
+
+As discussed in Section \ref{methods} the effect of a variable is not
+clear from its coefficient alone, since a change to a variable in a
+province influences, via the contagion effect, its neighbors. For this
+reason, the appropriate way to estimate the effects is to calculate both
+the own effect and the effect due to contagion, or in other words the
+direct and indirect effects, respectively. The total effect is the sum
+of the two. A summary of the effects appears in Table
 \ref{tab:summary-impacts-best-model}. All effects in the table are
 interpreted as percentage change in the incidence of COVID-19 as a
-consequence of a one percent change in the variable. The exception is
-Transit (which was not log-transformed), and this variable represents
-the percentage change in incidence between provinces without and with
-mass transit systems.
+consequence of a one percent change in the variable. The exception to
+this is Transit (which was not log-transformed). This variable instead
+represents the percentage change in incidence between provinces without
+and with mass transit systems.
 
@@
 effect of GDP per capita is to increase the incidence of COVID-19 by
-0.675\% for each percent increase of this variable (in €1,000s). In our
+0.449\% for each percent increase of this variable (in €1,000s). In our
 view, this is a measure of inertia, as provinces with higher GDP per
@@
 increase in this variable is associated with a relatively small
-reduction of the incidence of less than 1\%.
+reduction of the incidence of approximately 0.67\%.
 
@@
 between a 1\% and 2\% reduction in the incidence of the disease. This
-effect is amplified via contagion, as seen in the central panel in the
+effect is compounded via contagion, as seen in the central panel in the
 figure, and the indirect effect can further reduce the incidence by up
-to 1.5\%. The effect of humidity is also to reduce the incidence: each
+to 0.75\%. The effect of humidity is also to reduce the incidence: each
 percent point of increase in humidity is associated with a reduction of
-up to 1.5\% in incidence. With the addition of the indirect effect, the
+up to 2\% in incidence. With the addition of the indirect effect, the
 total effect of a 1\% increase in humidity is to reduce incidence by up
 to 3\%. As seen in the figure, the indirect (i.e., contagion) effects
-are almost as strong as the direct effects immediately before and at the
-beginning of the lockdown period. Nonetheless, by the end of the
-lockdown period, the indirect effects have weakened considerably. The
-effect of density goes from positive (+0.334) to negative (-0.244). It
-is crucial to keep in mind here that this is the effect of density over
-a period of \emph{lockdown} and possibly subject to risk compensation,
-as discussed above which suggests, in turn, that efforts should be made
-to reduce or eliminate risk compensation in less densely populated
-regions.
+are stronger at the before and at the beginning of the lockdown period.
+Nonetheless, by the end of the period under study, the indirect effects
+have weakened considerably.
 
@@
 inter-regional contagion is reduced, as expected, and the total effects
-of the variables tend towards their direct effects. In those first few
+of the variables tend towards their direct effects. In the first few
 days covered by our analysis the total effect of all variables is
-greater due to the spatial contagion effect. This also makes analysis
+greater due to the spatial contagion effect. Contagion makes analysis
 and intervention more complex: the contagion effect essentially acts
@@
 
-First, the analysis was conducted under a situation of lockdown, and
-therefore, besides the results at the beginning of the period examined,
-one must exercise caution when trying to extrapolate the findings to
-other situations. Secondly, it is well known that there is in many
+First, the analysis was conducted mostly under a situation of lockdown,
+and therefore, besides first days of the period examined, one must
+exercise caution when trying to extrapolate the findings to a situations
+without lockdown. Secondly, it is well known that there is in many
 countries substantial underreporting of cases of COVID-19 due to limited
@@
 reporting, and as long as the underreporting is consistent across units
-of analysis, we do not expect biased results - even if we are aware that
-the number of true cases is likely higher. Finally, we defined
-neighborhoods based on adjacency. It would be interesting to compare
-other connectivity criteria, for instance based on domestic
-transportation instrastructure and services. We flag this as a matter
-for future research.
+of analysis, we do not expect biased results; it is still important,
+however, to remain aware that the number of true cases is likely higher.
+Finally, we defined neighborhoods based on adjacency. It would be
+interesting to compare adjacency to other connectivity criteria, for
+instance based on domestic transportation instrastructure and services.
+We flag this as a matter for future research.
 
@@
 
-\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
+\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag8: lagged 8-day moving average)}
 \centering
@@
 \multicolumn{5}{l}{\textbf{Direct Effects}}\\
-\hspace{1em}log(GDPpc) & 0.694 & 0.694 & 0.694 & Constrained\\
-\hspace{1em}log(Older) & -0.930 & -0.930 & -0.930 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.213 & -0.102 & 0.110 & Non-constrained\\
-\hspace{1em}Transit & 0.351 & 0.496 & 0.602 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -2.351 & -0.475 & 0.194 & Non-constrained\\
-\hspace{1em}log(Temperature) & -2.184 & -1.528 & -0.990 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.275 & 0.077 & 0.180 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.457 & 0.457 & 0.457 & Constrained\\
+\hspace{1em}log(Older) & -0.688 & -0.688 & -0.688 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.213 & -0.106 & 0.145 & Non-constrained\\
+\hspace{1em}Transit & 0.349 & 0.532 & 0.608 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.971 & -0.440 & 0.054 & Non-constrained\\
+\hspace{1em}log(Temperature) & -1.939 & -1.245 & -0.825 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.191 & 0.099 & 0.190 & Non-constrained\\
 \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Indirect Effects}}\\
-\hspace{1em}log(GDPpc) & 0.718 & 0.718 & 0.718 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.962 & -0.962 & -0.962 & Constrained\\
-\hspace{1em}log(Density) & -0.115 & -0.020 & 0.114 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}Transit & 0.035 & 0.158 & 0.447 & Non-constrained\\
-\hspace{1em}log(Humidity) & -2.431 & -0.279 & 0.023 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -2.259 & -0.538 & -0.095 & Non-constrained\\
-\hspace{1em}log(Sunshine) & -0.284 & 0.001 & 0.059 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.165 & 0.165 & 0.165 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.249 & -0.249 & -0.249 & Constrained\\
+\hspace{1em}log(Density) & -0.075 & -0.015 & 0.052 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.097 & 0.240 & Non-constrained\\
+\hspace{1em}log(Humidity) & -0.712 & -0.104 & 0.001 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -0.700 & -0.238 & -0.013 & Non-constrained\\
+\hspace{1em}log(Sunshine) & -0.069 & 0.017 & 0.063 & Non-constrained\\
 \rowcolor{gray!6}  \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Total Effects}}\\
-\hspace{1em}log(GDPpc) & 1.411 & 1.411 & 1.411 & Constrained\\
-\hspace{1em}log(Older) & -1.892 & -1.892 & -1.892 & Constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.241 & -0.122 & 0.224 & Non-constrained\\
-\hspace{1em}Transit & 0.530 & 0.655 & 0.905 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -4.782 & -0.754 & 0.217 & Non-constrained\\
-\hspace{1em}log(Temperature) & -4.442 & -2.066 & -1.087 & Non-constrained\\
-\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.559 & 0.078 & 0.238 & Non-constrained\\
+\hspace{1em}log(GDPpc) & 0.622 & 0.622 & 0.622 & Constrained\\
+\hspace{1em}log(Older) & -0.937 & -0.937 & -0.937 & Constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.265 & -0.121 & 0.198 & Non-constrained\\
+\hspace{1em}Transit & 0.466 & 0.629 & 0.847 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -2.683 & -0.543 & 0.055 & Non-constrained\\
+\hspace{1em}log(Temperature) & -2.640 & -1.483 & -0.870 & Non-constrained\\
+\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.259 & 0.116 & 0.235 & Non-constrained\\
 \bottomrule
@@
 
-In this paper we have presented a spatio-temporal analysis of incidence
-of COVID-19 in Spain. The analysis covers a 30-day period that begins
+In this paper we presented a spatio-temporal analysis of incidence of
+COVID-19 in Spain. The analysis covers a 30-day period that begins
 immediately before the state of emergency was declared in the country.
@@
 COVID-19 at the population level is lower at higher temperatures and
-levels of humidity: the estiamated effect is a reduction in the
+levels of humidity: the estimated effect is a reduction in the
 neighborhood of 3\% percent in incidence per each 1\% increase in
@@
 humidity \emph{under conditions of contagion}. These reductions are
-estimated to be lower when lockdown conditions are in place. The
+lower when contagion has ceased (i.e., due to lockdown conditions). The
 question here seems to be whether these environmental variables can
 yield a bigger reduction of \emph{more} cases, or a smaller reduction of
-\emph{fewer} cases. Our control variables also offer some interesting
-insights. In particular, there is evidence of behavioral adaptations in
-the form of risk compensation (density) and compliance with the lockdown
-(sunshine). These results offer a cautionary tale, with regards to the
-effectiveness of the lockdown in more dense areas, and also the
+\emph{fewer} cases.
+
+Our control variables also offer some interesting insights. In
+particular, there is evidence of behavioral adaptations during lockdown
+in the form of risk compensation (density) and compliance with the
+lockdown (sunshine). These results offer a cautionary tale with regards
+to the effectiveness of the lockdown in more dense areas, and also the
 implications for compliance with stay-at-home orders as the northern
 hemisphere moves towards summer and more hours of sunshine during the
-day.
+day. If risk compensation is a factor, then efforts should be made to
+reduce or eliminate risk compensation in less densely populated regions.
 
@@
 
-Needless to say, the analysis presented here is ecological, and at the
-level of population health. For this reason, the analysis does not make
-any claims with respect to the effect of ultraviolet light on the virus,
-but rather about transmission of the virus in the population. In this
-way, the analysis does not say that the virus moves less effectively
-where more people live in close proximity to each other, but rather that
-humans are more contagious when they feel safe in less dense regions.
-Similarly, more sunshine does not mean that the virus thrives, but
-rather that humans are more contagious to each other when their behavior
-adapts to this environmental condition.
+Needless to say, the analysis presented here is at the level of
+population health. For this reason, the analysis does not make any
+claims with respect to the effect of ultraviolet light on the virus, but
+rather about transmission of the virus in the population. For example,
+the analysis does not imply that the virus moves less effectively in
+places where more people live in close proximity to each other, but
+rather that humans are more contagious when they feel safe in less dense
+regions. Similarly, more sunshine does not mean that the virus thrives,
+but rather that humans are more contagious to each other when their
+behavior adapts to this environmental condition.
 
@@
 addition, the environmental variables examined here relate to
-meteorological conditions, and did not include other environmental
+meteorological conditions only, and did not include other environmental
 factors that may incide in the transmission of the virus, such as
-pollution. Other factors could be incorporated in future studies.
+pollution. These other factors should be incorporated in future studies.
 

### 2020-05-07 Environmental-Correlates-of-COVID19-Spain/Response-to-Reviewers-1.Rmd

@@
 > *We changed the scale for the colors to be consistent for all maps.
-We tested the insets for the Canary Islands, but they crowded the figures and added relatively little information.
-We have decided instead to exclude them from the maps.
-Once this is done, the line for Madrid is less obtrusive, and overall the figures are easier to read.*
+After your comment about the connectivity of the islands, we decided to exclude them, and therefore the maps now show only the coterminous provinces.
+After this, the line for Madrid is less obtrusive, and overall the figures are easier to read.*
 

### 2020-05-07 README.Rmd

--- 2020-04-20 README.Rmd
@@
 
-Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health systems and the economy.
-Public health interventions have slowed the spread of the virus, albeit at a high cost.
+The novel SARS-CoV2 has disrupted health systems and the economy and public health interventions to slow its spread have been costly.
 How and when to ease restrictions to movement hinges in part on whether SARS-CoV2 will display seasonality due to variations in temperature, humidity, and hours of sunshine.
-Here, we address this by means of a spatio-temporal analysis of the incidence of COVID-19 in Spain.
+Here, we address this question by means of a spatio-temporal analysis in Spain of the incidence of  COVID-19, the disease caused by the virus.
 Use of spatial Seemingly Unrelated Regressions (SUR) allows us to model the incidence of reported cases of the disease per 100,000 population as an interregional contagion process, in addition to a function of temperature, humidity, and sunshine.
@@
 
-From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus has threatened to overrun health systems the world over.
-In efforts to contain the spread, numerous governments in many nations and regions have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place.
-There are encouraging signs that these measures have arrested the spread of the virus where they have been implemented, and have thus helped to keep a bad situation from becoming even worse [e.g., -@Lancastle2020impact].
-However, this has come at a high cost, and the consequences for all spheres of the economy, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how].
+From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus and COVID-19, the disease it causes, have threatened to overrun health systems around the world.
+In efforts to contain the spread of the disease, numerous governments in many regions and nations have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place.
+There are encouraging signs that these measures have mitigated the spread of the virus [e.g., @Lancastle2020impact; @Lewnard2020scientific; @Wilder2020isolation].
+Even so, this has come at a high cost, and the consequences for all spheres of economic, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how].
 As a result, there is a sense of urgency to anticipate the progression of the pandemic, in order to plan for progressive lifting of restrictions to movement and social contact [e.g., @Kissler2020projecting].
@@
 
-A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations.
+A salient question in the debate about how and when to ease social distancing measures is whether the virus will display seasonal variations.
+Existing research on similar pathogens suggests that the virus could be more stable and potentially easier to transmit in conditions of low temperature and low humidity.
+While this is encouraging, it is important to keep in mind that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" [@deangel2020weathering, section 4].
+This urges caution when extrapolating from known viruses.
+The evidence in this respect is as yet inconclusive, and although easing restrictions as the weather warms may appear tempting, doing so prematurely could well undo weeks or possibly months of costly measures.
+
+It is not surprising, given the stakes involved, that this issue has already triggered a lively debate.
+The current state of knowledge was well-summarized by the National Academy of Sciences, Engineering, and Medicine in the U.S.
+in a recent report [see @National2020rapid].
+Engaged by the Office of the Executive for guidance on this matter, this organization concluded that: "[some] limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV2 pandemic unfolds could shed more light on the effects of climate on transmission" (p.
+6).
+To further complicate matters, much of the relevant work has yet to be peer-reviewed [see for instance the challenge of @Harbert2020spatial; to @Araujo2020spread].
+
+With the above considerations in mind, our objective with this paper is to investigate the influence of environmental factors, concretely temperature, humidity, and sunshine, on the progression of the pandemic.
+We adopt a population health approach, and report results from a spatio-temporal model of the incidence of COVID-19 in the coterminous provinces in Spain, one of the countries hardest hit by the pandemic.
+We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
+We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables.
+These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a contagion process.
+
+The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
+The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
+Our control variables also provide some intriguing insights.
+Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
+The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
+
+Please note that this paper is prepared as a reproducible research document.
+The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{\url{https://drive.google.com/drive/u/1/folders/1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_}}.
+
+Background {#background}
+============
+
+The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
+In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
+Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
+As noted in the introduction, the focus of the paper is on environmental variables, concretely three related to meteorological conditions: temperature, humidity, and sunshine.
+
+Much of what is known about the potential seasonality of SARS-CoV2 is a result of research on other pathogens.
 Earlier, diverse studies have shown the effect of temperature and humidity on the incidence of influenza [e.g., @Makinen2009cold; @Jaakkola2014decline; @Kudo2019low].
@@
 [-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low.
-After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV.
-In this way, Casanova et al.
-[-@Casanova2010effects] used two surrogates, namely the gastroenteritis (TGEV) and mouse hepatitis viruses (MHV), to find that virus inactivation was more rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
-In a similar vein, but working directly with SARS-CoV, Chan et al.
-[-@Chan2011effects] found that viability of the virus was lost at temperatures higher than 38C and relative humidity superior to 95%.
-
-While existing research on similar pathogens suggests that SARS-CoV is more stable and potentially easier to transmit in conditions of low temperature and low humidity, it is far from certain that this will also be the case with the novel SARS-CoV2.
-If such is the case, there is the possibility of easing restrictions to social contact as the weather warms; however, weeks or possibly months of costly measures could become undone if not, and the restrictions are lifted prematurely.
-Not surprisingly, given the stakes involved, this issue has already triggered a lively debate.
-
-Some of what is thought about the possible seasonality of COVID-19 is based on analogies to the patterns of other known respiratory viruses.
-However, de Ángel Solá et al.
-[-@deangel2020weathering] note that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" (section 4).
-This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive.
+After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV [@Casanova2010effects; @Chan2011effects].
+Casanova et al.
+[-@Casanova2010effects], for instance, used surrogates to find that virus inactivation was likely more rapid at higher temperatures; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels.
+Chan et al.
+[-@Chan2011effects] also found that viability of the virus that causes SARS is also lost at higher temperatures (>38°C) and relative humidity superior to 95%.
+
+Whether results from laboratory experiments will hold when the virus circulates in the community remains uncertain.
 At a global scale, de Ángel Solá et al.
-[-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al.
+[-@deangel2020weathering] see less risk from SARS-CoV2 in the Caribean Basin; on the other hand, Coelho et al.
 [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate.
@@
 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).
-To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread).
-In the United States, the National Academy of Sciences, Engineering, and Medicine was engaged by the Office of the Executive for guidance on this matter [see @National2020rapid].
-Their conclusion summarizes the situation well (see p.
-6): "Some limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV-2 pandemic unfolds could shed more light on the effects of climate on transmission."
-
-With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature, humidity, and sunshine.
-We adopt a population health approach, and report results from a spatial model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic.
-We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days.
-We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables.
-These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a spatial contagion process.
-
-The results provide evidence of the effect of temperature, humidity, and sunshine on the incidence of COVID-19.
-The clearest result with respect to these variables is a lower incidence of COVID-19 at higher temperatures and levels of humidity, while the opposite happens with respect to hours of sunshine.
-Our control variables also provide some intriguing insights.
-Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
-The results of this analysis provide support to the hypothesis of seasonality of the novel SARS-CoV2, and should be of interest to public health officials and policy makers grappling with the question of the trajectory of the pandemic.
-
-Please note that this paper is prepared as a reproducible research document.
-The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following anonymous Drive folder:
-
-https://drive.google.com/open?id=1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_
+
+In addition to the environmental variables above, from a population health perspective it is also important to account for potential socio-economic and demographic confounders.
+
+To account for population-level factors, the first variable that we consider is GDP per capita.
+Much has been written about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
+A18)}.
+The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear but the plane has already departed [@Zhou2016accelerated].
+In reference to the earlier outbreak of SARS, van Wagner [-@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect.
+In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic.
+Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional].
+Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown.
+On the other hand, we cannot discount the possibility that less wealthy regions have a higher proportion of workers in manual occupations who cannot telework, and therefore have more difficulties complying with shelter-in-place orders.
+
+Secondly, we consider percentage of older adults (over 65) in a region.
+Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g.
+@Novel2020epidemiological].
+However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission rates of the infection.
+The tool of choice in containing the spread of the disease has been social distancing.
+In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility than most everyone, except the youngest members of the public [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
+In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation.
+Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age].
+Since the age-structured matrix of social contact in Spain is similar to Italy [see @Prem2017projecting], our expectation is that populations with higher percentages of older adults will tend to have lower levels of social contact and hence of incidence.
+
+Population density is also relevant since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
+The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g.
+cumulative incidence in urban areas like NYC).
+For this reason, we anticipate a positive relationship between population density and the incidence of the disease.
+
+The last variable that we consider as a control is the presence of mass transit systems in a province.
+Every province in Spain offers some form of public transportation, but only five provinces have higher order systems of mass mobility (e.g.
+metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
+Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
+After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.
 
 Context and Data {#context-and-data}
-============
+===================
 
@@
 
-The first reported case of COVID-19 in Spain was on January 31th, when a German tourist in the Canary Islands tested positive for the virus.
-However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
-In a short period of time, after this relatively slow start, the outbreak flared.
+The first reported case of COVID-19 in Spain was on January 31st, 2020, when a German tourist in the Canary Islands tested positive for the virus.
+After this case, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia).
+In a short period of time, as testing started to ramp up, it became clear that an outbreak was flaring.
 By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic.
 This declaration marked a turning point for the public in Spain too.
-As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe, after Italy.
+As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe after Italy.
 In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th.
@@
 
-## Selection of Variables {#variables}
-
-The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global].
-In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife.
-Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important.
-As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.
-These variables have been implicated in the viability and ease of transmission of similar viruses.
-In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.
-
-The first variable that we consider is GDP per capita.
-Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p.
-A18)}.
-The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated].
-In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect.
-In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic.
-Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional].
-Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown.
-On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.
-
-The percentaje of older adults (over 65) is the second variable that we consider as a control.
-Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g.
-@Novel2020epidemiological].
-However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection.
-The tool of choice in containing the spread of the disease has been social distancing.
-In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility].
-In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation.
-Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age].
-Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].
-
-The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling].
-The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g.
-cumulative incidence in urban areas like NYC).
-For this reason, we anticipate a positive relationship between population density and the incidence of the disease.
-
-The last control variable is the presence of mass transit systems in a province.
-Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g.
-metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
-Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing.
-After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.
-
 ## Data {#data}
 
-Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
-For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
-The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España.
+Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat terminology) for the period between March 13th and April 11th, inclusive.
+For our purposes, we consider positive cases reported, but exclude symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test.
+The Spanish National Government publishes periodic updates at the regional level (NUTS2) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España.
 This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
-In addition, we consider two sets of explanatory variables.
-The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food).
-The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute).
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.
+We consider two sets of explanatory variables.
+The first one, and the focus of this research, are the three environmental variables, collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food).
+The second set provides some relevant controls as discussed above, and are also collected from official sources, i.e., INE, the national statistics institute.
+Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data used in this research.
 
@@
                       "AEMET",
-                      "MAPA")) %>%
+                      "MAPA",
+                      "AEMET")) %>%
   kable(#"latex",
@@
 <!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/ -->
+
   ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
@@
 week11.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week12.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
         plot.margin = margin(0, 0, 0, 0, "cm")) +
-  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 9, y = madrid_xy[2] + 0.6) +
-  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])
+  geom_sf_text(label = "Madrid", 
+               x = madrid_xy[1] - 4.7, 
+               y = madrid_xy[2] + 0.6) +
+  geom_segment(x = madrid_xy[1] - 3.4,
+               y = madrid_xy[2] + 0.6, 
+               xend = madrid_xy[1], 
+               yend = madrid_xy[2])
 
 week13.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week14.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 week15.plot <- covid19_spain %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(province, week = isoweek(Date)) %>% 
@@
 Each box-and-whisker in the figure represents the distribution of the variable for a community over the 30-day period.
-In the plot, the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost.
+In the plot, the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Andalucia the southernmost.
 As seen in the figure, there is a relatively wide range of values both within and between provinces over the 30-day period examined.
 The top panel of the figure shows the distribution of mean temperatures.
-The lowest mean temperature for a community on any given day was 2.8C, and the highest 22.4C, for a range of approximately 20 degrees.
-Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6.
+The lowest mean temperature for a community on any given day was approximately 3°C, and the highest about 20°C, for a range of approximately 17 degrees.
+Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is approximately 48% and the highest is close to 100%.
 Finally, the bottom panel displays mean daily hours of sunshine in the community.
@@
                              GDPpc > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
                              Older > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
                              Density > q[4] ~ "4th")) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
                                      "Yes"))) %>%
-  #filter(CCAA != "Canarias") %>%
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   ggplot() +
@@
 Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive.
+Overall for the period under examination, the pairwise correlations  between these variables and incidence are significant at $p<0.05$, with the exception of the three Sunshine_hours variables.
 
@@
 covid19_spain %>% 
+  filter(CCAA != "Canarias", province != "Baleares") %>%
   group_by(Date) %>%
-  summarize(Older = cor(log(Older), log(Incidence)),
-            GDPpc = cor(log(GDPpc), log(Incidence)),
+  summarize(GDPpc = cor(log(GDPpc), log(Incidence)),
+            Older = cor(log(Older), log(Incidence)),
             Density = cor(log(Density), log(Incidence)),
@@
 To implement the SUR approach, we must define a matrix of spatial weights $W$.
-In this case, we consider neighborhoods based on adjacency, based on the well-known rook criterion (two provinces are adjacent if they share a boundary, but not if they touch only at a vertex).
-We modify this criterion in two cases.
-First, we make an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are adjacent between each other.
-And secondly, we assume that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello).
-After matrix $W$ has been specified, estimation of the model can proceed as usual.
+In this case, we consider neighborhoods based on adjacency, based on the well-known queen criterion (two provinces are adjacent if they share a boundary or touch at a vertex).
+The analysis is of the coterminous provinces\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and included the three islands in the sample.
+In this case we made an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), and assumed that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello) The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.
 
 For estimation, we log-transform the dependent and quantitative independent variables.
-The only variable that is not transformed is the categorical variable for transity systems.
+The only variable that is not transformed is the categorical variable for transit systems.
 A log-log transformation is appropriate to capture non-linear relationships between variables and provides a straightforward interpretation of the coefficients as percentage change.
@@
 
-Results and Discussion {#results-and-discussion}
+Discussion {#discussion}
 ===================
 
-Table \ref{tab:summary-best-model} presents a summary of the results of Model 2 (*lag11*).
+Table \ref{tab:summary-best-model} presents a summary of the results of Model 1 (*lag8*).
 Recall that two coefficients were constrained and are estimated only for the first equation of the system, and thus are assumed to be constant over time.
-These are the coefficients corresponding to GDP per capita and percentage of older adults.
-Both coefficients are significant at $p\leq0.05$.
+These are the coefficients corresponding to GDP per capita ($p\leq0.10$) and percentage of older adults ($p\leq0.05$).
 The sign of the coefficient for GDP per capita is positive, which indicates that wealthier regions tend to have a higher incidence of COVID-19.
@@
 We begin our discussion with the evolution of the spatial autocorrelation coefficient ($\rho$) in Figure \ref{fig:results-time-1} (left panel).
-It is interesting to note that the magnitude of the spatial autocorrelation coefficient $\rho$ declines over the period under analysis, and is not significant for some days.
+We notice that the magnitude of the spatial autocorrelation coefficient $\rho$ declines over the period under analysis, and is not significant for some days.
 This is an interesting result: immediately prior to the declaration of the state of emergency, there appears to have been a strong inter-provincial contagion effect.
-Keeping in mind that the incubation period ranges between 2 and 11 days with a median of 5, it is reasonable to expect that the effect of the lockdown will be observed with some delay.
-Indeed, as seen in the figure, the autocorrelation coefficient remains high for three days, then declines on March 16, and continues to weaken over time.
-At the end of the period under examination, the strength of this effect is much diminished.
-We would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero.
-The fact that it is not is evidence of some weak contagion effect due to restrictions to the movement between provinces.
+Keeping in mind that the incubation period ranges between 2 and 12 days with a median of 5, it is reasonable to expect that the effect of the lockdown will be observed with some delay.
+Indeed, as seen in the figure, the autocorrelation coefficient remains high for several days, then begins to decline around March 23, and continues to weaken over time.
+At the end of the period under examination, the strength of this effect is much diminished and we would expect that under full compliance with strict lockdown conditions (meaning no inter-provincial mobility) the spatial contagion effect would be zero - as seems to be the case.
 
@@
 
-The preceding discussion helps to establish the inferential contributions of the analysis.
+The preceding discussion helps to establish the inferential contributions of the analysis, and indicate which variables display significant statistical associations with incidence of COVID-19.
 The remaining question is, what are the implications.
-As discussed in Section \ref{methods} the question of the magnitude of the effects is not clear from the coefficients alone, since a change to a variable in a province influences, via the contagion effect, its neighbors.
-For this reason, the appropriate way to estimate the effects is to calculate both the own and contagion effects (in other words the direct and indirect effects, respectively), as well as their total.
-A summary of the different variable effects appears in Table \ref{tab:summary-impacts-best-model}.
+
+As discussed in Section \ref{methods} the effect of a variable is not clear from its coefficient alone, since a change to a variable in a province influences, via the contagion effect, its neighbors.
+For this reason, the appropriate way to estimate the effects is to calculate both the own effect and the effect due to contagion, or in other words the direct and indirect effects, respectively.
+The total effect is the sum of the two.
+A summary of the effects appears in Table \ref{tab:summary-impacts-best-model}.
 All effects in the table are interpreted as percentage change in the incidence of COVID-19 as a consequence of a one percent change in the variable.
-The exception is Transit (which was not log-transformed), and this variable represents the percentage change in incidence between provinces without and with mass transit systems.
+The exception to this is Transit (which was not log-transformed).
+This variable instead represents the percentage change in incidence between provinces without and with mass transit systems.
 
 Two variables had temporally constrained coefficients.
-The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.675% for each percentual increase of this variable (in €1,000s).
-In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemia.
-Percentage of older adults has a negative effect, and each percentual increase in this variable is associated with a relatively small reduction of the incidence of less than 1%.
+The estimated effect of GDP per capita is to increase the incidence of COVID-19 by 0.449% for each percent increase of this variable (in €1,000s).
+In our view, this is a measure of inertia, as provinces with higher GDP per capita where among the first to see exponential growth in the pandemic.
+Percentage of older adults has a negative effect, and each percent increase in this variable is associated with a relatively small reduction of the incidence of approximately 0.67%.
 
@@
 The largest positive direct effect is Transit, and the largest direct negative effects are temperature and humidity.
-The direct effect of these variables is as follows: for each percentual point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
-This effect is amplified via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 1.5%.
-The effect of humidity is also to reduce the incidence: each percentual point of increase in humidity is associated with a reduction of up to 1.5% in incidence.
+The direct effects of these variables are as follows: for each percent point increase in temperature, there is between a 1% and 2% reduction in the incidence of the disease.
+This effect is compounded via contagion, as seen in the central panel in the figure, and the indirect effect can further reduce the incidence by up to 0.75%.
+The effect of humidity is also to reduce the incidence: each percent point of increase in humidity is associated with a reduction of up to 2% in incidence.
 With the addition of the indirect effect, the total effect of a 1% increase in humidity is to reduce incidence by up to 3%.
-As seen in the figure, the indirect (i.e., contagion) effects are almost as strong as the direct effects immediately before and at the beginning of the lockdown period.
-However, by the end of the lockdown period, the indirect effects have weakened considerably.
-The effect of density goes from positive (+0.334) to negative (-0.244).
-It is crucial to keep in mind here that this is the effect of density over a period of *lockdown* and possibly subject to risk compensation, as discussed above.
-In other words, the lockdown seems to be more effective in places with higher density.
-Accordingly, cities should not rush to ease lockdowns, but rather efforts should be made to reduce or eliminate risk compensation in less densely populated regions.
+As seen in the figure, the indirect (i.e., contagion) effects are stronger at the before and at the beginning of the lockdown period.
+Nonetheless, by the end of the period under study, the indirect effects have weakened considerably.
 
 What do these effects mean?
-Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the effect of temperature is estimated to reduce the incidence of COVID-19 by between one and two percent with each percentual increase of this variable.
-Without a lockdown, or a relaxed lockdown, the effect is even greater, but only because the disease is transmitted inter-regionally!
-Imagine a region that is closed to its neighbors: increases in temperature are associated with smaller reductions in incidence - however, it does not import or export cases.
-A province that is not closed and not under lockdown will see a greater effect of temperature, but at the cost of importing/exporting cases.
+Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the total effects of the variables tend towards their direct effects.
+In the first few days covered by our analysis the total effect of all variables is greater due to the spatial contagion effect.
+Contagion makes analysis and intervention more complex: the contagion effect essentially acts like a multiplier, whereby developments in each province spill over to their neighbors.
+Once the contagion effect has been tamed, each province can be "treated" independently from its neighbors.
+
+It is important, before concluding our discussion, to highlight some limitations of this study.
+
+First, the analysis was conducted mostly under a situation of lockdown, and therefore, besides first days of the period examined, one must exercise caution when trying to extrapolate the findings to a situations without lockdown.
+Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
+In this case we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results; it is still important, however, to remain aware that the number of true cases is likely higher.
+Finally, we defined neighborhoods based on adjacency.
+It would be interesting to compare adjacency to other connectivity criteria, for instance based on domestic transportation instrastructure and services.
+We flag this as a matter for future research.
 
@@
         col.names = c("Variable", "Min", "Mean", "Max", "Note"),
-        caption = "\\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)") %>%
+        caption = "\\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag8: lagged 8-day moving average)") %>%
   #kable_styling(latex_options = c("striped")) %>%
@@
 
-Needless to say, the analysis presented here is ecological, and at the level of population health.
+Needless to say, the analysis presented here is at the level of population health.
 For this reason, the analysis does not make any claims with respect to the effect of ultraviolet light on the virus, but rather about transmission of the virus in the population.
-In this way, the analysis does not say that the virus moves less effectively where more people live in close proximity to each other, but rather that humans are more contagious when they feel safe in less dense regions.
+For example, the analysis does not imply that the virus moves less effectively in places where more people live in close proximity to each other, but rather that humans are more contagious when they feel safe in less dense regions.
 Similarly, more sunshine does not mean that the virus thrives, but rather that humans are more contagious to each other when their behavior adapts to this environmental condition.
 
-It is important, in this respect, to highlight some limitations of this study.
-
-First, the analysis was conducted under a situation of lockdown, and therefore, besides the results are the beginning of the period examined, one must exercise caution when trying to extrapolate the findings to other situations.
-Secondly, it is well known that there is in many countries substantial underreporting of cases of COVID-19 due to limited testing.
-However, we do not suspect systematic provincial bias in reporting, and as long as the underreporting is consistent across units of analysis, we do not expect biased results - even if we are aware that the number of true cases is likely higher.
-Thirdly, all environmental data are based on a single station in a province.
-Here, we made an effort to select the most representative station based on the distribution of the population.
-An alternative approach would be to calculate the environmental variables based on the average of multiple stations within each province: this would likely smooth to some extent these variables.
-Finally, we defined spatial neighborhoods based on adjacency.
-It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services.
-We flag this as a matter for future research.
+Some directions for future research include investigating other modelling frameworks, such as geographically and temporally weighted regression and/or space-time conditional autoregressive models.
+In addition, the environmental variables examined here relate to meteorological conditions only, and did not include other environmental factors that may incide in the transmission of the virus, such as pollution.
+These other factors should be incorporated in future studies.
 

### 2020-05-07 covid19env/data_processing/02-Data-Preprocessing-2.Rmd

--- 2020-04-18 covid19env/data_processing/01-Data-Preprocessing.Rmd
@@
 
-Load climate data (ignore the humidity data in this file; it was a calculated variable):
+Load climate data (This is the output of `01-Preprocess-meteo-data.Rmd`):
 ```{r}
-load("COVID-19-Datos_Climaticos_Spain-Extendido.RData")
+load("meteo_data_province.RData")
 ```
@@
 
-Prepare climatic factors:
+Arrange humidity data:
 ```{r}
-# Maximum temperature
-TMax <- TMax %>% select(-ID, -provincia) %>%
-  pivot_longer(starts_with("T"), 
-               names_to = "Date",
-               values_to = "Max_Temp") %>%
-  mutate(Date = rep(seq(dmy("01-03-2020"), 
-                        dmy("11-04-2020"), 
-                        by = "days"), 
-                    50))
-
-# Minimum temperature
-TMin <- TMin %>% select(-ID, -provincia) %>%
-  pivot_longer(starts_with("T"), 
-               names_to = "Date",
-               values_to = "Min_Temp") %>%
-  mutate(Date = rep(seq(dmy("01-03-2020"), 
-                        dmy("11-04-2020"), 
-                        by = "days"), 
-                    50))
-
-# Mean temperature
-TMed <- TMed %>% select(-ID, -provincia) %>%
-  pivot_longer(starts_with("T"), 
-               names_to = "Date",
-               values_to = "Mean_Temp") %>%
-  mutate(Date = rep(seq(dmy("01-03-2020"), 
-                        dmy("11-04-2020"), 
-                        by = "days"), 
-                    50))
-
-# Hours of sunshine 
-Sun <- Sun %>% select(-ID, -provincia) %>%
-  pivot_longer(starts_with("S"), 
-               names_to = "Date",
-               values_to = "Sunshine_Hours") %>%
-  mutate(Date = rep(seq(dmy("01-03-2020"), 
-                        dmy("11-04-2020"), 
-                        by = "days"), 
-                    50))
-
-# Precipitation 
-Plu <- Plu %>% select(-ID, -provincia) %>%
-  pivot_longer(c(starts_with("P")), 
-               names_to = "Date",
-               values_to = "Precipitation") %>%
-  mutate(Date = rep(seq(dmy("01-03-2020"), 
-                        dmy("11-04-2020"),
-                        by = "days"), 
-                    50)) %>%
-  replace_na(list(Precipitation = 0)) # Replace two missing values in Malaga on 2020-03-01 and 2020-03-02
-
 # Humidity 
@@
 ```{r}
-climatic_vars <- TMax %>%
-  left_join(TMin, by = c("ID_INE", "Date")) %>%
-  left_join(TMed, by = c("ID_INE", "Date")) %>%
-  left_join(Sun, by = c("ID_INE", "Date")) %>%
-  left_join(Plu, by = c("ID_INE", "Date"))  %>%
-  left_join(Hum, by = c("ID_INE", "Date"))
+meteo_data_province <- meteo_data_province %>%
+  left_join(Hum, by = c("ID_INE", "date" = "Date")) %>%
+  rename(Max_Temp = MaxTemperature,
+         Min_Temp = MinTemperature,
+         Mean_Temp = MeanTemperature,
+         Sunshine_Hours = SunshineHours) %>% 
+  filter(date >= "2020-03-01" & date <= "2020-04-11")
 ```
@@
 # Lagged mean temperature: 8 day average
-TMed_lag8 <- TMed %>% 
+TMed_lag8 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
   transmute(ID_INE, 
-            Date = rep(seq(dmy("01-03-2020"),
+            date = rep(seq(dmy("01-03-2020"),
                            dmy("11-04-2020"), 
@@
 # Lagged sunshine hours: 8 day average
-Sun_lag8 <- Sun %>% 
+Sun_lag8 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged precipitation: 8 day average
-Plu_lag8 <- Plu %>% 
+Plu_lag8 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged humidity: 8 day average
-Hum_lag8 <- Hum %>% 
+Hum_lag8 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged mean temperature: 8 day average
-TMed_lag11 <- TMed %>% 
+TMed_lag11 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged sunshine hours: 8 day average
-Sun_lag11 <- Sun %>% 
+Sun_lag11 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged precipitation: 8 day average
-Plu_lag11 <- Plu %>% 
+Plu_lag11 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged humidity: 8 day average
-Hum_lag11 <- Hum %>% 
+Hum_lag11 <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged mean temperature: 8 day average
-TMed_lag11w <- TMed %>% 
+TMed_lag11w <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged sunshine hours: 8 day average
-Sun_lag11w <- Sun %>% 
+Sun_lag11w <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged precipitation: 8 day average
-Plu_lag11w <- Plu %>% 
+Plu_lag11w <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 # Lagged humidity: 8 day average
-Hum_lag11w <- Hum %>% 
+Hum_lag11w <- meteo_data_province %>% 
   group_by(ID_INE) %>%
@@
 ```{r}
-climatic_vars <- climatic_vars %>%
-  left_join(TMed_lag8, by = c("ID_INE", "Date")) %>%
-  left_join(Sun_lag8, by = c("ID_INE", "Date")) %>%
-  left_join(Plu_lag8, by = c("ID_INE", "Date")) %>%
-  left_join(Hum_lag8, by = c("ID_INE", "Date")) %>%
-  left_join(TMed_lag11, by = c("ID_INE", "Date")) %>%
-  left_join(Sun_lag11, by = c("ID_INE", "Date")) %>%
-  left_join(Plu_lag11, by = c("ID_INE", "Date")) %>%
-  left_join(Hum_lag11, by = c("ID_INE", "Date")) %>%
-  left_join(TMed_lag11w, by = c("ID_INE", "Date")) %>%
-  left_join(Sun_lag11w, by = c("ID_INE", "Date")) %>%
-  left_join(Plu_lag11w, by = c("ID_INE", "Date")) %>%
-  left_join(Hum_lag11w, by = c("ID_INE", "Date"))
+meteo_data_province <- meteo_data_province %>%
+  left_join(TMed_lag8, by = c("ID_INE", "date")) %>%
+  left_join(Sun_lag8, by = c("ID_INE", "date")) %>%
+  left_join(Plu_lag8, by = c("ID_INE", "date")) %>%
+  left_join(Hum_lag8, by = c("ID_INE", "date")) %>%
+  left_join(TMed_lag11, by = c("ID_INE", "date")) %>%
+  left_join(Sun_lag11, by = c("ID_INE", "date")) %>%
+  left_join(Plu_lag11, by = c("ID_INE", "date")) %>%
+  left_join(Hum_lag11, by = c("ID_INE", "date")) %>%
+  left_join(TMed_lag11w, by = c("ID_INE", "date")) %>%
+  left_join(Sun_lag11w, by = c("ID_INE", "date")) %>%
+  left_join(Plu_lag11w, by = c("ID_INE", "date")) %>%
+  left_join(Hum_lag11w, by = c("ID_INE", "date"))
 ```
@@
 covid19_spain <- covid19_spain %>%
-  full_join(climatic_vars, by = c("ID_INE", "Date"))
+  full_join(meteo_data_province, by = c("ID_INE", "Date" = "date"))
 ```
@@
 
-## Plot two control variables Ratio Male/Female; Median_Age
-
-Ratio of male to female in province:
-```{r}
-ggplot(data = provinces_spain) + 
-  geom_sf(aes(fill = Male2Female)) + 
-  scale_fill_distiller(palette = "Blues", direction = 1) + 
-  theme_tufte()
-```
-
-Median age of population in province:
-```{r}
-ggplot(data = provinces_spain) + 
-  geom_sf(aes(fill = Median_Age)) + 
-  scale_fill_distiller(palette = "Blues", direction = 1) + 
-  theme_tufte()
-```
-
-### Incidence progress
-
-Plot the progression of the incidence of COVID-19 cases by province:
-```{r}
-p1 <- ggplot(data = covid19_spain %>%
-               filter(Date == "2020-03-19") %>% 
-         left_join(provinces_spain, by = "ID_INE") %>% 
-         st_as_sf()) + 
-  geom_sf(aes(fill = Incidence)) + 
-  scale_fill_distiller(palette = "Reds", direction = 1) + 
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom") + 
-  facet_wrap(~ Date, nrow = 3)
-
-p2 <- ggplot(data = covid19_spain %>%
-               filter(Date == "2020-03-28") %>% 
-         left_join(provinces_spain, by = "ID_INE") %>% 
-         st_as_sf()) + 
-  geom_sf(aes(fill = Incidence)) + 
-  scale_fill_distiller(palette = "Reds", direction = 1) + 
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom") + 
-  facet_wrap(~ Date, nrow = 3)
-
-p3 <- ggplot(data = covid19_spain %>%
-               filter(Date == "2020-04-03") %>% 
-         left_join(provinces_spain, by = "ID_INE") %>% 
-         st_as_sf()) + 
-  geom_sf(aes(fill = Incidence)) + 
-  scale_fill_distiller(palette = "Reds", direction = 1) + 
-  theme_tufte() +
-  theme(axis.text = element_blank(),
-        legend.position = "bottom") +
-  facet_wrap(~ Date, nrow = 3)
-
-gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
-```
-
 ## Save data
@@
 ```{r}
-covid19_spain <- covid19_spain %>%
+covid19_spain_2 <- covid19_spain %>%
   filter(!is.na(province)) %>% # Remove records that are missing data on the province
-  filter(!is.na(Max_Temp)) # Remove records that are missing climatic variables
+  filter(!is.na(Mean_Temp)) # Remove records that are missing climatic variables
 
@@
 ```{r}
-summary(covid19_spain)
+summary(covid19_spain_2)
 ```
@@
 # Dataframe
-save(covid19_spain, file = "covid19_spain.RData", compress = "xz")
+save(covid19_spain_2, file = "covid19_spain_2.RData", compress = "xz")
 

### 2020-05-12 README.Rmd
Updated README.Rmd after acceptance by Geographical Analysis

@@
 Maira Galdino da Rocha Pitta (Universidade Federal de Pernambuco)
+
+**Forthcoming in Geographical Analysis**
 
@@
 <!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/ -->
-
 ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)

### 2020-05-12 COVID-19 Municipalities v0.Rmd

@@
 ```
-** Cataluña 30 de Marzo**
+
+**Cataluña 30 de Marzo**
 ```{r cat-p1}
@@
 ```
-** Cataluña 30 de Abril**
+
+**Cataluña 30 de Abril**
 ```{r cat-p2}
@@
 
-** País Vasco 20 de Marzo - 20 Abril**
+**País Vasco 20 de Marzo**
 ```{r pv-p1}
@@
 ```
-** País Vasco 26 de Abril**
+
+**País Vasco 20 de Abril**
 ```{r pv-p2}
@@
 ```
+
 **Madrid de 8 de Abril**
@@
 ```
+
 **Madrid de 4 de Mayo**

### 2020-05-12 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Added missing reference to aemet

@@
 
-The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
+The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [@Rodriguez2018aemet], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].
 

### 2020-05-12 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Added missing reference to aemet

@@
 investigation and the authors wish to acknowledge their developers:
-\texttt{aemet} {[}{]}, \texttt{ggthemes} (Arnold, 2019),
-\texttt{gridExtra} (Auguie, 2017), \texttt{kableExtra} (Zhu, 2019),
-\texttt{knitr} (Xie, 2015, 2014), \texttt{lubridate} (Grolemund and
-Wickham, 2011), \texttt{plm} (Millo, 2017), \texttt{rticles} (Allaire et
-al., 2020), \texttt{sf} (Pebesma, 2018), \texttt{spdep} (Bivand et al.,
-2013), spsur (Angulo et al., 2020) \texttt{tidyverse} (Wickham et al.,
-2019), \texttt{units} (Pebesma et al., 2016).
+\texttt{aemet} (Rodriguez-Sanchez et al., 2018), \texttt{ggthemes}
+(Arnold, 2019), \texttt{gridExtra} (Auguie, 2017), \texttt{kableExtra}
+(Zhu, 2019), \texttt{knitr} (Xie, 2015, 2014), \texttt{lubridate}
+(Grolemund and Wickham, 2011), \texttt{plm} (Millo, 2017),
+\texttt{rticles} (Allaire et al., 2020), \texttt{sf} (Pebesma, 2018),
+\texttt{spdep} (Bivand et al., 2013), spsur (Angulo et al., 2020)
+\texttt{tidyverse} (Wickham et al., 2019), \texttt{units} (Pebesma et
+al., 2016).
 
@@
 
+\leavevmode\hypertarget{ref-Rodriguez2018aemet}{}%
+Rodriguez-Sanchez, F., Balao, F., Gámez, D., 2018. Aemet: Obtain
+climatic and meteorological data from spanish meteorological agency
+(aemet).
+
 \leavevmode\hypertarget{ref-Roorda2010trip}{}%

### 2020-05-15 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Updated link to repository and acknowledgments before production.

@@
 Please note that this paper is prepared as a reproducible research document.
-The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{\url{https://drive.google.com/drive/u/1/folders/1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_}}.
+The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{\url{https://github.com/paezha/covid19-environmental-correlates}}.
 
@@
 
-Add acknowledgments as appropriate in final draft.
-
+Fernando A.
+López is supported by the program Groups of Excellence of the Region of Murcia, Fundación Séneca, Science and Technology Agency of the region of Murcia project 19884/GERM/15.
+Tatiane Menezes, Renata Cavalcanti, and Maira Galdino da Rocha Pitta are supported by Brazil's Conselho Nacional de Desenvolvimento Científico e Tecnológico (CNPq) and Coordenação de Aperfeiçoamento de Pessoal de Nível Superior (Fundação Capes).
+Antonio Páez is not supported by Canada's Research Councils.
 The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [@Rodriguez2018aemet], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].

### 2020-05-15 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Updated link to repository and acknowledgments before production.

@@
 from a public
-repository\footnote{\url{https://drive.google.com/drive/u/1/folders/1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_}}.
+repository\footnote{\url{https://github.com/paezha/covid19-environmental-correlates}}.
 
@@
 
-Add acknowledgments as appropriate in final draft.
-
-The following \texttt{R} packages were used in the course of this
+Fernando A. López is supported by the program Groups of Excellence of
+the Region of Murcia, Fundación Séneca, Science and Technology Agency of
+the region of Murcia project 19884/GERM/15. Tatiane Menezes, Renata
+Cavalcanti, and Maira Galdino da Rocha Pitta are supported by Brazil's
+Conselho Nacional de Desenvolvimento Científico e Tecnológico (CNPq) and
+Coordenação de Aperfeiçoamento de Pessoal de Nível Superior (Fundação
+Capes). Antonio Páez is not supported by Canada's Research Councils. The
+following \texttt{R} packages were used in the course of this
 investigation and the authors wish to acknowledge their developers:

### 2020-05-29 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Added some missing references flagged during proof-reading for typesetting.

@@
 [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19.
-This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al.
-2016; Dąbrowska-Leonik et al.
-2018; Dinlen et al.
-2016; Mathyssen et al.
-2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).
+This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections [@Zittermann2016vitamin; @Dąbrowska2018vitamin; @Dinlen2016association; @Mathyssen2017vitamin; @Esposito2015vitamin; @Jat2017vitamin; @Moriyama2020seasonality].
 
@@
 As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial].
-When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation].
+When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial].
 In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor.

### 2020-05-29 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Added some missing references flagged during proof-reading for typesetting.

@@
 to higher prevalence and severity of acute respiratory tract infections
-(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
-al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
-Moriyama, Hugentobler, and Iwasaki 2020).
+(Dąbrowska-Leonik et al., 2018; Dinlen et al., 2016; Esposito and Lelii,
+2015; Jat, 2017; Mathyssen et al., 2017; Moriyama et al., 2020;
+Zittermann et al., 2016).
 
@@
 is rejected, several alternative specifications have been proposed to
-include spatial effects (Anselin, 1988, see also 2016). In this paper we
-consider a spatial SUR model that incorporates a spatial lag of the
-dependent variable as an explanatory factor. Spatial analytical
-approaches were used to understand contagion-difussion processes in the
-case of infectious disease in general (e.g., Cliff et al., 1998) and the
+include spatial effects (Anselin, 1988). In this paper we consider a
+spatial SUR model that incorporates a spatial lag of the dependent
+variable as an explanatory factor. Spatial analytical approaches were
+used to understand contagion-difussion processes in the case of
+infectious disease in general (e.g., Cliff et al., 1998) and the
 2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
@@
 
-\leavevmode\hypertarget{ref-Anselin2016estimation}{}%
-Anselin, L., 2016. Estimation and testing in the spatial seemingly
-unrelated regression (sur). Geoda Center for Geospatial Analysis;
-Computation, Arizona State University. Working Paper 2016-01.
-
 \leavevmode\hypertarget{ref-Araujo2020spread}{}%
@@
 
+\leavevmode\hypertarget{ref-Dux105browska2018vitamin}{}%
+Dąbrowska-Leonik, N., Bernatowska, E., Pac, M., Filipiuk, W., Mulawka,
+J., Pietrucha, B., Heropolitańska-Pliszka, E., Bernat-Sitarz, K.,
+Wolska-Kuśnierz, B., Mikołuć, B., 2018. Vitamin d deficiency in children
+with recurrent respiratory infections, with or without immunoglobulin
+deficiency. Advances in medical sciences 63, 173--178.
+
 \leavevmode\hypertarget{ref-Desjardins2020rapid}{}%
@@
 
+\leavevmode\hypertarget{ref-Dinlen2016association}{}%
+Dinlen, N., Zenciroglu, A., Beken, S., Dursun, A., Dilli, D., Okumus,
+N., 2016. Association of vitamin d deficiency with acute lower
+respiratory tract infections in newborns. The Journal of Maternal-Fetal
+\& Neonatal Medicine 29, 928--932.
+
 \leavevmode\hypertarget{ref-Elhorst2014spatial}{}%
@@
 spatial panels. Springer.
+
+\leavevmode\hypertarget{ref-Esposito2015vitamin}{}%
+Esposito, S., Lelii, M., 2015. Vitamin d and respiratory tract
+infections in childhood. BMC infectious diseases 15, 487.
 
@@
 influenza in cold climate. Environmental Health 13, 22.
+
+\leavevmode\hypertarget{ref-Jat2017vitamin}{}%
+Jat, K.R., 2017. Vitamin d deficiency and lower respiratory tract
+infections in children: A systematic review and meta-analysis of
+observational studies. Tropical doctor 47, 77--84.
 
@@
 
+\leavevmode\hypertarget{ref-Mathyssen2017vitamin}{}%
+Mathyssen, C., Gayan-Ramirez, G., Bouillon, R., Janssens, W., 2017.
+Vitamin d supplementation in respiratory diseases: Evidence from
+randomized controlled trials. Polish archives of internal medicine 127,
+775--784.
+
 \leavevmode\hypertarget{ref-Makinen2009cold}{}%
@@
 Geography 19, 39--50.
+
+\leavevmode\hypertarget{ref-Moriyama2020seasonality}{}%
+Moriyama, M., Hugentobler, W.J., Iwasaki, A., 2020. Seasonality of
+respiratory viral infections. Annual Review of Virology 7.
 
@@
 
+\leavevmode\hypertarget{ref-Zittermann2016vitamin}{}%
+Zittermann, A., Pilz, S., Hoffmann, H., März, W., 2016. Vitamin d and
+airway infections: A european perspective. European journal of medical
+research 21, 14.
+
 \end{document}

### 2020-05-29 README.Rmd
Added some missing references flagged during proof-reading for typesetting.

@@
 
-The novel SARS-CoV2 has disrupted health systems and the economy and public health interventions to slow its spread have been costly.
+The novel SARS-CoV2 has disrupted health systems and the economy, and public health interventions to slow its spread have been costly.
 How and when to ease restrictions to movement hinges in part on whether SARS-CoV2 will display seasonality due to variations in temperature, humidity, and hours of sunshine.

### 2020-06-08 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
RELEASE. Updated preprint with names of authors and added doi information to README.Rmd

@@
 author:
-  - name: Author A
-    email: author.a@example.com
-    affiliation: Some School
+  - name: Antonio Paez
+    email: paezha@mcmaster.ca
+    affiliation: McMaster University
     footnote: 1
-  - name: Author B
-    email: author.b@example.com
-    affiliation: Some Institute
+  - name: Fernando A. Lopez
+    email: fernando.lopez@upct.es
+    affiliation: Universidad Politecnica de Cartagena
+  - name: Tatiane Menezes
+    email: tatiane.menezes@ufpe.br
+    affiliation: Departamento de Economia
+  - name: Renata Cavalcanti
+    email: renata.vcsantos@gmail.com
+    affiliation: Nucleo de Pesquisa
+  - name: Maira Galdino da Rocha Pitta
+    email: mgrpitta@ufpe.br
+    affiliation: Nucleo de Pesquisa
 address:
-  - code: Some School
-    address: School of the Things
-  - code: Some Institute
-    address: Institute of Everything
+  - code: McMaster University
+    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
+  - code: Universidad Politecnica de Cartagena
+    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
   - code: Departamento de Economia
+    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
+  - code: Nucleo de Pesquisa
+    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
 footnote:

### 2020-06-08 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
RELEASE. Updated preprint with names of authors and added doi information to README.Rmd

@@
 incidence in Spain}
-    \author[Some School]{Author A\corref{1}}
-   \ead{author.a@example.com} 
-    \author[Some Institute]{Author B}
-   \ead{author.b@example.com} 
-      \address[Some School]{School of the Things}
-    \address[Some Institute]{Institute of Everything}
-    \address[Departamento de Economia]{}
+    \author[McMaster University]{Antonio Paez\corref{1}}
+   \ead{paezha@mcmaster.ca} 
+    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
+   \ead{fernando.lopez@upct.es} 
+    \author[Departamento de Economia]{Tatiane Menezes}
+   \ead{tatiane.menezes@ufpe.br} 
+    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
+   \ead{renata.vcsantos@gmail.com} 
+    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
+   \ead{mgrpitta@ufpe.br} 
+      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
+St W, Hamilton, ON, L8S 4K1, Canada}
+    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
+Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
+30201, Cartagena, Murcia, Spain}
+    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
+Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
+    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
+Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
+50670-901, Brazil}
       \cortext[1]{Corresponding Author}

### 2020-06-08 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
RELEASE. Updated preprint with doi information.

@@
 Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.
-journal: "Geographical Analysis"
+journal: "Geographical Analysis (https://doi.org/10.1111/gean.12241)"
 date: "`r Sys.Date()`"
